### 2024-04-28 paper.tex
Import, add section 2, notes on abstract

\documentclass[english,submission]{programming}
%\usepackage[backend=biber]{biblatex}
%\addbibresource{paper.bib}

\usepackage{multicol}
\usepackage{ccicons}
\usepackage{natbib}

\paperdetails{
  perspective=engineering,
  area={Programming Systems},
}

\begin{document}

\title{Schema Change: Challenge Problems}

\author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
\authorinfo{is TBD. Contact him at \email{jonathanmedwards@gmail.com}.}
\affiliation[a]{Independent, Boston, MA, USA}

\author[b]{Tomas Petricek}%[0000-0002-7242-2208]
\authorinfo{is TBD. Contact him at \email{tomas@tomasp.net}.}
\affiliation[b]{Charles University, Prague, Czechia}

\author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
\authorinfo{is TBD. Contact him at \email{storm@cwi.nl}.}
\affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
\affiliation[d]{University of Groningen, Groningen, Netherlands}

\author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
\authorinfo{is TBD. Contact him at \email{gklitt@gmail.com}.}
\affiliation[b]{Ink \& Switch, Planet Earth}

\authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}

\keywords{schema change, live programming, local-first programming}

% TODO: Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification System

\maketitle

\begin{abstract}
Schema change is an unsolved problem in both live programming and local-first software. We include
in schema change any change to the expected shape of data, whether that is expressed explicitly
in a database schema or type system, or whether those expectations are implicit in the behavior
of the code.

Schema changes during live programming can create a mismatch between the code and
data in the running environment. Correspondingly, Schema changes in local-first programming can
create mismatches between data in different replicas, and between data in a replica and the code
colocated with it. In all of these situations the problem of schema change is to migrate or
translate existing data in coordination with changes to the code.

This paper contributes a set of concrete scenarios involving schema change that are offered as
challenge problems to the live programming and local-first communities. We hope that these
problems will spur progress by providing concrete objectives and a basis for comparing alternative
solutions.

~

~

\noindent\textbf{Context}: \emph{What is the broad context of the work? What is the importance of the general research area?}
Schema change is unsolved problem in live and local-first programming. It is also poorly understood
problem that calls for a more fundamental analysis. It is particularly important as more programming
is done in programming systems where programs may be changed by their users and as more programming
is local-first, meaning that such changes need to be synchronized.

\noindent\textbf{Inquiry}: \emph{What problem or question does the paper address? How has this problem or question been addressed?}
Specific systems have specific solutions, but what is lacking is a general framework for thinking
about schema change and also a good suite of challenge problems to motivate and evaluate new solutions.

\noindent\textbf{Approach}: \emph{What was done that unveiled new knowledge?}
We collect challenge problems from a range of different kinds of (primarily) live and local-first
application contexts. We develop a framework for thinking about program change and use it to
present our challenge problems in a systematic way.

\noindent\textbf{Knowledge}: \emph{What new facts were uncovered? What new capabilities are enabled by the work?}
We understand change in programs in a novel way, using a two-dimensional framwork that looks at
the different pace layers in a program and at the different ways in which change is propagated
across program variants.

\noindent\textbf{Grounding}: \emph{What argument, feasibility proof, artifacts, or results and evaluation support this work?}
We use the framework to provide a framing for a range of challenge problems derived from both
existing literature and our own expertise.

\noindent\textbf{Importance}: \emph{Why does this work matter?}
This paper contributes a set of concrete scenarios involving schema change that are offered as
challenge problems to the live programming and local-first communities. We hope that these
problems will spur progress by providing concrete objectives and a basis for comparing alternative
solutions.

\end{abstract}

\section{Introduction}

Schema change won't go away. Changing requirements and changing code lead to changing the schema of a database. Schema change, also called schema migration, is the problem of migrating existing data from the old schema to the new. This often involves custom migration programs or specialized Domain Specific Langiuages. The migration must be carefully coordinated with upgrading the application code and associated artifacts that assume the new schema. Schema change on database servers is often delegated to Database Administrators and DevOps. Live programming~\cite{tanimoto90,hancock03} and local-first software~\cite{localfirst} both move data away from the server, eliminating those jobs, but not eliminating the need to do schema change, and indeed increasing the need to do it automatically. Recently \citet{Cambria} spotlighted these problems and proposed an approach using lenses~\cite{Foster2007}, but otherwise there has been surprisingly little research. To promote further progress we offer a set of challenge problems to the live programming and local-first communities, inviting them to propose and compare solutions.

Live programming seeks to erase the boundary between editing and running programs. In order to do so program data must be kept around while the program is being edited. Classic Lisp and Smalltalk systems integrated code and data into a persistent \textit{image}~\cite{Sandewall78, Goldberg80}. An interactive shell or \textit{Read Eval Print Loop (REPL)}~\cite{Deutsch64} is more transient than an image, but still builds up a context of data over long-lived programming sessions the loss of which disrupts the programmer's workflow. In all of these environments programs eventually get changed to create and expect data in a form incompatible with extant data. This happens whether or not the form of the data is specified explicitly in a type system or database schema, or whether it is left implicit in the behavior of the code. Some languages can tolerate a larger set of such changes, most notably Smalltalk which will automatically insert and delete members in existing instances when a class definition is changed~\cite[pp.252-272]{Goldberg80}.\footnote{Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a sophisticated schema change API~\cite{Gemstone}. Schema change won't go away.} But there is still a large class of changes that create incompatibilities with existing data, whether it is inside an image, programming environment, REPL, or an external database. Some live programming environments generate data with unit tests, but that only shifts the problem of schema change to adapting those tests. One way or another, stopping everything to manually deal with schema change contradicts the goals of live programming. Schema change won't go away.

Local-first software faces related problems. Its goal is to empower users by moving code and data from the cloud to the user's own devices. Distributed programming techniques like Convergent Replicated Data Types (CRDTs)~\cite{Shapiro11} are used to coordinate data changes peer-to-peer. Unfortunately these techniques so far have not addressed schema change nor code deployment The traditional techniques of schema change used in centrally managed databases are complicated by the distributed and intermittently connected nature of local-first data. Schema change won't go away.

In the following sections we present a series of challenge problems dealing with schema change in the context of live programming and local-first software. These problems are necessarily expressed using established idioms or conventions, but nevertheless we welcome solutions that translate the spirit of the problem into other contexts.

~

~

~

TODO: Need to explain that we take the programming system view - program is not something
built centrally, but something that exists and can be modified

\newpage
\section{Dimensions of change}

\begin{figure}[t]
\vspace{-1em}
\centering
\includegraphics[width=0.5\textwidth]{figures/layers.png}
\caption{Pace layers from Stewart Brand's classic \emph{How Buildings Learn} \cite{Brand95}.
  The outer layers rarely change, while the inner layers change frequently. The geographic
  site of a building is eternal, its structure remains stable for decades, space plan
  needs to change every couple of years. ``Because of the different rates of change of its
  components, a building is always tearing itself apart.''}
\label{fig:layers}
\end{figure}

In the broadest sense, this paper is concerned with the difficulties posed by a change
in a program. The first dimension of difficulties that we consider in this paper is
local. When something in a program changes, other parts of the program typically
need to be brought to sync with the changed part. The complexity of the problem varies.
If nothing else in the program depends on the changed part, for example when changing
a runtime value of an object field, the change poses no difficulty.
If many other parts of the program depend on the changed part, the difficulty is greater.
For example, if we change the schema of data structures used in the program, we also need to
change code that works with it and existing data that the users created. To structure our
discussion about this part of the problem, we will use the idea of pace layers introduced by
Brand \cite{Brand95,Brand18} (Figure~\ref{fig:layers}).

The second dimension that we consider in this paper is non-local. A program may have multiple
variants that exist independently and need to be synchronized. The difficulty of the problem
depends on what kind of synchronization we want to support. In the easier case, all variants
of the program converge, i.e.~all variants should eventually adopt all changes. In the more
challenging case, variants of a program may diverge, i.e.~a user may want to apply only certain
changes from other users. Note the non-local dimension may interact with the pace layers of the
local dimension. A typical case may be that the code and schema of a program should converge,
while data and the transient state of a program may diverge.

\paragraph{Dimensions.}
The framework for talking about schema change used in this paper thus distinguishes between
two dimensions:

\begin{itemize}
  \item The \emph{local dimension} is concerned with difficulties that a change causes
    across different program pace layers that exist within a single program.
  \item The \emph{non-local dimension} is concerned with the difficulties that a change causes
    across different program variants that exist independently.
\end{itemize}

\paragraph{Program pace layers.} As with buildings (Figure~\ref{fig:layers}), programs consist
of multiple pace layers that change at a different rate. As with buildings, making a change
to a more permanent layer also typically affects all the less permanent layers. Making a change
at a more permanent layer is thus more challenging. Alas, changing more permanent structures in
programming is frequently necessary. In this paper, we distinguish between four layers.

\begin{itemize}
\item \emph{State} -- transient state of a program. This is typically never shared across
  multiple variants of a program and it can often be discarded when a change occurs.
\item \emph{Data} -- data collected by the program. This represents a more permanent structure
  that is typically stored in a database and outlives program restarts.
\item \emph{Code} -- program logic. This layer changes whenever any aspect of the program logic
  is modified. A change typically affects transient state, but may not affect data.
\item \emph{Schema} -- the structures of program data. A change at this layer typically
  requires adapting code to work with the new structure and also updating data and state.
\end{itemize}

\paragraph{Program variants.} We are interested in programs that are distributed across multiple
environments as, for example, in local-first software. As discussed earlier, one aspect of this
dimension is whether all variants of a program eventually converge or whether it is possible to
maintain a diverging version. It is also useful to distinguish the directions in which a
change may flow, i.e., can a change originate in any variant or is there a single centralized
source of change. Note that each of this characteristics may  be associated with individual pace
layers of a program.

\begin{itemize}
\item \emph{Convergence} vs. \emph{Divergence}. In the convergence model, all program variants
  eventually adopt all changes. It may not be possible to adopt a particular change before adopting
  other changes it depends on. In the divergence model, a user may choose only particular changes
  they want to adopt.
\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (at a
  specific layer) can only originate from a particular source. In the decentralized model,
  changes can be done on any of the multiple co-existing variants of a program.
\end{itemize}

\paragraph{Challenge problems.}
The framework outlined above offers a wide range of configurations and one could conceive of
a challenge problem for many of those configurations. We can start with a change at each of the
four layers, consider each of the four configurations with respect to program variants (and also
consider different behaviour at all the affected less permanent layers). The aim of this paper
is, however, not to enumerate all options. Instead, our goal is to present a number of typical
and often encountered challenge problems. We use the above framework to more precisely define
the nature of the challenge problem. We also often use the above framework to discuss variations
on the particular challenge.

\newpage

\section{Extract Entity}

\subsection{Context}
In this challenge we consider a typical example of how the design of a database evolves as new needs are discovered. Much has been written on the subject of data modeling methodologies and the theory of relational database normalization~\cite{Molina08}. But that work often focuses on determining the correct data model up front before the system goes live, which would indeed be the optimal solution if only we were omniscient. The Extract Entity challenge confronts the practical reality of evolving a relational data model in flight.

\subsection{Example}
The Acme Corporation needs to record orders from various customers for various products. The simplest implementation is a spreadsheet with a row for each order and columns for information about the customer and product. It might look like this:

\begin{table}[h!]
\centering
  \begin{tabular}{ |l|c|c|l|l|}
   \hline
   Item & Quantity & Ship Date & Customer name & Customer address \\
   \hline \hline
   Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
   \hline
   Dynamite & 2 & & Daffy Duck & White Rock Lake \\
   \hline
   Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
   \hline
  \end{tabular}
\caption{Orders}
\end{table}

The shipping department filters this table on blank ship dates to see what they need to ship. But after a while the orders department realizes they are wasting effort duplicating the address for new order from an old customer. And when the customer's address changes they have to go back and edit all of their orders. What is needed is two tables, one with orders that links to one with customers, like this:

\begin{table}[h!]
  \centering
    \begin{tabular}{ |l|l|}
      \hline
      Name & Address \\
      \hline \hline
      Wile E Coyote & 123 Desert Station \\
      \hline
      Daffy Duck & White Rock Lake \\
      \hline
    \end{tabular}
  \caption{Customers}
\end{table}

\begin{table}[h!]
  \centering
    \begin{tabular}{ |l|c|c|l|}
     \hline
     Item & Quantity & Ship Date & Customer name \\
     \hline \hline
     Anvil & 1 & 2/3/23 & Wile E Coyote \\
     \hline
     Dynamite & 2 & & Daffy Duck  \\
     \hline
     Bird Seed & 1 & & Wile E Coyote \\
     \hline
    \end{tabular}
  \caption{Orders linking to Customers}
\end{table}

We often encounter this sort of design change when we realize that attributes of one type of entity should actually belong to a new type of entity that will be associated with the first one. Often the motivation is to centralize changes to those attributes in one place. We call this operation \textit{Extract Entity}.

In a SQL database, \texttt{Customers.Name} would be a primary key and \texttt{Orders.Customer} would be a foreign key. But before \texttt{Customers.Name} can be made a primary key the two occurrences of \texttt{Wile E Coyote} must be deduplicated one way or another~\cite{dedupe}. Given the dominance of SQL databases it is not surprising that there are many tools for performing schema change on them~\cite{RailsMigrations, liquibase}. We do find it surprising how low-level these tools are. There is support for the basic SQL DDL commands like adding and dropping tables, columns, or constraints, but no higher level operations that would help solve our problem. One is left to write custom SQL, which can become difficult if the migration needs to be performed online without bring down the database. There is a rich theory of relational database \textit{normalization} based on \textit{functional dependencies}~\cite{Molina08} but that theory has not been implemented into schema change operations in databases -- all you get is SQL. \citet{ambler06} define higher-level patterns of schema change like \textit{Split Table} that sketch SQL implementations but don't provide everything required for Extract Entity. These difficulties fairly beg for new linguistic abstractions, though programming patterns and APIs are reasonable pragmatic solutions to our challenge.

The Extract Entity schema change is straightforward in a spreadsheet. The \texttt{Customer Name} and \texttt{Customer Address} columns are copied and pasted into a new Customers table. The \textit{Remove Duplicates} command is used to deduplicate the customers. To preserve the original view for the shipping department the \texttt{Customer Address} column is redefined with a \texttt{VLOOKUP} formula to pull the address out of the customer table based on the customer's name. A spreadsheet won't maintain referential integrity between the two tables, but end-user database tools can do that~\cite{airtable, notion}.

\subsection{Challenge}
The Extract Entity challenge for live programming is to provide interactive operations in the programming environment that will perform the schema change on live code and data. The data can be in any form desired, including relations, objects, or JSON. The code can be a simple CRUD UI that also provides the shipping department its view of unshipped orders. The central problem is to coordinate the schema change with code edits/refactorings to keep the system executing live and correctly. Solutions should try to minimize:

\begin{enumerate}
  \item The number and complexity of commands or UI affordances that must be introduced.
  \item Interruption to live execution of the code.
\end{enumerate}

The Extract Entity challenge for local-first software is similar to the live programming challenge except that the context is an application that offers similar functionality with peer-to-peer collaboration. In this context the schema change does not need to be performed interactively — a developer can take some time to build, test, and package a solution, perhaps using an API or DSL. The hard part comes when the schema change is to be deployed across all the replicas. This deployment must synchronize code and data upgrades (or decouple them as in Cambria~\cite{Cambria}). The deployment must also deal with migrating ``in flight'' operations so that all replicas converge on the same state without data loss, and ideally without centralized coordination. A radical approach could try to incorporate schema change operations into the underlying datastore/CRDT itself and integrate code as well, but layered solutions are welcome too.

\subsection{Goals}
Solutions should try to minimize:

\begin{enumerate}
  \item The need for users to manually intervene.
  \item Possibilities that developer-written code is subtly incorrect in edge cases, preferably by reducing the need for developer-written code.
  \item The need for a centralized coordinator.
  \item How long replicas might be partitioned until resynchronization is acquired.
\end{enumerate}

\section{Divergence Control}

\subsection{Context}
Schema change is usually concerned with migrating old data into a new schema. But sometimes we also want to migrate in the reverse direction: new data into an old schema. \citet{Cambria} discuss various ways that code changes can require bidirectional migration. Instead in this problem we focus on changes made by end-users. Users frequently make copies of structured documents like spreadsheets and then diverge them by altering both data content and schema. In the case of spreadsheets schema changes include rearrangements to the structure of rows and columns as well as changes to formulas. The problem is that the users want to transfer such data and schema changes between these divergent copies, and they want to pick and choose which of these changes to transfer. In practice such transfer is done manually through copy \& paste. \citet{Basman19} has documented an ecology of emailed spreadsheets. \citet{Burnett14} distilled field observations of these practices in the story of Frieda, which we quote here:

\begin{quotation}

For example, consider ``Frieda'', an office manager in charge of her department's budget tracking. (Frieda was a participant in a set of interviews with spreadsheet users that the first author conducted. Frieda is not her real name.) Every year, the company she works for produces an updated budget tracking spreadsheet with the newest reporting requirements embedded in its structure and formulas. But this spreadsheet is not a perfect fit to the kinds of projects and sub-budgets she manages, so every year Frieda needs to change it. She does this by working with four variants of the spreadsheet at once: the one the company sent out last year (we will call it Official-lastYear), the one she derived from that one to fit her department's needs (Dept-lastYear), the one the company sent out this year (Official-thisYear), and the one she is trying to put together for this year (Dept-thisYear).

Using these four variants, Frieda exploratively mixes reverse engineering, reuse, programming, testing, and debugging, mostly by trial-and-error. She begins this process by reminding herself of ways she changed last year's by reverse engineering a few of the differences between Official-lastYear and Dept-lastYear. She then looks at the same portions of Official-thisYear to see if those same changes can easily be made, given her department's current needs.

She can reuse some of these same changes this year, but copying them into Dept-thisYear is troublesome, with some of the formulas automatically adjusting themselves to refer to Dept-lastYear. She patches these up (if she notices them), then tries out some new columns or sections of Dept-thisYear to reflect her new projects. She mixes in ``testing'' along the way by entering some of the budget values for this year and eyeballing the values that come out, then debugs if she notices something amiss. At some point, she moves on to another set of related columns, repeating the cycle for these. Frieda has learned over the years to save some of her spreadsheet variants along the way (using a different filename for each), because she might decide that the way she did some of her changes was a bad idea, and she wants to revert to try a different way she had started before.
\end{quotation}

\subsection{Example}
The Divergence Control challenge instantiates the problem of divergent state with the example from the Extract Entity challenge. Every month the orders department sends its spreadsheet to the accounting department, which adds the data to a version of the spreadsheet that it has customized for financial tracking. But when the orders department migrates its spreadsheet to extract out customers the accounting department does not want to conform. They could manually convert incoming data in the new schema into their variant of the old schema. But they shouldn't have to. It should be possible to run new data through the schema change in reverse, converting it back into the format the accounting department is used to ingesting. Note that this is not a matter of synchronizing the divergent spreadsheets -- they maintain differently evolved schema.

\subsection{Challenge}
What is needed is a user-friendly mechanism for transferring data changes from one schema to another bidirectionally. In one direction we want to transfer changes made in the new schema into a variant of the old schema. The opposite direction might be needed, for example, if the accounting department makes corrections to customer addresses, which ought to be pushed through into the orders department's new schema.

Bidirectional transfer of changes between divergent copies is similar in some ways to source code version control as in Git~\cite{ProGit}. There is \textit{forking} of long-lived divergent copies. We want to \textit{diff} these forks to see exactly how they have diverged. We want to partially \textit{merge} them by \textit{cherry picking} certain differences. Yet there are also many dissimilarities with Git: our data is more richly structured than lines of text; our schema changes are higher-level transformations on these structures than inserting and deleting characters; and we expect that end-users be able to understand it~\cite{gitless}. The Divergence Control challenge is in a sense to provide ``version control for schema change'' meeting these criteria, with the key technical challenge being the ability to transfer selected differences bidirectionally through schema changes as independently as possible.

\subsection{Goals}
This challenge invites a change of perspective in both live programming and local-first software. Live programming must move from being a solitary activity to a collaborative workflow, and one where the collaboration is not just on editing source code but on live integrated code and data, as in the Smalltalk/Lisp images of old.\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!} For its part, local-first software must move from automatically converging data replicas to also interactively managing long-term divergence. That implies either application state is being directly exposed to the user, or the application code is using an API to present version control affordances. We encourage both communities to think outside the box of Git, which has proven to baffle not only end-users but also a substantial fraction of developers.

\section{Conference Organizer: Merging structural document edits}

\subsection{Context}
In this challenge, we consider a document-based format akin to that used, for example, by Notion. We focus on schema change applied to data, but the document-based format can also be extended to include programming as done, for example in Boxer. We assume that the document can be concurrently edited by multiple users in the same fashion as in local-first software, i.e., users may edit a local copy that then needs to be synchronized with changes made by other users. The edits to the document can be done in a rich-text editor, or by using a more structure-aware structure editor that can, for example, record changes as a sequence of high-level edit operations.

\subsection{Example}
In this challenge, we consider the problem of collaborative conference planning. A number of co-organizers are planning a conference that will feature talks by several invited speakers. The organizers need to agree on a list of speakers to invite, coordinate who contacts whom and track the acceptance of invitations. They start with a document shown in Figure~\ref{fig:conf-orig}.

\begin{figure}[h!]
\fbox{\parbox{24em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}
\begin{itemize}
 \item Adele Goldberg, adele@xerox.com
 \item Margaret Hamilton, hamilton@mit.com
 \item Betty Jean Jennings, betty@rand.com
\end{itemize}
}}
\caption{Conference organization. Initial state of the document.}
\label{fig:conf-orig}
\end{figure}

\begin{figure}
\fbox{\parbox{24em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}
\begin{itemize}
 \item Ada Lovelace, lovelace@rsoc.ac.uk
 \item Adele Goldberg, adele@xerox.com
 \item Betty Jean Jennings, betty@rand.com
 \item Margaret Hamilton, hamilton@mit.com
\end{itemize}
}}
\caption{Conference organization. Added a speaker and sorted list.}
\label{fig:conf-add}
\end{figure}

\begin{figure}
\fbox{\parbox{24em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}

\begin{tabular}{| l | l | l |}
  \hline
  \textbf{Name} & \textbf{Email} & \textbf{Organizer} \\
  \hline \hline
  Adele Goldberg & adele@xerox.com & TP\\
  \hline
  Margaret Hamilton & hamilton@mit.com & JE\\
  \hline
  Betty Jean Jennings & betty@rand.com & JE\\
  \hline
\end{tabular}
}}\caption{Conference organization. Refactored list into a table.}
\label{fig:conf-tab}
\end{figure}

\begin{figure}
\fbox{\parbox{24em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}

\begin{tabular}{| l | l | l |}
  \hline
  \textbf{Name} & \textbf{Email} & \textbf{Organizer} \\
  \hline \hline
  Ada Lovelace & lovelace@rsoc.ac.uk & \\
  \hline
  Adele Goldberg & adele@xerox.com & TP\\
  \hline
  Betty Jean Jennings & betty@rand.com & JE\\
  \hline
  Margaret Hamilton & hamilton@mit.com & JE\\
  \hline
\end{tabular}
}}\caption{Conference organization. Resulting document after change merging.}
\label{fig:conf-fin}
\end{figure}

\subsection{Challenge}
The challenge is to merge document edits that are done locally and independently by the co-organizers of the conference. Specifically, consider the following two edits:

\begin{enumerate}
\item The first organizer adds an additional speaker to the list and sorts the list of speakers alphabetically by their first name. (Figure~\ref{fig:conf-add})
\item The second organizer refactors the list into a table. They split the single textual value into a name and an email (using a comma as the separator) and add an additional column for tracking which of the organizers should contact the speaker.
\end{enumerate}

The system should be able to merge the changes and produce a final table shown in Figure~\ref{fig:conf-fin}. The resulting table needs to include the additional speaker created by the first organizer, use the order specified by the first organizer and use the format defined by the second organizer. The newly added speaker should be reformatted into the new format. For the newly added ``Organizer'' column, the second user may specify default value or the newly added row may use an empty value.

\subsection{Goals}
A solution to the challenge should strive to satisfy the following goals:

\begin{itemize}
\item \emph{Completeness.} It should be possible to semi-automatically merge any two sequences of edits. In other words, the merging should always be defined, regardless of what edits the users perform.
\item \emph{Commutativity.} The order of edits should not matter. If users $A$ and $B$ perform edits independently starting from the same initial document, the resulting document should be the same if the system treats edits from $A$ as occurring before the edits of $B$ and vice versa.
\item \emph{Conflict resolution.} It may not always be possible to merge conflicting edits. In this case, the system should interactively ask the user for guidance or, possibly, use default behavior specified by the user.
\item \emph{Convergence.} The system does not need to consider the case where one users does not want to adopt changes created by another user. We assume that all users want to eventually see the same final document produced by applying all edits.
\item \emph{Structure editing.} The system may use a structure editor that provides high-level commands for operations such as reordering of elements or refactoring of a list into a table. In other words, we assume that the challenge may not be solvable in a system based on plain text editing of document source code.
\end{itemize}

\subsection{Remarks}
It is worth noting that this challenge does not discuss explicit distinction between the schema (or ``type'') of the document and data (or ``values''). The implementing system may or may not maintain such distinction. In our example, adding a new speaker and sorting the list is arguably a mere change of the data, but the refactoring of a list into a table is a schema change. The system may leverage knowledge about such distinction and, for example, handle the merging of schema-related and data-related changes differently.

\section{Conference Organizer: Merging edits in the presence of code}

\subsection{Context}
Consider the document-based format as in the previous challenge, but extended with the support for computed values. A computed value is a node in the document whose value is specified by a formula akin to those used in spreadsheets. The equations can refer to other nodes in the document, possibly using either an absolute or relative path and may involve a range of built-in functions, for example to count the number of elements or to sum a sequence of numerical values. We now also assume that elements in the document can have IDs that are not displayed in the default view, but may be used to refer to them in equations.

\begin{figure}
\fbox{\parbox{28em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}
\begin{itemize}
 \item Adele Goldberg, adele@xerox.com
 \item Margaret Hamilton, hamilton@mit.com
 \item Betty Jean Jennings, betty@rand.com
\end{itemize}
\vspace{0.5em}
\textbf{Conference budget}\\
Travel cost per speaker:\\
\hspace*{2em} \$1200 \\
Number of speakers: \\
\hspace*{2em} \texttt{=COUNT(/ul[id='speakers']/li)} \\
Travel expenses:\\
\hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
}}
\caption{Conference organization. Initial state of the document.}
\label{fig:conf-budget}
\end{figure}

\begin{figure}
\fbox{\parbox{28em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}

\begin{tabular}{| l | l | l |}
  \hline
  \textbf{Name} & \textbf{Email} & \textbf{Organizer} \\
  \hline \hline
  Ada Lovelace & lovelace@rsoc.ac.uk & \\
  \hline
  Adele Goldberg & adele@xerox.com & TP\\
  \hline
  Betty Jean Jennings & betty@rand.com & JE\\
  \hline
  Margaret Hamilton & hamilton@mit.com & JE\\
  \hline
\end{tabular}

\vspace{0.5em}
\textbf{Conference budget}\\
Travel cost per speaker:\\
\hspace*{2em} \$1200 \\
Number of speakers: \\
\hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
Travel expenses:\\
\hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
}}\caption{Conference organization. Resulting document after change merging.}
\label{fig:conf-finfin}
\end{figure}

\subsection{Example}
This challenge extends the previous problem. The conference organizers also want to use the document to manage the conference budget. In a very simplified form, one aspect of this is calculating the estimated travel expenses for all the speakers. For this, they will want to add a computed value to the document that counts the number of speakers and multiplies the result by a fixed cost per speaker elsewhere in the document.

The example is shown in Figure~\ref{fig:conf-budget}. The additional section of the document contains one fixed value (travel cost per speaker) and two computed values. The first selects all \texttt{li} elements of a \texttt{ul} element with ID \texttt{speakers} and counts their number. The second computed value then multiplies the constant travel cost per speaker by the number of speakers. Here, we assume the new section uses the HTML definition list \texttt{dl} and the equation selects its first and second \texttt{dd} item, respectively.

\subsection{Challenge}
The challenge is, again, to merge the edits done independently by the co-organizers. Assume two co-organizers start with the same original document shown in Figure~\ref{fig:conf-orig}.
\begin{enumerate}
    \item The first co-organizer performs the edits outlined previously that result in the document shown in Figure~\ref{fig:conf-fin}, i.e., edits the speakers and refactors the structure to use a table.
    \item The second co-organizer adds the budget calculation as shown in Figure~\ref{fig:conf-budget}.
\end{enumerate}
As before, the system should be able to merge the edits made to the document. Crucially, this requires that the refactoring of the document structure done in (1), is also reflected in the code of the equations added in (2). As shown in Figure~\ref{fig:conf-finfin}, in this specific case, the system changes the equation \texttt{COUNT(/ul[id='speakers']/li)} to \texttt{COUNT(/table[id='speakers']/tbody/tr)}. This is the case because the refactoring consists of four steps that each transform the document and also need to update the code of the equation accordingly:
\begin{itemize}
    \item[--] Change the type of the \texttt{ul} element to \texttt{table}
    \item[--] Wrap the body of the \texttt{table} element with \texttt{tbody}
    \item[--] Change the type of all children of \texttt{tbody} from \texttt{li} to \texttt{tr}
    \item[--] Further split the body into two \texttt{td} elements
\end{itemize}

\subsection{Goals}
A solution to the challenge should meet the goals of the preceding challenge, i.e., completeness (merging should be always defined), commutativity (order does not matter), conflict resolution (ask for guidance if needed), convergence (everyone eventually wants the same document) and structure editing (the system may capture high-level edits).

\subsection{Extensions}
The basic challenge outlined above focuses primarily on the local-first software scenario, but it can be extended to also apply to the live programming scenario. In particular, when the document contains computed values, those may be evaluated either explicitly by the user or automatically on-the-fly. An edit can then invalidate some of the values (either by changing the data that a computed value depends on, or by changing the equation used to compute the value). A live programming system should be able to detect which computed values are affected by an edit and, either invalidate those (requiring an explicit re-evaluation by the user) or automatically re-evaluate them (and possibly highlight the affected values to inform the user).

\section{Live Modeling Languages Require Run-Time State Migration}
%Related work
%``Slogan'': editing a program is diverging from run-time schema.

\subsection{Context}
In stateful live programming, a change in the program can make the run-time state of the execution out of date, hence we want to migrate the run-time state to keep on running.

One can see a program itself as an instance of a static schema (its AST type), that in turn determines (defines/implies/induces) a run-time schema: the run-time structures of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as the structure of the run-time state. Whereas the run-time \textit{code} structures do not typically change while using an application, the run-time \textit{state} constantly changes.
Live programming, however, requires reconciling changes to the program with the run-time structures of both the code and the state. This typically means that at a certain point (a quiescent point) during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.

\subsection{Example}
Consider the example of a simple state machine language with on-entry actions, and typed global variables (loosely inspired by the SML language of \citet{vanRozen19}). An example state machine and an excerpt of its abstract syntax schema is shown in Figure~\ref{LST:statemachines}. Depicted on the left, an actual statemachine modeling the opening and closing of a door, with one global boolean variable, \lstinline{isClosed}, which is flipped according to transitioning between the \lstinline{closed} and \lstinline{opened} states. For the sake of brevity, the abstract syntax schema (similar, to e.g., and Ecore metamodel~\cite{EMF}) shown on the right of the figure omits classes for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).

\begin{figure}[t]
\centering
\begin{minipage}[t]{0.4\textwidth}
\begin{lstlisting}[language=java,morekeywords={machine,on,state,var}]
machine Doors

var isClosed: bool = true

state closed
  isClosed := true
  on open => opened

state opened
  isClosed := false
  on close => closed
\end{lstlisting}
\end{minipage}
\hspace*{2pt}\vline
\begin{minipage}[t]{0.4\textwidth}
\begin{lstlisting}[language=java,morekeywords={on}]
  class Machine
    name: string
    states: State*
    vars: Decl*

  class State
    name: string
    onEntry: Stmt*
    transitions: Trans*

  class Trans
    event: string
    target: State

  class Decl
    name: string
    type: Type
  \end{lstlisting}
\end{minipage}
\caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
\label{LST:statemachines}
\end{figure}

In the following we assume that the state machine is executed by an interpreter that process an incoming stream of events, and traversing run-time pointers representing transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program. The structure that the interpreter operates on can be described as derived  schema from the abstract syntax scheme shown on the right of Figure~\ref{LST:statemachines}, but specialized for a particular machine, e.g., \lstinline{Doors}.
This schema will have additional fields and associations modeling the run-time data required for execution. For the \lstinline{Doors} statemachine, this means adding the current state to the \lstinline{Machine} type (required for any machine) and the machine specific data fields, namely \lstinline{isClosed}. The run-time schema for \lstinline{Doors} is then the following\footnote{It is possible to represent the global variables at run-time using an untyped table or hashmap, but for the sake of schema migration complexity, the encoding as fields in classes is more instructive.}:
\begin{quote}
\begin{lstlisting}[language=java,morekeywords={on},mathescape=true]
class Machine$^{++}$
  states: State*
  current: State
  isClosed: bool
\end{lstlisting}
\end{quote}
This \textit{extended} class $\text{\lstinline{Machine}}^{++}$ also represents state machines, but this time at run time.

\begin{figure}[t]
  \centering
\includegraphics[width=0.5\textwidth]{figures/doorsmachine.pdf}
\caption{UML object diagram of the run-time structures of the Doors statemachine (excluding code objects)}
\label{FIG:doorsRuntime}
\end{figure}

An instance of a running state machine consists of an object graph conforming to the run-time schema of state machines, updating the \lstinline{current} state pointer in response to events. This object graph is graphically depicted in Figure~\ref{FIG:doorsRuntime} as a UML object diagram, excluding the statement objects representing on-entry actions. As the diagram shows, the running machine has a current state (\lstinline{closed}) and the \lstinline{isClosed} field has an actual value (\lstinline{true}). It is instructive to note that the diagram encode both static and dynamic aspects of the state machine language. Live editing the state machine then requires to \textit{patch}~\cite{SemanticDeltas} this run-time structure without shutting it down, possibly requiring migration of run-time state.

\subsection{Challenge}

\begin{table}[t]
  \centering
\begin{tabular}{lll}\toprule
\textsc{Change} & \textsc{What} & \textsc{How}\\\midrule
add & state & simple  \\
remove & state & structurally simple, but heuristic if state is current \\
rename & state & simple\\
add & variable & migrate class and initialize field \\
remove & variable & migrate class (NB: assumes variable is unused)\\
rename & variable & rename in class, preserving value\\
type change & variable & migrate class and convert or reinitialize value \\
add & transition & simple \\
remove & transition & structurally simple, but heuristic for pending events\\
event change & transition & structurally simple, but heuristic for pending events\\
add/remove & statement & hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\caption{Possible program changes and how to deal with them at run time}
\label{TBL:changes}
\end{table}

The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{TBL:changes}.
The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:
\begin{itemize}
\item all rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care needs is required if the subject of removal is the current state. In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
\item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
\item removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale. Strategies to deal with this situation include: simply dropping the events, or requiring that such edits cannot be patched at run time when there are pending events.
\item the type change edit of variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion. For instance, if the type change is from boolean to integer, then true could become 1, and false could become 0.
\item note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
\end{itemize}

\subsection{Goals}

The goals for this challenge are twofold: from the end-user perspective and from the language engineering perspective. A particular challenge from the end-user perspective is to ``do minimal harm'': it is essential for fluid programming experience that the automatically triggered migrations are in a sense as near as possible to the previous application state, to not surprise or confuse the user/developer.
One approach  considered in earlier work~\cite{RuntimeConstraint} employs the constraint solver Z3 to find a ``nearest'' run-time instance compatible with a source change. Nevertheless, the patching of the run-time state should be quick enough so as to not disrupt the programmer experience.

From the language engineering perspective the goal is to employ techniques, formalisms, and tools, to make the construction of such languages easier. The above example state machine DSL is derived from earlier work~\cite{vanRozen19}, where the authors manually implemented the run-time patch operation and concluded that even for such a simple language (simpler even than the example above) it is a complex and error-pone endeavor. Furthermore, the field of software language engineering studies and develops generic and reusable techniques to improve the development of DSLs and programming languages, for instance, in the context of language workbenches~\cite{ERDWEG201524}. The development of live programming languages, however, is currently still out of reach for all existing language workbenches. The aforementioned approach using a constraint solver is an example of such a \textit{language parametric} technique, in that it operates on the (extended) abstract syntax metamodel of a language, and does not assume anything further about the language itself. Another approach is the \textsc{Cascade} metamodeling formalism which has builtin support for run-time patching~\cite{Cascade}. However, further research is needed to design better principled language engineering approaches that solve the problem in a way that is both declarative and fast.

\subsection{Extensions}

While the above example is arguably simple, the problem becomes much more challenging when the programs themselves define data types, classes, records, structures etc. Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program. Typically a refactoring should be behavior preserving, but can it also preserve run-time data? The minimal example is the consistent variable rename rename in Table~\ref{TBL:changes}, which should not have any effect on run-time state.  A more complex refactoring is described in Section~\ref{SECT:elm}.

\section{Live programming in the context of Elm architecture}
\label{SECT:elm}
\subsection{Context}
Another challenge involving stateful live programming can be drawn from the programming model based on the Elm architecture. In this model, a reactive (web) application is structured in terms of current state and events that affect the state. The implementation then consists of two functions that we may call \texttt{update} and \texttt{render}:

\begin{quote}
\begin{lstlisting}[language=ml,morekeywords={on}]
type State = { .. }
type Event = .. | ..

val update : State -> Event -> State
val render : State -> Html
\end{lstlisting}
\end{quote}

The programming model works as follows:
\begin{itemize}
    \item \texttt{State} represents the entire application state, i.e., everything that the user can work with.
    \item \texttt{Event} represents all events that the user can trigger by interacting with the application.
    \item \texttt{update} is called whenever an event happens. It takes the current state, the event and computes a new state.
    \item \texttt{render} takes the current state and produces a representation of what should be displayed on the screen (e.g., an HTML tree).
\end{itemize}

Traditionally, when Elm applications are developed, they are restarted each time the code is modified and any previous state is discarded. However, a more effective programming model based on live programming would allow live updates to both the code of the two functions and the structure of the two types.

\subsection{Example}

As an example, consider the uninspiring, but well-known, TODO list application. The types that capture the state and events in the application may look as follows:

\begin{quote}
\begin{lstlisting}[language=ml,morekeywords={on}]
type Item = { id : id; title : string; completed : bool }
type State = { items : Item list }
type Event =
  | SetCompleted of id * bool
  | SetTitle of id * string
  | Remove of id
  | Add of string
\end{lstlisting}
\end{quote}

The application state consists of a list of items. Each item has a unique ID alongside with a title and a flag indicating whether it is completed. The events represent edits to the items, deletion and addition. The implementation of the \texttt{update} and \texttt{render} functions is simple and not important for the challenge.

\subsection{Challenge}
Now, imagine that we have a running TODO list application with the above state and events. To test the application, the programmer has already created a number of items and so there is a single value of the \texttt{State} type that represents a current state of the application such as:

\begin{quote}
\begin{lstlisting}[language=ml,morekeywords={on}]
{ items : [
   [ { id = 1; title = "check twitter"; completed = true  }
     { id = 1; title = "Write paper"; completed = false } ] }
\end{lstlisting}
\end{quote}

As above, there are a number of edits to the code and types that the programmer may want to do without restarting the application. Those are summarized in Table~\ref{TBL:elmchanges}. Modifying the code is simple and only requires waiting until the current execution completes. Modifying the \texttt{Event} type is also simple, but it may lead to unused code or missing case in \texttt{update} that needs to be addressed. Finally, modifying \texttt{State} ranges from relatively simple problems (adding a new field) to challenging case when the structure of \texttt{State} is changed.

For example, imagine that the programmer would want to edit the structure of the state and migrate the previous definition of \texttt{State} to the following new definition where individual fields are stored in separate lists:

\begin{quote}
\begin{lstlisting}[language=ml,morekeywords={on}]
type State =
  { ids : id list
    titles string list
    completes : bool list }
\end{lstlisting}
\end{quote}

This representation is semantically equivalent (assuming the lists have the same length) to the original one. It should thus, in principle, be possible to migrate the original state value to a value using the new structure. Moreover, it should, in principle, be also possible to automatically transform the implementations of \texttt{update} and \texttt{render} to work as before, but using the new state structure.

\begin{table}[t]
  \centering
\begin{tabular}{lll}\toprule
\textsc{Change} & \textsc{What} & \textsc{How}\\\midrule
modify & \texttt{render} & simple, but wait until current execution finishes  \\
modify & \texttt{update} & simple, but wait until current execution finishes  \\
add & case to \texttt{Event} & requires adding corresponding case to \texttt{update} \\
remove & case from \texttt{Event} & remove unused code from \texttt{update} \\
add & field to \texttt{State} & migrate state value and initialize field \\
remove & field from \texttt{State} & migrate state (assuming field unused) \\
modify & structure of \texttt{State} & migrate state value and edit code accordingly \\
\bottomrule
\end{tabular}
\caption{Possible program changes and how to deal with them at run time}
\label{TBL:elmchanges}
\end{table}

\subsection{Goals}
As above, the key requirement from the live programming perspective is to “do minimal harm”. In this case, this results in the following goals:

\begin{itemize}
\item When migrating the application state, this needs to be done automatically and the system should strive to produce new state that is as near as possible to the previous state.
\item The solution may rely on structure editing so that the system has access to a high-level logical description of the edits performed by the user.
\item The system needs to handle the case when the type structure diverges from the code structure. This can be addressed in various ways (transform code, add error handlers, etc.) but it is desirable to avoid "breaking" the \texttt{render} function as this would make the new application state impossible to see.
\end{itemize}

\subsection{Remarks}
It is worth noting that the particularly difficult aspect of this challenge, i.e., the case where the structure of state is refactored to an equivalent one, is related to the Extract Entity challenge.

\begin{comment}
\section{Rename field; Add new required field}
Starter problems suggested by Geoffrey. These are already directly supported in SQL, so there is good support in SQL schema change tools, but they don't address how to synchronize the change with live code refactoring or local-first code deployment.
\end{comment}

\begin{comment}
\section{Cambria (Geoffrey/pvh)}
\citet{Cambria} exhumed and studied the old problem of schema change, primarily in the context of local-first software. In this paper we add more scenarios and the complementary problems of live programming. The primary goal of local-first software is user autonomy, which could include letting users stay on older versions of application code while continuing to collaborate with newer versions. This problem is a nice counterpart to that of Divergence Control. They both require a kind of bidirectional schema change, with Cambria building upon research on bidirectional programming with lenses, while Divergence Control is more closely connected to version control techniques. Appendix III of \cite{Cambria} discusses the tricky problem of converting between scalars and arrays (which in data modeling is called a multiplicity change). That could be the challenge problem right there.

The Cambria paper also presents motivating examples of schema change problems in Stripe, Kafka, and Mastodon. Perhaps one of those examples could be crafted into a challenge problem.
\end{comment}

\begin{comment}
\section{Denormalization (Matt Tantaman)}
Matt points out Extract Entity is a normalization so we ought to complement it with a denormalization. Note that copying/duplication is a technical problem in some formalisms like lenses. Denormalization is often done for performance reasons but that isn't an issue in our small-scale scenarios. What is a small-scale use case for denormalization?

There are good cases for denormalizing from 1NF, i.e. nesting structures. Local-first software seems focused on JSON data rather than relational. An extension of Extract Entity is \textit{Nest Entity} where a table of orders are nested inside each customer.
\end{comment}

\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

% Local Variables:
% TeX-engine: luatex
% End:

### 2024-04-29 paper.tex
Revising the conference organizer section

@@
 
+\usepackage{changepage}
 \usepackage{multicol}
@@
 \usepackage{natbib}
+\usepackage[many]{tcolorbox}
+
+\newcounter{challengen}
+\newcommand{\challenge}[1]{\subsection*{\addtocounter{challengen}{1} Challenge \#\thechallengen: #1}}
+
+\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
+\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
+  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
+  outer arc=0pt]\setlength{\parskip}{0.5em}\setlength{\parindent}{0em}\textbf{\sffamily Framework perspective.}\quad #2
+\end{tcolorbox}}
 
@@
 
-TODO: Need to explain that we take the programming system view - program is not something
-built centrally, but something that exists and can be modified
+\textcolor{red}{TODO: Need to explain that we take the programming system view - program is not something
+built centrally, but something that exists and can be modified}
+
+~
+
+\textcolor{red}{TODO: Maybe the right way to frame this is to say that the challenges are intended
+less for evaluating particular solutions but more for framing and making sense of different
+kinds of schema-chaange-related problems.  You can read this paper and when you face a schema-change-related
+problem, think about it in terms of our dimensions. Maybe compare it to our challenges.
+We do not necessarilly expect that people will directly solve our challenges though. (Although
+some might and that would be nice..)}
+
+~
+
+\textcolor{red}{TODO: I also guess this is more about just ``change'' than ``schema change''
+which is one particular (most challenging) instance - i.e., change at the most permanent
+pace layer.}
 
@@
 
-\section{Conference Organizer: Merging structural document edits}
-
-\subsection{Context}
-In this challenge, we consider a document-based format akin to that used, for example, by Notion.
-We focus on schema change applied to data, but the document-based format can also be extended to include programming as done, for example in Boxer.
-We assume that the document can be concurrently edited by multiple users in the same fashion as in local-first software, i.e., users may edit a local copy that then needs to be synchronized with changes made by other users.
-The edits to the document can be done in a rich-text editor, or by using a more structure-aware structure editor that can, for example, record changes as a sequence of high-level edit operations.
-
-\subsection{Example}
-In this challenge, we consider the problem of collaborative conference planning.
-A number of co-organizers are planning a conference that will feature talks by several invited speakers.
-The organizers need to agree on a list of speakers to invite, coordinate who contacts whom and track the acceptance of invitations.
-They start with a document shown in Figure~\ref{fig:conf-orig}.
-
-\begin{figure}[h!]
-\fbox{\parbox{24em}{
+\newpage
+
+\begin{figure}
+\begin{adjustwidth}{-5.5em}{-5.5em}
+\centering
+\begin{subfigure}[b]{23em}\vspace{0pt}
+   \fbox{\parbox{22.6em}{
    \textbf{PROGRAMMING CONFERENCE 2023}\\
@@
    \end{itemize}
+   \\~
    }}
-\caption{Conference organization. Initial state of the document.}
+   \caption{Initial state of the document.}
    \label{fig:conf-orig}
-\end{figure}
-
-\begin{figure}
-\fbox{\parbox{24em}{
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{23em}\vspace{0pt}
+   \fbox{\parbox{22.6em}{
    \textbf{PROGRAMMING CONFERENCE 2023}\\
@@
    \begin{itemize}
- \item Ada Lovelace, lovelace@rsoc.ac.uk
+    \item Ada Lovelace, ada@rsoc.ac.uk
     \item Adele Goldberg, adele@xerox.com
@@
    }}
-\caption{Conference organization. Added a speaker and sorted list.}
+   \caption{Added a speaker and sorted the list.}
    \label{fig:conf-add}
-\end{figure}
-
-\begin{figure}
-\fbox{\parbox{24em}{
+\end{subfigure}
+\\~\\
+\begin{subfigure}[b]{23em}\vspace{0pt}
+   \fbox{\parbox{22.6em}{
    \textbf{PROGRAMMING CONFERENCE 2023}\\
@@
      \hline
-  \textbf{Name} & \textbf{Email} & \textbf{Organizer} \\
+     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
      \hline \hline
@@
    \end{tabular}
-}}\caption{Conference organization. Refactored list into a table.}
+   \\~
+   }}
+   \caption{Refactored list into a table.}
    \label{fig:conf-tab}
-\end{figure}
-
-\begin{figure}
-\fbox{\parbox{24em}{
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{23em}\vspace{0pt}
+   \fbox{\parbox{22.6em}{
    \textbf{PROGRAMMING CONFERENCE 2023}\\
@@
      \hline
-  \textbf{Name} & \textbf{Email} & \textbf{Organizer} \\
+     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
      \hline \hline
-  Ada Lovelace & lovelace@rsoc.ac.uk & \\
+     Ada Lovelace & ada@rsoc.ac.uk & \\
      \hline
@@
    \end{tabular}
-}}\caption{Conference organization. Resulting document after change merging.}
+   }}
+   \caption{Resulting document after change merging.}
    \label{fig:conf-fin}
+\end{subfigure}
+\end{adjustwidth}
+\vspace{0.5em}
+\caption{Conference organizer. Initial version of the document with two independent changes
+and the final version of the document after the two changes are merged.}
+\label{fig:conf}
+\vspace{0.5em}
 \end{figure}
 
-\subsection{Challenge}
-The challenge is to merge document edits that are done locally and independently by the co-organizers of the conference. Specifically, consider the following two edits:
+\section{Conference Organizer: Merging edits in computational documents}
+In this section, we consider challenges related to working with a document-based format akin to HTML.
+We discuss two variants of the format. In the first format, a document
+contains only data. In the second variant, the format is extended with computed values.
+A computed value is a node in the document whose value is specified by a formula akin to those
+used in spreadsheets. The equations can refer to other nodes in the document, possibly using
+an absolute or a relative path and may involve a range of built-in functions, for example to
+count the number of elements or to sum a sequence of numerical values.
+
+We assume that the document can be
+concurrently edited by multiple users in the same fashion as in local-first software, i.e.,
+users may edit a local copy that then needs to be synchronized with changes made by other users.
+The edits to the document can be done in a rich-text editor, or by using a more structure-aware
+structure editor that can, for example, record changes as a sequence of high-level edit operations.
+
+The concrete challenges are centered around the problem of collaborative conference planning
+as illustrated in Figure~\ref{fig:conf}. A number of co-organizers are planning a conference
+that will feature talks by several invited speakers. They need to agree on the speaker list,
+contact speakers and calculate the conference budget.
+
+\frameworkbox{
+In this scenario, the only transient \emph{state} is the result of computed values. The content
+of the document corresponds to the \emph{data} pace layer. A change in the data layer may invalidate
+the computed value at the state layer. Formulas correspond to the \emph{code} layer and changing them,
+again, invalidates the state. The \emph{schema} pace layer is implicit. A change to the document that
+modifies content is a mere data change, but restructuring the document is an implicit schema
+change. Such schema change that may affect the code pace layer.
+
+The scenario assumes the \emph{decentralized} model. There are multiple program (document)
+variants and it should be possible to propagate a change at any pace layer to other program
+variants. The challenge may be solved using both the easier \emph{convergence} and the more
+challenging \emph{divergence} model.}
+
+\paragraph{Applications.}
+One of the earliest systems to combine documents with computational elements
+is Boxer \cite{diSessa86}. The challenges in this section apply to a wide range of
+systems that can be seen as its successors. The specific scenario discussed here is
+inspired by the use of Notion~\cite{Notion24}, which is a collaborative structured document
+editing system. Treating documents as end-user programs has recently
+been explored in Potluck~\cite{Litt2023}, which is local-only but includes formulas similar
+to those in our challenge.
+
+The challenges in this section are also relevant to work on computational scientific notebooks.
+Jupyter \cite{Kluyver2016} uses a simple representation as a sequence of cells and does not have
+a mechanism for embedding data in the document and referring to it, but numerous
+project such as Nextjournal \cite{Nextjournal21} advance the format. The idea of producing
+interactive, computational secientific documents is an active research area. Recent contributions
+include Nota~\cite{Crichton2021} and Living Papers \cite{Heer2023}.
+
+\challenge{Merging structured document edits}
+
+First, the organizers need to agree on a list of speakers to invite, coordinate who contacts
+whom and track the acceptance of invitations. They start with a document shown in
+Figure~\ref{fig:conf-orig}. The challenge is to merge document edits that are done locally
+and independently by two co-organizers. Specifically, consider the following two edits:
 
 \begin{enumerate}
-\item The first organizer adds an additional speaker to the list and sorts the list of speakers alphabetically by their first name. (Figure~\ref{fig:conf-add})
-\item The second organizer refactors the list into a table.
-They split the single textual value into a name and an email (using a comma as the separator) and add an additional column for tracking which of the organizers should contact the speaker.
+\item The first organizer adds an additional speaker to the list and sorts the list of speakers
+  alphabetically by their first name. (Figure~\ref{fig:conf-add})
+\item The second organizer refactors the list into a table. They split the single textual value
+  into a name and an email (using a comma as the separator) and add an additional column for
+  tracking which of the organizers should contact the speaker.
 \end{enumerate}
 
-The system should be able to merge the changes and produce a final table shown in Figure~\ref{fig:conf-fin}.
-The resulting table needs to include the additional speaker created by the first organizer, use the order specified by the first organizer and use the format defined by the second organizer.
-The newly added speaker should be reformatted into the new format.
-For the newly added ``Organizer'' column, the second user may specify default value or the newly added row may use an empty value.
-
-\subsection{Goals}
+\noindent
+The system should be able to merge the changes and produce a final table shown in
+Figure~\ref{fig:conf-fin}. The resulting table needs to include the additional speaker
+created by the first organizer, use the order specified by the first organizer and use
+the format defined by the second organizer. The newly added speaker should be reformatted
+into the new format. For the newly added ``Organizer'' column, the second user may
+specify default value or the newly added row may use an empty value.
+
+\frameworkbox{
+In this challenge, the change done by the first organizer is at the \emph{data} layer,
+while the change done by the second organizer spans the \emph{schema} and \emph{data} layers.
+In implementing the schema change, the second organizer transforms the affected document data.
+The challenge is to apply the same transformation to the data independently added by the
+second organizer.
+}
+
+\paragraph{Requirements.}
+The challenge is concerned with merging two sequences edits to a base document.
 A solution to the challenge should strive to satisfy the following goals:
@@
 \begin{itemize}
-\item \emph{Completeness.} It should be possible to semi-automatically merge any two sequences of edits.
-In other words, the merging should always be defined, regardless of what edits the users perform.
-\item \emph{Commutativity.} The order of edits should not matter.
-If users $A$ and $B$ perform edits independently starting from the same initial document, the resulting document should be the same if the system treats edits from $A$ as occurring before the edits of $B$ and vice versa.
+\item \emph{Completeness.} It should be possible to semi-automatically merge any two sequences
+  of edits. In other words, the merging should always be defined, regardless of what edits the
+  users perform, although user input may sometimes be necessary.
+\item \emph{Commutativity.} The order of edits should not matter, regardless of their pace layer.
+  If users $A$ and $B$ perform edits independently starting from the same initial document, the
+  resulting document should be the same if the system treats edits from $A$ as occurring before
+  the edits of $B$ and vice versa.
 \item \emph{Conflict resolution.} It may not always be possible to merge conflicting edits.
-In this case, the system should interactively ask the user for guidance or, possibly, use default behavior specified by the user.
-\item \emph{Convergence.} The system does not need to consider the case where one users does not want to adopt changes created by another user.
-We assume that all users want to eventually see the same final document produced by applying all edits.
-\item \emph{Structure editing.} The system may use a structure editor that provides high-level commands for operations such as reordering of elements or refactoring of a list into a table.
-In other words, we assume that the challenge may not be solvable in a system based on plain text editing of document source code.
+  In this case, the system should interactively ask the user for guidance or, possibly, use
+  default behavior specified by the user.
+\item \emph{Structure editing.} The system may use a structure editor that provides high-level
+  commands for operations such as reordering of elements or refactoring of a list into a table.
+  In other words, we recognise that the challenge may not be solvable in a system based on plain
+  text editing of document source code.
 \end{itemize}
 
-\subsection{Remarks}
-It is worth noting that this challenge does not discuss explicit distinction between the schema (or ``type'') of the document and data (or ``values'').
-The implementing system may or may not maintain such distinction.
-In our example, adding a new speaker and sorting the list is arguably a mere change of the data, but the refactoring of a list into a table is a schema change.
-The system may leverage knowledge about such distinction and, for example, handle the merging of schema-related and data-related changes differently.
-
-\section{Conference Organizer: Merging edits in the presence of code}
-
-\subsection{Context}
-Consider the document-based format as in the previous challenge, but extended with the support for computed values.
-A computed value is a node in the document whose value is specified by a formula akin to those used in spreadsheets.
-The equations can refer to other nodes in the document, possibly using either an absolute or relative path and may involve a range of built-in functions, for example to count the number of elements or to sum a sequence of numerical values.
-We now also assume that elements in the document can have IDs that are not displayed in the default view, but may be used to refer to them in equations.
+\paragraph{Remarks.}
+As discussed earlier, this challenge does not make an explicit distinction between the schema
+of the document and data. The implementing system may or may not maintain such distinction.
+In our example, adding a new speaker and sorting the list is a mere change of the data, but the
+refactoring of a list into a table is a schema change. The system may offer different user interface
+for changes at different layers and leverage knowledge about such distinction, for example,
+by handling the merging of schema-related and data-related changes differently.
 
 \begin{figure}
-\fbox{\parbox{28em}{
+\begin{adjustwidth}{-5.5em}{-5.5em}
+\centering
+\begin{subfigure}[b]{23em}\vspace{0pt}
+   \fbox{\parbox{22.6em}{
    \textbf{PROGRAMMING CONFERENCE 2023}\\
@@
    \end{itemize}
-\vspace{0.5em}
+   \vspace{1.5em}
    \textbf{Conference budget}\\
@@
    }}
-\caption{Conference organization. Initial state of the document.}
+   \caption{Initial state of the document with formulas.}
    \label{fig:conf-budget}
-\end{figure}
-
-\begin{figure}
-\fbox{\parbox{28em}{
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{23em}\vspace{0pt}
+   \fbox{\parbox{22.6em}{
    \textbf{PROGRAMMING CONFERENCE 2023}\\
@@
    \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
-}}\caption{Conference organization. Resulting document after change merging.}
+   }}\caption{Resulting document after schema change merging.}
    \label{fig:conf-finfin}
+\end{subfigure}
+\caption{Conference organizer. Formulas to calculate the budget have been added to the
+original document (Figure~\ref{fig:conf-orig}) and those have to be merged with the
+refactoring of a list into a table (Fig~\ref{fig:conf-tab}). Formulas in the final version
+are updated to match with the new document structure.}
+\label{fig:conf-calc}
+\end{adjustwidth}
 \end{figure}
 
-\subsection{Example}
-This challenge extends the previous problem.
-The conference organizers also want to use the document to manage the conference budget.
-In a very simplified form, one aspect of this is calculating the estimated travel expenses for all the speakers.
-For this, they will want to add a computed value to the document that counts the number of speakers and multiplies the result by a fixed cost per speaker elsewhere in the document.
-
-The example is shown in Figure~\ref{fig:conf-budget}.
-The additional section of the document contains one fixed value (travel cost per speaker) and two computed values.
-The first selects all \texttt{li} elements of a \texttt{ul} element with ID \texttt{speakers} and counts their number.
-The second computed value then multiplies the constant travel cost per speaker by the number of speakers.
-Here, we assume the new section uses the HTML definition list \texttt{dl} and the equation selects its first and second \texttt{dd} item, respectively.
-
-\subsection{Challenge}
-The challenge is, again, to merge the edits done independently by the co-organizers. Assume two co-organizers start with the same original document shown in Figure~\ref{fig:conf-orig}.
+\challenge{Applying schema edits in presence of code}
+
+This challenge extends the previous problem. The conference organizers now also want to use
+the document to manage the conference budget. In a very simplified form, one aspect
+of this is calculating the estimated travel expenses for the speakers. For this,
+they add a computed value to the document that counts the number of speakers
+and multiplies the result by a fixed cost per speaker elsewhere in the document.
+
+The scenario is shown in Figure~\ref{fig:conf-budget}. Assume two co-organizers start with
+the same original document shown in Figure~\ref{fig:conf-orig}. One adds an additional
+section with computed values, while the other performs the edits discussed earlier.
+The challenge is, again, to merge the edits done independently by the co-organizers.
+
 \begin{enumerate}
-    \item The first co-organizer performs the edits outlined previously that result in the document shown in Figure~\ref{fig:conf-fin}, i.e., edits the speakers and refactors the structure to use a table.
-    \item The second co-organizer adds the budget calculation as shown in Figure~\ref{fig:conf-budget}.
+\item The first co-organizer adds the budget calculation as shown in Figure~\ref{fig:conf-budget}.
+  This includes two formulas. The first selects all \texttt{li} elements of a \texttt{ul} element
+  with ID \texttt{speakers} (not visible in the UI) and counts their number. The second formula
+  multiplies the constant travel cost per speaker by the number of speakers. Here, we assume the
+  new section uses the HTML definition list \texttt{dl} and the formula selects its first and
+  second \texttt{dd} item, respectively.
+
+\item The second co-organizer performs the refactoring and data edits discussed previously.
+  They edit the speakers and refactor the structure to use a table, resulting in the document
+  shown in Figure~\ref{fig:conf-fin}.
 \end{enumerate}
+
+\noindent
 As before, the system should be able to merge the edits made to the document.
-Crucially, this requires that the refactoring of the document structure done in (1), is also reflected in the code of the equations added in (2).
-As shown in Figure~\ref{fig:conf-finfin}, in this specific case, the system changes the equation \texttt{COUNT(/ul[id='speakers']/li)} to \texttt{COUNT(/table[id='speakers']/tbody/tr)}.
-This is the case because the refactoring consists of four steps that each transform the document and also need to update the code of the equation accordingly:
+
+\frameworkbox{
+In this challenge, the change done by the first organizer is at the \emph{code} layer, while
+the change done by the second organizer is both at the \emph{data} layer and, implicitly,
+at the \emph{schema} layer. A change solely at the \emph{data} layer would not affect code,
+but a change at the \emph{schema} layer requires adapting some aspects of the code.
+}
+
+\paragraph{Remarks.}
+Addressing the challenge requires that the refactoring of the document structure done in (1),
+is also reflected in the code of the equations added in (2). As shown in
+Figure~\ref{fig:conf-finfin}, in this specific case, the system needs to change the equation
+\texttt{COUNT(/ul[id='speakers']/li)} to \texttt{COUNT(/table[id='speakers']/tbody/tr)}.
+This is the case because the refactoring consists of four steps that each transform the
+document and also need to update the code of the equation accordingly:
+
 \begin{itemize}
@@
 
-\subsection{Goals}
-A solution to the challenge should meet the goals of the preceding challenge, i.e., completeness (merging should be always defined), commutativity (order does not matter), conflict resolution (ask for guidance if needed), convergence (everyone eventually wants the same document) and structure editing (the system may capture high-level edits).
-
-\subsection{Extensions}
-The basic challenge outlined above focuses primarily on the local-first software scenario, but it can be extended to also apply to the live programming scenario.
-In particular, when the document contains computed values, those may be evaluated either explicitly by the user or automatically on-the-fly.
-An edit can then invalidate some of the values (either by changing the data that a computed value depends on, or by changing the equation used to compute the value).
-A live programming system should be able to detect which computed values are affected by an edit and, either invalidate those (requiring an explicit re-evaluation by the user) or automatically re-evaluate them (and possibly highlight the affected values to inform the user).
+\noindent
+It is expected that the system solving the challenge will need to be aware of those high-level
+edits. In particular, they may be explicitly specified by the user (through an interactive
+user interface for editing the document) or semi-automatically inferred.
+
+\paragraph{Requirements.}
+A solution to the challenge should meet the goals of the preceding challenge, i.e., completeness
+(merging should be always defined), commutativity (order does not matter), conflict resolution
+(ask for guidance if needed), convergence (everyone eventually wants the same document) and
+structure editing (the system may capture high-level edits).
+
+\subsection*{Challenge Extensions: Liveness and divergence}
+So far, we assumed that the system follows the \emph{convergence} model, i.e., all variants
+of the program eventually adopt all changes. A more difficult variant on the challenge is to also
+support the \emph{divergence} model. In this case, some of the users continue using their variant
+of the program without adopting all of the changes done by other users.
+
+\frameworkbox{
+In the \emph{divergence} model, it may be expected that users will want to adopt all changes
+done at the \emph{data} pace layer. They may selectively choose some of the changes at the
+\emph{code} and \emph{schema} layers. The challenge is maintaining the same data across multiple
+program variants and merging changes done to the data based on different document schema.
+}
+
+The basic challenge outlined above focuses primarily on the local-first software scenario, but
+it can be extended to also apply to the live programming scenario. In particular, when the
+document contains computed values, those may be evaluated either explicitly by the user or
+automatically on-the-fly. An edit can then invalidate some of the values (either by changing
+the data that a computed value depends on, or by changing the equation used to compute the value).
+A live programming system should be able to detect which computed values are affected by an edit
+and, either invalidate those (requiring an explicit re-evaluation by the user) or automatically
+re-evaluate them (and possibly highlight the affected values to inform the user).
+
+\frameworkbox{
+In the live extension, we also consider the transient \emph{state} pace layer. A system should
+automatically detect when a change at the \emph{data} or \emph{code} layers affects the computed
+value and forces it to be recomputed. Interestingly, a change at the \emph{schema} layer may not
+always affect the \emph{state} layer, even if it schema change requires modifying formulas that
+compute the state.
+}
+
+\newpage
 

### 2024-04-30 paper.tex
Replace \\~ with vspace
My latex installation was erroring with "There's no line here to end."
Using a vspace instead for the space seems to achieve the same
end result without an error.
draft of multiplicity problem

@@
    \end{itemize}
-   \\~
+   \vspace{1em}
    }}
@@
 
+\section{Multiplicity change}
+
+\subsection{Context}
+
+A common kind of schema change is \textit{multiplicity change}: when a field changes from storing a single value to storing multiple values.
+For example, we might start by storing a single address for each contact in an address book, before realizing that we need to store multiple addresses for a single contact; or we might assign each todo in a list to one person before realizing we want the ability to assign a todo to multiple people.
+
+In a relational schema, solving this problem might require extracting a normalized table for the linked entity; some of the challenges of this approach are covered in the Extract Entity challenge.
+In this section we will instead consider a document schema with support for arrays; in this context, we can have a multiplicity change by turning a scalar value into an array value.
+
+Multiplicity change clearly requires changing both the data and the code.
+It becomes particularly challenging to handle when there is a need to support ongoing writes to both the scalar and array schemas.
+This need can arise in any situation where the code and data can't all be updated atomically together—e.g., in a zero-downtime upgrade in a web architecture with stateless servers and a database, or in a decentralized setting like a public API where some consumers of the API may be expecting an outdated schema.
+
+\subsection{Example}
+
+Consider the following schema for a todo list item, in which each item has a single assignee, represented by a user ID:
+
+\begin{quote}
+\begin{lstlisting}[language=ml]
+type Item = { id : id; title : string; assignee : string }
+\end{lstlisting}
+\end{quote}
+
+Now, we change the schema so that each item has a list of assignees:
+
+\begin{quote}
+\begin{lstlisting}[language=ml]
+type Item = { id : id; title : string; assignees : Array<string> }
+\end{lstlisting}
+\end{quote}
+
+Our goal is to preserve the ability for actors in the system to read and write to a shared todo list in either the old or new schema—for example, they should be able to write to either the scalar assignee field or to the list of assignees.
+
+A natural invariant to preserve across these schemas is that the value of the scalar field should equal the first element of the list field (and if the list is empty, the scalar field should be null).
+If we were to write code for a one-time data migration from the scalar to list schema, we could easily satisfy this invariant:
+
+\begin{quote}
+\begin{lstlisting}[language=ml]
+item.assignees = if (item.assignee == null) then [] else [item.assignee]
+\end{lstlisting}
+\end{quote}
+
+Ongoing edits to the array schema can also be handled in a straightforward way.
+After edits are made, the value of the scalar field should be set to the first element of the new array (or null if the new array is empty.)
+
+However, handling edits to the scalar schema presents more of a challenge.
+Consider the following todo with two assignees, presented in terms of the scalar and array schemas.
+A write is made in the scalar schema to set the new assignee to C.
+
+\begin{quote}
+\begin{lstlisting}[language=ml]
+todoScalar = { id: 1, title: "Foo", assignee: A }
+todoArray = { id: 1, title: "Foo", assignees: [A, B] }
+
+todoScalar.assignee = C
+\end{lstlisting}
+\end{quote}
+
+What should the array value become after this edit to the scalar schema?
+To satisfy our invariant, we know that C must become the first element of the array, but this leaves open several options with different tradeoffs:
+
+\begin{enumerate}
+  \item \texttt{[C]}: "Only C should be assigned." This option produces an array that corresponds directly to the resulting scalar.
+But it has the downside of deleting data that wasn't even visible in the scalar schema.
+  \item \texttt{[C, B]}: "Replace A with C." If the writer wanted to remove A from the assignment and add C, this option performs that intent.
+However, there is data remaining in the list which was not visible to the scalar writer.
+  \item \texttt{[C, A, B]}: "Add C to the list." Perhaps the scalar writer wanted to add C to the assignment without removing anyone; this option satisfies that intent.
+But it preserves data not visible to the scalar schema.
+\end{enumerate}
+
+Because the intent of the writer to the scalar schema cannot be unambiguously interpreted from the write alone, it is impossible to make a perfect choice among these options.
+
+\subsection{Challenge}
+
+Multiplicity change illustrates a general problem for handling ongoing writes between diverging schemas.
+Different schemas may expose partial information, and writers in those schemas have to operate without total knowledge of the information available in other schemas.
+As a result, synchronizing writes across the schemas requires making difficult tradeoffs, such as choosing to either preserve or destroy hidden data not visible to the writer.
+
+Although we see no silver bullet to navigate these tradeoffs, a good solution to this problem would give developers or users tools to manage these tradeoffs.
+For example, a system might allow developers to specify the desired behavior for a particular pair of schemas based on the requirements of the domain.
+Or a system might even allow users to manually disambiguate their intent for a given write.
+
+\subsection{Goals}
+
+TODO
+
 \begin{comment}

### 2024-05-01 paper.tex
use biblatex, comments
natbib isn't working. Journal template uses biblatex so switched to that. still an error about inputenx

@@
 \documentclass[english,submission]{programming}
-%\usepackage[backend=biber]{biblatex}
-%\addbibresource{paper.bib}
+
+\usepackage{csquotes}
+\usepackage{inputenx}
+\usepackage[backend=biber]{biblatex}
+%\usepackage{natbib}
+\addbibresource{paper.bib}
 
@@
 \usepackage{ccicons}
-\usepackage{natbib}
 \usepackage[many]{tcolorbox}
@@
 
+% define natbib citet
+\newcommand{\citet}[1]{\citeauthor*{#1} \cite{#1}}
+
 \title{Schema Change: Challenge Problems}
@@
 \authorinfo{is TBD. Contact him at \email{gklitt@gmail.com}.}
-\affiliation[b]{Ink \& Switch, Planet Earth}
+\affiliation[e]{Ink \& Switch, Planet Earth}
 
@@
 
-\keywords{schema change, live programming, local-first programming}
+\keywords{schema change, database migration, live programming, local-first programming}
 
@@
 \noindent\textbf{Context}: \emph{What is the broad context of the work? What is the importance of the general research area?}
-Schema change is unsolved problem in live and local-first programming. It is also poorly understood
-problem that calls for a more fundamental analysis. It is particularly important as more programming
+Schema change is an unsolved problem in live and local-first programming that is poorly understood and calls for a more fundamental analysis. It is particularly important as more programming
 is done in programming systems where programs may be changed by their users and as more programming
@@
 Schema change on database servers is often delegated to Database Administrators and DevOps.
-Live programming~\cite{tanimoto90,hancock03} and local-first software~\cite{localfirst} both move data away from the server, eliminating those jobs, but not eliminating the need to do schema change, and indeed increasing the need to do it automatically.
+Live programming~\cite{tanimoto90,Hancock03} and local-first software~\cite{localfirst} both move data away from the server, eliminating those jobs, but not eliminating the need to do schema change, and indeed increasing the need to do it automatically.
 Recently \citet{Cambria} spotlighted these problems and proposed an approach using lenses~\cite{Foster2007}, but otherwise there has been surprisingly little research.
@@
 ~
+\textcolor{red}{[JE] We could expand our context to include conventional DB systems not just live/local.
+It is a widespread problem, attested by the many tools to assiste with it.
+We could include in our problems a description of a typical solution with a standard DB as a sort of baseline.
+}
 
@@
 less for evaluating particular solutions but more for framing and making sense of different
-kinds of schema-chaange-related problems.  You can read this paper and when you face a schema-change-related
+kinds of schema-change-related problems.  You can read this paper and when you face a schema-change-related
 problem, think about it in terms of our dimensions. Maybe compare it to our challenges.
@@
   needs to change every couple of years. ``Because of the different rates of change of its
-  components, a building is always tearing itself apart.''}
+  components, a building is always tearing itself apart.''
+  \textcolor{red}{[JE] Except the outer skin changes faster than the structure.}
+  }
 \label{fig:layers}
 \end{figure}
+
+\textcolor{red}{[JE] `change in a program' might be misunderstood as changes to the source code. Perhaps `system' or `software system' might be better than `program'? }
 
@@
 
-\bibliographystyle{abbrv}
-\bibliography{paper}
+%\bibliographystyle{abbrv}
+%\bibliography{paper}
+\printbibliography
 

### 2024-05-02 paper.tex
comment
We could expand the scope to include conventional DBs, asking for a more declarative solution than hand-coded SQL.

@@
 ~
-\textcolor{red}{[JE] We could expand our context to include conventional DB systems not just live/local.
-It is a widespread problem, attested by the many tools to assiste with it.
-We could include in our problems a description of a typical solution with a standard DB as a sort of baseline.
+\textcolor{red}{[JE] We could expand our context to include schema change for conventional DB systems, not just live/local.
+It is a widespread problem, attested by the many tools that assist with it, but all of them come down to manually coding SQL.
+We could call for a higher-level declarative solution.
 }

### 2024-05-03 paper.tex
Start of Related Work

@@
 
+\section{Related Work}
+
+Schema change and data migration are a widespread problem in practice for SQL databases, attested by the many tools that have been created to help.
+Some tools, like Rails Migrations~\cite{RailsMigrations} provide a Domain Specific Language (DSL) for specifying and executing schema changes.
+Other tools, such as those in Liquibase~\cite{liquibase} or PlanetScale\cite{planetscale} are based on textually comparing SQL Data Definition Language (DDL) statements that define the database schema.
+The differences between these DDL schema can be used to infer a delta of DDL statements to change one into the other.
+Unfortunately these inferences must sometimes guess at the user's intent: for example has a column been renamed or has it been deleted and a new one created?
+The difference matters a lot for the data in that column!
+EdgeDB~\cite{edgedb} asks the developer to resolve such ambiguities with a series of questions.
+
+All of these tools end up emitting SQL DDL statements to change the schema.
+Unfortunately that does not migrate the data.
+Some databases add limited features for migration.
+For example SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
+MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data.
+But apart from such special cases, SQL databases have no general purpose support for schema migration.
+A developer must write custom SQL code to transform data from the old to the new schema.
+This custom migration code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
+In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema change patterns, including typical strategies for online migration and example SQL code.
+
+There is surprisingly little discussion of schema migration in the database literature nor support from the major database vendors.
+There is indeed a rich theory of database normalization~\cite{Molina08} but that is presumed to happen before the database is populated with data.
+To a large extent this entire problem area has been relegated to application developers and `dev ops'.
+Yet prima facia, schema migration can only be done throughly with direct support inside the database.
+EdgeDB~cite{edgedb} stands out for its improved support of schema migration through a proprietary language of data transformations but is still limited by the capabilties of its underlying PostgresSQL database.
+
+\begin{comment}
+
+Newly discovered thread of DB research on schema change:
+https://en.wikipedia.org/wiki/Schema_evolution
+https://www.researchgate.net/publication/220710372_Schema_Evolution_in_Wikipedia_-_Toward_a_Web_Information_System_Benchmark
+https://dl.acm.org/doi/10.14778/1453856.1453939
+https://dl.acm.org/doi/10.1145/3448016.3452842
+https://dl.acm.org/doi/10.14778/3565816.3565818
+https://link.springer.com/chapter/10.1007/978-3-030-58135-0_10
+
+Schema change IDEs
+https://www.red-gate.com/solutions/overview
+https://www.jetbrains.com/datagrip/features/
+
+Smalltalk images
+	depends on dynamic types and adding new fields to objects
+	Gemstone added complex API
+
+“unstructured data”: JSON and RDF used with untyped PL like Javascript or Python
+Like Smalltallk, small set of changes can be handled gradecfully, but in general need to write custom code
+
+dataframe diff
+https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
+
+Tijs on Live programming and model-driven engineering
+https://ir.cwi.nl/pub/23770
+https://dl.acm.org/doi/10.1007/s10270-017-0608-7
+https://dl.acm.org/doi/10.1145/3276604.3276611
+https://ir.cwi.nl/pub/21858
+
+Geoffreu on Local-first
+https://www.inkandswitch.com/cambria/
+https://riffle.systems/papoc22.pdf
+
+Durable computing: pins runtime data frames to code versions. Doesn’t handle problem of global data schema change
+
+Version control for dbs
+https://people.eecs.berkeley.edu/~brewer/cs262b/update-conflicts.pdf
+https://github.com/dolthub/dolt
+https://github.com/dolthub/doltgresql
+https://irmin.org
+https://doc.replicache.dev/concepts/how-it-works
+
+\end{comment}
+
 \newpage
@@
 But before \texttt{Customers.Name} can be made a primary key the two occurrences of \texttt{Wile E Coyote} must be deduplicated one way or another~\cite{dedupe}.
+
+\textcolor{red}{this paragraph is now redundant with Related Owrk}
 Given the dominance of SQL databases it is not surprising that there are many tools for performing schema change on them~\cite{RailsMigrations, liquibase}.
@@
 systems that can be seen as its successors. The specific scenario discussed here is
-inspired by the use of Notion~\cite{Notion24}, which is a collaborative structured document
+inspired by the use of Notion~\cite{notion}, which is a collaborative structured document
 editing system. Treating documents as end-user programs has recently

### 2024-05-08 paper.tex
Related Work

--- 2024-05-01 paper.tex
@@
 % define natbib citet
-\newcommand{\citet}[1]{\citeauthor*{#1} \cite{#1}}
+\newcommand{\citet}[1]{\citeauthor*{#1}~\cite{#1}}
 
@@
 \section{Introduction}
+
+\textcolor{red}{[JE] We should adopt the term Schema Evolution, which is favored in the research literature, defined to include schema change, data migration, and query rewriting. }
 
@@
 ~
-
-~
-\textcolor{red}{[JE] We could expand our context to include conventional DB systems not just live/local.
-It is a widespread problem, attested by the many tools to assiste with it.
-We could include in our problems a description of a typical solution with a standard DB as a sort of baseline.
-}
 
@@
 
+\section{Related Work}
+
+Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help.
+SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
+MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data.
+But apart from such special cases, SQL databases have no general purpose support for data migration.
+As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data.
+Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
+In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
+
+\citet{bernstein07} observed in 2007 ``There are hardly any schema evolution tools today.
+This is rather surprising since there is a huge literature on schema evolution spanning more than two decades.'' There are now more tools available.
+Some tools such as Liquibase~\cite{liquibase} and PlanetScale\cite{planetscale} could be characterized as version control and continuous integration/deployment for schema.
+They track schema changes and can calculate diffs in the form of SQL DDL statements to convert one schema version to another, but do not help migrate data.
+Unfortunately comparing schemas can be ambiguous about the intention of changes.
+For example has a column been renamed or has it been deleted and a new one created?
+That distinction makes a big difference to the data in that column.
+EvolveDB\cite{evolvedb} addresses this ambiguity by reverse-engineering the schema into a richer data model and tracking the edits to that model within an IDE.
+This more precise edit history can be used to infer higher level intentions of a schema change, which then generate SQL scripts to evolve the database.
+EdgeDB~\cite{edgedb} resolves ambiguities by asking questions of the developer, with some answers supplying custom migration code in a proprietary query language.
+
+Some tools provide a Domain Specific Language (DSL) to describe schema evolution.
+Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions.
+However that falls short of the range of evolutions described by \citet{ambler06} which require the addition of custom Ruby or SQL code.
+\citet{curino08} initiated a large thread of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
+SMOs can also rewrite queries to accomodate schema changes.
+\citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17} that appears capable of handling our Extract Entity and Multiplicity Change challenges, including divergence.
+
+Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases.
+While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}.
+\citet{scherzinger13} define a set of operators like the relational SMOs discussed above.
+\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that appear capable of handling the data migration aspects of our Extract Entity and Multiplicity Change challenges.
+\citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON.
+None of the approaches to NoSQL evolution have yet extended like their SQL cousins into updating code or rewriting queries.
+
+Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}.
+Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with limited evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}.
+Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a sophisticated schema evolution API~\cite{Gemstone}.
+
+\begin{comment}
+
+dataframe diff
+https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
+
+Durable computing: pins runtime data frames to code versions. Doesn’t handle problem of global data schema change.
+https://restate.dev/blog/solving-durable-executions-immutability-problem/
+https://restate.dev/blog/code-that-sleeps-for-a-month/
+
+Tijs on Live programming and model-driven engineering
+https://ir.cwi.nl/pub/23770
+https://dl.acm.org/doi/10.1007/s10270-017-0608-7
+https://dl.acm.org/doi/10.1145/3276604.3276611
+https://ir.cwi.nl/pub/21858
+
+From Geoffrey:
+1) Finding Data Compatibility Bugs with JSON Subschema Checking
+https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
+I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant.
+It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
+2) Confluent Schema Registry:
+https://docs.confluent.io/platform/current/schema-registry/index.html
+An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
+\end{comment}
+
 \newpage

### 2024-05-11 paper.tex
harshen related work

@@
 Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions.
-However that falls short of the range of evolutions described by \citet{ambler06} which require the addition of custom Ruby or SQL code.
-\citet{curino08} initiated a large thread of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
+However that falls short of the range of evolutions described by \citet{ambler06}, requiring the addition of custom Ruby or SQL code.
+\citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
 SMOs can also rewrite queries to accomodate schema changes.
-\citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17} that appears capable of handling our Extract Entity and Multiplicity Change challenges, including divergence.
+\citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}.
+BiDEL appears capable of handling the basic requirements of our Extract Entity and Multiplicity Change challenges, though Split/Merge Entity is less clear.
+It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.
 
@@
 While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}.
+\citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON.
 \citet{scherzinger13} define a set of operators like the relational SMOs discussed above.
-\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that appear capable of handling the data migration aspects of our Extract Entity and Multiplicity Change challenges.
-\citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON.
-None of the approaches to NoSQL evolution have yet extended like their SQL cousins into updating code or rewriting queries.
+\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling Extract Entity and Multiplicity Change but not Split/Merge Entity nor divergence.
+None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.
 
 Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}.
-Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with limited evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}.
-Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a sophisticated schema evolution API~\cite{Gemstone}.
+Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}.
+Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.
 
@@
 
-dataframe diff
-https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
+Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations
+
+dataframe diff https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
 

### 2024-05-11 paper.tex
bibtek stopped working

@@
 
-\usepackage{csquotes}
-\usepackage{inputenx}
 \usepackage[backend=biber]{biblatex}
@@
 
+\usepackage{csquotes}
 \usepackage{changepage}
@@
 \usepackage[many]{tcolorbox}
+\lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
+  deletetexcs={title,author,bibliography},%
+  deletekeywords={tabular},
+  morekeywords={abstract},%
+  moretexcs={chapter},%
+  moretexcs=[2]{title,author,subtitle,keywords,maketitle,titlerunning,authorinfo,affiliation,authorrunning,paperdetails,acks,email},
+  moretexcs=[3]{addbibresource,printbibliography,bibliography},%
+}%
+\lstset{%
+  language={[programming]TeX},%
+  keywordstyle=\firamedium,
+  stringstyle=\color{RosyBrown},%
+  texcsstyle=*{\color{Purple}\mdseries},%
+  texcsstyle=*[2]{\color{Blue1}},%
+  texcsstyle=*[3]{\color{ForestGreen}},%
+  commentstyle={\color{FireBrick}},%
+  escapechar=`,}
+\newcommand*{\CTAN}[1]{\href{http://ctan.org/tex-archive/#1}{\nolinkurl{CTAN:#1}}}
+%%
 

### 2024-05-13 paper.tex
beef up Extract Entity

@@
 
+\usepackage{inputenx}
 \usepackage[backend=biber]{biblatex}
@@
 
-\section{Related Work}
-
-Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help.
-SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
-MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data.
-But apart from such special cases, SQL databases have no general purpose support for data migration.
-As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data.
-Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
-In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
-
-\citet{bernstein07} observed in 2007 ``There are hardly any schema evolution tools today.
-This is rather surprising since there is a huge literature on schema evolution spanning more than two decades.'' There are now more tools available.
-Some tools such as Liquibase~\cite{liquibase} and PlanetScale\cite{planetscale} could be characterized as version control and continuous integration/deployment for schema.
-They track schema changes and can calculate diffs in the form of SQL DDL statements to convert one schema version to another, but do not help migrate data.
-Unfortunately comparing schemas can be ambiguous about the intention of changes.
-For example has a column been renamed or has it been deleted and a new one created?
-That distinction makes a big difference to the data in that column.
-EvolveDB\cite{evolvedb} addresses this ambiguity by reverse-engineering the schema into a richer data model and tracking the edits to that model within an IDE.
-This more precise edit history can be used to infer higher level intentions of a schema change, which then generate SQL scripts to evolve the database.
-EdgeDB~\cite{edgedb} resolves ambiguities by asking questions of the developer, with some answers supplying custom migration code in a proprietary query language.
-
-Some tools provide a Domain Specific Language (DSL) to describe schema evolution.
-Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions.
-However that falls short of the range of evolutions described by \citet{ambler06}, requiring the addition of custom Ruby or SQL code.
-\citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
-SMOs can also rewrite queries to accomodate schema changes.
-\citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}.
-BiDEL appears capable of handling the basic requirements of our Extract Entity and Multiplicity Change challenges, though Split/Merge Entity is less clear.
-It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.
-
-Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases.
-While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}.
-\citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON.
-\citet{scherzinger13} define a set of operators like the relational SMOs discussed above.
-\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling Extract Entity and Multiplicity Change but not Split/Merge Entity nor divergence.
-None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.
-
-Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}.
-Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}.
-Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.
-
-\begin{comment}
-
-Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations
-
-dataframe diff https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
-
-Durable computing: pins runtime data frames to code versions. Doesn’t handle problem of global data schema change.
-https://restate.dev/blog/solving-durable-executions-immutability-problem/
-https://restate.dev/blog/code-that-sleeps-for-a-month/
-
-Tijs on Live programming and model-driven engineering
-https://ir.cwi.nl/pub/23770
-https://dl.acm.org/doi/10.1007/s10270-017-0608-7
-https://dl.acm.org/doi/10.1145/3276604.3276611
-https://ir.cwi.nl/pub/21858
-
-From Geoffrey:
-1) Finding Data Compatibility Bugs with JSON Subschema Checking
-https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
-I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant.
-It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
-2) Confluent Schema Registry:
-https://docs.confluent.io/platform/current/schema-registry/index.html
-An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
-\end{comment}
-
-\newpage
-
-\section{Extract Entity}
-
-\subsection{Context}
-In this challenge we consider a typical example of how the design of a database evolves as new needs are discovered.
-Much has been written on the subject of data modeling methodologies and the theory of relational database normalization~\cite{Molina08}.
-But that work often focuses on determining the correct data model up front before the system goes live, which would indeed be the optimal solution if only we were omniscient.
-The Extract Entity challenge confronts the practical reality of evolving a relational data model in flight.
-
-\subsection{Example}
-The Acme Corporation needs to record orders from various customers for various products.
-The simplest implementation is a spreadsheet with a row for each order and columns for information about the customer and product.
-It might look like this:
-
-\begin{table}[h!]
-\centering
-  \begin{tabular}{ |l|c|c|l|l|}
-   \hline
-   Item & Quantity & Ship Date & Customer name & Customer address \\
-   \hline \hline
-   Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
-   \hline
-   Dynamite & 2 & & Daffy Duck & White Rock Lake \\
-   \hline
-   Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
-   \hline
-  \end{tabular}
-\caption{Orders}
-\end{table}
-
-The shipping department filters this table on blank ship dates to see what they need to ship.
-But after a while the orders department realizes they are wasting effort duplicating the address for new order from an old customer.
-And when the customer's address changes they have to go back and edit all of their orders.
-What is needed is two tables, one with orders that links to one with customers, like this:
-
-\begin{table}[h!]
-  \centering
-    \begin{tabular}{ |l|l|}
-      \hline
-      Name & Address \\
-      \hline \hline
-      Wile E Coyote & 123 Desert Station \\
-      \hline
-      Daffy Duck & White Rock Lake \\
-      \hline
-    \end{tabular}
-  \caption{Customers}
-\end{table}
-
-\begin{table}[h!]
-  \centering
-    \begin{tabular}{ |l|c|c|l|}
-     \hline
-     Item & Quantity & Ship Date & Customer name \\
-     \hline \hline
-     Anvil & 1 & 2/3/23 & Wile E Coyote \\
-     \hline
-     Dynamite & 2 & & Daffy Duck  \\
-     \hline
-     Bird Seed & 1 & & Wile E Coyote \\
-     \hline
-    \end{tabular}
-  \caption{Orders linking to Customers}
-\end{table}
-
-We often encounter this sort of design change when we realize that attributes of one type of entity should actually belong to a new type of entity that will be associated with the first one.
-Often the motivation is to centralize changes to those attributes in one place.
-We call this operation \textit{Extract Entity}.
-
-In a SQL database, \texttt{Customers.Name} would be a primary key and \texttt{Orders.Customer} would be a foreign key.
-But before \texttt{Customers.Name} can be made a primary key the two occurrences of \texttt{Wile E Coyote} must be deduplicated one way or another~\cite{dedupe}.
-
-\textcolor{red}{this paragraph is now redundant with Related Owrk}
-Given the dominance of SQL databases it is not surprising that there are many tools for performing schema change on them~\cite{RailsMigrations, liquibase}.
-We do find it surprising how low-level these tools are.
-There is support for the basic SQL DDL commands like adding and dropping tables, columns, or constraints, but no higher level operations that would help solve our problem.
-One is left to write custom SQL, which can become difficult if the migration needs to be performed online without bring down the database.
-There is a rich theory of relational database \textit{normalization} based on \textit{functional dependencies}~\cite{Molina08} but that theory has not been implemented into schema change operations in databases -- all you get is SQL.
-\citet{ambler06} define higher-level patterns of schema change like \textit{Split Table} that sketch SQL implementations but don't provide everything required for Extract Entity.
-These difficulties fairly beg for new linguistic abstractions, though programming patterns and APIs are reasonable pragmatic solutions to our challenge.
-
-The Extract Entity schema change is straightforward in a spreadsheet.
-The \texttt{Customer Name} and \texttt{Customer Address} columns are copied and pasted into a new Customers table.
-The \textit{Remove Duplicates} command is used to deduplicate the customers.
-To preserve the original view for the shipping department the \texttt{Customer Address} column is redefined with a \texttt{VLOOKUP} formula to pull the address out of the customer table based on the customer's name.
-A spreadsheet won't maintain referential integrity between the two tables, but end-user database tools can do that~\cite{airtable, notion}.
-
-\subsection{Challenge}
-The Extract Entity challenge for live programming is to provide interactive operations in the programming environment that will perform the schema change on live code and data.
-The data can be in any form desired, including relations, objects, or JSON.
-The code can be a simple CRUD UI that also provides the shipping department its view of unshipped orders.
-The central problem is to coordinate the schema change with code edits/refactorings to keep the system executing live and correctly.
-Solutions should try to minimize:
-
-\begin{enumerate}
-  \item The number and complexity of commands or UI affordances that must be introduced.
-  \item Interruption to live execution of the code.
-\end{enumerate}
-
-The Extract Entity challenge for local-first software is similar to the live programming challenge except that the context is an application that offers similar functionality with peer-to-peer collaboration.
-In this context the schema change does not need to be performed interactively — a developer can take some time to build, test, and package a solution, perhaps using an API or DSL.
-The hard part comes when the schema change is to be deployed across all the replicas.
-This deployment must synchronize code and data upgrades (or decouple them as in Cambria~\cite{Cambria}).
-The deployment must also deal with migrating ``in flight'' operations so that all replicas converge on the same state without data loss, and ideally without centralized coordination.
-A radical approach could try to incorporate schema change operations into the underlying datastore/CRDT itself and integrate code as well, but layered solutions are welcome too.
-
-\subsection{Goals}
-Solutions should try to minimize:
-
-\begin{enumerate}
-  \item The need for users to manually intervene.
-  \item Possibilities that developer-written code is subtly incorrect in edge cases, preferably by reducing the need for developer-written code.
-  \item The need for a centralized coordinator.
-  \item How long replicas might be partitioned until resynchronization is acquired.
-\end{enumerate}
-
 \section{Divergence Control}
+
+\textcolor{red}{This section was an extension of the Extract Entity challenge but should be lifted up to the top-level discussion of divergence}
 
@@
 We encourage both communities to think outside the box of Git, which has proven to baffle not only end-users but also a substantial fraction of developers.
+
+\section{Challenge goals}
+\textcolor{red}{Text from the Extract Entity challenge that maybe should be lifted up to the top-level discussion of goals}
+
+The Extract Entity challenge for live programming is to provide interactive operations in the programming environment that will perform the schema change on live code and data.
+The data can be in any form desired, including relations, objects, or JSON.
+The code can be a simple CRUD UI that also provides the shipping department its view of unshipped orders.
+The central problem is to coordinate the schema change with code edits/refactorings to keep the system executing live and correctly.
+Solutions should try to minimize:
+
+\begin{enumerate}
+  \item The number and complexity of commands or UI affordances that must be introduced.
+  \item Interruption to live execution of the code.
+\end{enumerate}
+
+The Extract Entity challenge for local-first software is similar to the live programming challenge except that the context is an application that offers similar functionality with peer-to-peer collaboration.
+In this context the schema change does not need to be performed interactively — a developer can take some time to build, test, and package a solution, perhaps using an API or DSL.
+The hard part comes when the schema change is to be deployed across all the replicas.
+This deployment must synchronize code and data upgrades (or decouple them as in Cambria~\cite{Cambria}).
+The deployment must also deal with migrating ``in flight'' operations so that all replicas converge on the same state without data loss, and ideally without centralized coordination.
+A radical approach could try to incorporate schema change operations into the underlying datastore/CRDT itself and integrate code as well, but layered solutions are welcome too.
+
+\subsection{Goals}
+Solutions should try to minimize:
+
+\begin{enumerate}
+  \item The need for users to manually intervene.
+  \item Possibilities that developer-written code is subtly incorrect in edge cases, preferably by reducing the need for developer-written code.
+  \item The need for a centralized coordinator.
+  \item How long replicas might be partitioned until resynchronization is acquired.
+\end{enumerate}
+
+\section{Related Work}
+
+Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help.
+SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
+MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data.
+But apart from such special cases, SQL databases have no general purpose support for data migration.
+As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data.
+Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
+In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
+
+\citet{bernstein07} observed in 2007 ``There are hardly any schema evolution tools today.
+This is rather surprising since there is a huge literature on schema evolution spanning more than two decades.'' There are now more tools available.
+Some tools such as Liquibase~\cite{liquibase} and PlanetScale\cite{planetscale} could be characterized as version control and continuous integration/deployment for schema.
+They track schema changes and can calculate diffs in the form of SQL DDL statements to convert one schema version to another, but do not help migrate data.
+Unfortunately comparing schemas can be ambiguous about the intention of changes.
+For example has a column been renamed or has it been deleted and a new one created?
+That distinction makes a big difference to the data in that column.
+EvolveDB\cite{evolvedb} addresses this ambiguity by reverse-engineering the schema into a richer data model and tracking the edits to that model within an IDE.
+This more precise edit history can be used to infer higher level intentions of a schema change, which then generate SQL scripts to evolve the database.
+EdgeDB~\cite{edgedb} resolves ambiguities by asking questions of the developer, with some answers supplying custom migration code in a proprietary query language.
+
+Some tools provide a Domain Specific Language (DSL) to describe schema evolution.
+Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions.
+However that falls short of the range of evolutions described by \citet{ambler06}, requiring the addition of custom Ruby or SQL code.
+\citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
+SMOs can also rewrite queries to accomodate schema changes.
+\citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}.
+BiDEL appears capable of handling the basic requirements of our \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} challenges, though \textit{Split/Merge Entity} is less clear.
+It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.
+
+Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases.
+While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}.
+\citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON.
+\citet{scherzinger13} define a set of operators like the relational SMOs discussed above.
+\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} but not \textit{Split/Merge Entity} nor divergence.
+None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.
+
+Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}.
+Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}.
+Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.
+
+\begin{comment}
+
+Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations
+
+dataframe diff https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
+
+Durable computing: pins runtime data frames to code versions. Doesn’t handle problem of global data schema change.
+https://restate.dev/blog/solving-durable-executions-immutability-problem/
+https://restate.dev/blog/code-that-sleeps-for-a-month/
+
+Tijs on Live programming and model-driven engineering
+https://ir.cwi.nl/pub/23770
+https://dl.acm.org/doi/10.1007/s10270-017-0608-7
+https://dl.acm.org/doi/10.1145/3276604.3276611
+https://ir.cwi.nl/pub/21858
+
+From Geoffrey:
+1) Finding Data Compatibility Bugs with JSON Subschema Checking
+https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
+I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant.
+It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
+2) Confluent Schema Registry:
+https://docs.confluent.io/platform/current/schema-registry/index.html
+An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
+\end{comment}
+
+\newpage
+
+\section{Extract/Absorb/Split/Merge Entity}
+\textcolor{red}{Maybe called \textit{Individuation}? Maybe called \textit{Entity Evolution}?}
+
+The Acme Corporation needs to record orders from various customers for various products.
+The simplest implementation is a spreadsheet with a row for each order and columns for information about the customer and product.
+It might look like this:
+
+\begin{table}[h!]
+  \sffamily
+  \small
+  % \centering
+  \begin{tabular}{ |r|l|r|r|l|l|}
+   \hline
+   order\_id & item & quantity & ship\_date & customer\_name & customer\_address \\
+   \hline \hline
+   1 & Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
+   \hline
+   2 & Dynamite & 2 & & Daffy Duck & White Rock Lake \\
+   \hline
+   3 & Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
+   \hline
+  \end{tabular}
+  \caption{Orders}
+\end{table}
+
+The shipping department filters this table on blank ship dates to see what they need to ship.
+But after a while the orders department realizes they are wasting effort duplicating the address for new order from an old customer.
+And when the customer's address changes they have to go back and edit all of their orders.
+What is needed is two tables, one with orders that links to one with customers, like this:
+
+\begin{table}[h!]
+  \sffamily
+  \small
+  % \centering
+    \begin{tabular}{ |r|l|l|}
+      \hline
+      customer\_id & customer\_name & customer\_address \\
+      \hline \hline
+      1 & Wile E Coyote & 123 Desert Station \\
+      \hline
+      2 & Daffy Duck & White Rock Lake \\
+      \hline
+    \end{tabular}
+  \caption{Customers}
+\end{table}
+
+\begin{table}[h!]
+  \sffamily
+  \small
+  % \centering
+    \begin{tabular}{ |r|l|r|r|r|}
+     \hline
+     order\_id & item & quantity & ship\_date & customer\_id \\
+     \hline \hline
+     1 & Anvil & 1 & 2/3/23 & 1 \\
+     \hline
+     2 & Dynamite & 2 & & 2  \\
+     \hline
+     3 & Bird Seed & 1 & & 1 \\
+     \hline
+    \end{tabular}
+  \caption{Orders linking to Customers}
+\end{table}
+
+We often encounter this situation when we realize that the attributes of one type of entity should actually belong to a distinct type of entity that will be associated with the first one.
+Often the motivation is to centralize changes to those attributes in one place.
+That is what it means to be an \textit{entity}: a referenceable holder of mutable attributes.
+We call this operation \textit{Extract Entity}.
+
+Unique identifiers (here consecutive numbers) have been assigned to the customers and are referenced from the orders.
+This allows, for example, a spelling error in a customer address to be fixed in one place.
+It also allow a customer name or address to change without breaking the connection from all associated orders.
+Unique identifiers are the essence of entities, whether they are uniquely generated primary keys in a database or an abstraction of a memory address in a programming language.
+
+The \textit{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers, and 3) reference these identifiers from the orders.
+Our example took the simple approach of merging entities whose attributes are all equal.
+But that might not be correct in all cases.
+What if there was a typo in one of the addresses that made it look different from other instances of the customer?
+To repair that error we need to merge the falsely distinct customers into one and fix any references from orders.
+We call this operation \textit{Merge Entities} -- note that it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.
+
+Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
+End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links} between tables but do not provide operations that can do an \textit{Extract Entity}.
+Neither do the established database products (see Related Work).
+Acme will need to write code or do a big manual conversion.
+
+It is natural to expect every schema evolution to have an inverse, for the change in requirements that triggered an evolution might change back again, and we ought to pay off the technical debt of obsolete design decisions.
+Research on \textit{Divergence Control}~\cite{Foster2007, herrmann17, chillon22} further suggests operations should be invertible.
+The inverse of \textit{Extract Entity} is \textit{Absorb Entity}.
+It raises the question of how to handle \textit{one-to-many} relationships.
+For example should absorbing orders into customers do a relational join, returning us back to the starting point, or should it nest orders within customers, as a NoSQL database might do?
+
+The inverse of \textit{Merge Entities} is \textit{Split Entity}, but that can be awkward.
+Imagine that we didn't have addresses for customers to disambiguate them with and only knew their non-unique names.
+Then \textit{Extract Entity} will conflate them and discard the information needed to properly invert the operation.
+Should this information be retained?
+One could object that we shouldn't have taken orders for people we can't uniquely identify yet.
+And even if that was necessary it means the customers really aren't distinct entities and shouldn't have been extracted.
+These are arguments that we shouldn't need \textit{Split Entity}, contradicting the principle that evolution operations should be invertible.
+It is interesting to observe that evolution in nature can also fail to invert, leaving behind vestigial features.
+We consider this to be an unsettled issue.
+
+\subsection{Code Coevolution}
+A simple example of how code should evolve along with the schema is a SQL query to report all pending orders in the original schema:
+
+\begin{lstlisting}[language=SQL]
+SELECT order_id, item, quantity, customer_name, customer_address
+FROM Orders WHERE ship_date IS NULL;
+\end{lstlisting}
+
+The \textit{Extract Entity} operation described earlier should rewrite this query into:
+
+\begin{lstlisting}[language=SQL]
+SELECT order_id, item, quantity, customer_name, customer_address
+FROM Orders
+JOIN Customers ON Orders.customer_id = Customers.customer_id
+WHERE ship_date IS NULL;
+\end{lstlisting}
+
+Alternatively the \textit{Extract Entity} operation could create a view in which the original query is executed.
+
+\subsection{Goals}
+A solution to this challenge should:
+\begin{enumerate}
+  \item Define the \textit{Extract Entity} operation.
+  \item Define the \textit{Absorb Entity} operation and motivate the tradeoff between joining and nesting.
+  \item Define a \textit{Merge Entities} operation or if using automatic deduplication then discuss how to handle mistakes.
+  \item Define the \textit{Split Entity} operation or justify why it is not needed.
+  \item If proposing a user-facing solution explain the UI affordances invoking these operations.
+  \item Extend these operations to the Divergence Control scenario.
+\end{enumerate}
 

### 2024-05-13 paper.tex
reframe multiplicity change remarks around divergence control

@@
 
-Multiplicity change clearly requires changing both the data and the code.
-It becomes particularly challenging to handle when there is a need to support ongoing writes to both the scalar and array schemas.
-This need can arise in any situation where the code and data can't all be updated atomically together—e.g., in a zero-downtime upgrade in a web architecture with stateless servers and a database, or in a decentralized setting like a public API where some consumers of the API may be expecting an outdated schema.
+Multiplicity change clearly requires changing both the data and the code. It becomes particularly challenging to handle when there is ongoing divergence between multiple versions of the data.
 
@@
 
-\subsection{Challenge}
-
-Multiplicity change illustrates a general problem for handling ongoing writes between diverging schemas.
+\subsection{Remarks}
+
+If schemas, code and data can all be updated atomically together, then multiplicity change is relatively straightforward  to handle.
+Existing scalar data can be trivially migrated to a list, and the code using the data will need to change to accommodate the list data.
+
+However, ongoing divergence makes multiplicity change much more difficult, and exposes some general challenges.
 Different schemas may expose partial information, and writers in those schemas have to operate without total knowledge of the information available in other schemas.
@@
 
-Although we see no silver bullet to navigate these tradeoffs, a good solution to this problem would give developers or users tools to manage these tradeoffs.
+Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this problem would give developers or users tools to manage these tradeoffs.
 For example, a system might allow developers to specify the desired behavior for a particular pair of schemas based on the requirements of the domain.
 Or a system might even allow users to manually disambiguate their intent for a given write.
-
-\subsection{Goals}
-
-TODO
 

### 2024-05-13 paper.tex
rewrite divergence control explanation

@@
 \subsection{Context}
-Schema change is usually concerned with migrating old data into a new schema.
-But sometimes we also want to migrate in the reverse direction: new data into an old schema.
-\citet{Cambria} discuss various ways that code changes can require bidirectional migration.
-Instead in this problem we focus on changes made by end-users.
+Schema evolution becomes more challenging when data, code, and schema cannot all be updated together in an atomic way.
+In these situations, different versions must coexist and interoperate with one another: for example, edits made in one schema must be applied in another, or code written against one schema must run against data stored in another.
+
+Such situations are common in software engineering.
+Web backend architectures often put data and code on separate machines, which makes it impossible to update both atomically.
+To perform upgrades, teams must manually perform multi-step [zero-downtime deployment processes](https://planetscale.com/blog/zero-downtime-rails-migrations-planetscale-rails-gem) to step the database and the code servers forward gradually such that they are compatible with one another at any given step.
+One example would be deploying a change to add a database column before deploying the code that uses the column.
+
+In the simplest cases, divergence can last only for a brief moments in between zero-downtime deployment steps.
+But divergence can also occur across longer timescales, especially in situations where the different parts of a system are not all centrally controlled.
+For example, a public web API may need to support applications using old versions of the API for months or years, since it is not realistic to force all consumers to upgrade immediately.
+Divergence can even be indefinite—\citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.
+
+Divergence does not just occur in software engineering; it also appears in end-user workflows.
 Users frequently make copies of structured documents like spreadsheets and then diverge them by altering both data content and schema.

### 2024-05-14 paper.tex
Harsher on Rails

@@
 Some tools provide a Domain Specific Language (DSL) to describe schema evolution.
-Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions.
-However that falls short of the range of evolutions described by \citet{ambler06}, requiring the addition of custom Ruby or SQL code.
+Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code.
 \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.

### 2024-05-15 paper.tex
the paragraph (rough)

@@
 \usepackage[backend=biber]{biblatex}
-%\usepackage{natbib}
 \addbibresource{paper.bib}
@@
 
+\textcolor{red}{[TVDS] Super rough attempt at the paragraph:}
+
+Modern day software development is collaborative. Single user programming does not exist anymore.
+Just as document editing is becoming more collaborative through online systems like pioneered by Google Docs,
+programming would benefit form online collaboration. In a sense this is live programming++: whereas traditional
+live programming systems assume a single developer editing code and see their changes immediately reflected
+in the running system, online collaborative programming is the same but for multiple programmers at the same time.
+The key challenge here, we posit, the *schema change*. We take inspiration from database migration,
+model evolution, and version control, and identify benchmark problems. This is already evident in live programming:
+editing a program, consisting of type definitions defining the structure run-time state and (transient) procedures, functions
+etc. operating on that state, requires run-time state to be migrated, after the programmer edits their program.
+Adding the dimension of multiple programs brings us into the domain local-first, and decentralized version control, where 
+copies are created and edited, which eventually have to be reconciled again (e.g., through merging). 
+Whereas schema evolution and its associated migration problems has seen a lot of attention in the database world,
+the model-driven world, and the bidirectional transformation (BX) communities, the techniques are still too low-level,
+too manual, and too much focused on a single user perspectives. By identifying the similarities between analogous
+problems in live, multi-user programming systems, we hope to stimulate further research on more comprehensive, automatic, 
+high-level, and online solutions. So that the benefits of live programming can be enjoyed in a global and distributed setting
+involving multiple programmers at the same time.
+
 \textcolor{red}{[JE] We should adopt the term Schema Evolution, which is favored in the research literature, defined to include schema change, data migration, and query rewriting. }
@@
 For example, a public web API may need to support applications using old versions of the API for months or years, since it is not realistic to force all consumers to upgrade immediately.
-Divergence can even be indefinite—\citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.
+Divergence can even be indefinite -- \citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.
 
@@
 The Extract Entity challenge for local-first software is similar to the live programming challenge except that the context is an application that offers similar functionality with peer-to-peer collaboration.
-In this context the schema change does not need to be performed interactively — a developer can take some time to build, test, and package a solution, perhaps using an API or DSL.
+In this context the schema change does not need to be performed interactively -- a developer can take some time to build, test, and package a solution, perhaps using an API or DSL.
 The hard part comes when the schema change is to be deployed across all the replicas.
@@
 
-Durable computing: pins runtime data frames to code versions. Doesn’t handle problem of global data schema change.
+Durable computing: pins runtime data frames to code versions. Doesn't handle problem of global data schema change.
 https://restate.dev/blog/solving-durable-executions-immutability-problem/
@@
 \end{minipage}
-\hspace*{2pt}\vline
+\hspace*{5pt}\vline
 \begin{minipage}[t]{0.4\textwidth}
@@
 \item all rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
-\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care needs is required if the subject of removal is the current state.
+\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care is required if the subject of removal is the current state.
 In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
@@
 \subsection{Goals}
-As above, the key requirement from the live programming perspective is to “do minimal harm”. In this case, this results in the following goals:
+As above, the key requirement from the live programming perspective is to ``do minimal harm''. In this case, this results in the following goals:
 
@@
 
-Our goal is to preserve the ability for actors in the system to read and write to a shared todo list in either the old or new schema—for example, they should be able to write to either the scalar assignee field or to the list of assignees.
+Our goal is to preserve the ability for actors in the system to read and write to a shared todo list in either the old or new schema -- for example, they should be able to write to either the scalar assignee field or to the list of assignees.
 

### 2024-05-16 paper.tex
OK it's two paragraphs

--- 2024-05-15 paper.tex
@@
 
-\section{Introduction}
-
-\textcolor{red}{[TVDS] Super rough attempt at the paragraph:}
+\section{\textcolor{red}{[TVDS] Super rough attempt at the paragraph}}
 
@@
 involving multiple programmers at the same time.
+
+\section{\textcolor{red}{[JE] Type evolution for live programming}}
+
+Many past improvements in the practice of programming worked by speeding up feedback loops.
+Live Programming aspires to the optimum of immediate feedback.
+The problem is that modern software is built out of many layers of different technologies -- thoroughly live programming would propagate changes immediately up and down the entire stack of a running system, from the database to the code, and from the client to the server.
+We call this \textit{full-stack} live programming.
+Another problem is that modern software is built with complex collaborations involving development, testing, deployment, and suppport.
+Such workflows are themselves feedback loops -- throughly live programming would propagate changes between participants frictionlessly.
+We call this \textit{end-to-end} live programming.
+
+We propose a research agenda towards this greater vision of live programming.
+The starting point is database \textit{schema evolution}, which propagates database schema changes into data migrations and query rewriting.
+We call for generalized \textit{type evolution} that works immediately across the entire software stack, and automatically across the entire software workflow.
+This paper contributes a set of challenge problems to serve as targets for research on type evolution and as a basis of comparison for the results.
+
+\section{Introduction}
 

### 2024-05-16 paper.tex
add my short summary

@@
 This paper contributes a set of challenge problems to serve as targets for research on type evolution and as a basis of comparison for the results.
+
+\section{\textcolor{red}{[GL] Rough paragraph}}
+
+Most programming languages analysis treats state as ephemeral: tied to the lifetime of a running program, and reset when the program is edited.
+But when we take a broader view, most _programming systems_ (cite) have state that lasts beyond the lifetime of a single execution.
+This includes collaboration applications with state persisted in a database, image-based systems like Smalltalk, and live programming environments where the program can be edited without resetting all state.
+
+Although these programming systems exist in very different contexts, they all encounter similar problems related to the essential difficulty of preserving state across program edits.
+New versions of the program must correctly handle pre-existing data.
+When the data schema or type definitions change, existing data must be evolved to conform to the new schema.
+Changing data representations or types can also require changing corresponding parts of the program.
+These problems have been noticed, named, and studied in these different programming systems: "schema evolution" in the databases literature, "hot reloading" in live programming, etc.
+
+In this paper, we draw connections between these different contexts, pointing out essential shared structure in the hopes of motivating further research on this important problem and encouraging knowledge sharing across subfields.
+Our contributions are: 1) a shared framework for thinking about the relationship between schema, code, and data in a programming system, 2) Case studies of the framework in a highly diverse set of programming systems, motivated by real-world examples.
 

### 2024-05-16 paper.tex
Render GL paragraph

@@
 Another problem is that modern software is built with complex collaborations involving development, testing, deployment, and suppport.
-Such workflows are themselves feedback loops -- throughly live programming would propagate changes between participants frictionlessly.
+Such workflows are themselves feedback loops -- thoroughly live programming would propagate changes between participants frictionlessly.
 We call this \textit{end-to-end} live programming.
@@
 Most programming languages analysis treats state as ephemeral: tied to the lifetime of a running program, and reset when the program is edited.
-But when we take a broader view, most _programming systems_ (cite) have state that lasts beyond the lifetime of a single execution.
+But when we take a broader view, most \textit{programming systems} (cite) have state that lasts beyond the lifetime of a single execution.
 This includes collaboration applications with state persisted in a database, image-based systems like Smalltalk, and live programming environments where the program can be edited without resetting all state.
@@
 Changing data representations or types can also require changing corresponding parts of the program.
-These problems have been noticed, named, and studied in these different programming systems: "schema evolution" in the databases literature, "hot reloading" in live programming, etc.
+These problems have been noticed, named, and studied in these different programming systems: ``schema evolution'' in the databases literature, ``hot reloading'' in live programming, etc.
 

### 2024-05-21 paper.tex
Add TP rough sketch paragraph

@@
 Our contributions are: 1) a shared framework for thinking about the relationship between schema, code, and data in a programming system, 2) Case studies of the framework in a highly diverse set of programming systems, motivated by real-world examples.
+
+\section{\textcolor{red}{[TP] Rough paragraph}}
+
+Programming is increasingly collaborative and interactive. We welcome this trend, because
+greater collaboration and interactivity make programming more effective. Future programming
+environments enable programmers to collaborate at a more fine-grained level, adopting individual
+code changes created by their collaborators as needed. They also let programmers interactively
+edit code, often during program execution.
+
+In order to support this new way of working, programming systems increasingly have to tackle the
+problem of schema evolution. If a programmer imports a change that edits a schema, code and data
+that relies on such schema needs to be adapted to match the new schema. Similarly, if a programmer
+in a live programming environment modifies their type definition, current program state needs to
+be updated to match the new structure of types. This is a well-understood and well-studied problem
+in database community, but it has received little attention in work on programming systems.
+
+The aim of this paper is to provide a conceptual framework for talking about schema evolution in
+collaborative and live programming systems. Taking inspiration from Stewart Brand's analysis of
+pace layers, we propose to think of the problem as the problem of evolution of occuring at three
+layers, the most permanent schema layer, less permanent code layer and least permanent data layer.
+A change at the most permanent layer affects the less permanent layer, which then has to be
+updated to keep correspondence between the layers.
+
+We consider four case studies of such schema evolution in collaborative and live programming
+systems. We present those throught the perspective of our unified conceptual framework. The work
+both sheds a new light at how schema change needs to be handled in programming systems and also
+provides a range of challenges that show interesting problems that future programming systems
+will need to tackle.
 

### 2024-05-23 paper.tex
Add unification notes and diagrams
arrow diagrams

@@
 \usepackage[many]{tcolorbox}
+\usepackage{tikz-cd}
+
 \lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
@@
 
+% UNIFIED VERSION
+%
+% 1. past improvements & feedback loops
+% 2. for this we need to transcend treating state as ephemeral and adopt systems view
+% 3. two current directions - live and collaborative
+% 4. we welcome this development, but it poses challenges for programming systems
+% 5. schema change in databases, hot reloading in programming, dynamic software update
+% (maybe) pace layers and structure
+% 6. contributions are a (i) framework and (ii) challenges
+
 \section{Introduction}
@@
 \end{enumerate}
+
+\section{diagrams}
+
+\begin{figure}
+
+  \begin{tikzcd}[sep = huge]
+    schema_1  \arrow[r, "constrains"] \arrow[d, "change"{left},""{name=L}]& data_1 \arrow[d, ""{name=R, left}, "migration"{right}] \\
+    schema_2 \arrow[r, "constrains"{below}] & data_2
+    \arrow[Rightarrow, from=L, to=R, "propagate"]
+  \end{tikzcd}
+
+  \begin{tikzcd}[sep = huge]
+    schema_1  \arrow[r, "constrains"] \arrow[d, "change"{left},""{name=L}]& code_1 \arrow[d, ""{name=R, left}, "refactor"{right}] \\
+    schema_2 \arrow[r, "constrains"{below}] & code_2
+    \arrow[Rightarrow, from=L, to=R, "propagate"]
+  \end{tikzcd}
+
+  \caption{Commutative diagrams.}
+  \label{fig:diagrams}
+\end{figure}
 

### 2024-05-23 paper.tex
Edits in intro, update arrow sketches

--- 2024-05-21 paper.tex
@@
 \usepackage{inputenx}
-\usepackage[backend=biber]{biblatex}
+\usepackage[backend=bibtex]{biblatex}
 \addbibresource{paper.bib}
@@
 
-\title{Schema Change: Challenge Problems}
+\title{Type Evolution in Programming Systems}
 
@@
 
-\section{\textcolor{red}{[TVDS] Super rough attempt at the paragraph}}
-
-Modern day software development is collaborative. Single user programming does not exist anymore.
-Just as document editing is becoming more collaborative through online systems like pioneered by Google Docs,
-programming would benefit form online collaboration. In a sense this is live programming++: whereas traditional
-live programming systems assume a single developer editing code and see their changes immediately reflected
-in the running system, online collaborative programming is the same but for multiple programmers at the same time.
-The key challenge here, we posit, the *schema change*. We take inspiration from database migration,
-model evolution, and version control, and identify benchmark problems. This is already evident in live programming:
-editing a program, consisting of type definitions defining the structure run-time state and (transient) procedures, functions
-etc. operating on that state, requires run-time state to be migrated, after the programmer edits their program.
-Adding the dimension of multiple programs brings us into the domain local-first, and decentralized version control, where
-copies are created and edited, which eventually have to be reconciled again (e.g., through merging).
-Whereas schema evolution and its associated migration problems has seen a lot of attention in the database world,
-the model-driven world, and the bidirectional transformation (BX) communities, the techniques are still too low-level,
-too manual, and too much focused on a single user perspectives. By identifying the similarities between analogous
-problems in live, multi-user programming systems, we hope to stimulate further research on more comprehensive, automatic,
-high-level, and online solutions. So that the benefits of live programming can be enjoyed in a global and distributed setting
-involving multiple programmers at the same time.
-
-\section{\textcolor{red}{[JE] Type evolution for live programming}}
-
-Many past improvements in the practice of programming worked by speeding up feedback loops.
-Live Programming aspires to the optimum of immediate feedback.
-The problem is that modern software is built out of many layers of different technologies -- thoroughly live programming would propagate changes immediately up and down the entire stack of a running system, from the database to the code, and from the client to the server.
-We call this \textit{full-stack} live programming.
-Another problem is that modern software is built with complex collaborations involving development, testing, deployment, and suppport.
-Such workflows are themselves feedback loops -- thoroughly live programming would propagate changes between participants frictionlessly.
-We call this \textit{end-to-end} live programming.
-
-We propose a research agenda towards this greater vision of live programming.
-The starting point is database \textit{schema evolution}, which propagates database schema changes into data migrations and query rewriting.
-We call for generalized \textit{type evolution} that works immediately across the entire software stack, and automatically across the entire software workflow.
-This paper contributes a set of challenge problems to serve as targets for research on type evolution and as a basis of comparison for the results.
-
-\section{\textcolor{red}{[GL] Rough paragraph}}
-
-Most programming languages analysis treats state as ephemeral: tied to the lifetime of a running program, and reset when the program is edited.
-But when we take a broader view, most \textit{programming systems} (cite) have state that lasts beyond the lifetime of a single execution.
-This includes collaboration applications with state persisted in a database, image-based systems like Smalltalk, and live programming environments where the program can be edited without resetting all state.
-
-Although these programming systems exist in very different contexts, they all encounter similar problems related to the essential difficulty of preserving state across program edits.
-New versions of the program must correctly handle pre-existing data.
-When the data schema or type definitions change, existing data must be evolved to conform to the new schema.
-Changing data representations or types can also require changing corresponding parts of the program.
-These problems have been noticed, named, and studied in these different programming systems: ``schema evolution'' in the databases literature, ``hot reloading'' in live programming, etc.
-
-In this paper, we draw connections between these different contexts, pointing out essential shared structure in the hopes of motivating further research on this important problem and encouraging knowledge sharing across subfields.
-Our contributions are: 1) a shared framework for thinking about the relationship between schema, code, and data in a programming system, 2) Case studies of the framework in a highly diverse set of programming systems, motivated by real-world examples.
-
-\section{\textcolor{red}{[TP] Rough paragraph}}
-
-Programming is increasingly collaborative and interactive. We welcome this trend, because
-greater collaboration and interactivity make programming more effective. Future programming
-environments enable programmers to collaborate at a more fine-grained level, adopting individual
-code changes created by their collaborators as needed. They also let programmers interactively
-edit code, often during program execution.
-
-In order to support this new way of working, programming systems increasingly have to tackle the
-problem of schema evolution. If a programmer imports a change that edits a schema, code and data
-that relies on such schema needs to be adapted to match the new schema. Similarly, if a programmer
-in a live programming environment modifies their type definition, current program state needs to
-be updated to match the new structure of types. This is a well-understood and well-studied problem
-in database community, but it has received little attention in work on programming systems.
-
-The aim of this paper is to provide a conceptual framework for talking about schema evolution in
-collaborative and live programming systems. Taking inspiration from Stewart Brand's analysis of
-pace layers, we propose to think of the problem as the problem of evolution of occuring at three
-layers, the most permanent schema layer, less permanent code layer and least permanent data layer.
-A change at the most permanent layer affects the less permanent layer, which then has to be
-updated to keep correspondence between the layers.
-
-We consider four case studies of such schema evolution in collaborative and live programming
-systems. We present those throught the perspective of our unified conceptual framework. The work
-both sheds a new light at how schema change needs to be handled in programming systems and also
-provides a range of challenges that show interesting problems that future programming systems
-will need to tackle.
-
 \section{Introduction}
 
-\textcolor{red}{[JE] We should adopt the term Schema Evolution, which is favored in the research literature, defined to include schema change, data migration, and query rewriting. }
-
-Schema change won't go away.
-Changing requirements and changing code lead to changing the schema of a database.
-Schema change, also called schema migration, is the problem of migrating existing data from the old schema to the new.
-This often involves custom migration programs or specialized Domain Specific Langiuages.
-The migration must be carefully coordinated with upgrading the application code and associated artifacts that assume the new schema.
-Schema change on database servers is often delegated to Database Administrators and DevOps.
-Live programming~\cite{tanimoto90,Hancock03} and local-first software~\cite{localfirst} both move data away from the server, eliminating those jobs, but not eliminating the need to do schema change, and indeed increasing the need to do it automatically.
-Recently \citet{Cambria} spotlighted these problems and proposed an approach using lenses~\cite{Foster2007}, but otherwise there has been surprisingly little research.
-To promote further progress we offer a set of challenge problems to the live programming and local-first communities, inviting them to propose and compare solutions.
-
-Live programming seeks to erase the boundary between editing and running programs.
-In order to do so program data must be kept around while the program is being edited.
-Classic Lisp and Smalltalk systems integrated code and data into a persistent \textit{image}~\cite{Sandewall78, Goldberg80}.
-An interactive shell or \textit{Read Eval Print Loop (REPL)}~\cite{Deutsch64} is more transient than an image, but still builds up a context of data over long-lived programming sessions the loss of which disrupts the programmer's workflow.
-In all of these environments programs eventually get changed to create and expect data in a form incompatible with extant data.
-This happens whether or not the form of the data is specified explicitly in a type system or database schema, or whether it is left implicit in the behavior of the code.
-Some languages can tolerate a larger set of such changes, most notably Smalltalk which will automatically insert and delete members in existing instances when a class definition is changed~\cite[pp.252-272]{Goldberg80}.\footnote{Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a sophisticated schema change API~\cite{Gemstone}.
-Schema change won't go away.} But there is still a large class of changes that create incompatibilities with existing data, whether it is inside an image, programming environment, REPL, or an external database.
-Some live programming environments generate data with unit tests, but that only shifts the problem of schema change to adapting those tests.
-One way or another, stopping everything to manually deal with schema change contradicts the goals of live programming.
-Schema change won't go away.
-
-Local-first software faces related problems.
-Its goal is to empower users by moving code and data from the cloud to the user's own devices.
-Distributed programming techniques like Convergent Replicated Data Types (CRDTs)~\cite{Shapiro11} are used to coordinate data changes peer-to-peer.
-Unfortunately these techniques so far have not addressed schema change nor code deployment The traditional techniques of schema change used in centrally managed databases are complicated by the distributed and intermittently connected nature of local-first data.
-Schema change won't go away.
-
-In the following sections we present a series of challenge problems dealing with schema change in the context of live programming and local-first software.
-These problems are necessarily expressed using established idioms or conventions, but nevertheless we welcome solutions that translate the spirit of the problem into other contexts.
-
-~
-
-~
-
-\textcolor{red}{TODO: Need to explain that we take the programming system view - program is not something
-built centrally, but something that exists and can be modified}
-
-~
-
-\textcolor{red}{TODO: Maybe the right way to frame this is to say that the challenges are intended
-less for evaluating particular solutions but more for framing and making sense of different
-kinds of schema-change-related problems.  You can read this paper and when you face a schema-change-related
-problem, think about it in terms of our dimensions. Maybe compare it to our challenges.
-We do not necessarilly expect that people will directly solve our challenges though. (Although
-some might and that would be nice..)}
-
-~
-
-\textcolor{red}{TODO: I also guess this is more about just ``change'' than ``schema change''
-which is one particular (most challenging) instance - i.e., change at the most permanent
-pace layer.}
-
-\newpage
-\section{Dimensions of change}
+Many past improvements in the practice of programming worked by speeding up feedback loops,
+both in collaboration with other programmers (as in collaborative development tools
+\cite{goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
+programming environment (as in live programming \cite{rein2018exploratory}).
+
+To further tighten the feedback loop and make programming more effective, future programming
+systems will need to enable programmers to collaborate at a fine-grained level, letting them
+adopt individual changes created by their collaborators (as in local-first software
+\cite{localfirst}). To provide immediate feedback, future programming systems will need to
+transcend treating state as ephemeral, tied to the lifetime of a running program and reset when the
+program is edited.
+
+We need to shift our focus to \emph{programming systems} \cite{techdims} where program state
+is persistent and co-evolves with the behavior and data of the system. This is already the case
+in collaborative applications with state persistent in a database, image-based systems like
+Smalltalk and live programming environments with hot-reloading where the program can be edited
+without resetting all state.
+
+While we are firmly convinced that more collaborative and live programming systems are the
+future of programming, there are a number of problems that need to be resolved in order for
+such systems to become a reality. The key problem is of \emph{type evolution}, i.e., how to
+handle the change of schema of the data program works with in a way that preserves integrity
+constraints, existing data and the logic of code.
+
+Type evolution is not a new problem. It exists in various forms in a range of existing systems.
+In the database world, the problem of schema evolution is well-studied \cite{erhard06}.
+It typically focuses on ensuring data integrity, although co-evolution of database structure
+and code that manipulates the database has also been studied \cite{Cleve2006,wang2019synthesizing}.
+In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
+although it typically focuses on updating transient state and ignores the problem of more complex
+data migrations. The problem is equally studied in the bidirectional transformation (BX)
+communities \cite{czarnecki2009bidirectional} and in work on model-driven development
+\cite{metamodels,alanen2003difference}.
+
+Type evolution is an important challenge for future programming systems. But looking at the problem
+from the perspective of programming systems can also offer a new perspective that
+unfies the many different aspects of the problem partially explored in other areas. A programming
+system may have permanent data (as in a database) whose structure can change, requiring
+corresponding change to data and code. A programming system may have a transient state conforming
+to a type structure that may change during live programming. A programming system may also need
+to allow programmers to independently work on different versions of a program.
+
+In this paper, we draw connections between these different kinds of type evolution, pointing out
+essential shared structure. By doing so, we aim to motivate further research on this important
+problem and encouraging knowledge sharing across subfields. Our contributions are twofold.
+First, we present a shared conceptual framework for thinking about the relationship between
+schema, code, and data in a programming system. Second, we present five case studies that use
+the framework to study type evolution in a highly diverse set of real-world programming scenarios.
 
 \begin{figure}[t]
+\centering
 \vspace{-1em}
-\centering
-\includegraphics[width=0.5\textwidth]{figures/layers.png}
+\includegraphics[width=0.45\textwidth]{figures/layers.png}
 \caption{Pace layers from Stewart Brand's classic \emph{How Buildings Learn} \cite{Brand95}.
-  The outer layers rarely change, while the inner layers change frequently. The geographic
+  Different layers of a building change at different pace. The geographic
   site of a building is eternal, its structure remains stable for decades, space plan
-  needs to change every couple of years. ``Because of the different rates of change of its
-  components, a building is always tearing itself apart.''
-  \textcolor{red}{[JE] Except the outer skin changes faster than the structure.}
-  }
+  changes every couple of years.}
 \label{fig:layers}
@@
 
-\textcolor{red}{[JE] `change in a program' might be misunderstood as changes to the source code. Perhaps `system' or `software system' might be better than `program'? }
-
-In the broadest sense, this paper is concerned with the difficulties posed by a change
-in a program. The first dimension of difficulties that we consider in this paper is
-local. When something in a program changes, other parts of the program typically
-need to be brought to sync with the changed part. The complexity of the problem varies.
-If nothing else in the program depends on the changed part, for example when changing
-a runtime value of an object field, the change poses no difficulty.
-If many other parts of the program depend on the changed part, the difficulty is greater.
-For example, if we change the schema of data structures used in the program, we also need to
-change code that works with it and existing data that the users created. To structure our
-discussion about this part of the problem, we will use the idea of pace layers introduced by
-Brand \cite{Brand95,Brand18} (Figure~\ref{fig:layers}).
-
-The second dimension that we consider in this paper is non-local. A program may have multiple
+\section{Type Evolution}
+To paraphrase Stewart Brand \cite{Brand95}, ``almost no programs evolve well. But all programs
+(except for monuments) evolve anyway, however poorly, because the usages in and around them are
+changing constantly.'' Programs evolve both while they are being created and during their operation
+in response to new needs. As with buildings (Figure~\ref{fig:layers}), different aspects of
+program evolve at different pace. And ``because of the different rates of change of its components,
+a program is always tearing itself apart.''
+
+\subsection{Types, Code and Data}
+
+As is the case for buildings, making a change at some layers (skin or stuff) is relatively easy,
+but a change at a more basic layer (structure) has cascading effects on layers that depend on it.
+In the case of programs, a more basic layer is a structure that defines the shape of data and code.
+We refer to this structure as \emph{types}, although we do not use the term in an precise technical
+sense. In our conceptual framework, types may be data types that define the shape of data. They
+may be explicitly declared in code and checked statically, as well as dynamically checked or
+even implicit. The key property of types for us is that they define the shape of other layers
+of a program, such as data and code. As a result, a change of program types requires a
+corresponding change in program data and code.
+
+The situation is illustrated in Figure~\ref{fig:change} using two variations of a diagram that
+we will return to repeatedly in this paper. When a type changes from $T_1$ to $T_2$,
+we need to find and apply a corresponding change to other parts of the program whose shape
+is defined by the changed type. The problem of \emph{type evolution} is thus finding and
+applying changes that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$
+that correspond to the new type.
+
+The complexity of the problem varies. The first dimension of challenges that we consider
+is local. The complexity of the challenge depends on how many other parts of the program
+depend on the modified type. If both code and data depend on the type, the complexity of the
+program is greater. If there is no associated code or the existing data can be safely discarded,
+the complexity is lesser.
+
+The second dimension of challenges that we consider is non-local. A program may have multiple
 variants that exist independently and need to be synchronized. The difficulty of the problem
-depends on what kind of synchronization we want to support. In the easier case, all variants
-of the program converge, i.e.~all variants should eventually adopt all changes. In the more
-challenging case, variants of a program may diverge, i.e.~a user may want to apply only certain
-changes from other users. Note the non-local dimension may interact with the pace layers of the
-local dimension. A typical case may be that the code and schema of a program should converge,
-while data and the transient state of a program may diverge.
+depends on what kind of synchronization we want to support. If all variants of the program should
+eventually adopt all changes (converge), the complexity is lesser. If we allow user to apply
+only certain changes from other users (divergence), the complexity is greater. A typical case
+may be that the types and schema of a program should converge, but data may diverge.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.65\textwidth]{figures/arrows.png}
+\caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
+  in data and code whose shape is defined by type. The challenge is finding a suitable
+  corresponding transformation (semi-)automatically.}
+\label{fig:change}
+\end{figure}
+
+\subsection{Conceptual Framework}
+
+~
+
+\textcolor{red}{TODO: got as far as here so far...}
+
+~
 
@@
 
+% elm
+% database
+% divergence
+% cambria
+% documents
+% state machine
+
+% DIVERGENCE IN TRANSITION (because you cannot do big bang update) c.f. Tijs' example
+
+% typically people just blow away the state but Elm actually has demos
+% of people doing hot reload without losing state.
+% (maybe they never change the type?)
+
+% ELM EXAMPLE
+% current example ("completes") is contrived - it is equivalent data
+% more interesting is when you lose / add data
+% add "completed" with default instead
+
 \section{Divergence Control}
@@
 
+TODO: Maybe insert something about methodology in general?
+(ref the table challenges paper? although this paper is no longer so much about challenges)
+
+\paragraph{Schema evolution.}
 Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help.
@@
 Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.
+
+\paragraph{Live, MDD, etc. programming?}
+
+% Some references from Tomas
+%
+% collaborative programmign tools
+\cite{goldman2011real,kurniawan2015coder,replit}
+% exploratory/live review paper
+\cite{rein2018exploratory}
+% hot reload in live systems
+\cite{barenz2020essence,beckmann2021shortening}
+
+% Some references from Tijs
+%
+% SQL+code
+\cite{Cleve2006,wang2019synthesizing}
+% BX
+\cite{czarnecki2009bidirectional}
+% MDD
+\cite{metamodels,alanen2003difference}.
 
@@
 
+\appendix
+
+\section{\textcolor{red}{[TVDS] Super rough attempt at the paragraph}}
+
+Modern day software development is collaborative. Single user programming does not exist anymore.
+Just as document editing is becoming more collaborative through online systems like pioneered by Google Docs,
+programming would benefit form online collaboration. In a sense this is live programming++: whereas traditional
+live programming systems assume a single developer editing code and see their changes immediately reflected
+in the running system, online collaborative programming is the same but for multiple programmers at the same time.
+The key challenge here, we posit, the *schema change*. We take inspiration from database migration,
+model evolution, and version control, and identify benchmark problems. This is already evident in live programming:
+editing a program, consisting of type definitions defining the structure run-time state and (transient) procedures, functions
+etc. operating on that state, requires run-time state to be migrated, after the programmer edits their program.
+Adding the dimension of multiple programs brings us into the domain local-first, and decentralized version control, where
+copies are created and edited, which eventually have to be reconciled again (e.g., through merging).
+Whereas schema evolution and its associated migration problems has seen a lot of attention in the database world,
+the model-driven world, and the bidirectional transformation (BX) communities, the techniques are still too low-level,
+too manual, and too much focused on a single user perspectives. By identifying the similarities between analogous
+problems in live, multi-user programming systems, we hope to stimulate further research on more comprehensive, automatic,
+high-level, and online solutions. So that the benefits of live programming can be enjoyed in a global and distributed setting
+involving multiple programmers at the same time.
+
+\section{\textcolor{red}{[JE] Type evolution for live programming}}
+
+Many past improvements in the practice of programming worked by speeding up feedback loops.
+Live Programming aspires to the optimum of immediate feedback.
+The problem is that modern software is built out of many layers of different technologies -- thoroughly live programming would propagate changes immediately up and down the entire stack of a running system, from the database to the code, and from the client to the server.
+We call this \textit{full-stack} live programming.
+Another problem is that modern software is built with complex collaborations involving development, testing, deployment, and suppport.
+Such workflows are themselves feedback loops -- thoroughly live programming would propagate changes between participants frictionlessly.
+We call this \textit{end-to-end} live programming.
+
+We propose a research agenda towards this greater vision of live programming.
+The starting point is database \textit{schema evolution}, which propagates database schema changes into data migrations and query rewriting.
+We call for generalized \textit{type evolution} that works immediately across the entire software stack, and automatically across the entire software workflow.
+This paper contributes a set of challenge problems to serve as targets for research on type evolution and as a basis of comparison for the results.
+
+\section{\textcolor{red}{[GL] Rough paragraph}}
+
+Most programming languages analysis treats state as ephemeral: tied to the lifetime of a running program, and reset when the program is edited.
+But when we take a broader view, most \textit{programming systems} (cite) have state that lasts beyond the lifetime of a single execution.
+This includes collaboration applications with state persisted in a database, image-based systems like Smalltalk, and live programming environments where the program can be edited without resetting all state.
+
+Although these programming systems exist in very different contexts, they all encounter similar problems related to the essential difficulty of preserving state across program edits.
+New versions of the program must correctly handle pre-existing data.
+When the data schema or type definitions change, existing data must be evolved to conform to the new schema.
+Changing data representations or types can also require changing corresponding parts of the program.
+These problems have been noticed, named, and studied in these different programming systems: ``schema evolution'' in the databases literature, ``hot reloading'' in live programming, etc.
+
+In this paper, we draw connections between these different contexts, pointing out essential shared structure in the hopes of motivating further research on this important problem and encouraging knowledge sharing across subfields.
+Our contributions are: 1) a shared framework for thinking about the relationship between schema, code, and data in a programming system, 2) Case studies of the framework in a highly diverse set of programming systems, motivated by real-world examples.
+
+\section{\textcolor{red}{[TP] Rough paragraph}}
+
+Programming is increasingly collaborative and interactive. We welcome this trend, because
+greater collaboration and interactivity make programming more effective. Future programming
+environments enable programmers to collaborate at a more fine-grained level, adopting individual
+code changes created by their collaborators as needed. They also let programmers interactively
+edit code, often during program execution.
+
+In order to support this new way of working, programming systems increasingly have to tackle the
+problem of schema evolution. If a programmer imports a change that edits a schema, code and data
+that relies on such schema needs to be adapted to match the new schema. Similarly, if a programmer
+in a live programming environment modifies their type definition, current program state needs to
+be updated to match the new structure of types. This is a well-understood and well-studied problem
+in database community, but it has received little attention in work on programming systems.
+
+The aim of this paper is to provide a conceptual framework for talking about schema evolution in
+collaborative and live programming systems. Taking inspiration from Stewart Brand's analysis of
+pace layers, we propose to think of the problem as the problem of evolution of occuring at three
+layers, the most permanent schema layer, less permanent code layer and least permanent data layer.
+A change at the most permanent layer affects the less permanent layer, which then has to be
+updated to keep correspondence between the layers.
+
+We consider four case studies of such schema evolution in collaborative and live programming
+systems. We present those throught the perspective of our unified conceptual framework. The work
+both sheds a new light at how schema change needs to be handled in programming systems and also
+provides a range of challenges that show interesting problems that future programming systems
+will need to tackle.
+
+\section{Introduction}
+
+\textcolor{red}{[JE] We should adopt the term Schema Evolution, which is favored in the research literature, defined to include schema change, data migration, and query rewriting. }
+
+Schema change won't go away.
+Changing requirements and changing code lead to changing the schema of a database.
+Schema change, also called schema migration, is the problem of migrating existing data from the old schema to the new.
+This often involves custom migration programs or specialized Domain Specific Langiuages.
+The migration must be carefully coordinated with upgrading the application code and associated artifacts that assume the new schema.
+Schema change on database servers is often delegated to Database Administrators and DevOps.
+Live programming~\cite{tanimoto90,Hancock03} and local-first software~\cite{localfirst} both move data away from the server, eliminating those jobs, but not eliminating the need to do schema change, and indeed increasing the need to do it automatically.
+Recently \citet{Cambria} spotlighted these problems and proposed an approach using lenses~\cite{Foster2007}, but otherwise there has been surprisingly little research.
+To promote further progress we offer a set of challenge problems to the live programming and local-first communities, inviting them to propose and compare solutions.
+
+Live programming seeks to erase the boundary between editing and running programs.
+In order to do so program data must be kept around while the program is being edited.
+Classic Lisp and Smalltalk systems integrated code and data into a persistent \textit{image}~\cite{Sandewall78, Goldberg80}.
+An interactive shell or \textit{Read Eval Print Loop (REPL)}~\cite{Deutsch64} is more transient than an image, but still builds up a context of data over long-lived programming sessions the loss of which disrupts the programmer's workflow.
+In all of these environments programs eventually get changed to create and expect data in a form incompatible with extant data.
+This happens whether or not the form of the data is specified explicitly in a type system or database schema, or whether it is left implicit in the behavior of the code.
+Some languages can tolerate a larger set of such changes, most notably Smalltalk which will automatically insert and delete members in existing instances when a class definition is changed~\cite[pp.252-272]{Goldberg80}.\footnote{Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a sophisticated schema change API~\cite{Gemstone}.
+Schema change won't go away.} But there is still a large class of changes that create incompatibilities with existing data, whether it is inside an image, programming environment, REPL, or an external database.
+Some live programming environments generate data with unit tests, but that only shifts the problem of schema change to adapting those tests.
+One way or another, stopping everything to manually deal with schema change contradicts the goals of live programming.
+Schema change won't go away.
+
+Local-first software faces related problems.
+Its goal is to empower users by moving code and data from the cloud to the user's own devices.
+Distributed programming techniques like Convergent Replicated Data Types (CRDTs)~\cite{Shapiro11} are used to coordinate data changes peer-to-peer.
+Unfortunately these techniques so far have not addressed schema change nor code deployment The traditional techniques of schema change used in centrally managed databases are complicated by the distributed and intermittently connected nature of local-first data.
+Schema change won't go away.
+
+In the following sections we present a series of challenge problems dealing with schema change in the context of live programming and local-first software.
+These problems are necessarily expressed using established idioms or conventions, but nevertheless we welcome solutions that translate the spirit of the problem into other contexts.
+
+~
+
+~
+
+\textcolor{red}{TODO: Need to explain that we take the programming system view - program is not something
+built centrally, but something that exists and can be modified}
+
+~
+
+\textcolor{red}{TODO: Maybe the right way to frame this is to say that the challenges are intended
+less for evaluating particular solutions but more for framing and making sense of different
+kinds of schema-change-related problems.  You can read this paper and when you face a schema-change-related
+problem, think about it in terms of our dimensions. Maybe compare it to our challenges.
+We do not necessarilly expect that people will directly solve our challenges though. (Although
+some might and that would be nice..)}
+
+~
+
+\textcolor{red}{TODO: I also guess this is more about just ``change'' than ``schema change''
+which is one particular (most challenging) instance - i.e., change at the most permanent
+pace layer.}
+
 \end{document}

### 2024-05-23 paper.tex
combo diagram

--- 2024-05-23 paper.tex
@@
 
-\begin{figure}
-
   \begin{tikzcd}[sep = huge]
-    schema_1  \arrow[r, "constrains"] \arrow[d, "change"{left},""{name=L}]& data_1 \arrow[d, ""{name=R, left}, "migration"{right}] \\
-    schema_2 \arrow[r, "constrains"{below}] & data_2
-    \arrow[Rightarrow, from=L, to=R, "propagate"]
+    \textsf{code}_1
+    \arrow[d, "\textsf{refactor}"{left},""{name=L}]
+    &
+    \textsf{schema}_1
+    \arrow[l, "\textsf{constrains}"{above}]
+    \arrow[r, "\textsf{constrains}"]
+    \arrow[d, red, "\textsf{change}"{near start, description}, ""{name=CL, left}, ""{name=CR}]
+    &
+    \textsf{data}_1
+    \arrow[d, ""{name=R, left}, "\textsf{migrate}"{right}]
+    \\
+    \textsf{code}_2
+    &
+    \textsf{schema}_2
+    \arrow[l, "\textsf{constrains}"{below}]
+    \arrow[r, "\textsf{constrains}"{below}] &
+    \textsf{data}_2
+    \arrow[Rightarrow, from=CL, to=L, "\textsf{propagate}"{below}]
+    \arrow[Rightarrow, from=CR, to=R, "\textsf{propagate}"{below}]
   \end{tikzcd}
-
-  \begin{tikzcd}[sep = huge]
-    schema_1  \arrow[r, "constrains"] \arrow[d, "change"{left},""{name=L}]& code_1 \arrow[d, ""{name=R, left}, "refactor"{right}] \\
-    schema_2 \arrow[r, "constrains"{below}] & code_2
-    \arrow[Rightarrow, from=L, to=R, "propagate"]
-  \end{tikzcd}
-
-  \caption{Commutative diagrams.}
-  \label{fig:diagrams}
-\end{figure}
 

### 2024-05-24 paper.tex
3D divergence diagram

--- 2024-05-23 paper.tex
@@
 \usepackage[many]{tcolorbox}
+\usepackage{tikz-cd}
+\usepackage{float}
+
 \lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
@@
 \end{enumerate}
+
+\section{diagrams}
+
+\begin{figure}[H]
+  \begin{tikzcd}[sep = 6em]
+    \textsf{query}
+    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
+    &
+    \textsf{schema}
+    \arrow[l, "\textsf{constrains}"{above}]
+    \arrow[r, "\textsf{constrains}"{above}]
+    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
+    &
+    \textsf{data}
+    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
+    \\
+    \textsf{query}_\alpha
+    &
+    \textsf{schema}_\alpha
+    \arrow[l, "\textsf{constrains}"{above}]
+    \arrow[r, "\textsf{constrains}"{above}]
+    &
+    \textsf{data}_\alpha
+    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
+    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
+  \end{tikzcd}
+  \caption{Extract Entity}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{tikzcd}[column sep = large, row sep = 5em]
+    &
+    \textsf{schema}
+    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
+    \arrow[rr, "\textsf{constrains}"]
+    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
+    &
+    &
+    \textsf{data}
+    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
+    \arrow[Rightarrow, from=L, to=R]
+    \\
+    \textsf{schema}_\alpha
+    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
+    \arrow[rr, crossing over]
+    &
+    &
+    \textsf{data}_\alpha
+    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
+    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
+    &
+    \\
+    &
+    \textsf{schema}_\beta
+    \arrow[ld, red, "\mathit{merge}\left(\textsf{change}_\alpha {\,,\;} \textsf{change}_\beta\right)"{name=L4, near start, description}]
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_\beta
+    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
+    \arrow[Rightarrow, from=L4, to=R4]
+    \\
+    \textsf{schema}_{\alpha\beta}
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_{\alpha\beta}
+    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
+    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
+    &
+  \end{tikzcd}
+  \caption{Mergeing divergent changes}
+\end{figure}
 

### 2024-05-24 paper.tex
Rewrite framework, migrate some of Elm

--- 2024-05-23 paper.tex
@@
   keywordstyle=\firamedium,
-  stringstyle=\color{RosyBrown},%
-  texcsstyle=*{\color{Purple}\mdseries},%
-  texcsstyle=*[2]{\color{Blue1}},%
-  texcsstyle=*[3]{\color{ForestGreen}},%
-  commentstyle={\color{FireBrick}},%
+  belowskip=1em,
+  aboveskip=1em,
+  framesep=0.5em,
+  basicstyle=\normalsize\ttfamily,
+  %stringstyle=\color{RosyBrown},%
+  %texcsstyle=*{\color{Purple}\mdseries},%
+  %texcsstyle=*[2]{\color{Blue1}},%
+  %texcsstyle=*[3]{\color{ForestGreen}},%
+  %commentstyle={\color{FireBrick}},%
+  numbersep=1em,
+  xleftmargin=1.5em,
   escapechar=`,}
@@
 \newcounter{challengen}
-\newcommand{\challenge}[1]{\subsection*{\addtocounter{challengen}{1} Challenge \#\thechallengen: #1}}
+\newcommand{\challenge}[1]{\subsection*{\addtocounter{challengen}{1}Challenge \#\thechallengen: #1}}
 
@@
   width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
-  outer arc=0pt]\setlength{\parskip}{0.5em}\setlength{\parindent}{0em}\textbf{\sffamily Framework perspective.}\quad #2
+  outer arc=0pt]\setlength{\parskip}{0.5em}\setlength{\parindent}{0em}{\firamedium Framework perspective.}\quad #2
 \end{tcolorbox}}
@@
 Many past improvements in the practice of programming worked by speeding up feedback loops,
-both in collaboration with other programmers (as in collaborative development tools
+both in collaboration with other programmers (through collaborative development tools
 \cite{goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
-programming environment (as in live programming \cite{rein2018exploratory}).
-
-To further tighten the feedback loop and make programming more effective, future programming
-systems will need to enable programmers to collaborate at a fine-grained level, letting them
-adopt individual changes created by their collaborators (as in local-first software
-\cite{localfirst}). To provide immediate feedback, future programming systems will need to
+programming environment (through live programming \cite{rein2018exploratory}).
+
+To make programming more effective, future programming systems will need to further tighten
+the feedback loop. They will need to enable programmers to collaborate at a fine-grained level,
+letting them adopt individual changes created by their collaborators (as in local-first software
+\cite{localfirst}). To provide immediate feedback, future programming systems will also need to
 transcend treating state as ephemeral, tied to the lifetime of a running program and reset when the
@@
 
-We need to shift our focus to \emph{programming systems} \cite{techdims} where program state
-is persistent and co-evolves with the behavior and data of the system. This is already the case
-in collaborative applications with state persistent in a database, image-based systems like
-Smalltalk and live programming environments with hot-reloading where the program can be edited
-without resetting all state.
+To conceive of such systems, we need to shift our focus to \emph{programming systems} \cite{techdims}
+where program state is persistent and co-evolves with the behavior and data of the system. This is
+already the case in collaborative applications with state persistent in a database, image-based
+systems like Smalltalk and live programming environments with hot-reloading where the program can
+be edited without resetting all state.
 
@@
 future of programming, there are a number of problems that need to be resolved in order for
-such systems to become a reality. The key problem is of \emph{type evolution}, i.e., how to
+such systems to become a reality. The key problem is \emph{type evolution}, i.e., how to
 handle the change of schema of the data program works with in a way that preserves integrity
@@
 Type evolution is not a new problem. It exists in various forms in a range of existing systems.
-In the database world, the problem of schema evolution is well-studied \cite{erhard06}.
-It typically focuses on ensuring data integrity, although co-evolution of database structure
-and code that manipulates the database has also been studied \cite{Cleve2006,wang2019synthesizing}.
+The problem of schema evolution in databases is well-studied \cite{erhard06}.
+It typically considers data integrity, although co-evolution of database schema
+and code has also been studied \cite{Cleve2006,wang2019synthesizing}.
 In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
 although it typically focuses on updating transient state and ignores the problem of more complex
-data migrations. The problem is equally studied in the bidirectional transformation (BX)
-communities \cite{czarnecki2009bidirectional} and in work on model-driven development
-\cite{metamodels,alanen2003difference}.
-
-Type evolution is an important challenge for future programming systems. But looking at the problem
-from the perspective of programming systems can also offer a new perspective that
+data migrations. The problem is also studied in the bidirectional transformation (BX)
+community \cite{czarnecki2009bidirectional}, in work on model-driven development
+\cite{metamodels,alanen2003difference} and in more low-level work on dynamic software updating
+\cite{hicks2005dynamic}.
+
+Looking at type evolution from the perspective of programming systems can offer a new perspective that
 unfies the many different aspects of the problem partially explored in other areas. A programming
@@
 In this paper, we draw connections between these different kinds of type evolution, pointing out
-essential shared structure. By doing so, we aim to motivate further research on this important
-problem and encouraging knowledge sharing across subfields. Our contributions are twofold.
-First, we present a shared conceptual framework for thinking about the relationship between
-schema, code, and data in a programming system. Second, we present five case studies that use
-the framework to study type evolution in a highly diverse set of real-world programming scenarios.
+an essential shared structure. First, we present a shared conceptual
+framework for thinking about the relationship between schema, code, and data in a programming
+system. Second, we present five case studies that use the framework to study type evolution in a
+highly diverse set of real-world programming scenarios. We hope this will encourage knowledge
+sharing across different subfields, motivate further research on this important problem and
+offer inspiring challenges that developers of future live and collaborative programming systems
+will attempt to tackle.
 
@@
 \includegraphics[width=0.45\textwidth]{figures/layers.png}
-\caption{Pace layers from Stewart Brand's classic \emph{How Buildings Learn} \cite{Brand95}.
+\caption{Pace layers from Stewart Brand's \emph{How Buildings Learn} \cite{Brand95}.
   Different layers of a building change at different pace. The geographic
@@
 changing constantly.'' Programs evolve both while they are being created and during their operation
-in response to new needs. As with buildings (Figure~\ref{fig:layers}), different aspects of
-program evolve at different pace. And ``because of the different rates of change of its components,
+in response to new needs. As with buildings, different aspects of program evolve at different pace
+(Figure~\ref{fig:layers}). And ``because of the different rates of change of its components,
 a program is always tearing itself apart.''
 
-\subsection{Types, Code and Data}
-
-As is the case for buildings, making a change at some layers (skin or stuff) is relatively easy,
-but a change at a more basic layer (structure) has cascading effects on layers that depend on it.
-In the case of programs, a more basic layer is a structure that defines the shape of data and code.
+\subsection{Types, Data and Code}
+
+As is the case for buildings, making a change at surface layers (skin or stuff) is relatively easy,
+but a change at a more fundamental layer (structure) has cascading effects on layers that depend on it.
+In the case of programs, a more fundamental layer defines the shape of data and code.
 We refer to this structure as \emph{types}, although we do not use the term in an precise technical
-sense. In our conceptual framework, types may be data types that define the shape of data. They
+sense. In our conceptual framework, types may be data types that define the shape of data, they
 may be explicitly declared in code and checked statically, as well as dynamically checked or
-even implicit. The key property of types for us is that they define the shape of other layers
-of a program, such as data and code. As a result, a change of program types requires a
+even implicit. The key property of types for us is that they determine the shape of other layers
+of a program, namely data and code. As a result, a change of program types requires a
 corresponding change in program data and code.
@@
 only certain changes from other users (divergence), the complexity is greater. A typical case
-may be that the types and schema of a program should converge, but data may diverge.
+may be a combination of the two where types and schema of a program should converge, but data
+may diverge.
 
@@
 \centering
-\includegraphics[width=0.65\textwidth]{figures/arrows.png}
+\includegraphics[width=0.6\textwidth]{figures/arrows.png}
 \caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
@@
 
-~
-
-\textcolor{red}{TODO: got as far as here so far...}
-
-~
-
-\paragraph{Dimensions.}
-The framework for talking about schema change used in this paper thus distinguishes between
-two dimensions:
+We now describe a conceptual framework lets us view a range of diverse set of real-world
+type evolution problems in a unified way. We distinguish between two dimensions.
+The \emph{local dimension} is concerned with how type evolution affects data and code within
+a single program. The \emph{non-local dimension} is concerned with how type evolution
+propagates across different program variants that may exist independently.
+
+\paragraph{Program Layers.}
+We consider the more permanent layer of \emph{types} and two less permanent layers of code and
+data that are structured according to types. In specific case studies, the layers may map to
+different aspects of programs, some may be not present and some may be implicit.
 
 \begin{itemize}
-  \item The \emph{local dimension} is concerned with difficulties that a change causes
-    across different program pace layers that exist within a single program.
-  \item The \emph{non-local dimension} is concerned with the difficulties that a change causes
-    across different program variants that exist independently.
+\item \emph{Types} -- structure program data and determine aspects of code. A change in types
+  requires a corresponding change in data and code. Types may be explicit (database schema, type definition)
+  or implicit (shape of data structures, shape of a document).
+\item \emph{Data} -- information stored by the program. Data may be more permanent (data in a
+  database) or less permanent (state in live programming). When types evolve, data needs to be
+  updated or (in case of transient state) discarded.
+\item \emph{Code} -- or program logic, but not including source code that defines types.
+  When types evolve, code needs to change to consume/produce data of a correct shape.
 \end{itemize}
 
-\paragraph{Program pace layers.} As with buildings (Figure~\ref{fig:layers}), programs consist
-of multiple pace layers that change at a different rate. As with buildings, making a change
-to a more permanent layer also typically affects all the less permanent layers. Making a change
-at a more permanent layer is thus more challenging. Alas, changing more permanent structures in
-programming is frequently necessary. In this paper, we distinguish between four layers.
-
-\begin{itemize}
-\item \emph{State} -- transient state of a program. This is typically never shared across
-  multiple variants of a program and it can often be discarded when a change occurs.
-\item \emph{Data} -- data collected by the program. This represents a more permanent structure
-  that is typically stored in a database and outlives program restarts.
-\item \emph{Code} -- program logic. This layer changes whenever any aspect of the program logic
-  is modified. A change typically affects transient state, but may not affect data.
-\item \emph{Schema} -- the structures of program data. A change at this layer typically
-  requires adapting code to work with the new structure and also updating data and state.
-\end{itemize}
-
-\paragraph{Program variants.} We are interested in programs that are distributed across multiple
-environments as, for example, in local-first software. As discussed earlier, one aspect of this
-dimension is whether all variants of a program eventually converge or whether it is possible to
-maintain a diverging version. It is also useful to distinguish the directions in which a
-change may flow, i.e., can a change originate in any variant or is there a single centralized
-source of change. Note that each of this characteristics may  be associated with individual pace
-layers of a program.
+\paragraph{Program Variants.}
+In programming systems that enable collaboration, multiple variants of a program may exist
+concurrently and their types, data and code may need to be synchronized. Although collaboration
+is missing from some of our case studies, it adds an important dimension that is crucial for
+type evolution in future programming systems. We consider two characteristics.
 
@@
 \item \emph{Convergence} vs. \emph{Divergence}. In the convergence model, all program variants
-  eventually adopt all changes. It may not be possible to adopt a particular change before adopting
-  other changes it depends on. In the divergence model, a user may choose only particular changes
-  they want to adopt.
-\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (at a
-  specific layer) can only originate from a particular source. In the decentralized model,
+  eventually adopt all changes. It may not be possible to adopt a change without adopting
+  all earlier changes. In the divergence model, a user may choose only
+  particular changes they want to adopt, but keep other aspects of their custom design.
+\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (type
+  evolution) can only originate from a particular source. In the decentralized model,
   changes can be done on any of the multiple co-existing variants of a program.
@@
 
-\paragraph{Challenge problems.}
-The framework outlined above offers a wide range of configurations and one could conceive of
-a challenge problem for many of those configurations. We can start with a change at each of the
-four layers, consider each of the four configurations with respect to program variants (and also
-consider different behaviour at all the affected less permanent layers). The aim of this paper
-is, however, not to enumerate all options. Instead, our goal is to present a number of typical
-and often encountered challenge problems. We use the above framework to more precisely define
-the nature of the challenge problem. We also often use the above framework to discuss variations
-on the particular challenge.
+\paragraph{Challenge Problems.}
+The conceptual framework offers a wide range of configurations for challenge problems, but
+our aim is not to enumerate all options. We use it to recast a number of
+existing challenges in real-world programming systems as instances of the same problem of
+type evolution. The case studies show how well-known solutions or problems in one domain
+map to another domain and they also provide and inspiration and a benchmark for
+developers of future programming systems.
+
+\section{Live programming in the Elm architecture}
+\label{SECT:elm}
+
+The first challenge problem we recast in terms of our conceptual framework is live programming
+of user interfaces based on the Elm architecture. In this model, a reactive web application is
+structured in terms of current state and events that affect the state. The implementation then
+consists of two functions called \texttt{update} and \texttt{render}:
+
+\begin{lstlisting}[language=ml,morekeywords={on}]
+type State = { .. }
+type Event = .. | ..
+
+val update : State -> Event -> State
+val render : State -> Html
+\end{lstlisting}
+
+\noindent
+The programming model works as follows:
+
+\begin{itemize}
+\setlength\itemsep{0em}
+\item \texttt{State} represents the application state, i.e., everything that the user can work with.
+\item \texttt{Event} represents events that the user can trigger by interacting with the application.
+\item \texttt{update} is called when an event happens and computes a new program state.
+\item \texttt{render} takes the current state and produces a visual representation of the page.
+\end{itemize}
+
+\noindent
+As an example, consider the well-known TODO list app. Its state consists of a list of items,
+each with a unique ID, a title and a flag indicating whether it has been completed.
+The events represent changing of an item, deletion and addition:
+
+\begin{lstlisting}[language=ml]
+type Item = { id : id; title : string, completed : bool }
+type State = { items : Item list }
+type Event =
+  | SetTitle of id * string
+  | SetCompleted of id * bool
+  | Remove of id
+  | Add of string
+\end{lstlisting}
+
+\noindent
+Some programming systems that use the Elm architecture support hot reloading when the
+implementation of the \texttt{render} or \texttt{update} function changes, but they typically
+discard state and restart the application when types are changed. A more effective programming
+system based on live programming would allow updates to both the code of the two functions and
+the structure of the two types.
+
+\begin{table}[t]
+\begin{tabular}{lll}\toprule
+{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
+add & case to \texttt{Event} & requires adding corresponding case to \texttt{update} \\
+remove & case from \texttt{Event} & remove unused code from \texttt{update} \\
+add & field to \texttt{State} & migrate state value and initialize field \\
+remove & field from \texttt{State} & migrate state (assuming field unused) \\
+modify & structure of \texttt{State} & migrate state value and edit code accordingly \\
+\bottomrule
+\end{tabular}
+\vspace{0.3em}
+\caption{Changes to types and how a programming system should handle them}
+\label{TBL:elmchanges}
+\vspace{-1em}
+\end{table}
+
+\challenge{Live Type Evolution of State and Events}
+Assume that we have a running TODO list with the above state and events. To test the application,
+the programmer has already created a number of items and so there is a value of the
+\texttt{State} type that represents the current state of the application such as:
+
+\begin{lstlisting}[language=ml,morekeywords={true,false}]
+{ items : [
+   { id = 1; title = "Check Twitter"; completed = true }
+   { id = 2; title = "Write the paper"; completed = false } ] }
+\end{lstlisting}
+
+\noindent
+There is a number of edits to the types that the programmer may now want to do without
+restarting the application and losing the data. For example, they may want to change the
+\texttt{completed} field of \texttt{Item} to instead store optional completion time:
+
+\begin{lstlisting}[language=ml]
+type Item = { id : id; title : string; completed : DateTime option }
+type State = { items : Item list }
+type Event =
+  | .. | SetCompleted of id * DateTime option | ..
+\end{lstlisting}
+
+\noindent
+To apply the change without restarting the application, the existing data need to be migrated
+to match the new type definition. In this case, this likely cannot be done fully automatically
+and the programmer may need to provide mapping (\texttt{false} to \texttt{None}, \texttt{true}
+to \texttt{Some(DateTime(2024,5,1))}). If the programmer changes \texttt{Item} and \texttt{SetCompleted}
+at the same time, the \texttt{update} function likely does not need to change, but the rendering
+code needs to be modified (the programmer needs to specify how to map \texttt{DateTime option}
+back to a Boolean for a checkbox).
+
+More generally, there is a number of edits to the types that the programmer may want to do
+without restarting the application. The edits and the desired handling in the programming
+system are summarized in Table~\ref{TBL:elmchanges}. Modifying the \texttt{Event} type does not
+require data migration (assuming that the change does not happen while there are unprocessed
+events). It may result in unused code or missing cases in \texttt{update} that need to be addressed.
+Modifying \texttt{State} ranges from relatively simple problems (e.g., adding a new field with a
+default value) to challenging case. In particular, if a programer refactors \texttt{State} to
+semantically equivalent type, the programming system should, in principle, be able to automatically
+migrate both the original value and implementation of both functions to match the new type.
+
+~
+
+~
+
+\textcolor{red}{TODO: The rest needs more editing}
+
+~
+
+~
+
+\frameworkbox{
+In this
+}
+
+\subsection{Goals}
+As above, the key requirement from the live programming perspective is to ``do minimal harm''. In this case, this results in the following goals:
+
+\begin{itemize}
+\item When migrating the application state, this needs to be done automatically and the system should strive to produce new state that is as near as possible to the previous state.
+\item The solution may rely on structure editing so that the system has access to a high-level logical description of the edits performed by the user.
+\item The system needs to handle the case when the type structure diverges from the code structure.
+This can be addressed in various ways (transform code, add error handlers, etc.) but it is desirable to avoid "breaking" the \texttt{render} function as this would make the new application state impossible to see.
+\end{itemize}
+
+\subsection{Remarks}
+It is worth noting that the particularly difficult aspect of this challenge, i.e., the case where the structure of state is refactored to an equivalent one, is related to the Extract Entity challenge.
+
+\newpage
+\textcolor{red}{TODO: The rest needs more editing}
+
+~
+
+~
 
@@
 \end{enumerate}
+
+\section{diagrams}
+
+  \begin{tikzcd}[sep = huge]
+    \textsf{code}_1
+    \arrow[d, "\textsf{refactor}"{left},""{name=L}]
+    &
+    \textsf{schema}_1
+    \arrow[l, "\textsf{constrains}"{above}]
+    \arrow[r, "\textsf{constrains}"]
+    \arrow[d, red, "\textsf{change}"{near start, description}, ""{name=CL, left}, ""{name=CR}]
+    &
+    \textsf{data}_1
+    \arrow[d, ""{name=R, left}, "\textsf{migrate}"{right}]
+    \\
+    \textsf{code}_2
+    &
+    \textsf{schema}_2
+    \arrow[l, "\textsf{constrains}"{below}]
+    \arrow[r, "\textsf{constrains}"{below}] &
+    \textsf{data}_2
+    \arrow[Rightarrow, from=CL, to=L, "\textsf{propagate}"{below}]
+    \arrow[Rightarrow, from=CR, to=R, "\textsf{propagate}"{below}]
+  \end{tikzcd}
 
@@
 The run-time schema for \lstinline{Doors} is then the following\footnote{It is possible to represent the global variables at run-time using an untyped table or hashmap, but for the sake of schema migration complexity, the encoding as fields in classes is more instructive.}:
-\begin{quote}
+
 \begin{lstlisting}[language=java,morekeywords={on},mathescape=true]
@@
 \end{lstlisting}
-\end{quote}
+
 This \textit{extended} class $\text{\lstinline{Machine}}^{++}$ also represents state machines, but this time at run time.
@@
 
-\section{Live programming in the context of Elm architecture}
-\label{SECT:elm}
-\subsection{Context}
-Another challenge involving stateful live programming can be drawn from the programming model based on the Elm architecture.
-In this model, a reactive (web) application is structured in terms of current state and events that affect the state.
-The implementation then consists of two functions that we may call \texttt{update} and \texttt{render}:
-
-\begin{quote}
-\begin{lstlisting}[language=ml,morekeywords={on}]
-type State = { .. }
-type Event = .. | ..
-
-val update : State -> Event -> State
-val render : State -> Html
-\end{lstlisting}
-\end{quote}
-
-The programming model works as follows:
-\begin{itemize}
-    \item \texttt{State} represents the entire application state, i.e., everything that the user can work with.
-    \item \texttt{Event} represents all events that the user can trigger by interacting with the application.
-    \item \texttt{update} is called whenever an event happens. It takes the current state, the event and computes a new state.
-    \item \texttt{render} takes the current state and produces a representation of what should be displayed on the screen (e.g., an HTML tree).
-\end{itemize}
-
-Traditionally, when Elm applications are developed, they are restarted each time the code is modified and any previous state is discarded.
-However, a more effective programming model based on live programming would allow live updates to both the code of the two functions and the structure of the two types.
-
-\subsection{Example}
-
-As an example, consider the uninspiring, but well-known, TODO list application. The types that capture the state and events in the application may look as follows:
-
-\begin{quote}
-\begin{lstlisting}[language=ml,morekeywords={on}]
-type Item = { id : id; title : string; completed : bool }
-type State = { items : Item list }
-type Event =
-  | SetCompleted of id * bool
-  | SetTitle of id * string
-  | Remove of id
-  | Add of string
-\end{lstlisting}
-\end{quote}
-
-The application state consists of a list of items.
-Each item has a unique ID alongside with a title and a flag indicating whether it is completed.
-The events represent edits to the items, deletion and addition.
-The implementation of the \texttt{update} and \texttt{render} functions is simple and not important for the challenge.
-
-\subsection{Challenge}
-Now, imagine that we have a running TODO list application with the above state and events.
-To test the application, the programmer has already created a number of items and so there is a single value of the \texttt{State} type that represents a current state of the application such as:
-
-\begin{quote}
-\begin{lstlisting}[language=ml,morekeywords={on}]
-{ items : [
-   [ { id = 1; title = "check twitter"; completed = true  }
-     { id = 1; title = "Write paper"; completed = false } ] }
-\end{lstlisting}
-\end{quote}
-
-As above, there are a number of edits to the code and types that the programmer may want to do without restarting the application.
-Those are summarized in Table~\ref{TBL:elmchanges}.
-Modifying the code is simple and only requires waiting until the current execution completes.
-Modifying the \texttt{Event} type is also simple, but it may lead to unused code or missing case in \texttt{update} that needs to be addressed.
-Finally, modifying \texttt{State} ranges from relatively simple problems (adding a new field) to challenging case when the structure of \texttt{State} is changed.
-
-For example, imagine that the programmer would want to edit the structure of the state and migrate the previous definition of \texttt{State} to the following new definition where individual fields are stored in separate lists:
-
-\begin{quote}
-\begin{lstlisting}[language=ml,morekeywords={on}]
-type State =
-  { ids : id list
-    titles string list
-    completes : bool list }
-\end{lstlisting}
-\end{quote}
-
-This representation is semantically equivalent (assuming the lists have the same length) to the original one.
-It should thus, in principle, be possible to migrate the original state value to a value using the new structure.
-Moreover, it should, in principle, be also possible to automatically transform the implementations of \texttt{update} and \texttt{render} to work as before, but using the new state structure.
-
-\begin{table}[t]
-  \centering
-\begin{tabular}{lll}\toprule
-\textsc{Change} & \textsc{What} & \textsc{How}\\\midrule
-modify & \texttt{render} & simple, but wait until current execution finishes  \\
-modify & \texttt{update} & simple, but wait until current execution finishes  \\
-add & case to \texttt{Event} & requires adding corresponding case to \texttt{update} \\
-remove & case from \texttt{Event} & remove unused code from \texttt{update} \\
-add & field to \texttt{State} & migrate state value and initialize field \\
-remove & field from \texttt{State} & migrate state (assuming field unused) \\
-modify & structure of \texttt{State} & migrate state value and edit code accordingly \\
-\bottomrule
-\end{tabular}
-\caption{Possible program changes and how to deal with them at run time}
-\label{TBL:elmchanges}
-\end{table}
-
-\subsection{Goals}
-As above, the key requirement from the live programming perspective is to ``do minimal harm''. In this case, this results in the following goals:
-
-\begin{itemize}
-\item When migrating the application state, this needs to be done automatically and the system should strive to produce new state that is as near as possible to the previous state.
-\item The solution may rely on structure editing so that the system has access to a high-level logical description of the edits performed by the user.
-\item The system needs to handle the case when the type structure diverges from the code structure.
-This can be addressed in various ways (transform code, add error handlers, etc.) but it is desirable to avoid "breaking" the \texttt{render} function as this would make the new application state impossible to see.
-\end{itemize}
-
-\subsection{Remarks}
-It is worth noting that the particularly difficult aspect of this challenge, i.e., the case where the structure of state is refactored to an equivalent one, is related to the Extract Entity challenge.
-
 \section{Multiplicity change}
@@
 
-\begin{quote}
 \begin{lstlisting}[language=ml]
@@
 \end{lstlisting}
-\end{quote}
 

### 2024-05-25 paper.tex

@@
 \usepackage{tikz-cd}
+\usepackage{float}
 
@@
 
-  \begin{tikzcd}[sep = huge]
-    \textsf{code}_1
-    \arrow[d, "\textsf{refactor}"{left},""{name=L}]
+\begin{figure}[H]
+  \begin{tikzcd}[sep = 6em]
+    \textsf{query}
+    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
     &
-    \textsf{schema}_1
+    \textsf{schema}
     \arrow[l, "\textsf{constrains}"{above}]
-    \arrow[r, "\textsf{constrains}"]
-    \arrow[d, red, "\textsf{change}"{near start, description}, ""{name=CL, left}, ""{name=CR}]
+    \arrow[r, "\textsf{constrains}"{above}]
+    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
     &
-    \textsf{data}_1
-    \arrow[d, ""{name=R, left}, "\textsf{migrate}"{right}]
+    \textsf{data}
+    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
     \\
-    \textsf{code}_2
+    \textsf{query}_\alpha
     &
-    \textsf{schema}_2
-    \arrow[l, "\textsf{constrains}"{below}]
-    \arrow[r, "\textsf{constrains}"{below}] &
-    \textsf{data}_2
-    \arrow[Rightarrow, from=CL, to=L, "\textsf{propagate}"{below}]
-    \arrow[Rightarrow, from=CR, to=R, "\textsf{propagate}"{below}]
+    \textsf{schema}_\alpha
+    \arrow[l, "\textsf{constrains}"{above}]
+    \arrow[r, "\textsf{constrains}"{above}]
+    &
+    \textsf{data}_\alpha
+    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
+    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
   \end{tikzcd}
+  \caption{Extract Entity}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{tikzcd}[column sep = large, row sep = 5em]
+    &
+    \textsf{schema}
+    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
+    \arrow[rr, "\textsf{constrains}"]
+    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
+    &
+    &
+    \textsf{data}
+    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
+    \arrow[Rightarrow, from=L, to=R]
+    \\
+    \textsf{schema}_\alpha
+    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
+    \arrow[rr, crossing over]
+    &
+    &
+    \textsf{data}_\alpha
+    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
+    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
+    &
+    \\
+    &
+    \textsf{schema}_\beta
+    \arrow[ld, red, "\mathit{merge}\left(\textsf{change}_\alpha {\,,\;} \textsf{change}_\beta\right)"{name=L4, near start, description}]
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_\beta
+    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
+    \arrow[Rightarrow, from=L4, to=R4]
+    \\
+    \textsf{schema}_{\alpha\beta}
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_{\alpha\beta}
+    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
+    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
+    &
+  \end{tikzcd}
+  \caption{Mergeing divergent changes}
+\end{figure}
 

### 2024-05-25 paper.tex
Conf organizer diagram

@@
 \begin{figure}[H]
+  \centering
   \begin{tikzcd}[sep = 6em]
@@
 \begin{figure}[H]
+  \centering
   \begin{tikzcd}[column sep = large, row sep = 5em]
@@
     \textsf{schema}_\beta
-    \arrow[ld, red, "\mathit{merge}\left(\textsf{change}_\alpha {\,,\;} \textsf{change}_\beta\right)"{name=L4, near start, description}]
+    \arrow[ld, "\mathit{rebase}\left(\textsf{change}_\alpha {\,,\;} \textsf{change}_\beta\right)"{name=L4, near start, description}]
     \arrow[rr]
@@
   \caption{Mergeing divergent changes}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \begin{tikzcd}[column sep = large, row sep = 5em]
+    &
+    \textsf{schema}
+    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
+    \arrow[rr, "\textsf{constrains}"]
+    \arrow[dd, "\textsf{identity}"{near start, description}]
+    &
+    &
+    \textsf{data}
+    \arrow[dd, red, "\textsf{update}_\beta"{near start, description}]
+    \\
+    \textsf{schema}_\alpha
+    \arrow[rr, crossing over]
+    &
+    &
+    \textsf{data}_\alpha
+    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
+    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
+    \arrow[rr, red, crossing over, "\textsf{update}_\gamma"{near start, description}]
+    &
+    &
+    \textsf{data}_{\alpha\gamma}
+    \arrow[dd, "\mathit{rebase}\left(\textsf{update}_\beta {\,,\;} \textsf{update}_\gamma\right)"{near start, description}]
+    \\
+    &
+    \textsf{schema}
+    \arrow[ld, "\textsf{change}_\alpha"{name=L4, near start, description}]
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_\beta
+    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
+    \arrow[Rightarrow, from=L4, to=R4]
+    \\
+    \textsf{schema}_\alpha
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_{\alpha\beta}
+    &
+    &
+    \textsf{data}_{\alpha\beta\gamma}
+  \end{tikzcd}
+  \caption{Conference Organizer merging schema and divergent data updates}
 \end{figure}

### 2024-05-25 paper.tex
diagram doubts
rebase operator

@@
     \textsf{schema}_\beta
-    \arrow[ld, "\mathit{rebase}\left(\textsf{change}_\alpha {\,,\;} \textsf{change}_\beta\right)"{name=L4, near start, description}]
+    \arrow[ld, "\textsf{change}_\alpha \succ\, \textsf{change}_\beta"{name=L4, near start, description}]
     \arrow[rr]
@@
   \end{tikzcd}
-  \caption{Mergeing divergent changes}
+  \caption{Mergeing divergent changes with rebase ($\succ$)}
 \end{figure}
@@
     \textsf{data}_{\alpha\gamma}
-    \arrow[dd, "\mathit{rebase}\left(\textsf{update}_\beta {\,,\;} \textsf{update}_\gamma\right)"{near start, description}]
+    \arrow[dd, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha \succ\, \textsf{update}_\gamma"{near start, description}]
     \\
@@
     \textsf{data}_\beta
-    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
+    \arrow[ld, "\textsf{migrate}_\alpha"{name=R4, near start, description}]
     \arrow[Rightarrow, from=L4, to=R4]
@@
     \textsf{data}_{\alpha\beta}
+    \arrow[from=uu, crossing over, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha"{near start, description}]
     &
@@
 \end{figure}
+
+\textcolor{red}{[JE] These diagrams are getting hairy. Are they actually helpful?}
 

### 2024-05-25 paper.tex
move related work to end

@@
 \textcolor{red}{[JE] These diagrams are getting hairy. Are they actually helpful?}
-
-\section{Related Work}
-
-TODO: Maybe insert something about methodology in general?
-(ref the table challenges paper? although this paper is no longer so much about challenges)
-
-\paragraph{Schema evolution.}
-Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help.
-SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
-MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data.
-But apart from such special cases, SQL databases have no general purpose support for data migration.
-As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data.
-Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
-In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
-
-\citet{bernstein07} observed in 2007 ``There are hardly any schema evolution tools today.
-This is rather surprising since there is a huge literature on schema evolution spanning more than two decades.'' There are now more tools available.
-Some tools such as Liquibase~\cite{liquibase} and PlanetScale\cite{planetscale} could be characterized as version control and continuous integration/deployment for schema.
-They track schema changes and can calculate diffs in the form of SQL DDL statements to convert one schema version to another, but do not help migrate data.
-Unfortunately comparing schemas can be ambiguous about the intention of changes.
-For example has a column been renamed or has it been deleted and a new one created?
-That distinction makes a big difference to the data in that column.
-EvolveDB\cite{evolvedb} addresses this ambiguity by reverse-engineering the schema into a richer data model and tracking the edits to that model within an IDE.
-This more precise edit history can be used to infer higher level intentions of a schema change, which then generate SQL scripts to evolve the database.
-EdgeDB~\cite{edgedb} resolves ambiguities by asking questions of the developer, with some answers supplying custom migration code in a proprietary query language.
-
-Some tools provide a Domain Specific Language (DSL) to describe schema evolution.
-Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code.
-\citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
-SMOs can also rewrite queries to accomodate schema changes.
-\citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}.
-BiDEL appears capable of handling the basic requirements of our \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} challenges, though \textit{Split/Merge Entity} is less clear.
-It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.
-
-Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases.
-While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}.
-\citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON.
-\citet{scherzinger13} define a set of operators like the relational SMOs discussed above.
-\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} but not \textit{Split/Merge Entity} nor divergence.
-None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.
-
-Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}.
-Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}.
-Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.
-
-\paragraph{Live, MDD, etc. programming?}
-
-% Some references from Tomas
-%
-% collaborative programmign tools
-\cite{goldman2011real,kurniawan2015coder,replit}
-% exploratory/live review paper
-\cite{rein2018exploratory}
-% hot reload in live systems
-\cite{barenz2020essence,beckmann2021shortening}
-
-% Some references from Tijs
-%
-% SQL+code
-\cite{Cleve2006,wang2019synthesizing}
-% BX
-\cite{czarnecki2009bidirectional}
-% MDD
-\cite{metamodels,alanen2003difference}.
-
-\begin{comment}
-
-Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations
-
-dataframe diff https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
-
-Durable computing: pins runtime data frames to code versions. Doesn't handle problem of global data schema change.
-https://restate.dev/blog/solving-durable-executions-immutability-problem/
-https://restate.dev/blog/code-that-sleeps-for-a-month/
-
-Tijs on Live programming and model-driven engineering
-https://ir.cwi.nl/pub/23770
-https://dl.acm.org/doi/10.1007/s10270-017-0608-7
-https://dl.acm.org/doi/10.1145/3276604.3276611
-https://ir.cwi.nl/pub/21858
-
-From Geoffrey:
-1) Finding Data Compatibility Bugs with JSON Subschema Checking
-https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
-I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant.
-It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
-2) Confluent Schema Registry:
-https://docs.confluent.io/platform/current/schema-registry/index.html
-An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
-\end{comment}
 
@@
 
+\section{Related Work}
+
+TODO: Maybe insert something about methodology in general?
+(ref the table challenges paper? although this paper is no longer so much about challenges)
+
+\paragraph{Schema evolution.}
+Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help.
+SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
+MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data.
+But apart from such special cases, SQL databases have no general purpose support for data migration.
+As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data.
+Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
+In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
+
+\citet{bernstein07} observed in 2007 ``There are hardly any schema evolution tools today.
+This is rather surprising since there is a huge literature on schema evolution spanning more than two decades.'' There are now more tools available.
+Some tools such as Liquibase~\cite{liquibase} and PlanetScale\cite{planetscale} could be characterized as version control and continuous integration/deployment for schema.
+They track schema changes and can calculate diffs in the form of SQL DDL statements to convert one schema version to another, but do not help migrate data.
+Unfortunately comparing schemas can be ambiguous about the intention of changes.
+For example has a column been renamed or has it been deleted and a new one created?
+That distinction makes a big difference to the data in that column.
+EvolveDB\cite{evolvedb} addresses this ambiguity by reverse-engineering the schema into a richer data model and tracking the edits to that model within an IDE.
+This more precise edit history can be used to infer higher level intentions of a schema change, which then generate SQL scripts to evolve the database.
+EdgeDB~\cite{edgedb} resolves ambiguities by asking questions of the developer, with some answers supplying custom migration code in a proprietary query language.
+
+Some tools provide a Domain Specific Language (DSL) to describe schema evolution.
+Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code.
+\citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
+SMOs can also rewrite queries to accomodate schema changes.
+\citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}.
+BiDEL appears capable of handling the basic requirements of our \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} challenges, though \textit{Split/Merge Entity} is less clear.
+It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.
+
+Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases.
+While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}.
+\citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON.
+\citet{scherzinger13} define a set of operators like the relational SMOs discussed above.
+\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} but not \textit{Split/Merge Entity} nor divergence.
+None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.
+
+Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}.
+Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}.
+Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.
+
+\paragraph{Live, MDD, etc. programming?}
+
+% Some references from Tomas
+%
+% collaborative programmign tools
+\cite{goldman2011real,kurniawan2015coder,replit}
+% exploratory/live review paper
+\cite{rein2018exploratory}
+% hot reload in live systems
+\cite{barenz2020essence,beckmann2021shortening}
+
+% Some references from Tijs
+%
+% SQL+code
+\cite{Cleve2006,wang2019synthesizing}
+% BX
+\cite{czarnecki2009bidirectional}
+% MDD
+\cite{metamodels,alanen2003difference}.
+
 \begin{comment}
-\section{Rename field; Add new required field}
-Starter problems suggested by Geoffrey.
-These are already directly supported in SQL, so there is good support in SQL schema change tools, but they don't address how to synchronize the change with live code refactoring or local-first code deployment.
-\end{comment}
-
-\begin{comment}
-\section{Cambria (Geoffrey/pvh)}
-\citet{Cambria} exhumed and studied the old problem of schema change, primarily in the context of local-first software.
-In this paper we add more scenarios and the complementary problems of live programming.
-The primary goal of local-first software is user autonomy, which could include letting users stay on older versions of application code while continuing to collaborate with newer versions.
-This problem is a nice counterpart to that of Divergence Control.
-They both require a kind of bidirectional schema change, with Cambria building upon research on bidirectional programming with lenses, while Divergence Control is more closely connected to version control techniques.
-Appendix III of \cite{Cambria} discusses the tricky problem of converting between scalars and arrays (which in data modeling is called a multiplicity change).
-That could be the challenge problem right there.
-
-The Cambria paper also presents motivating examples of schema change problems in Stripe, Kafka, and Mastodon. Perhaps one of those examples could be crafted into a challenge problem.
-\end{comment}
-
-\begin{comment}
-\section{Denormalization (Matt Tantaman)}
-Matt points out Extract Entity is a normalization so we ought to complement it with a denormalization.
-Note that copying/duplication is a technical problem in some formalisms like lenses.
-Denormalization is often done for performance reasons but that isn't an issue in our small-scale scenarios.
-What is a small-scale use case for denormalization?
-
-There are good cases for denormalizing from 1NF, i.e.
-nesting structures.
-Local-first software seems focused on JSON data rather than relational.
-An extension of Extract Entity is \textit{Nest Entity} where a table of orders are nested inside each customer.
+From Tijs:
+
+BX community:
+https://www.cs.cornell.edu/~jnfoster/papers/grace-report.pdf
+(overview paper)
+
+Isil Dilig's stuff;
+https://www.cs.utexas.edu/~isil/pldi19b.pdf
+https://www.cs.utexas.edu/~isil/vldb20.pdf
+
+Eelco Visser/Sander Vermolen
+https://dblp.org/pid/10/6098.html
+(we’ve discussed the catalog before)
+
+On the code-database co-evolution problem:
+https://link.springer.com/chapter/10.1007/11877028_17
+(little less technical than Isil’s stuff)
+
+Coupled transformations:
+https://dl.acm.org/doi/10.1007/978-3-540-69611-7_19
+https://www.sciencedirect.com/science/article/pii/S1571066108003058
+https://link.springer.com/chapter/10.1007/978-3-540-87875-9_45
+
+(from the MDD angle, only the tip of the ice berg…)
+
+Divergence in model-driven:
+https://dl.acm.org/doi/10.1145/2000410.2000414
+
+The seminal model-diffing paper:
+http://btn1x4.inf.uni-bayreuth.de/publications/dotor_buchmann/SCM/ChefRepo/DiffUndMerge/Alanen2003.pdf
+(Diffing, versioning, collaborative editing, it’s all well researched in MDD)
+
+Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations
+
+dataframe diff https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
+
+Durable computing: pins runtime data frames to code versions. Doesn't handle problem of global data schema change.
+https://restate.dev/blog/solving-durable-executions-immutability-problem/
+https://restate.dev/blog/code-that-sleeps-for-a-month/
+
+Tijs on Live programming and model-driven engineering
+https://ir.cwi.nl/pub/23770
+https://dl.acm.org/doi/10.1007/s10270-017-0608-7
+https://dl.acm.org/doi/10.1145/3276604.3276611
+https://ir.cwi.nl/pub/21858
+
+From Geoffrey:
+1) Finding Data Compatibility Bugs with JSON Subschema Checking
+https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
+I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant.
+It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
+2) Confluent Schema Registry:
+https://docs.confluent.io/platform/current/schema-registry/index.html
+An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
 \end{comment}

### 2024-05-25 paper.tex
Finish Elm migration, Entity evolution

--- 2024-05-25 paper.tex
@@
 \usepackage{tikz-cd}
-\usepackage{float}
 
@@
 
-\keywords{schema change, database migration, live programming, local-first programming}
+\keywords{Type Evolution, Schema Evolution, Live Programming, Local-first Programming}
 
@@
 \begin{abstract}
-Schema change is an unsolved problem in both live programming and local-first software. We include
-in schema change any change to the expected shape of data, whether that is expressed explicitly
-in a database schema or type system, or whether those expectations are implicit in the behavior
-of the code.
-
-Schema changes during live programming can create a mismatch between the code and
-data in the running environment. Correspondingly, Schema changes in local-first programming can
-create mismatches between data in different replicas, and between data in a replica and the code
-colocated with it. In all of these situations the problem of schema change is to migrate or
-translate existing data in coordination with changes to the code.
-
-This paper contributes a set of concrete scenarios involving schema change that are offered as
-challenge problems to the live programming and local-first communities. We hope that these
-problems will spur progress by providing concrete objectives and a basis for comparing alternative
-solutions.
-
-~
-
-~
-
-\noindent\textbf{Context}: \emph{What is the broad context of the work? What is the importance of the general research area?}
-Schema change is an unsolved problem in live and local-first programming that is poorly understood and calls for a more fundamental analysis. It is particularly important as more programming
-is done in programming systems where programs may be changed by their users and as more programming
-is local-first, meaning that such changes need to be synchronized.
-
-\noindent\textbf{Inquiry}: \emph{What problem or question does the paper address? How has this problem or question been addressed?}
-Specific systems have specific solutions, but what is lacking is a general framework for thinking
-about schema change and also a good suite of challenge problems to motivate and evaluate new solutions.
-
-\noindent\textbf{Approach}: \emph{What was done that unveiled new knowledge?}
-We collect challenge problems from a range of different kinds of (primarily) live and local-first
-application contexts. We develop a framework for thinking about program change and use it to
-present our challenge problems in a systematic way.
-
-\noindent\textbf{Knowledge}: \emph{What new facts were uncovered? What new capabilities are enabled by the work?}
-We understand change in programs in a novel way, using a two-dimensional framwork that looks at
-the different pace layers in a program and at the different ways in which change is propagated
-across program variants.
-
-\noindent\textbf{Grounding}: \emph{What argument, feasibility proof, artifacts, or results and evaluation support this work?}
-We use the framework to provide a framing for a range of challenge problems derived from both
-existing literature and our own expertise.
-
-\noindent\textbf{Importance}: \emph{Why does this work matter?}
-This paper contributes a set of concrete scenarios involving schema change that are offered as
-challenge problems to the live programming and local-first communities. We hope that these
-problems will spur progress by providing concrete objectives and a basis for comparing alternative
-solutions.
+
+% Context: What is the broad context of the work? What is the importance of the general research area?
+Many recent programming improvements have been enabled by shorter feedback loops. To further
+advance the state of programming, we need programming systems that enable rapid collaboration
+and make programming more live across the entire development stack. To do this, programming
+systems need to preserve state during development and interaction with the programmer.
+
+% Inquiry: What problem or question does the paper address? How has this problem or question been addressed?
+The key obstacle to live collaborative programming systems is the problem of \emph{type evolution}.
+When programmer changes the structure of a type, programming system needs to co-evolve
+existing live data and adapt code that implements the program behaviour to match the new structure.
+This problem has been studied in isolation as schema evolution in databases, hot code reloading
+in live programming and dynamic software updating, but we lack a unified comprehensive perspective.
+
+% Approach: What was done that unveiled new knowledge?
+In this paper, we develop a unfied conceptual framework for analysing the problem of type evolution.
+We use the framework to study type evolution in five diverse case studies.
+%
+% Knowledge: What new facts were uncovered? What new capabilities are enabled by the work?
+By looking at the case studies through a unified perspective, we hope to inspire new thinking
+about the problem of type evolution and also encourage knowledge transfer across sub-fields.
+
+% Grounding: What argument, feasibility proof, artifacts, or results and evaluation support this work?
+Our work takes the form of a sequence of challenge problems drawn from real-world programming
+scenario in diverse field such as live programming, database programming and model-driven
+development.
+%
+% Importance: Why does this work matter?
+Tackling the problem of type evolution in a unified way, enabled by our new conceptual framework,
+will make it possible to design and build a future generation of effective, live and collaborative
+programming systems.
 
 \end{abstract}
+
+% ==================================================================================================
+% INTRODUCTION
+% ==================================================================================================
 
@@
 
+% ==================================================================================================
+% FRAMEWORK
+% ==================================================================================================
+
 \section{Type Evolution}
@@
 
-\section{Live programming in the Elm architecture}
-\label{SECT:elm}
-
-The first challenge problem we recast in terms of our conceptual framework is live programming
+% ==================================================================================================
+% ELM
+% ==================================================================================================
+
+\section{Live Programming for the Elm Architecture}
+\label{sec:elm}
+
+The first challenge problem that we present in terms of our conceptual framework is live programming
 of user interfaces based on the Elm architecture. In this model, a reactive web application is
@@
 implementation of the \texttt{render} or \texttt{update} function changes, but they typically
-discard state and restart the application when types are changed. A more effective programming
-system based on live programming would allow updates to both the code of the two functions and
-the structure of the two types.
+discard state and restart the application when the \texttt{State} type changes. A more effective
+programming system based on live programming would allow updates to both the code of the two
+functions and the structure of the two types.
 
@@
 {\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
-add & case to \texttt{Event} & requires adding corresponding case to \texttt{update} \\
-remove & case from \texttt{Event} & remove unused code from \texttt{update} \\
-add & field to \texttt{State} & migrate state value and initialize field \\
-remove & field from \texttt{State} & migrate state (assuming field unused) \\
-modify & structure of \texttt{State} & migrate state value and edit code accordingly \\
+Add & case to \texttt{Event} & Requires adding corresponding case to \texttt{update} \\
+Remove & case from \texttt{Event} & Remove unused code from \texttt{update} \\
+Add & field to \texttt{State} & Migrate state value and initialize field \\
+Remove & field from \texttt{State} & Migrate state (assuming field unused) \\
+Modify & structure of \texttt{State} & Migrate state value and edit code accordingly \\
 \bottomrule
@@
 \caption{Changes to types and how a programming system should handle them}
-\label{TBL:elmchanges}
+\label{tbl:elmchanges}
 \vspace{-1em}
@@
 
-\challenge{Live Type Evolution of State and Events}
+\challenge{Live State Type Evolution}
 Assume that we have a running TODO list with the above state and events. To test the application,
@@
 without restarting the application. The edits and the desired handling in the programming
-system are summarized in Table~\ref{TBL:elmchanges}. Modifying the \texttt{Event} type does not
-require data migration (assuming that the change does not happen while there are unprocessed
+system are summarized in Table~\ref{tbl:elmchanges}. Modifying the \texttt{Event} type does not
+require data migration (assuming that the change does not happen when there are unprocessed
 events). It may result in unused code or missing cases in \texttt{update} that need to be addressed.
 Modifying \texttt{State} ranges from relatively simple problems (e.g., adding a new field with a
-default value) to challenging case. In particular, if a programer refactors \texttt{State} to
+default value) to challenging case. In particular, if the programer refactors \texttt{State} to a
 semantically equivalent type, the programming system should, in principle, be able to automatically
@@
 
+\frameworkbox{
+The evolving \emph{type} in this case is the \texttt{State} type, \emph{code} is the
+program logic implemented in \texttt{update} and \texttt{render} and \emph{data} is the current
+state. To support live programming, a local type change requires synchronizing local data to
+match the new type, and also a change in code so that it continues working against the new type
+structure. In a non-local scenario, programmers would likely want to synchronize types and code,
+but not necessarily data. A collaborative programming system would need to support decentralized
+and possibly divergent type evolution, but without data synchronization.
+}
+
+\subsection{Remarks}
+The key requirement from the live programming perspective is to ``do minimal harm''.
+The migration of \emph{data} needs to be done automatically and the system should strive
+to produce a new state that is as near as possible to the previous state. Migrating
+\emph{code} can be done in various ways (transform code, trigger error condition, etc.),
+but it is desirable to avoid breaking the \texttt{render} function as this would make the
+new application state impossible to visualize.
+
+The specific case where the type is refactored to a semantically equivalent type is related to
+the Extract Entity challenge discussed below. Finally, a programming system tackling this
+challenge may rely on structure editing so that the system has access to a high-level logical
+description of the edits performed by the user.
+
+% ==================================================================================================
+% DATABASES
+% ==================================================================================================
+
+\section{Entity Evolution in Data-Centric Systems}
+Many problems studied in the context of database systems equally apply to programming systems
+if we think of them as stateful environments where programs co-exist with their persistent data.
+Although schema evolution is well-studied in databses (Section~\ref{sec:related}), past work
+does not resolve all problems in our second challenge.
+
+As an example, consider the Acme Corporation that records orders from various customers for
+various products. They use a spreadsheet (Figure~\ref{fig:db-orders}) with a row for each order
+and columns for information about the customer and product. The shipping department filters this
+table on blank ship dates to see what they need to ship. But after a while the orders department
+realizes they are wasting effort duplicating the address for new order from an old customer. And
+when the customer's address changes they have to go back and edit all of their orders.
+
+\begin{figure}
+\begin{adjustwidth}{-2em}{-3em}
+\begin{subfigure}[b]{35em}\vspace{0pt}
+  \sffamily
+  \small
+  \begin{tabular}{ |r|l|r|r|l|l|}
+     \hline
+     oid & item & quantity & ship\_date & customer\_name & customer\_address \\
+     \hline \hline
+     1 & Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
+     \hline
+     2 & Dynamite & 2 & & Daffy Duck & White Rock Lake \\
+     \hline
+     3 & Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
+     \hline
+  \end{tabular}
+  \caption{Spreadsheet recording orders with all associated information}
+  \label{fig:db-orders}
+\end{subfigure}
+\hfill
+\\[-0.5em]~\\
+\begin{subfigure}[b]{20em}\vspace{0pt}
+  \sffamily
+  \small
+  \begin{tabular}{ |r|l|l|}
+    \hline
+    cid & customer\_name & customer\_address \\
+    \hline \hline
+    1 & Wile E Coyote & 123 Desert Station \\
+    \hline
+    2 & Daffy Duck & White Rock Lake \\
+    \hline
+  \end{tabular}
+  \caption{Customers table after type evolution}
+  \label{fig:db-customers}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{20em}\vspace{0pt}
+  \sffamily
+  \small
+  \begin{tabular}{ |r|l|r|r|r|}
+     \hline
+     oid & item & quantity & ship\_date & cid \\
+     \hline \hline
+     1 & Anvil & 1 & 2/3/23 & 1 \\
+     \hline
+     2 & Dynamite & 2 & & 2  \\
+     \hline
+     3 & Bird Seed & 1 & & 1 \\
+     \hline
+    \end{tabular}
+  \caption{Orders linking to Customers, after type evolution}
+ \label{fig:db-orders-link}
+\end{subfigure}
+\end{adjustwidth}
+\vspace{0.25em}
+\caption{Types and data before and after performing Extract Entity}
+\label{fig:db}
+\end{figure}
+
+\challenge{Extract Entity}
+Acme needs to evolve the above type and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
+one with orders that links to one with customers.  We refer to this as \emph{Extract Entity}
+operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
+The operation is needed when we realize that some attributes of one type of entity should belong to
+a  distinct type of entity that will be associated with the first one.
+
+The system needs to provide a mechanism for referencing entities, such as by using unique
+identifiers (here consecutive numbers). This allows, for example, a spelling error in a customer
+address to be fixed in one place. It also allow a customer name or address to change without
+breaking the connection from all associated orders. Unique identifiers are the essence of
+entities, whether they are uniquely generated primary keys in a database or an abstraction
+of a memory address in a programming language.
+
+The \textit{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers,
+and 3) reference these identifiers from the orders. Our example took the simple approach of
+merging entities whose attributes are all equal. But that might not be correct in all cases.
+What if there was a typo in one of the addresses that made it look different from other instances
+of the customer? To repair that error we need to merge the falsely distinct customers into one
+and fix any references from orders. We call this operation \textit{Merge Entities} -- note that
+it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.
+
+More generally, entity evolution can happen through operations such as those listed in
+Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
+change in requirements that triggered an evolution might change back again. Inverse
+is also required in research on \emph{divergence control}~\cite{Foster2007, herrmann17, chillon22}.
+The inverse of \textit{Extract Entity} is \textit{Absorb Entity}. It raises the question of how
+to handle \textit{one-to-many} relationships. For example should absorbing orders into customers
+do a relational join, returning us back to the starting point, or should it nest orders within
+customers, as a NoSQL database might do?
+
+The inverse of \textit{Merge Entities} is \textit{Split Entity}, but that can be awkward.
+Imagine that we didn't have addresses for customers to disambiguate them with and only knew
+their non-unique names. Then \textit{Extract Entity} will conflate them and discard the
+information needed to properly invert the operation. Should this information be retained?
+One could object that we shouldn't have taken orders for people we can't uniquely identify
+yet. And even if that was necessary it means the customers really aren't distinct entities
+and shouldn't have been extracted. These are arguments that we shouldn't need \textit{Split
+Entity}, contradicting the principle that evolution operations should be invertible.
+It is interesting to observe that evolution in nature can also fail to invert, leaving
+behind vestigial features. We consider this to be an open issue.
+
+\begin{table}[t]
+\begin{tabular}{lll}\toprule
+{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
+Extract & new type of entity & Migrate data to a new table and add links \\
+Absorb & a linked type of entity & Migrate data and remove unneeded table \\
+Merge & multiple data items & Automatic de-duplication with mistake handling \\
+Split & multiple data items & Interactively, following user guidance\\
+\bottomrule
+\end{tabular}
+\vspace{0.3em}
+\caption{Changes to types and how a programming system should handle them}
+\label{tbl:dbchanges}
+\vspace{-1em}
+\end{table}
+
+\frameworkbox{
+In a programming system, the \emph{types} to which entity evolution applies can be data tables,
+but also objects or JSON. When the type evolves, the system needs to update corresponding
+live \emph{data} and \emph{code}. In a non-local scenario, entity evolution
+typically concerns permanent data. As a result, the system needs to synchronize types,
+data and code across all replicas. Doing so in a way that allows divergence is discussed
+in the next section.
+}
+
+\subsection{Remarks}
+The central problem is to coordinate the type evolution with code edits to keep the system
+executing live and correctly. Solutions may infer the desired type evolution by analysing
+source code edits, but would typically offer UI affordances for the commands. They should
+try to minimize the number and complexity of commands that must be introduced and also the
+interruption to live execution of the code.
+
+The Extract Entity challenge in the non-local case (for local-first software) is similar to the
+local live programming challenge except that the context is an application that offers similar
+functionality with peer-to-peer collaboration. In this context the schema change does not need
+to be performed interactively -- a developer can take some time to build, test, and package a
+solution, perhaps using an API or DSL. The hard part comes when the schema change is to
+be deployed across all the replicas. This deployment must synchronize code and data upgrades
+(or decouple them as in Cambria~\cite{Cambria}). The deployment must also deal with
+migrating ``in flight'' operations so that all replicas converge on the same state
+without data loss, and ideally without centralized coordination. A radical approach
+could try to incorporate schema change operations into the underlying datastore
+itself (possibly a CRDT) and integrate code as well, but layered solutions are welcome too.
+
+\challenge{Code Co-evolution for Extract Entity}
+
+In a programming system, there is typically also some code for programmatic manipulation
+of entities. The code can, for example, implement a simple CRUD UI that also provides the
+Acme shipping department its view of unshipped orders. The holistic perspective of programming
+systems lets us see this code not as separate from the system, but as its integral part.
+The code thus also needs to co-evolve when the entity evolves. An example is a SQL
+query to report all pending orders in the original schema:
+
+\begin{lstlisting}[language=SQL]
+SELECT order_id, item, quantity, customer_name, customer_address
+FROM Orders WHERE ship_date IS NULL;
+\end{lstlisting}
+
+\noindent
+After applying the \textit{Extract Entity} operation as described earlier, the code should
+be rewritten into the following query that joins the two linked tables:
+
+\begin{lstlisting}[language=SQL]
+SELECT order_id, item, quantity, customer_name, customer_address
+FROM Orders
+JOIN Customers ON Orders.customer_id = Customers.customer_id
+WHERE ship_date IS NULL;
+\end{lstlisting}
+
+\noindent
+The same code transformation would need to be applied to queries written over other
+data representations such as JSON. The system can also use mechanism such as a database
+view and execute the original query over the view. When applying code migrations,
+the system should minimize the need for users to manaually intervene. It should also
+minimize the possibilities that developer-written code is subtly incorrect in edge cases,
+preferably by reducing the need for developer-written code.
+
+\newpage
 ~
-
-~
-
-\textcolor{red}{TODO: The rest needs more editing}
-
-~
-
-~
-
-\frameworkbox{
-In this
-}
-
-\subsection{Goals}
-As above, the key requirement from the live programming perspective is to ``do minimal harm''. In this case, this results in the following goals:
-
-\begin{itemize}
-\item When migrating the application state, this needs to be done automatically and the system should strive to produce new state that is as near as possible to the previous state.
-\item The solution may rely on structure editing so that the system has access to a high-level logical description of the edits performed by the user.
-\item The system needs to handle the case when the type structure diverges from the code structure.
-This can be addressed in various ways (transform code, add error handlers, etc.) but it is desirable to avoid "breaking" the \texttt{render} function as this would make the new application state impossible to see.
-\end{itemize}
-
-\subsection{Remarks}
-It is worth noting that the particularly difficult aspect of this challenge, i.e., the case where the structure of state is refactored to an equivalent one, is related to the Extract Entity challenge.
-
 \newpage
-\textcolor{red}{TODO: The rest needs more editing}
-
-~
-
-~
 
@@
 
-\section{Challenge goals}
-\textcolor{red}{Text from the Extract Entity challenge that maybe should be lifted up to the top-level discussion of goals}
-
-The Extract Entity challenge for live programming is to provide interactive operations in the programming environment that will perform the schema change on live code and data.
-The data can be in any form desired, including relations, objects, or JSON.
-The code can be a simple CRUD UI that also provides the shipping department its view of unshipped orders.
-The central problem is to coordinate the schema change with code edits/refactorings to keep the system executing live and correctly.
-Solutions should try to minimize:
-
-\begin{enumerate}
-  \item The number and complexity of commands or UI affordances that must be introduced.
-  \item Interruption to live execution of the code.
-\end{enumerate}
-
-The Extract Entity challenge for local-first software is similar to the live programming challenge except that the context is an application that offers similar functionality with peer-to-peer collaboration.
-In this context the schema change does not need to be performed interactively -- a developer can take some time to build, test, and package a solution, perhaps using an API or DSL.
-The hard part comes when the schema change is to be deployed across all the replicas.
-This deployment must synchronize code and data upgrades (or decouple them as in Cambria~\cite{Cambria}).
-The deployment must also deal with migrating ``in flight'' operations so that all replicas converge on the same state without data loss, and ideally without centralized coordination.
-A radical approach could try to incorporate schema change operations into the underlying datastore/CRDT itself and integrate code as well, but layered solutions are welcome too.
-
-\subsection{Goals}
-Solutions should try to minimize:
-
-\begin{enumerate}
-  \item The need for users to manually intervene.
-  \item Possibilities that developer-written code is subtly incorrect in edge cases, preferably by reducing the need for developer-written code.
-  \item The need for a centralized coordinator.
-  \item How long replicas might be partitioned until resynchronization is acquired.
-\end{enumerate}
-
 \section{diagrams}
 
-\begin{figure}[H]
-  \centering
-  \begin{tikzcd}[sep = 6em]
-    \textsf{query}
-    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
+  \begin{tikzcd}[sep = huge]
+    \textsf{code}_1
+    \arrow[d, "\textsf{refactor}"{left},""{name=L}]
     &
-    \textsf{schema}
+    \textsf{schema}_1
     \arrow[l, "\textsf{constrains}"{above}]
-    \arrow[r, "\textsf{constrains}"{above}]
-    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
+    \arrow[r, "\textsf{constrains}"]
+    \arrow[d, red, "\textsf{change}"{near start, description}, ""{name=CL, left}, ""{name=CR}]
     &
-    \textsf{data}
-    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
+    \textsf{data}_1
+    \arrow[d, ""{name=R, left}, "\textsf{migrate}"{right}]
     \\
-    \textsf{query}_\alpha
+    \textsf{code}_2
     &
-    \textsf{schema}_\alpha
-    \arrow[l, "\textsf{constrains}"{above}]
-    \arrow[r, "\textsf{constrains}"{above}]
-    &
-    \textsf{data}_\alpha
-    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
-    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
+    \textsf{schema}_2
+    \arrow[l, "\textsf{constrains}"{below}]
+    \arrow[r, "\textsf{constrains}"{below}] &
+    \textsf{data}_2
+    \arrow[Rightarrow, from=CL, to=L, "\textsf{propagate}"{below}]
+    \arrow[Rightarrow, from=CR, to=R, "\textsf{propagate}"{below}]
   \end{tikzcd}
-  \caption{Extract Entity}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \begin{tikzcd}[column sep = large, row sep = 5em]
-    &
-    \textsf{schema}
-    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
-    \arrow[rr, "\textsf{constrains}"]
-    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
-    &
-    &
-    \textsf{data}
-    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
-    \arrow[Rightarrow, from=L, to=R]
-    \\
-    \textsf{schema}_\alpha
-    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
-    \arrow[rr, crossing over]
-    &
-    &
-    \textsf{data}_\alpha
-    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
-    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
-    &
-    \\
-    &
-    \textsf{schema}_\beta
-    \arrow[ld, "\mathit{rebase}\left(\textsf{change}_\alpha {\,,\;} \textsf{change}_\beta\right)"{name=L4, near start, description}]
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_\beta
-    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
-    \arrow[Rightarrow, from=L4, to=R4]
-    \\
-    \textsf{schema}_{\alpha\beta}
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_{\alpha\beta}
-    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
-    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
-    &
-  \end{tikzcd}
-  \caption{Mergeing divergent changes}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \begin{tikzcd}[column sep = large, row sep = 5em]
-    &
-    \textsf{schema}
-    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
-    \arrow[rr, "\textsf{constrains}"]
-    \arrow[dd, "\textsf{identity}"{near start, description}]
-    &
-    &
-    \textsf{data}
-    \arrow[dd, red, "\textsf{update}_\beta"{near start, description}]
-    \\
-    \textsf{schema}_\alpha
-    \arrow[rr, crossing over]
-    &
-    &
-    \textsf{data}_\alpha
-    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
-    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
-    \arrow[rr, red, crossing over, "\textsf{update}_\gamma"{near start, description}]
-    &
-    &
-    \textsf{data}_{\alpha\gamma}
-    \arrow[dd, "\mathit{rebase}\left(\textsf{update}_\beta {\,,\;} \textsf{update}_\gamma\right)"{near start, description}]
-    \\
-    &
-    \textsf{schema}
-    \arrow[ld, "\textsf{change}_\alpha"{name=L4, near start, description}]
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_\beta
-    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
-    \arrow[Rightarrow, from=L4, to=R4]
-    \\
-    \textsf{schema}_\alpha
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_{\alpha\beta}
-    &
-    &
-    \textsf{data}_{\alpha\beta\gamma}
-  \end{tikzcd}
-  \caption{Conference Organizer merging schema and divergent data updates}
-\end{figure}
 
 \section{Related Work}
+\label{sec:related}
+
+Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
+End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
+between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
+the established database products (see Related Work). Acme will need to write code or do a big manual conversion.
 
@@
 \newpage
-
-\section{Extract/Absorb/Split/Merge Entity}
-\textcolor{red}{Maybe called \textit{Individuation}? Maybe called \textit{Entity Evolution}?}
-
-The Acme Corporation needs to record orders from various customers for various products.
-The simplest implementation is a spreadsheet with a row for each order and columns for information about the customer and product.
-It might look like this:
-
-\begin{table}[h!]
-  \sffamily
-  \small
-  % \centering
-  \begin{tabular}{ |r|l|r|r|l|l|}
-   \hline
-   order\_id & item & quantity & ship\_date & customer\_name & customer\_address \\
-   \hline \hline
-   1 & Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
-   \hline
-   2 & Dynamite & 2 & & Daffy Duck & White Rock Lake \\
-   \hline
-   3 & Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
-   \hline
-  \end{tabular}
-  \caption{Orders}
-\end{table}
-
-The shipping department filters this table on blank ship dates to see what they need to ship.
-But after a while the orders department realizes they are wasting effort duplicating the address for new order from an old customer.
-And when the customer's address changes they have to go back and edit all of their orders.
-What is needed is two tables, one with orders that links to one with customers, like this:
-
-\begin{table}[h!]
-  \sffamily
-  \small
-  % \centering
-    \begin{tabular}{ |r|l|l|}
-      \hline
-      customer\_id & customer\_name & customer\_address \\
-      \hline \hline
-      1 & Wile E Coyote & 123 Desert Station \\
-      \hline
-      2 & Daffy Duck & White Rock Lake \\
-      \hline
-    \end{tabular}
-  \caption{Customers}
-\end{table}
-
-\begin{table}[h!]
-  \sffamily
-  \small
-  % \centering
-    \begin{tabular}{ |r|l|r|r|r|}
-     \hline
-     order\_id & item & quantity & ship\_date & customer\_id \\
-     \hline \hline
-     1 & Anvil & 1 & 2/3/23 & 1 \\
-     \hline
-     2 & Dynamite & 2 & & 2  \\
-     \hline
-     3 & Bird Seed & 1 & & 1 \\
-     \hline
-    \end{tabular}
-  \caption{Orders linking to Customers}
-\end{table}
-
-We often encounter this situation when we realize that the attributes of one type of entity should actually belong to a distinct type of entity that will be associated with the first one.
-Often the motivation is to centralize changes to those attributes in one place.
-That is what it means to be an \textit{entity}: a referenceable holder of mutable attributes.
-We call this operation \textit{Extract Entity}.
-
-Unique identifiers (here consecutive numbers) have been assigned to the customers and are referenced from the orders.
-This allows, for example, a spelling error in a customer address to be fixed in one place.
-It also allow a customer name or address to change without breaking the connection from all associated orders.
-Unique identifiers are the essence of entities, whether they are uniquely generated primary keys in a database or an abstraction of a memory address in a programming language.
-
-The \textit{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers, and 3) reference these identifiers from the orders.
-Our example took the simple approach of merging entities whose attributes are all equal.
-But that might not be correct in all cases.
-What if there was a typo in one of the addresses that made it look different from other instances of the customer?
-To repair that error we need to merge the falsely distinct customers into one and fix any references from orders.
-We call this operation \textit{Merge Entities} -- note that it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.
-
-Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
-End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links} between tables but do not provide operations that can do an \textit{Extract Entity}.
-Neither do the established database products (see Related Work).
-Acme will need to write code or do a big manual conversion.
-
-It is natural to expect every schema evolution to have an inverse, for the change in requirements that triggered an evolution might change back again, and we ought to pay off the technical debt of obsolete design decisions.
-Research on \textit{Divergence Control}~\cite{Foster2007, herrmann17, chillon22} further suggests operations should be invertible.
-The inverse of \textit{Extract Entity} is \textit{Absorb Entity}.
-It raises the question of how to handle \textit{one-to-many} relationships.
-For example should absorbing orders into customers do a relational join, returning us back to the starting point, or should it nest orders within customers, as a NoSQL database might do?
-
-The inverse of \textit{Merge Entities} is \textit{Split Entity}, but that can be awkward.
-Imagine that we didn't have addresses for customers to disambiguate them with and only knew their non-unique names.
-Then \textit{Extract Entity} will conflate them and discard the information needed to properly invert the operation.
-Should this information be retained?
-One could object that we shouldn't have taken orders for people we can't uniquely identify yet.
-And even if that was necessary it means the customers really aren't distinct entities and shouldn't have been extracted.
-These are arguments that we shouldn't need \textit{Split Entity}, contradicting the principle that evolution operations should be invertible.
-It is interesting to observe that evolution in nature can also fail to invert, leaving behind vestigial features.
-We consider this to be an unsettled issue.
-
-\subsection{Code Coevolution}
-A simple example of how code should evolve along with the schema is a SQL query to report all pending orders in the original schema:
-
-\begin{lstlisting}[language=SQL]
-SELECT order_id, item, quantity, customer_name, customer_address
-FROM Orders WHERE ship_date IS NULL;
-\end{lstlisting}
-
-The \textit{Extract Entity} operation described earlier should rewrite this query into:
-
-\begin{lstlisting}[language=SQL]
-SELECT order_id, item, quantity, customer_name, customer_address
-FROM Orders
-JOIN Customers ON Orders.customer_id = Customers.customer_id
-WHERE ship_date IS NULL;
-\end{lstlisting}
-
-Alternatively the \textit{Extract Entity} operation could create a view in which the original query is executed.
-
-\subsection{Goals}
-A solution to this challenge should:
-\begin{enumerate}
-  \item Define the \textit{Extract Entity} operation.
-  \item Define the \textit{Absorb Entity} operation and motivate the tradeoff between joining and nesting.
-  \item Define a \textit{Merge Entities} operation or if using automatic deduplication then discuss how to handle mistakes.
-  \item Define the \textit{Split Entity} operation or justify why it is not needed.
-  \item If proposing a user-facing solution explain the UI affordances invoking these operations.
-  \item Extend these operations to the Divergence Control scenario.
-\end{enumerate}
 
@@
 \caption{Possible program changes and how to deal with them at run time}
-\label{TBL:changes}
+\label{tbl:smchanges}
 \end{table}
 
-The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{TBL:changes}.
+The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{tbl:smchanges}.
 The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:
@@
 Typically a refactoring should be behavior preserving, but can it also preserve run-time data?
-The minimal example is the consistent variable rename rename in Table~\ref{TBL:changes}, which should not have any effect on run-time state.
-A more complex refactoring is described in Section~\ref{SECT:elm}.
+The minimal example is the consistent variable rename rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.
+A more complex refactoring is described in Section~\ref{sec:elm}.
 

### 2024-05-25 paper.tex

@@
 
-  \begin{tikzcd}[sep = huge]
-    \textsf{code}_1
-    \arrow[d, "\textsf{refactor}"{left},""{name=L}]
-    &
-    \textsf{schema}_1
+\begin{figure}[H]
+  \centering
+  \begin{tikzcd}[sep = 6em]
+    \textsf{query}
+    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
+    &
+    \textsf{schema}
     \arrow[l, "\textsf{constrains}"{above}]
-    \arrow[r, "\textsf{constrains}"]
-    \arrow[d, red, "\textsf{change}"{near start, description}, ""{name=CL, left}, ""{name=CR}]
-    &
-    \textsf{data}_1
-    \arrow[d, ""{name=R, left}, "\textsf{migrate}"{right}]
+    \arrow[r, "\textsf{constrains}"{above}]
+    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
+    &
+    \textsf{data}
+    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
     \\
-    \textsf{code}_2
-    &
-    \textsf{schema}_2
-    \arrow[l, "\textsf{constrains}"{below}]
-    \arrow[r, "\textsf{constrains}"{below}] &
-    \textsf{data}_2
-    \arrow[Rightarrow, from=CL, to=L, "\textsf{propagate}"{below}]
-    \arrow[Rightarrow, from=CR, to=R, "\textsf{propagate}"{below}]
+    \textsf{query}_\alpha
+    &
+    \textsf{schema}_\alpha
+    \arrow[l, "\textsf{constrains}"{above}]
+    \arrow[r, "\textsf{constrains}"{above}]
+    &
+    \textsf{data}_\alpha
+    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
+    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
   \end{tikzcd}
+  \caption{Extract Entity}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \begin{tikzcd}[column sep = large, row sep = 5em]
+    &
+    \textsf{schema}
+    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
+    \arrow[rr, "\textsf{constrains}"]
+    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
+    &
+    &
+    \textsf{data}
+    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
+    \arrow[Rightarrow, from=L, to=R]
+    \\
+    \textsf{schema}_\alpha
+    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
+    \arrow[rr, crossing over]
+    &
+    &
+    \textsf{data}_\alpha
+    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
+    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
+    &
+    \\
+    &
+    \textsf{schema}_\beta
+    \arrow[ld, "\textsf{change}_\alpha \succ\, \textsf{change}_\beta"{name=L4, near start, description}]
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_\beta
+    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
+    \arrow[Rightarrow, from=L4, to=R4]
+    \\
+    \textsf{schema}_{\alpha\beta}
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_{\alpha\beta}
+    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
+    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
+    &
+  \end{tikzcd}
+  \caption{Mergeing divergent changes with rebase ($\succ$)}
+\end{figure}
 
@@
 
-TODO: Maybe insert something about methodology in general?
-(ref the table challenges paper? although this paper is no longer so much about challenges)
-
-\paragraph{Schema evolution.}
-Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help.
-SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
-MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data.
-But apart from such special cases, SQL databases have no general purpose support for data migration.
-As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data.
-Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
-In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
-
-\citet{bernstein07} observed in 2007 ``There are hardly any schema evolution tools today.
-This is rather surprising since there is a huge literature on schema evolution spanning more than two decades.'' There are now more tools available.
-Some tools such as Liquibase~\cite{liquibase} and PlanetScale\cite{planetscale} could be characterized as version control and continuous integration/deployment for schema.
-They track schema changes and can calculate diffs in the form of SQL DDL statements to convert one schema version to another, but do not help migrate data.
-Unfortunately comparing schemas can be ambiguous about the intention of changes.
-For example has a column been renamed or has it been deleted and a new one created?
-That distinction makes a big difference to the data in that column.
-EvolveDB\cite{evolvedb} addresses this ambiguity by reverse-engineering the schema into a richer data model and tracking the edits to that model within an IDE.
-This more precise edit history can be used to infer higher level intentions of a schema change, which then generate SQL scripts to evolve the database.
-EdgeDB~\cite{edgedb} resolves ambiguities by asking questions of the developer, with some answers supplying custom migration code in a proprietary query language.
-
-Some tools provide a Domain Specific Language (DSL) to describe schema evolution.
-Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code.
-\citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
-SMOs can also rewrite queries to accomodate schema changes.
-\citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}.
-BiDEL appears capable of handling the basic requirements of our \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} challenges, though \textit{Split/Merge Entity} is less clear.
-It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.
-
-Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases.
-While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}.
-\citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON.
-\citet{scherzinger13} define a set of operators like the relational SMOs discussed above.
-\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} but not \textit{Split/Merge Entity} nor divergence.
-None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.
-
-Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}.
-Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}.
-Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.
-
-\paragraph{Live, MDD, etc. programming?}
-
-% Some references from Tomas
-%
-% collaborative programmign tools
-\cite{goldman2011real,kurniawan2015coder,replit}
-% exploratory/live review paper
-\cite{rein2018exploratory}
-% hot reload in live systems
-\cite{barenz2020essence,beckmann2021shortening}
-
-% Some references from Tijs
-%
-% SQL+code
-\cite{Cleve2006,wang2019synthesizing}
-% BX
-\cite{czarnecki2009bidirectional}
-% MDD
-\cite{metamodels,alanen2003difference}.
-
-\begin{comment}
-
-Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations
-
-dataframe diff https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
-
-Durable computing: pins runtime data frames to code versions. Doesn't handle problem of global data schema change.
-https://restate.dev/blog/solving-durable-executions-immutability-problem/
-https://restate.dev/blog/code-that-sleeps-for-a-month/
-
-Tijs on Live programming and model-driven engineering
-https://ir.cwi.nl/pub/23770
-https://dl.acm.org/doi/10.1007/s10270-017-0608-7
-https://dl.acm.org/doi/10.1145/3276604.3276611
-https://ir.cwi.nl/pub/21858
-
-From Geoffrey:
-1) Finding Data Compatibility Bugs with JSON Subschema Checking
-https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
-I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant.
-It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
-2) Confluent Schema Registry:
-https://docs.confluent.io/platform/current/schema-registry/index.html
-An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
-\end{comment}
+\begin{figure}[H]
+  \centering
+  \begin{tikzcd}[column sep = large, row sep = 5em]
+    &
+    \textsf{schema}
+    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
+    \arrow[rr, "\textsf{constrains}"]
+    \arrow[dd, "\textsf{identity}"{near start, description}]
+    &
+    &
+    \textsf{data}
+    \arrow[dd, red, "\textsf{update}_\beta"{near start, description}]
+    \\
+    \textsf{schema}_\alpha
+    \arrow[rr, crossing over]
+    &
+    &
+    \textsf{data}_\alpha
+    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
+    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
+    \arrow[rr, red, crossing over, "\textsf{update}_\gamma"{near start, description}]
+    &
+    &
+    \textsf{data}_{\alpha\gamma}
+    \arrow[dd, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha \succ\, \textsf{update}_\gamma"{near start, description}]
+    \\
+    &
+    \textsf{schema}
+    \arrow[ld, "\textsf{change}_\alpha"{name=L4, near start, description}]
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_\beta
+    \arrow[ld, "\textsf{migrate}_\alpha"{name=R4, near start, description}]
+    \arrow[Rightarrow, from=L4, to=R4]
+    \\
+    \textsf{schema}_\alpha
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_{\alpha\beta}
+    \arrow[from=uu, crossing over, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha"{near start, description}]
+    &
+    &
+    \textsf{data}_{\alpha\beta\gamma}
+  \end{tikzcd}
+  \caption{Conference Organizer merging schema and divergent data updates}
+\end{figure}
+
+\textcolor{red}{[JE] These diagrams are getting hairy. Are they actually helpful?}
 

### 2024-05-27 paper.tex
Better merge conflict fix
So much more related work!
make it stop

--- 2024-05-25 paper.tex
@@
 It typically considers data integrity, although co-evolution of database schema
-and code has also been studied \cite{Cleve2006,wang2019synthesizing}.
+and code has also been studied \cite{Cleve2006,wang19}.
 In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
@@
 community \cite{czarnecki2009bidirectional}, in work on model-driven development
-\cite{metamodels,alanen2003difference} and in more low-level work on dynamic software updating
+\cite{Cicchetti11,alanen2003} and in more low-level work on dynamic software updating
 \cite{hicks2005dynamic}.
@@
 
+Programming language theory undergirds research on type-driven transformations.
+Bidirectional Transformations~\cite{czarnecki2009bidirectional} have been used for data format conversion and view update.
+Coupled Transformations~\cite{Berdaguer07, alcino06, Cleve2006} express transformations coordinated between types and their instances by encoding them into functions on Haskell GADTs built with strategy combinators.
+\citet{JVisser08} extends the encoding to transform queries written in point-free style.
+\citet{lammel16} recapitulates coupled transformations within logic programming which extends it to transform logic programs.
+It is not clear how bidirectional transformations can handle generating the unique IDs required in some evolutions, nor the information loss and duplication which can arise in some divergence scenarios.
+
+Model Driven Engineering has also studied evolution.
+Unlike the textual artifacts involved in other domains, models typically assign unique indentifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
+Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}.
+\citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
+\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database.
+They consider the evolution of OO-style subclass hierarchies which goes beyond most other work.
+Their running example includes both our Extract Entity and Multiplicty Change examples, and would make for a good follow-on problem.
+However they do not address code/query rewriting.
+
+\citet{wang19} use program synthesis to rewrite SQL programs to adapt to a schema change.
+This process is driven by finding possible correspondences between columns in the old and new schema guided by hueristics on textual similarity.
+It is assumed the migration does not change datatypes or lose any information visible to the programs yet those are real needs.
+It is not clear how pathologically ambiguous schema changes are resolved (for example renaming equityped columns).
+
+Programming by example has been used to infer schema migration programs~\cite{wang20, Alexe11}.
+This work has dealt with the reality that schema live in the database and so examples must be supplied separately, but we wonder why not incorporate examples into a schema definition language?
+Then purely textual edits that properly adapted the examples could be used to infer migrations and resolve the ambiguities from comparing pure schema.
+
+\paragraph{Tijs} \cite{vanRozen15, vanRozen19, RuntimeConstraint, SemanticDeltas}
+
 \paragraph{Live, MDD, etc. programming?}
@@
 % Some references from Tomas
-%
-% collaborative programmign tools
+\paragraph{collaborative programmign tools}
 \cite{goldman2011real,kurniawan2015coder,replit}
-% exploratory/live review paper
+\paragraph{exploratory/live review paper}
 \cite{rein2018exploratory}
-% hot reload in live systems
+\paragraph{hot reload in live systems}
 \cite{barenz2020essence,beckmann2021shortening}
 
-% Some references from Tijs
-%
-% SQL+code
-\cite{Cleve2006,wang2019synthesizing}
-% BX
-\cite{czarnecki2009bidirectional}
-% MDD
-\cite{metamodels,alanen2003difference}.
-
 \begin{comment}
-From Tijs:
-
-BX community:
-https://www.cs.cornell.edu/~jnfoster/papers/grace-report.pdf
-(overview paper)
-
-Isil Dilig's stuff;
-https://www.cs.utexas.edu/~isil/pldi19b.pdf
-https://www.cs.utexas.edu/~isil/vldb20.pdf
-
-Eelco Visser/Sander Vermolen
-https://dblp.org/pid/10/6098.html
-(we’ve discussed the catalog before)
-
-On the code-database co-evolution problem:
-https://link.springer.com/chapter/10.1007/11877028_17
-(little less technical than Isil’s stuff)
-
-Coupled transformations:
-https://dl.acm.org/doi/10.1007/978-3-540-69611-7_19
-https://www.sciencedirect.com/science/article/pii/S1571066108003058
-https://link.springer.com/chapter/10.1007/978-3-540-87875-9_45
-
-(from the MDD angle, only the tip of the ice berg…)
-
-Divergence in model-driven:
-https://dl.acm.org/doi/10.1145/2000410.2000414
-
-The seminal model-diffing paper:
-http://btn1x4.inf.uni-bayreuth.de/publications/dotor_buchmann/SCM/ChefRepo/DiffUndMerge/Alanen2003.pdf
-(Diffing, versioning, collaborative editing, it’s all well researched in MDD)
 
@@
 https://restate.dev/blog/code-that-sleeps-for-a-month/
-
-Tijs on Live programming and model-driven engineering
-https://ir.cwi.nl/pub/23770
-https://dl.acm.org/doi/10.1007/s10270-017-0608-7
-https://dl.acm.org/doi/10.1145/3276604.3276611
-https://ir.cwi.nl/pub/21858
 

### 2024-05-27 paper.tex
Paragraph headings

@@
 
+\paragraph{Type-driven transformations}
 Programming language theory undergirds research on type-driven transformations.
@@
 
+\paragraph{Model Driven Engineering}
 Model Driven Engineering has also studied evolution.
@@
 
+\paragraph{Program synthesis}
 \citet{wang19} use program synthesis to rewrite SQL programs to adapt to a schema change.
@@
 
+\paragraph{Migration by example}
 Programming by example has been used to infer schema migration programs~\cite{wang20, Alexe11}.

### 2024-05-27 paper.tex
Phrasing

--- 2024-05-27 paper.tex
@@
 letting them adopt individual changes created by their collaborators (as in local-first software
-\cite{localfirst}). To provide immediate feedback, future programming systems will also need to
-transcend treating state as ephemeral, tied to the lifetime of a running program and reset when the
-program is edited.
+\cite{localfirst}). To provide live feedback, future programming systems will also need to
+transcend treating state as ephemeral and reset when the
+program is edited and rerun.
 
@@
 future of programming, there are a number of problems that need to be resolved in order for
-such systems to become a reality. The key problem is \emph{type evolution}, i.e., how to
-handle the change of schema of the data program works with in a way that preserves integrity
+such systems to become a reality. A key problem is \emph{type evolution}, i.e., how to
+gracefully adapt the shape of data used in a program while preserving integrity
 constraints, existing data and the logic of code.
@@
 In the case of programs, a more fundamental layer defines the shape of data and code.
-We refer to this structure as \emph{types}, although we do not use the term in an precise technical
+We refer to this structure as \emph{types}, although we do not use the term in a precise technical
 sense. In our conceptual framework, types may be data types that define the shape of data, they
 may be explicitly declared in code and checked statically, as well as dynamically checked or
-even implicit. The key property of types for us is that they determine the shape of other layers
+even implicit. The key property of types for us is that they constrain the shape of other layers
 of a program, namely data and code. As a result, a change of program types requires a
@@
 
-We now describe a conceptual framework lets us view a range of diverse set of real-world
+We now describe a conceptual framework that lets us view a diverse set of real-world
 type evolution problems in a unified way. We distinguish between two dimensions.
@@
 We consider the more permanent layer of \emph{types} and two less permanent layers of code and
-data that are structured according to types. In specific case studies, the layers may map to
+data that are structured by types. In specific case studies, the layers may map to
 different aspects of programs, some may be not present and some may be implicit.
@@
 
+\paragraph{Type-driven transformations.}
 Programming language theory undergirds research on type-driven transformations.
@@
 
+\paragraph{Model Driven Engineering.}
 Model Driven Engineering has also studied evolution.
@@
 
+\paragraph{Program synthesis.}
 \citet{wang19} use program synthesis to rewrite SQL programs to adapt to a schema change.
-This process is driven by finding possible correspondences between columns in the old and new schema guided by hueristics on textual similarity.
+This process is driven by finding possible correspondences between columns in the old and new schema guided by heuristics on textual similarity.
 It is assumed the migration does not change datatypes or lose any information visible to the programs yet those are real needs.
@@
 
+\paragraph{Migration by example.}
 Programming by example has been used to infer schema migration programs~\cite{wang20, Alexe11}.

### 2024-05-27 paper.tex
Possibly complete related work

@@
 \section{Live Modeling Languages Require Run-Time State Migration}
+\label{livemodeling}
 %Related work
@@
 \paragraph{Schema evolution.}
-Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help.
-SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
-MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data.
+Schema evolution has long been a major problem for SQL databases, which provide surprisingly little help.
+SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
+MYSql \cite{mysqlAlterTable} can reorder columns without destroying their data.
 But apart from such special cases, SQL databases have no general purpose support for data migration.
@@
 Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
-In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
+In \emph{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
 
@@
 
+\citet{wang19} use program synthesis to rewrite SQL programs to adapt to a schema change.
+This process is driven by finding possible correspondences between columns in the old and new schema guided by heuristics on textual similarity and confirmed by the existence of a synthesized equivalent program.
+\begin{comment}
+  I hate this paper - incredibly sophisticated techniques applied to an over-idealized problem.
+It is assumed the migration does not change datatypes or lose any information visible to the programs.
+It assumes the database has full code coverage.
+It ignores without explanation ambiguous schema changes (for example renaming equityped columns).
+\end{comment}
+
 Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases.
@@
 None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.
+
+\emph{Local-first software}~\cite{localfirst} adds the divergence dimension to NoSQL evolution, because the code running on different replicas can evolve at different paces.
+\citet{Cambria} applies lenses~\cite{Foster2007} to this problem, specifically discussing the \emph{Multiplicity Change} problem.
 
@@
 \paragraph{Model Driven Engineering.}
-Model Driven Engineering has also studied evolution.
+Model Driven Engineering must also handle evolution.
 Unlike the textual artifacts involved in other domains, models typically assign unique indentifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
@@
 
-\paragraph{Program synthesis.}
-\citet{wang19} use program synthesis to rewrite SQL programs to adapt to a schema change.
-This process is driven by finding possible correspondences between columns in the old and new schema guided by heuristics on textual similarity.
-It is assumed the migration does not change datatypes or lose any information visible to the programs yet those are real needs.
-It is not clear how pathologically ambiguous schema changes are resolved (for example renaming equityped columns).
-
 \paragraph{Migration by example.}
-Programming by example has been used to infer schema migration programs~\cite{wang20, Alexe11}.
-This work has dealt with the reality that schema live in the database and so examples must be supplied separately, but we wonder why not incorporate examples into a schema definition language?
-Then purely textual edits that properly adapted the examples could be used to infer migrations and resolve the ambiguities from comparing pure schema.
-
-\paragraph{Tijs} \cite{vanRozen15, vanRozen19, RuntimeConstraint, SemanticDeltas}
-
-\paragraph{Live, MDD, etc. programming?}
-
-% Some references from Tomas
-\paragraph{collaborative programmign tools}
-\cite{goldman2011real,kurniawan2015coder,replit}
-\paragraph{exploratory/live review paper}
-\cite{rein2018exploratory}
-\paragraph{hot reload in live systems}
-\cite{barenz2020essence,beckmann2021shortening}
-
-\begin{comment}
-
-Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations
-
-dataframe diff https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf
-
-Durable computing: pins runtime data frames to code versions. Doesn't handle problem of global data schema change.
-https://restate.dev/blog/solving-durable-executions-immutability-problem/
-https://restate.dev/blog/code-that-sleeps-for-a-month/
-
-From Geoffrey:
-1) Finding Data Compatibility Bugs with JSON Subschema Checking
-https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
-I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant.
-It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
-2) Confluent Schema Registry:
-https://docs.confluent.io/platform/current/schema-registry/index.html
-An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
-\end{comment}
+Programming by example has been used to infer schema migration programs from examples specified by the developer~\cite{wang20, Alexe11}.
+This work deals with the reality that schema live in the database and so examples must be supplied separately, but we wonder why not incorporate examples into a schema definition language?
+Textual edits to schema in this language would also adapt the examples, which then could be used to infer an abstract migration, thus avoiding the ambiguities of textually comparing pure schema.
+
+\paragraph{Live programming.}
+Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (mostly solo) programming.
+% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
+% A special case of live programming is \emph{hot reloading} (also called \emph{hot swapping})~\cite{barenz2020essence, } that installs code changes into a running system preserving state but not attempting to migrate that state through type changes.
+\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
+\citet{RuntimeConstraint} study the live modeling problem in Section \ref{livemodeling}.
+They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution.
+\citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
+%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.
+
+%\paragraph{collaborative programming tools}\cite{goldman2011real,kurniawan2015coder,replit}
 

### 2024-05-28 paper.tex
Emphasize SQL DDL limitations

@@
 
+There has been a lot of research on evolving interdependent layers of software, much of it guided by schema and types.
+But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system.
+Nevertheless it is important to understand these batch techniques to envision more interactive ones.
+
 \paragraph{Schema evolution.}
-Schema evolution has long been a major problem for SQL databases, which provide surprisingly little help.
+Schema evolution has long been a major problem for SQL databases because
+the SQL Data Definition Language (DDL) has limited capabilties to redefine existing data.
+It can rename tables and columns.
 SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
@@
 But apart from such special cases, SQL databases have no general purpose support for data migration.
-As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data.
+As a result in practice schema evolution is mostly done by writing custom Data Manipulation Language (DML) code to migrate the data.
 Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.

### 2024-05-28 paper.tex
Work in progress on conf organizer

--- 2024-05-27 paper.tex
@@
 \usepackage{float}
+\usepackage{wrapfig}
 
@@
   width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
-  outer arc=0pt]\setlength{\parskip}{0.5em}\setlength{\parindent}{0em}{\firamedium Framework perspective.}\quad #2
+  outer arc=0pt]\setlength{\parskip}{0.5em}#2
 \end{tcolorbox}}
+\newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}
 
@@
 \centering
-\includegraphics[width=0.6\textwidth]{figures/arrows.png}
+\includegraphics[scale=0.175]{figures/diag-data-code.png}
 \caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
@@
 \frameworkbox{
-The evolving \emph{type} in this case is the \texttt{State} type, \emph{code} is the
-program logic implemented in \texttt{update} and \texttt{render} and \emph{data} is the current
-state. To support live programming, a local type change requires synchronizing local data to
-match the new type, and also a change in code so that it continues working against the new type
-structure. In a non-local scenario, programmers would likely want to synchronize types and code,
-but not necessarily data. A collaborative programming system would need to support decentralized
-and possibly divergent type evolution, but without data synchronization.
+\begin{wrapfigure}{r}{14em}
+\vspace{1em}
+\includegraphics[scale=0.175]{figures/diag-both.png}
+\end{wrapfigure}
+\frameworkboxtitle{Code-data Co-evolution}
+In this case study, the evolving \emph{type} is the \texttt{State} type. When the type
+evolves, corresponding change needs to be applied to both \emph{code} and \emph{data}.
+The former is the program logic implemented in \texttt{update} and \texttt{render},
+while the latter is the current state (a live value of type \texttt{State}). The scenario
+illustrates that code and data need to co-evolve as illustrated by a revised diagram on the right.
 }
@@
 \frameworkbox{
-In a programming system, the \emph{types} to which entity evolution applies can be data tables,
-but also objects or JSON. When the type evolves, the system needs to update corresponding
-live \emph{data} and \emph{code}. In a non-local scenario, entity evolution
-typically concerns permanent data. As a result, the system needs to synchronize types,
-data and code across all replicas. Doing so in a way that allows divergence is discussed
-in the next section.
+\begin{wrapfigure}{r}{8em}
+\includegraphics[scale=0.175]{figures/diag-variants.png}
+\end{wrapfigure}
+\frameworkboxtitle{Primitive Type Transformations}
+In this example, \emph{type} refers to data tables, but the same scenario applies to
+objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
+that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
+The evolution can be done through a small set of primitive type transformations (illustrated on the
+right) with corresponding data transformation for each. The system may either expose the
+primitive transformations through a user interface, or infer them from textual edits made by the
+programmer, e.g. to a database table schema definition.
 }
@@
 
-\newpage
-~
-\newpage
-
-% elm
-% database
-% divergence
-% cambria
-% documents
-% state machine
-
-% DIVERGENCE IN TRANSITION (because you cannot do big bang update) c.f. Tijs' example
-
-% typically people just blow away the state but Elm actually has demos
-% of people doing hot reload without losing state.
-% (maybe they never change the type?)
-
-% ELM EXAMPLE
-% current example ("completes") is contrived - it is equivalent data
-% more interesting is when you lose / add data
-% add "completed" with default instead
-
-\section{Divergence Control}
-
-\textcolor{red}{This section was an extension of the Extract Entity challenge but should be lifted up to the top-level discussion of divergence}
-
-\subsection{Context}
-Schema evolution becomes more challenging when data, code, and schema cannot all be updated together in an atomic way.
-In these situations, different versions must coexist and interoperate with one another: for example, edits made in one schema must be applied in another, or code written against one schema must run against data stored in another.
-
-Such situations are common in software engineering.
-Web backend architectures often put data and code on separate machines, which makes it impossible to update both atomically.
-To perform upgrades, teams must manually perform multi-step [zero-downtime deployment processes](https://planetscale.com/blog/zero-downtime-rails-migrations-planetscale-rails-gem) to step the database and the code servers forward gradually such that they are compatible with one another at any given step.
-One example would be deploying a change to add a database column before deploying the code that uses the column.
-
-In the simplest cases, divergence can last only for a brief moments in between zero-downtime deployment steps.
-But divergence can also occur across longer timescales, especially in situations where the different parts of a system are not all centrally controlled.
-For example, a public web API may need to support applications using old versions of the API for months or years, since it is not realistic to force all consumers to upgrade immediately.
-Divergence can even be indefinite -- \citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.
-
-Divergence does not just occur in software engineering; it also appears in end-user workflows.
-Users frequently make copies of structured documents like spreadsheets and then diverge them by altering both data content and schema.
-In the case of spreadsheets schema changes include rearrangements to the structure of rows and columns as well as changes to formulas.
-The problem is that the users want to transfer such data and schema changes between these divergent copies, and they want to pick and choose which of these changes to transfer.
-In practice such transfer is done manually through copy \& paste.
-\citet{Basman19} has documented an ecology of emailed spreadsheets.
-\citet{Burnett14} distilled field observations of these practices in the story of Frieda, which we quote here:
-
-\begin{quotation}
-
-For example, consider ``Frieda'', an office manager in charge of her department's budget tracking.
-(Frieda was a participant in a set of interviews with spreadsheet users that the first author conducted.
-Frieda is not her real name.) Every year, the company she works for produces an updated budget tracking spreadsheet with the newest reporting requirements embedded in its structure and formulas.
-But this spreadsheet is not a perfect fit to the kinds of projects and sub-budgets she manages, so every year Frieda needs to change it.
-She does this by working with four variants of the spreadsheet at once: the one the company sent out last year (we will call it Official-lastYear), the one she derived from that one to fit her department's needs (Dept-lastYear), the one the company sent out this year (Official-thisYear), and the one she is trying to put together for this year (Dept-thisYear).
-
-Using these four variants, Frieda exploratively mixes reverse engineering, reuse, programming, testing, and debugging, mostly by trial-and-error.
-She begins this process by reminding herself of ways she changed last year's by reverse engineering a few of the differences between Official-lastYear and Dept-lastYear.
-She then looks at the same portions of Official-thisYear to see if those same changes can easily be made, given her department's current needs.
-
-She can reuse some of these same changes this year, but copying them into Dept-thisYear is troublesome, with some of the formulas automatically adjusting themselves to refer to Dept-lastYear.
-She patches these up (if she notices them), then tries out some new columns or sections of Dept-thisYear to reflect her new projects.
-She mixes in ``testing'' along the way by entering some of the budget values for this year and eyeballing the values that come out, then debugs if she notices something amiss.
-At some point, she moves on to another set of related columns, repeating the cycle for these.
-Frieda has learned over the years to save some of her spreadsheet variants along the way (using a different filename for each), because she might decide that the way she did some of her changes was a bad idea, and she wants to revert to try a different way she had started before.
-\end{quotation}
-
-\subsection{Example}
-The Divergence Control challenge instantiates the problem of divergent state with the example from the Extract Entity challenge.
-Every month the orders department sends its spreadsheet to the accounting department, which adds the data to a version of the spreadsheet that it has customized for financial tracking.
-But when the orders department migrates its spreadsheet to extract out customers the accounting department does not want to conform.
-They could manually convert incoming data in the new schema into their variant of the old schema.
-But they shouldn't have to.
-It should be possible to run new data through the schema change in reverse, converting it back into the format the accounting department is used to ingesting.
-Note that this is not a matter of synchronizing the divergent spreadsheets -- they maintain differently evolved schema.
-
-\subsection{Challenge}
-What is needed is a user-friendly mechanism for transferring data changes from one schema to another bidirectionally.
-In one direction we want to transfer changes made in the new schema into a variant of the old schema.
-The opposite direction might be needed, for example, if the accounting department makes corrections to customer addresses, which ought to be pushed through into the orders department's new schema.
-
-Bidirectional transfer of changes between divergent copies is similar in some ways to source code version control as in Git~\cite{ProGit}.
-There is \textit{forking} of long-lived divergent copies.
-We want to \textit{diff} these forks to see exactly how they have diverged.
-We want to partially \textit{merge} them by \textit{cherry picking} certain differences.
-Yet there are also many dissimilarities with Git: our data is more richly structured than lines of text; our schema changes are higher-level transformations on these structures than inserting and deleting characters; and we expect that end-users be able to understand it~\cite{gitless}.
-The Divergence Control challenge is in a sense to provide ``version control for schema change'' meeting these criteria, with the key technical challenge being the ability to transfer selected differences bidirectionally through schema changes as independently as possible.
-
-\subsection{Goals}
-This challenge invites a change of perspective in both live programming and local-first software.
-Live programming must move from being a solitary activity to a collaborative workflow, and one where the collaboration is not just on editing source code but on live integrated code and data, as in the Smalltalk/Lisp images of old.\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!} For its part, local-first software must move from automatically converging data replicas to also interactively managing long-term divergence.
-That implies either application state is being directly exposed to the user, or the application code is using an API to present version control affordances.
-We encourage both communities to think outside the box of Git, which has proven to baffle not only end-users but also a substantial fraction of developers.
-
-\section{diagrams}
-
-\begin{figure}[H]
+% ==================================================================================================
+% DOCUMENTS
+% ==================================================================================================
+
+\section{Evolution of Computational Documents}
+A different kind of type evolution is present in programming systems based on documents.
+The idea, dating back to Boxer \cite{diSessa86}, is to represent programs as documents
+with embedded computed values. In our model, a computed value is a document node whose value is
+specified by a simple formula (akin to those in Potluck~\cite{Litt2023} or spreadsheets) that
+may refer to other nodes of the document. We assume that documents can be concurrently edited
+by multiple users in a local-first manner, i.e., local edits are later synchronized with peers.
+
+Our concrete example is collaborative conference planning, illustred in Figure~\ref{fig:conf}
+and inspired by a typical use of Notion~\cite{notion}. A number of co-organizers are planning
+a conference that will feature talks by several invited speakers. They need to agree on the
+speaker list, contact speakers and calculate the conference budget.
+%
+The ideas in this section also apply to work on computational scientific notebooks
+that combine code and data such as Jupyter \cite{Kluyver2016}. We hope to contribute to ongoinig
+work on advancing the format as done by Nextjournal \cite{Nextjournal21}, Nota~\cite{Crichton2021}
+and Living Papers~\cite{Heer2023}.
+
+\begin{figure}
+\begin{adjustwidth}{-2.5em}{-2.5em}
 \centering
-  \begin{tikzcd}[sep = 6em]
-    \textsf{query}
-    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
-    &
-    \textsf{schema}
-    \arrow[l, "\textsf{constrains}"{above}]
-    \arrow[r, "\textsf{constrains}"{above}]
-    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
-    &
-    \textsf{data}
-    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
-    \\
-    \textsf{query}_\alpha
-    &
-    \textsf{schema}_\alpha
-    \arrow[l, "\textsf{constrains}"{above}]
-    \arrow[r, "\textsf{constrains}"{above}]
-    &
-    \textsf{data}_\alpha
-    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
-    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
-  \end{tikzcd}
-  \caption{Extract Entity}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \begin{tikzcd}[column sep = large, row sep = 5em]
-    &
-    \textsf{schema}
-    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
-    \arrow[rr, "\textsf{constrains}"]
-    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
-    &
-    &
-    \textsf{data}
-    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
-    \arrow[Rightarrow, from=L, to=R]
-    \\
-    \textsf{schema}_\alpha
-    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
-    \arrow[rr, crossing over]
-    &
-    &
-    \textsf{data}_\alpha
-    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
-    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
-    &
-    \\
-    &
-    \textsf{schema}_\beta
-    \arrow[ld, "\textsf{change}_\alpha \succ\, \textsf{change}_\beta"{name=L4, near start, description}]
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_\beta
-    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
-    \arrow[Rightarrow, from=L4, to=R4]
-    \\
-    \textsf{schema}_{\alpha\beta}
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_{\alpha\beta}
-    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
-    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
-    &
-  \end{tikzcd}
-  \caption{Mergeing divergent changes with rebase ($\succ$)}
-\end{figure}
-
-\section{Related Work}
-\label{sec:related}
-
-Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
-End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
-between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
-the established database products (see Related Work). Acme will need to write code or do a big manual conversion.
-
-\begin{figure}[H]
-  \centering
-  \begin{tikzcd}[column sep = large, row sep = 5em]
-    &
-    \textsf{schema}
-    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
-    \arrow[rr, "\textsf{constrains}"]
-    \arrow[dd, "\textsf{identity}"{near start, description}]
-    &
-    &
-    \textsf{data}
-    \arrow[dd, red, "\textsf{update}_\beta"{near start, description}]
-    \\
-    \textsf{schema}_\alpha
-    \arrow[rr, crossing over]
-    &
-    &
-    \textsf{data}_\alpha
-    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
-    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
-    \arrow[rr, red, crossing over, "\textsf{update}_\gamma"{near start, description}]
-    &
-    &
-    \textsf{data}_{\alpha\gamma}
-    \arrow[dd, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha \succ\, \textsf{update}_\gamma"{near start, description}]
-    \\
-    &
-    \textsf{schema}
-    \arrow[ld, "\textsf{change}_\alpha"{name=L4, near start, description}]
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_\beta
-    \arrow[ld, "\textsf{migrate}_\alpha"{name=R4, near start, description}]
-    \arrow[Rightarrow, from=L4, to=R4]
-    \\
-    \textsf{schema}_\alpha
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_{\alpha\beta}
-    \arrow[from=uu, crossing over, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha"{near start, description}]
-    &
-    &
-    \textsf{data}_{\alpha\beta\gamma}
-  \end{tikzcd}
-  \caption{Conference Organizer merging schema and divergent data updates}
-\end{figure}
-
-\textcolor{red}{[JE] These diagrams are getting hairy. Are they actually helpful?}
-
-\newpage
-
-\newpage
-
-\begin{figure}
-\begin{adjustwidth}{-5.5em}{-5.5em}
-\centering
-\begin{subfigure}[b]{23em}\vspace{0pt}
-   \fbox{\parbox{22.6em}{
-   \textbf{PROGRAMMING CONFERENCE 2023}\\
-   \textbf{Invited speakers}
-   \begin{itemize}
-    \item Adele Goldberg, adele@xerox.com
-    \item Margaret Hamilton, hamilton@mit.com
-    \item Betty Jean Jennings, betty@rand.com
-   \end{itemize}
-   \vspace{1em}
-   }}
-   \caption{Initial state of the document.}
-   \label{fig:conf-orig}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{23em}\vspace{0pt}
-   \fbox{\parbox{22.6em}{
+\begin{subfigure}[b]{18em}\vspace{0pt}
+   \fbox{\parbox{17.6em}{\small
    \textbf{PROGRAMMING CONFERENCE 2023}\\
@@
    }}
-   \caption{Added a speaker and sorted the list.}
+   \caption{Jonathan adds a new speaker and sorts the list.}
    \label{fig:conf-add}
 \end{subfigure}
-\\~\\
-\begin{subfigure}[b]{23em}\vspace{0pt}
-   \fbox{\parbox{22.6em}{
+\hfill
+\begin{subfigure}[b]{22em}\vspace{0pt}
+   \fbox{\parbox{21.6em}{\small
    \textbf{PROGRAMMING CONFERENCE 2023}\\
    \textbf{Invited speakers}
+   \vspace{0.3em}
 
@@
    }}
-   \caption{Refactored list into a table.}
+   \caption{Tomas refactors the list into a table.}
    \label{fig:conf-tab}
 \end{subfigure}
+\\~\\
+\begin{subfigure}[b]{18em}\vspace{0pt}
+   \fbox{\parbox{17.6em}{\small
+   \textbf{PROGRAMMING CONFERENCE 2023}\\
+   ~\\[-0.15em]
+   \textbf{Invited speakers}
+   \begin{itemize}
+    \item Adele Goldberg, adele@xerox.com
+    \item Margaret Hamilton, hamilton@mit.com
+    \item Betty Jean Jennings, betty@rand.com
+   \end{itemize}
+   \vspace{1.5em}
+   \textbf{Conference budget}\\
+   Travel cost per speaker:\\
+   \hspace*{2em} \$1200 \\
+   Number of speakers: \\
+   \hspace*{2em} \texttt{=COUNT(/ul[id='speakers']/li)} \\
+   Travel expenses:\\
+   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
+   }}
+   \caption{Tijs adds formulas to compute the total cost.}
+   \label{fig:conf-budget}
+\end{subfigure}
 \hfill
-\begin{subfigure}[b]{23em}\vspace{0pt}
-   \fbox{\parbox{22.6em}{
+\begin{subfigure}[b]{22em}\vspace{0pt}
+   \fbox{\parbox{21.6em}{\small
    \textbf{PROGRAMMING CONFERENCE 2023}\\
@@
    \end{tabular}
-   }}
-   \caption{Resulting document after change merging.}
-   \label{fig:conf-fin}
+
+   \vspace{0.5em}
+   \textbf{Conference budget}\\
+   Travel cost per speaker:\\
+   \hspace*{2em} \$1200 \\
+   Number of speakers: \\
+   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
+   Travel expenses:\\
+   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
+   }}\caption{Geoffrey adopts changes from all three co-organizers.}
+   \label{fig:conf-finfin}
 \end{subfigure}
 \end{adjustwidth}
-\vspace{0.5em}
-\caption{Conference organizer. Initial version of the document with two independent changes
-and the final version of the document after the two changes are merged.}
+\caption{
+Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
+Jennings). The first challenge involves merging two independent edits to the data
+(Fig.~\ref{fig:conf-add} and Fig.~\ref{fig:conf-tab}); the second merging structural change
+with code (Fig.~\ref{fig:conf-budget} and Fig.~\ref{fig:conf-tab}). Formulas in the final version
+(Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
 \label{fig:conf}
-\vspace{0.5em}
 \end{figure}
 
-\section{Conference Organizer: Merging edits in computational documents}
-In this section, we consider challenges related to working with a document-based format akin to HTML.
-We discuss two variants of the format. In the first format, a document
-contains only data. In the second variant, the format is extended with computed values.
-A computed value is a node in the document whose value is specified by a formula akin to those
-used in spreadsheets. The equations can refer to other nodes in the document, possibly using
-an absolute or a relative path and may involve a range of built-in functions, for example to
-count the number of elements or to sum a sequence of numerical values.
-
-We assume that the document can be
-concurrently edited by multiple users in the same fashion as in local-first software, i.e.,
-users may edit a local copy that then needs to be synchronized with changes made by other users.
-The edits to the document can be done in a rich-text editor, or by using a more structure-aware
-structure editor that can, for example, record changes as a sequence of high-level edit operations.
-
-The concrete challenges are centered around the problem of collaborative conference planning
-as illustrated in Figure~\ref{fig:conf}. A number of co-organizers are planning a conference
-that will feature talks by several invited speakers. They need to agree on the speaker list,
-contact speakers and calculate the conference budget.
+\challenge{Merging Structured Document Edits}
+
+The conference organizers first need to agree on a list of speakers to invite, coordinate who
+contacts whom and track the acceptance of invitations. They start with a list of three
+speakers (not shown). The first challenge is to merge document edits done locally
+by Jonathan and Tomas and change the list and the structure of the document:
+
+\begin{enumerate}
+\item Jonathan adds an additional speaker to the list and sorts the list of speakers
+  alphabetically by their first name. (Figure~\ref{fig:conf-add})
+\item Tomas refactors the list into a table. He splits the single textual value
+  into a name and an email (using a comma as the separator) and adds an additional column for
+  tracking which of the organizers should contact the speaker.
+\end{enumerate}
+
+\noindent
+The system should be able to merge the changes and produce a table containing the new data
+in a tabular format (not shown). The result needs to include the additional speaker
+created by Jonathan, use the order specified by Jonathan, but use the format defined by
+Tomas. The newly added speaker should be reformatted into the new format. For the newly added
+``Organizer'' column, Tomas may specify default value or the newly added row may use an
+empty value.
 
 \frameworkbox{
+\frameworkboxtitle{Framework perspective}
 In this scenario, the only transient \emph{state} is the result of computed values. The content
@@
 
-\paragraph{Applications.}
-One of the earliest systems to combine documents with computational elements
-is Boxer \cite{diSessa86}. The challenges in this section apply to a wide range of
-systems that can be seen as its successors. The specific scenario discussed here is
-inspired by the use of Notion~\cite{notion}, which is a collaborative structured document
-editing system. Treating documents as end-user programs has recently
-been explored in Potluck~\cite{Litt2023}, which is local-only but includes formulas similar
-to those in our challenge.
-
-The challenges in this section are also relevant to work on computational scientific notebooks.
-Jupyter \cite{Kluyver2016} uses a simple representation as a sequence of cells and does not have
-a mechanism for embedding data in the document and referring to it, but numerous
-project such as Nextjournal \cite{Nextjournal21} advance the format. The idea of producing
-interactive, computational secientific documents is an active research area. Recent contributions
-include Nota~\cite{Crichton2021} and Living Papers \cite{Heer2023}.
-
-\challenge{Merging structured document edits}
-
-First, the organizers need to agree on a list of speakers to invite, coordinate who contacts
-whom and track the acceptance of invitations. They start with a document shown in
-Figure~\ref{fig:conf-orig}. The challenge is to merge document edits that are done locally
-and independently by two co-organizers. Specifically, consider the following two edits:
-
-\begin{enumerate}
-\item The first organizer adds an additional speaker to the list and sorts the list of speakers
-  alphabetically by their first name. (Figure~\ref{fig:conf-add})
-\item The second organizer refactors the list into a table. They split the single textual value
-  into a name and an email (using a comma as the separator) and add an additional column for
-  tracking which of the organizers should contact the speaker.
-\end{enumerate}
-
-\noindent
-The system should be able to merge the changes and produce a final table shown in
-Figure~\ref{fig:conf-fin}. The resulting table needs to include the additional speaker
-created by the first organizer, use the order specified by the first organizer and use
-the format defined by the second organizer. The newly added speaker should be reformatted
-into the new format. For the newly added ``Organizer'' column, the second user may
-specify default value or the newly added row may use an empty value.
-
 \frameworkbox{
+\frameworkboxtitle{Framework perspective}
 In this challenge, the change done by the first organizer is at the \emph{data} layer,
@@
 
-\begin{figure}
-\begin{adjustwidth}{-5.5em}{-5.5em}
-\centering
-\begin{subfigure}[b]{23em}\vspace{0pt}
-   \fbox{\parbox{22.6em}{
-   \textbf{PROGRAMMING CONFERENCE 2023}\\
-   \textbf{Invited speakers}
-   \begin{itemize}
-    \item Adele Goldberg, adele@xerox.com
-    \item Margaret Hamilton, hamilton@mit.com
-    \item Betty Jean Jennings, betty@rand.com
-   \end{itemize}
-   \vspace{1.5em}
-   \textbf{Conference budget}\\
-   Travel cost per speaker:\\
-   \hspace*{2em} \$1200 \\
-   Number of speakers: \\
-   \hspace*{2em} \texttt{=COUNT(/ul[id='speakers']/li)} \\
-   Travel expenses:\\
-   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
-   }}
-   \caption{Initial state of the document with formulas.}
-   \label{fig:conf-budget}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{23em}\vspace{0pt}
-   \fbox{\parbox{22.6em}{
-   \textbf{PROGRAMMING CONFERENCE 2023}\\
-   \textbf{Invited speakers}
-
-   \begin{tabular}{| l | l | l |}
-     \hline
-     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
-     \hline \hline
-     Ada Lovelace & ada@rsoc.ac.uk & \\
-     \hline
-     Adele Goldberg & adele@xerox.com & TP\\
-     \hline
-     Betty Jean Jennings & betty@rand.com & JE\\
-     \hline
-     Margaret Hamilton & hamilton@mit.com & JE\\
-     \hline
-   \end{tabular}
-
-   \vspace{0.5em}
-   \textbf{Conference budget}\\
-   Travel cost per speaker:\\
-   \hspace*{2em} \$1200 \\
-   Number of speakers: \\
-   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
-   Travel expenses:\\
-   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
-   }}\caption{Resulting document after schema change merging.}
-   \label{fig:conf-finfin}
-\end{subfigure}
-\caption{Conference organizer. Formulas to calculate the budget have been added to the
-original document (Figure~\ref{fig:conf-orig}) and those have to be merged with the
-refactoring of a list into a table (Fig~\ref{fig:conf-tab}). Formulas in the final version
-are updated to match with the new document structure.}
-\label{fig:conf-calc}
-\end{adjustwidth}
-\end{figure}
-
 \challenge{Applying schema edits in presence of code}
@@
 \frameworkbox{
+\frameworkboxtitle{Framework perspective}
 In this challenge, the change done by the first organizer is at the \emph{code} layer, while
@@
 \frameworkbox{
+\frameworkboxtitle{Framework perspective}
 In the \emph{divergence} model, it may be expected that users will want to adopt all changes
@@
 \frameworkbox{
+\frameworkboxtitle{Framework perspective}
 In the live extension, we also consider the transient \emph{state} pace layer. A system should
@@
 }
+
+\newpage
+~
+\newpage
+
+% elm
+% database
+% divergence
+% cambria
+% documents
+% state machine
+
+% DIVERGENCE IN TRANSITION (because you cannot do big bang update) c.f. Tijs' example
+
+% typically people just blow away the state but Elm actually has demos
+% of people doing hot reload without losing state.
+% (maybe they never change the type?)
+
+% ELM EXAMPLE
+% current example ("completes") is contrived - it is equivalent data
+% more interesting is when you lose / add data
+% add "completed" with default instead
+
+\section{Divergence Control}
+
+\textcolor{red}{This section was an extension of the Extract Entity challenge but should be lifted up to the top-level discussion of divergence}
+
+\subsection{Context}
+Schema evolution becomes more challenging when data, code, and schema cannot all be updated together in an atomic way.
+In these situations, different versions must coexist and interoperate with one another: for example, edits made in one schema must be applied in another, or code written against one schema must run against data stored in another.
+
+Such situations are common in software engineering.
+Web backend architectures often put data and code on separate machines, which makes it impossible to update both atomically.
+To perform upgrades, teams must manually perform multi-step [zero-downtime deployment processes](https://planetscale.com/blog/zero-downtime-rails-migrations-planetscale-rails-gem) to step the database and the code servers forward gradually such that they are compatible with one another at any given step.
+One example would be deploying a change to add a database column before deploying the code that uses the column.
+
+In the simplest cases, divergence can last only for a brief moments in between zero-downtime deployment steps.
+But divergence can also occur across longer timescales, especially in situations where the different parts of a system are not all centrally controlled.
+For example, a public web API may need to support applications using old versions of the API for months or years, since it is not realistic to force all consumers to upgrade immediately.
+Divergence can even be indefinite -- \citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.
+
+Divergence does not just occur in software engineering; it also appears in end-user workflows.
+Users frequently make copies of structured documents like spreadsheets and then diverge them by altering both data content and schema.
+In the case of spreadsheets schema changes include rearrangements to the structure of rows and columns as well as changes to formulas.
+The problem is that the users want to transfer such data and schema changes between these divergent copies, and they want to pick and choose which of these changes to transfer.
+In practice such transfer is done manually through copy \& paste.
+\citet{Basman19} has documented an ecology of emailed spreadsheets.
+\citet{Burnett14} distilled field observations of these practices in the story of Frieda, which we quote here:
+
+\begin{quotation}
+
+For example, consider ``Frieda'', an office manager in charge of her department's budget tracking.
+(Frieda was a participant in a set of interviews with spreadsheet users that the first author conducted.
+Frieda is not her real name.) Every year, the company she works for produces an updated budget tracking spreadsheet with the newest reporting requirements embedded in its structure and formulas.
+But this spreadsheet is not a perfect fit to the kinds of projects and sub-budgets she manages, so every year Frieda needs to change it.
+She does this by working with four variants of the spreadsheet at once: the one the company sent out last year (we will call it Official-lastYear), the one she derived from that one to fit her department's needs (Dept-lastYear), the one the company sent out this year (Official-thisYear), and the one she is trying to put together for this year (Dept-thisYear).
+
+Using these four variants, Frieda exploratively mixes reverse engineering, reuse, programming, testing, and debugging, mostly by trial-and-error.
+She begins this process by reminding herself of ways she changed last year's by reverse engineering a few of the differences between Official-lastYear and Dept-lastYear.
+She then looks at the same portions of Official-thisYear to see if those same changes can easily be made, given her department's current needs.
+
+She can reuse some of these same changes this year, but copying them into Dept-thisYear is troublesome, with some of the formulas automatically adjusting themselves to refer to Dept-lastYear.
+She patches these up (if she notices them), then tries out some new columns or sections of Dept-thisYear to reflect her new projects.
+She mixes in ``testing'' along the way by entering some of the budget values for this year and eyeballing the values that come out, then debugs if she notices something amiss.
+At some point, she moves on to another set of related columns, repeating the cycle for these.
+Frieda has learned over the years to save some of her spreadsheet variants along the way (using a different filename for each), because she might decide that the way she did some of her changes was a bad idea, and she wants to revert to try a different way she had started before.
+\end{quotation}
+
+\subsection{Example}
+The Divergence Control challenge instantiates the problem of divergent state with the example from the Extract Entity challenge.
+Every month the orders department sends its spreadsheet to the accounting department, which adds the data to a version of the spreadsheet that it has customized for financial tracking.
+But when the orders department migrates its spreadsheet to extract out customers the accounting department does not want to conform.
+They could manually convert incoming data in the new schema into their variant of the old schema.
+But they shouldn't have to.
+It should be possible to run new data through the schema change in reverse, converting it back into the format the accounting department is used to ingesting.
+Note that this is not a matter of synchronizing the divergent spreadsheets -- they maintain differently evolved schema.
+
+\subsection{Challenge}
+What is needed is a user-friendly mechanism for transferring data changes from one schema to another bidirectionally.
+In one direction we want to transfer changes made in the new schema into a variant of the old schema.
+The opposite direction might be needed, for example, if the accounting department makes corrections to customer addresses, which ought to be pushed through into the orders department's new schema.
+
+Bidirectional transfer of changes between divergent copies is similar in some ways to source code version control as in Git~\cite{ProGit}.
+There is \textit{forking} of long-lived divergent copies.
+We want to \textit{diff} these forks to see exactly how they have diverged.
+We want to partially \textit{merge} them by \textit{cherry picking} certain differences.
+Yet there are also many dissimilarities with Git: our data is more richly structured than lines of text; our schema changes are higher-level transformations on these structures than inserting and deleting characters; and we expect that end-users be able to understand it~\cite{gitless}.
+The Divergence Control challenge is in a sense to provide ``version control for schema change'' meeting these criteria, with the key technical challenge being the ability to transfer selected differences bidirectionally through schema changes as independently as possible.
+
+\subsection{Goals}
+This challenge invites a change of perspective in both live programming and local-first software.
+Live programming must move from being a solitary activity to a collaborative workflow, and one where the collaboration is not just on editing source code but on live integrated code and data, as in the Smalltalk/Lisp images of old.\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!} For its part, local-first software must move from automatically converging data replicas to also interactively managing long-term divergence.
+That implies either application state is being directly exposed to the user, or the application code is using an API to present version control affordances.
+We encourage both communities to think outside the box of Git, which has proven to baffle not only end-users but also a substantial fraction of developers.
+
+\section{diagrams}
+
+\begin{figure}[H]
+  \centering
+  \begin{tikzcd}[sep = 6em]
+    \textsf{query}
+    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
+    &
+    \textsf{schema}
+    \arrow[l, "\textsf{constrains}"{above}]
+    \arrow[r, "\textsf{constrains}"{above}]
+    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
+    &
+    \textsf{data}
+    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
+    \\
+    \textsf{query}_\alpha
+    &
+    \textsf{schema}_\alpha
+    \arrow[l, "\textsf{constrains}"{above}]
+    \arrow[r, "\textsf{constrains}"{above}]
+    &
+    \textsf{data}_\alpha
+    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
+    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
+  \end{tikzcd}
+  \caption{Extract Entity}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \begin{tikzcd}[column sep = large, row sep = 5em]
+    &
+    \textsf{schema}
+    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
+    \arrow[rr, "\textsf{constrains}"]
+    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
+    &
+    &
+    \textsf{data}
+    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
+    \arrow[Rightarrow, from=L, to=R]
+    \\
+    \textsf{schema}_\alpha
+    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
+    \arrow[rr, crossing over]
+    &
+    &
+    \textsf{data}_\alpha
+    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
+    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
+    &
+    \\
+    &
+    \textsf{schema}_\beta
+    \arrow[ld, "\textsf{change}_\alpha \succ\, \textsf{change}_\beta"{name=L4, near start, description}]
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_\beta
+    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
+    \arrow[Rightarrow, from=L4, to=R4]
+    \\
+    \textsf{schema}_{\alpha\beta}
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_{\alpha\beta}
+    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
+    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
+    &
+  \end{tikzcd}
+  \caption{Mergeing divergent changes with rebase ($\succ$)}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \begin{tikzcd}[column sep = large, row sep = 5em]
+    &
+    \textsf{schema}
+    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
+    \arrow[rr, "\textsf{constrains}"]
+    \arrow[dd, "\textsf{identity}"{near start, description}]
+    &
+    &
+    \textsf{data}
+    \arrow[dd, red, "\textsf{update}_\beta"{near start, description}]
+    \\
+    \textsf{schema}_\alpha
+    \arrow[rr, crossing over]
+    &
+    &
+    \textsf{data}_\alpha
+    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
+    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
+    \arrow[rr, red, crossing over, "\textsf{update}_\gamma"{near start, description}]
+    &
+    &
+    \textsf{data}_{\alpha\gamma}
+    \arrow[dd, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha \succ\, \textsf{update}_\gamma"{near start, description}]
+    \\
+    &
+    \textsf{schema}
+    \arrow[ld, "\textsf{change}_\alpha"{name=L4, near start, description}]
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_\beta
+    \arrow[ld, "\textsf{migrate}_\alpha"{name=R4, near start, description}]
+    \arrow[Rightarrow, from=L4, to=R4]
+    \\
+    \textsf{schema}_\alpha
+    \arrow[rr]
+    &
+    &
+    \textsf{data}_{\alpha\beta}
+    \arrow[from=uu, crossing over, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha"{near start, description}]
+    &
+    &
+    \textsf{data}_{\alpha\beta\gamma}
+  \end{tikzcd}
+  \caption{Conference Organizer merging schema and divergent data updates}
+\end{figure}
+
+\textcolor{red}{[JE] These diagrams are getting hairy. Are they actually helpful?}
+
+\newpage
+
+\newpage
 
@@
 \section{Related Work}
+\label{sec:related}
+
+Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
+End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
+between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
+the established database products (see Related Work). Acme will need to write code or do a big manual conversion.
 

### 2024-05-28 paper.tex
orphaned text in extract entity

@@
 both in collaboration with other programmers (through collaborative development tools
-\cite{goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
+\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
 programming environment (through live programming \cite{rein2018exploratory}).
@@
 of a memory address in a programming language.
+
+\textcolor{red}{[JE. Found this snippet in the related work]
+Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
+End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
+between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
+the established database products (see Related Work). Acme will need to write code or do a big manual conversion.
+}
 
@@
 
-Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
-End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
-between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
-the established database products (see Related Work). Acme will need to write code or do a big manual conversion.
-
-TODO: Maybe insert something about methodology in general?
-(ref the table challenges paper? although this paper is no longer so much about challenges)
+\textcolor{red}{Maybe insert something about methodology in general?
+(ref the table challenges paper? although this paper is no longer so much about challenges)}
 
@@
 % \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
-% A special case of live programming is \emph{hot reloading} (also called \emph{hot swapping})~\cite{barenz2020essence, } that installs code changes into a running system preserving state but not attempting to migrate that state through type changes.
+% A special case of live programming is \emph{hot reloading} or \emph{hot swapping})~\cite{barenz2020essence, hicks2005dynamic} that installs code changes into a running system preserving state but not attempting to migrate that state through type changes.
 \citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.

### 2024-05-28 paper.tex
Replace hand-drawn diagrams

@@
 The situation is illustrated in Figure~\ref{fig:change} using two variations of a diagram that
-we will return to repeatedly in this paper. When a type changes from $T_1$ to $T_2$,
+we will return to repeatedly in this paper. When a type changes from $\textsf{type}_1$ to $\textsf{type}_2$,
 we need to find and apply a corresponding change to other parts of the program whose shape
 is defined by the changed type. The problem of \emph{type evolution} is thus finding and
-applying changes that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$
+applying changes that will update $\textsf{data}_1$ and $\textsf{code}_1$ into new versions $\textsf{data}_1$ and $\textsf{code}_2$
 that correspond to the new type.
@@
 \centering
-\includegraphics[scale=0.175]{figures/diag-data-code.png}
-\caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
-  in data and code whose shape is defined by type. The challenge is finding a suitable
+%\includegraphics[scale=0.175]{figures/diag-data-code.png}
+\begin{tikzcd}[sep = 6em]
+  \textsf{code}_1
+  \arrow[d, "\textsf{refactor}"{name=L, description}]
+  &
+  \textsf{type}_1
+  \arrow[l, "\textsf{constrains}"{above}]
+  \arrow[r, "\textsf{constrains}"{above}]
+  \arrow[d, red, "\textsf{change}"{name=C, description}]
+  &
+  \textsf{data}_1
+  \arrow[d, "\textsf{migrate}"{name=R, description}]
+  \\
+  \textsf{code}_2
+  &
+  \textsf{type}_2
+  \arrow[l, "\textsf{constrains}"{above}]
+  \arrow[r, "\textsf{constrains}"{above}]
+  &
+  \textsf{data}_2
+  \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
+  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
+\end{tikzcd}
+\caption{When a type changes from $\textsf{type}_1$ to $\textsf{type}_2$, a corresponding change needs to take place
+  in data and code constrained by the type. The challenge is finding a suitable
   corresponding transformation (semi-)automatically.}
@@
 \frameworkbox{
-\begin{wrapfigure}{r}{14em}
-\vspace{1em}
-\includegraphics[scale=0.175]{figures/diag-both.png}
-\end{wrapfigure}
+% \begin{wrapfigure}{r}{14em}
+% \vspace{1em}
+% \includegraphics[scale=0.175]{figures/diag-both.png}
+% \end{wrapfigure}
 \frameworkboxtitle{Code-data Co-evolution}
-In this case study, the evolving \emph{type} is the \texttt{State} type. When the type
+In this case study, the evolving type in Figure~\ref{fig:change} is the \texttt{State} type. When the type
 evolves, corresponding change needs to be applied to both \emph{code} and \emph{data}.
@@
 
+\begin{figure}[H]
+\begin{tikzcd}[sep = 6em]
+  \textsf{type}_1
+  \arrow[r, "\textsf{constrains}"{above}]
+  \arrow[d, red, "\textsf{change}_1"{name=C, description}]
+  &
+  \textsf{data}_1
+  \arrow[d, "\textsf{migrate}_1"{name=R, description}]
+  \\
+  \textsf{type}_2
+  \arrow[r, "\textsf{constrains}"{above}]
+  \arrow[d, red, "\textsf{change}_2"{name=C2, description}]
+  &
+  \textsf{data}_2
+  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
+  \arrow[d, "\textsf{migrate}_2"{name=R2, description}]
+  \\
+  \textsf{type}_3
+  \arrow[r, "\textsf{constrains}"{above}]
+  &
+  \textsf{data}_3
+  \arrow[Rightarrow, from=C2, to=R2, "\textsf{propagate}"{above}]
+\end{tikzcd}
+\caption{Multiversion type evolution.}
+\label{fig:multiversion}
+\end{figure}
+
 \frameworkbox{
-\begin{wrapfigure}{r}{8em}
-\includegraphics[scale=0.175]{figures/diag-variants.png}
-\end{wrapfigure}
+% wrapfigure doesn't seem to work with tikz
+%\begin{wrapfigure}{r}{8em}
+% \includegraphics[scale=0.175]{figures/diag-variants.png}
+%\end{wrapfigure}
 \frameworkboxtitle{Primitive Type Transformations}
-In this example, \emph{type} refers to data tables, but the same scenario applies to
+In this example, \textsf{type} in Figure~\ref{fig:multiversion} refers to data tables, but the same scenario applies to
 objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
@@
 \newpage
-~
-\newpage
 

### 2024-05-28 paper.tex
Finish documents, integrating divergence and multiplicity

--- 2024-05-28 paper.tex
@@
 \noindent
-As an example, consider the well-known TODO list app. Its state consists of a list of items,
+As an example, consider the well-known todo list app. Its state consists of a list of items,
 each with a unique ID, a title and a flag indicating whether it has been completed.
@@
 \challenge{Live State Type Evolution}
-Assume that we have a running TODO list with the above state and events. To test the application,
+Assume that we have a running todo list with the above state and events. To test the application,
 the programmer has already created a number of items and so there is a value of the
@@
 
-\subsection{Remarks}
+\subsection*{Remarks: Requirements and Implementation}
 The key requirement from the live programming perspective is to ``do minimal harm''.
@@
 \frameworkbox{
-\begin{wrapfigure}{r}{8em}
-\includegraphics[scale=0.175]{figures/diag-variants.png}
+\begin{wrapfigure}{r}{9em}
+\includegraphics[scale=0.2]{figures/diag-variants.png}
 \end{wrapfigure}
@@
 
-\subsection{Remarks}
+\subsection*{Remarks: Representation and Replication}
 The central problem is to coordinate the type evolution with code edits to keep the system
@@
 The idea, dating back to Boxer \cite{diSessa86}, is to represent programs as documents
-with embedded computed values. In our model, a computed value is a document node whose value is
-specified by a simple formula (akin to those in Potluck~\cite{Litt2023} or spreadsheets) that
-may refer to other nodes of the document. We assume that documents can be concurrently edited
-by multiple users in a local-first manner, i.e., local edits are later synchronized with peers.
-
-Our concrete example is collaborative conference planning, illustred in Figure~\ref{fig:conf}
+with embedded computations. In our model, document can contain simple formulas (akin to those
+in Potluck~\cite{Litt2023} or spreadsheets) that may refer to other nodes of the document.
+We assume that documents can be concurrently edited by multiple users in a local-first manner,
+i.e., local edits are later synchronized with peers.
+
+The challenge focuses on collaborative conference planning, illustred in Figure~\ref{fig:conf}
 and inspired by a typical use of Notion~\cite{notion}. A number of co-organizers are planning
@@
 
-\challenge{Merging Structured Document Edits}
-
+\challenge{Structured Document Edits}
 The conference organizers first need to agree on a list of speakers to invite, coordinate who
@@
 \item Jonathan adds an additional speaker to the list and sorts the list of speakers
-  alphabetically by their first name. (Figure~\ref{fig:conf-add})
+  alphabetically by their first name. This changes data, but not the structure (type) of
+  the document (Figure~\ref{fig:conf-add}).
 \item Tomas refactors the list into a table. He splits the single textual value
   into a name and an email (using a comma as the separator) and adds an additional column for
-  tracking which of the organizers should contact the speaker.
+  tracking which of the organizers should contact the speaker (Figure~\ref{fig:conf-tab}).
+  This changes the document structure (type).
 \end{enumerate}
@@
 created by Jonathan, use the order specified by Jonathan, but use the format defined by
-Tomas. The newly added speaker should be reformatted into the new format. For the newly added
-``Organizer'' column, Tomas may specify default value or the newly added row may use an
-empty value.
+Tomas. The newly added speaker should be reformatted into the new format; for the
+``Organizer'' column, the system may use a suitable default value or ask the user interactively.
 
 \frameworkbox{
-\frameworkboxtitle{Framework perspective}
-In this scenario, the only transient \emph{state} is the result of computed values. The content
-of the document corresponds to the \emph{data} pace layer. A change in the data layer may invalidate
-the computed value at the state layer. Formulas correspond to the \emph{code} layer and changing them,
-again, invalidates the state. The \emph{schema} pace layer is implicit. A change to the document that
-modifies content is a mere data change, but restructuring the document is an implicit schema
-change. Such schema change that may affect the code pace layer.
-
-The scenario assumes the \emph{decentralized} model. There are multiple program (document)
-variants and it should be possible to propagate a change at any pace layer to other program
-variants. The challenge may be solved using both the easier \emph{convergence} and the more
-challenging \emph{divergence} model.}
-
-\frameworkbox{
-\frameworkboxtitle{Framework perspective}
-In this challenge, the change done by the first organizer is at the \emph{data} layer,
-while the change done by the second organizer spans the \emph{schema} and \emph{data} layers.
-In implementing the schema change, the second organizer transforms the affected document data.
-The challenge is to apply the same transformation to the data independently added by the
-second organizer.
-}
-
-\paragraph{Requirements.}
-The challenge is concerned with merging two sequences edits to a base document.
-A solution to the challenge should strive to satisfy the following goals:
-
-\begin{itemize}
-\item \emph{Completeness.} It should be possible to semi-automatically merge any two sequences
-  of edits. In other words, the merging should always be defined, regardless of what edits the
-  users perform, although user input may sometimes be necessary.
-\item \emph{Commutativity.} The order of edits should not matter, regardless of their pace layer.
-  If users $A$ and $B$ perform edits independently starting from the same initial document, the
-  resulting document should be the same if the system treats edits from $A$ as occurring before
-  the edits of $B$ and vice versa.
-\item \emph{Conflict resolution.} It may not always be possible to merge conflicting edits.
-  In this case, the system should interactively ask the user for guidance or, possibly, use
-  default behavior specified by the user.
-\item \emph{Structure editing.} The system may use a structure editor that provides high-level
-  commands for operations such as reordering of elements or refactoring of a list into a table.
-  In other words, we recognise that the challenge may not be solvable in a system based on plain
-  text editing of document source code.
-\end{itemize}
-
-\paragraph{Remarks.}
-As discussed earlier, this challenge does not make an explicit distinction between the schema
-of the document and data. The implementing system may or may not maintain such distinction.
-In our example, adding a new speaker and sorting the list is a mere change of the data, but the
-refactoring of a list into a table is a schema change. The system may offer different user interface
-for changes at different layers and leverage knowledge about such distinction, for example,
-by handling the merging of schema-related and data-related changes differently.
-
-\challenge{Applying schema edits in presence of code}
-
-This challenge extends the previous problem. The conference organizers now also want to use
-the document to manage the conference budget. In a very simplified form, one aspect
-of this is calculating the estimated travel expenses for the speakers. For this,
-they add a computed value to the document that counts the number of speakers
-and multiplies the result by a fixed cost per speaker elsewhere in the document.
-
-The scenario is shown in Figure~\ref{fig:conf-budget}. Assume two co-organizers start with
-the same original document shown in Figure~\ref{fig:conf-orig}. One adds an additional
-section with computed values, while the other performs the edits discussed earlier.
+\begin{wrapfigure}{r}{8em}
+\includegraphics[scale=0.25]{figures/diag-data-change.png}
+\end{wrapfigure}
+\frameworkboxtitle{Independent Data and Code Edits}
+In this example, the \emph{type} is implicit and represents the structure of the document.
+A change to the \emph{type} done by one organizer needs to be applied to \emph{data} that
+has been modified by the other organizer in the meantime, as illustrated on the right.
+In the next section, we will see similar pattern, but with formulas, or \emph{code}.
+We assume the \emph{decentralized} scenario where \emph{type} evolution can occur in any variant
+and \emph{covergence} model where all variants eventually adopt all changes. As discussed
+later, the system could also operate in the more challenging \emph{divergence} model.}
+
+\subsection*{Remarks: Requirements and Representation}
+It should be possible to automatically merge any two sequences of edits (changing both
+the document structure and data) and the order in which the edits are applied should also not
+matter. This may not always be possible and the system may ask the user for resolution or,
+possibly, use a default behavior specified by the user.
+
+The type and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
+lists edits appearing in our examples (but not a complete list). A system may
+use a structure editor that provides high-level commands for triggering those edits.
+Recognising user intent from plain text editing of source code may make the challenge harder.
+
+\challenge{Code Co-Evolution for Structured Document Edits}
+The conference organizers now also use the document to calculate the conference budget.
+They add a formula that multiplies a fixed value by the number of speakers, obtained
+by counting the number of element of a specified list in the document.
+
+\begin{enumerate}
+\item Tijs adds budget calculation to the original document (Figure~\ref{fig:conf-budget}).
+  This includes two formulas. The first selects all \texttt{li} elements of an \texttt{ul} element
+  with ID \texttt{speakers} (not visible in the user interface) and counts their number. The second
+  formula multiplies the constant travel cost per speaker by the number of speakers. Here, we assume
+  the new section uses the HTML definition list \texttt{dl} and the formula selects its first and
+  second \texttt{dd} item, respectively.
+
+\item At the same time, Jonathan updates the list of speakers (Figure~\ref{fig:conf-add})
+  and Tomas refactors the data from list into a table as discussed previously (Figure~\ref{fig:conf-tab}).
+\end{enumerate}
+
+\noindent
 The challenge is, again, to merge the edits done independently by the co-organizers.
-
-\begin{enumerate}
-\item The first co-organizer adds the budget calculation as shown in Figure~\ref{fig:conf-budget}.
-  This includes two formulas. The first selects all \texttt{li} elements of a \texttt{ul} element
-  with ID \texttt{speakers} (not visible in the UI) and counts their number. The second formula
-  multiplies the constant travel cost per speaker by the number of speakers. Here, we assume the
-  new section uses the HTML definition list \texttt{dl} and the formula selects its first and
-  second \texttt{dd} item, respectively.
-
-\item The second co-organizer performs the refactoring and data edits discussed previously.
-  They edit the speakers and refactor the structure to use a table, resulting in the document
-  shown in Figure~\ref{fig:conf-fin}.
-\end{enumerate}
-
-\noindent
-As before, the system should be able to merge the edits made to the document.
-
-\frameworkbox{
-\frameworkboxtitle{Framework perspective}
-In this challenge, the change done by the first organizer is at the \emph{code} layer, while
-the change done by the second organizer is both at the \emph{data} layer and, implicitly,
-at the \emph{schema} layer. A change solely at the \emph{data} layer would not affect code,
-but a change at the \emph{schema} layer requires adapting some aspects of the code.
-}
-
-\paragraph{Remarks.}
-Addressing the challenge requires that the refactoring of the document structure done in (1),
-is also reflected in the code of the equations added in (2). As shown in
-Figure~\ref{fig:conf-finfin}, in this specific case, the system needs to change the equation
+The challenge extends the previous one by adding code. The change to the document structure
+(\emph{type}) thus has to co-evolve with both the \emph{data} and the \emph{code}.
+As shown in Figure~\ref{fig:conf-finfin}, the system needs to change the equation
 \texttt{COUNT(/ul[id='speakers']/li)} to \texttt{COUNT(/table[id='speakers']/tbody/tr)}.
-This is the case because the refactoring consists of four steps that each transform the
-document and also need to update the code of the equation accordingly:
-
-\begin{itemize}
-    \item[--] Change the type of the \texttt{ul} element to \texttt{table}
-    \item[--] Wrap the body of the \texttt{table} element with \texttt{tbody}
-    \item[--] Change the type of all children of \texttt{tbody} from \texttt{li} to \texttt{tr}
-    \item[--] Further split the body into two \texttt{td} elements
-\end{itemize}
-
-\noindent
-It is expected that the system solving the challenge will need to be aware of those high-level
-edits. In particular, they may be explicitly specified by the user (through an interactive
-user interface for editing the document) or semi-automatically inferred.
-
-\paragraph{Requirements.}
-A solution to the challenge should meet the goals of the preceding challenge, i.e., completeness
-(merging should be always defined), commutativity (order does not matter), conflict resolution
-(ask for guidance if needed), convergence (everyone eventually wants the same document) and
-structure editing (the system may capture high-level edits).
-
-\subsection*{Challenge Extensions: Liveness and divergence}
-So far, we assumed that the system follows the \emph{convergence} model, i.e., all variants
+If the edit is recorded as a list of high-level edits shown in Figure~\ref{tbl:docchanges}
+(wrapping element, changing their types, etc.), then each edit needs to perform corresponding
+edit to selectors in formulas in the document.
+
+\begin{table}[t]
+\begin{adjustwidth}{-2em}{-2em}
+\begin{tabular}{lll}\toprule
+{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
+Change & Tag of an element & Apply to all matching elements and affected formulas\\
+Wrap & Selected element(s) & Wrap elements and update selectors in affected formulas  \\
+Split & Selected element(s) & Split matching elements using a rule (regular expression)\\
+Add & New item to a list & Add data without affecting document structure (type)\\
+Reorder & List items & Reorder data without affecting document structure (type)\\
+\bottomrule
+\end{tabular}
+\vspace{0.3em}
+\caption{Changes to type and data and how the programming system should respond}
+\label{tbl:docchanges}
+\end{adjustwidth}
+\vspace{-1em}
+\end{table}
+
+\subsection*{Remarks: Liveness and Divergence}
+The above challenge focuses on the local-first software scenario, but it can be extended to also
+apply to live programming. If formulas in the document are evaluated on-the-fly, the system
+needs to identify edits that invalide some of the values (by changing the data that a formula
+depends on, or by changing the equation). It should then automatically re-evaluate or invalidate
+the computed results (requiring explicit re-evaluation). In this scenario, we distinguish between
+\emph{permanent data} (information in the document) and transient \emph{data} (computed by code).
+Whereas permanent data should be synchronized and co-evolve with the document structure (type),
+transient data can be discarded.
+
+So far, we also assumed that the system follows the \emph{convergence} model, i.e., all variants
 of the program eventually adopt all changes. A more difficult variant on the challenge is to also
 support the \emph{divergence} model. In this case, some of the users continue using their variant
-of the program without adopting all of the changes done by other users.
+of the program without adopting all of the changes done by other users. In this model, users
+may want to adopt all changes done at the \emph{data} layer, but they may selectively choose
+some of the changes at the \emph{code} and \emph{schema} layers. The challenge is maintaining the
+same data across multiple program variants and merging changes done to the data based on
+different document schema. This is the subject of the case studies discussed in the next two
+sections.
+
+\section{Divergence Control in Spreadsheets and Databases}
+Type evolution becomes more challenging when data, code, and types cannot all be updated
+together in an atomic way. In these situations, different versions must coexist and interoperate
+with one another. For example, edits done against data of one type must be applied to
+data of another type, or code written against one type must run against data stored in another.
+
+Such situations are common in software engineering. Web backend architectures often put data and
+code on separate machines, which makes it impossible to update both atomically. To perform
+upgrades, teams must manually perform multi-step zero-downtime deployment processes~\cite{planetscale}
+to step the database and the code servers forward gradually such that they are compatible with one
+another at any given step. One example would be deploying a change to add a database column
+before deploying the code that uses the column.
+
+In the simplest cases, divergence can last only for a brief moment in between zero-downtime
+deployment steps. But divergence can also occur across longer timescales, especially when
+the different parts of a system are not all centrally controlled. For example,
+a public web API may need to support applications using old versions of the API for months or
+years, since it is not realistic to force all consumers to upgrade immediately. Divergence can
+even be indefinite -- Cambria \cite{Cambria} addresses collaborative applications where different
+users may want to collaborate on shared data through different tools using different schemas,
+with no intention of ever converging.
 
 \frameworkbox{
-\frameworkboxtitle{Framework perspective}
-In the \emph{divergence} model, it may be expected that users will want to adopt all changes
-done at the \emph{data} pace layer. They may selectively choose some of the changes at the
-\emph{code} and \emph{schema} layers. The challenge is maintaining the same data across multiple
-program variants and merging changes done to the data based on different document schema.
-}
-
-The basic challenge outlined above focuses primarily on the local-first software scenario, but
-it can be extended to also apply to the live programming scenario. In particular, when the
-document contains computed values, those may be evaluated either explicitly by the user or
-automatically on-the-fly. An edit can then invalidate some of the values (either by changing
-the data that a computed value depends on, or by changing the equation used to compute the value).
-A live programming system should be able to detect which computed values are affected by an edit
-and, either invalidate those (requiring an explicit re-evaluation by the user) or automatically
-re-evaluate them (and possibly highlight the affected values to inform the user).
-
-\frameworkbox{
-\frameworkboxtitle{Framework perspective}
-In the live extension, we also consider the transient \emph{state} pace layer. A system should
-automatically detect when a change at the \emph{data} or \emph{code} layers affects the computed
-value and forces it to be recomputed. Interestingly, a change at the \emph{schema} layer may not
-always affect the \emph{state} layer, even if it schema change requires modifying formulas that
-compute the state.
-}
+\begin{wrapfigure}{r}{12em}
+\vspace{0.5em}
+\includegraphics[scale=0.25]{figures/diag-divergence.png}
+\end{wrapfigure}
+\frameworkboxtitle{Type Branching with Divergence}
+The divergence scenario adds another dimension to the problem. There are now multiple
+versions of the system with different \emph{types} that may co-exist as multiple branches.
+Edits to \emph{data} are done against a specific type, but they then need to be migrated
+to versions with different types. Consequently, any type evolution needs to be accompanied
+both with forward projection (transforming data to a new version) and backward projection
+(transforming new data to the old version).}
+
+Divergence does not just occur in software engineering; it also appears in end-user
+workflows. Users frequently make copies of structured documents like spreadsheets and then
+diverge them by altering both data content and schema. In the case of spreadsheets schema
+changes include rearrangements to the structure of rows and columns as well as changes to
+formulas. The problem is that the users want to transfer such data and schema changes between
+these divergent copies, and they want to pick and choose which of these changes to transfer.
+In practice such transfer is done manually through copy \& paste. \citet{Basman19} has
+documented an ecology of emailed spreadsheets. \citet{Burnett14} distilled field observations
+of these practices in the story of Frieda, which we quote here:
+
+\begin{quotation}
+\setlength{\parskip}{0.5em}
+\setlength{\parindent}{0em}
+
+\noindent
+\emph{For example, consider ``Frieda'', an office manager in charge of her department's budget
+tracking. (\ldots) Every year, the company she works for produces an updated budget tracking
+spreadsheet with the newest reporting requirements embedded in its structure and formulas.
+But this spreadsheet is not a perfect fit to the kinds of projects and sub-budgets she manages,
+so every year Frieda needs to change it. She does this by working with four variants of the
+spreadsheet at once: the one the company sent out last year (we will call it Official-lastYear),
+the one she derived from that one to fit her department's needs (Dept-lastYear), the one the
+company sent out this year (Official-thisYear), and the one she is trying to put together
+for this year (Dept-thisYear).}
+
+\emph{Using these four variants, Frieda exploratively mixes reverse engineering, reuse,
+programming, testing, and debugging, mostly by trial-and-error. She begins this process by
+reminding herself of ways she changed last year's by reverse engineering a few of the
+differences between Official-lastYear and Dept-lastYear. She then looks at the same portions
+of Official-thisYear to see if those same changes can easily be made, given her department's
+current needs.
+(\ldots)
+She can reuse some of these same changes this year, but copying them into Dept-thisYear
+is troublesome, with some of the formulas automatically adjusting themselves to refer to
+Dept-lastYear. (\ldots) Frieda has learned over the years to
+save some of her spreadsheet variants along the way (using a different filename for each),
+because she might decide that the way she did some of her changes was a bad idea, and she
+wants to revert to try a different way she had started before.}
+\end{quotation}
+
+\challenge{Divergence Control for Extract Entity}
+
+As a concrete example, we combine the problem of divergence control with the example from the
+earlier Extract Entity challenge. Every month the orders department sends its spreadsheet to
+the accounting department, which adds the data to a version of the spreadsheet that it has
+customized for financial tracking. But when the orders department migrates its spreadsheet to
+extract out customers, the accounting department does not want to conform. They could manually
+convert incoming data in the new schema into their variant of the old schema. But they
+shouldn't have to. It should be possible to run new data through the schema change in reverse,
+converting it back into the format the accounting department is used to ingesting. Note that
+this is not a matter of synchronizing the divergent spreadsheets -- they maintain differently
+evolved schema.
+
+What is needed is a user-friendly mechanism for transferring data changes from one schema (type) to
+another bidirectionally. In one direction we want to transfer changes made in the new schema (type)
+into a variant of the old schema (type). The opposite direction might be needed, for example,
+if the accounting department makes corrections to customer addresses, which ought to be pushed
+through into the orders department's new schema.
+
+\subsection*{Remarks: Source Code Version Control}
+
+Bidirectional transfer of changes between divergent copies is similar to source code version
+control as in Git~\cite{ProGit}. There is \textit{forking} of long-lived divergent copies.
+We want to \textit{diff} these forks to see exactly how they have diverged. We want to
+partially \textit{merge} them by \textit{cherry picking} certain differences. Yet there are
+also many dissimilarities with Git: our data is more richly structured than lines of text; our
+schema changes are higher-level transformations on these structures than inserting and
+deleting characters; and we expect that end-users be able to understand it~\cite{gitless}.
+The Divergence Control challenge is in a sense to provide ``version control for schema
+change'' meeting these criteria, with the key technical challenge being the ability to
+transfer selected differences bidirectionally through schema changes as independently as possible.
+
+This challenge invites a change of perspective in both live programming and local-first software.
+Live programming must move from being a solitary activity to a collaborative workflow, and one
+where the collaboration is not just on editing source code but on live integrated code and data,
+as in the Smalltalk/Lisp images of old.
+%\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!}
+For its part, local-first software must move from automatically converging data replicas
+to also interactively managing long-term divergence. That implies either application state
+is being directly exposed to the user, or the application code is using an API to present
+version control affordances. We encourage both communities to think outside the box of Git,
+which has proven to baffle not only end-users but also a substantial fraction of developers.
+
+\section{Multiplicity Change in Data Formats}
+
+A common kind of schema change is \textit{multiplicity change}: when a field changes from storing
+a single value to storing multiple values. For example, we might start by storing a single
+address for each contact in an address book, before realizing that we need to store multiple
+addresses for a single contact; or we might assign each todo in a list to one person before
+realizing we want the ability to assign a todo to multiple people.
+
+In a relational schema, solving this problem might require extracting a normalized table for
+the linked entity; some of the challenges of this approach are covered in the Extract Entity
+challenge. In this section we will instead consider a document schema with support for arrays;
+in this context, we can have a multiplicity change by turning a scalar value into an array value.
+
+Multiplicity change clearly requires changing both the data and the code. It becomes particularly
+challenging to handle when there is ongoing divergence between multiple versions of the data.
+
+\challenge{Multiplicity Change}
+
+Consider the following schema for a todo list item, in which each item has a single assignee, represented by a user ID:
+
+\newpage
+\begin{lstlisting}[language=ml]
+type Item = { id : id; title : string; assignee : string }
+\end{lstlisting}
+
+Now, we change the schema so that each item has a list of assignees:
+
+\begin{lstlisting}[language=ml]
+type Item = { id : id; title : string; assignees : Array<string> }
+\end{lstlisting}
+
+Our goal is to preserve the ability for actors in the system to read and write to a shared todo list in either the old or new schema -- for example, they should be able to write to either the scalar assignee field or to the list of assignees.
+
+A natural invariant to preserve across these schemas is that the value of the scalar field should equal the first element of the list field (and if the list is empty, the scalar field should be null).
+If we were to write code for a one-time data migration from the scalar to list schema, we could easily satisfy this invariant:
+
+\begin{lstlisting}[language=ml]
+item.assignees = if (item.assignee == null) then [] else [item.assignee]
+\end{lstlisting}
+
+Ongoing edits to the array schema can also be handled in a straightforward way.
+After edits are made, the value of the scalar field should be set to the first element of the new array (or null if the new array is empty.)
+
+However, handling edits to the scalar schema presents more of a challenge.
+Consider the following todo with two assignees, presented in terms of the scalar and array schemas.
+A write is made in the scalar schema to set the new assignee to C.
+
+\begin{lstlisting}[language=ml]
+todoScalar = { id: 1, title: "Foo", assignee: A }
+todoArray = { id: 1, title: "Foo", assignees: [A, B] }
+
+todoScalar.assignee = C
+\end{lstlisting}
+
+What should the array value become after this edit to the scalar schema?
+To satisfy our invariant, we know that C must become the first element of the array, but this leaves open several options with different tradeoffs:
+
+\begin{enumerate}
+  \item \texttt{[C]}: "Only C should be assigned." This option produces an array that corresponds directly to the resulting scalar.
+But it has the downside of deleting data that wasn't even visible in the scalar schema.
+  \item \texttt{[C, B]}: "Replace A with C." If the writer wanted to remove A from the assignment and add C, this option performs that intent.
+However, there is data remaining in the list which was not visible to the scalar writer.
+  \item \texttt{[C, A, B]}: "Add C to the list." Perhaps the scalar writer wanted to add C to the assignment without removing anyone; this option satisfies that intent.
+But it preserves data not visible to the scalar schema.
+\end{enumerate}
+
+Because the intent of the writer to the scalar schema cannot be unambiguously interpreted from the write alone, it is impossible to make a perfect choice among these options.
+
+\subsection*{Remarks: Managing Design Tradeoffs}
+
+If schemas, code and data can all be updated atomically together, then multiplicity change is relatively straightforward  to handle.
+Existing scalar data can be trivially migrated to a list, and the code using the data will need to change to accommodate the list data.
+
+However, ongoing divergence makes multiplicity change much more difficult, and exposes some general challenges.
+Different schemas may expose partial information, and writers in those schemas have to operate without total knowledge of the information available in other schemas.
+As a result, synchronizing writes across the schemas requires making difficult tradeoffs, such as choosing to either preserve or destroy hidden data not visible to the writer.
+
+Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this problem would give developers or users tools to manage these tradeoffs.
+For example, a system might allow developers to specify the desired behavior for a particular pair of schemas based on the requirements of the domain.
+Or a system might even allow users to manually disambiguate their intent for a given write.
 
@@
 \newpage
-
-% elm
-% database
-% divergence
-% cambria
-% documents
-% state machine
-
-% DIVERGENCE IN TRANSITION (because you cannot do big bang update) c.f. Tijs' example
-
-% typically people just blow away the state but Elm actually has demos
-% of people doing hot reload without losing state.
-% (maybe they never change the type?)
-
-% ELM EXAMPLE
-% current example ("completes") is contrived - it is equivalent data
-% more interesting is when you lose / add data
-% add "completed" with default instead
-
-\section{Divergence Control}
-
-\textcolor{red}{This section was an extension of the Extract Entity challenge but should be lifted up to the top-level discussion of divergence}
-
-\subsection{Context}
-Schema evolution becomes more challenging when data, code, and schema cannot all be updated together in an atomic way.
-In these situations, different versions must coexist and interoperate with one another: for example, edits made in one schema must be applied in another, or code written against one schema must run against data stored in another.
-
-Such situations are common in software engineering.
-Web backend architectures often put data and code on separate machines, which makes it impossible to update both atomically.
-To perform upgrades, teams must manually perform multi-step [zero-downtime deployment processes](https://planetscale.com/blog/zero-downtime-rails-migrations-planetscale-rails-gem) to step the database and the code servers forward gradually such that they are compatible with one another at any given step.
-One example would be deploying a change to add a database column before deploying the code that uses the column.
-
-In the simplest cases, divergence can last only for a brief moments in between zero-downtime deployment steps.
-But divergence can also occur across longer timescales, especially in situations where the different parts of a system are not all centrally controlled.
-For example, a public web API may need to support applications using old versions of the API for months or years, since it is not realistic to force all consumers to upgrade immediately.
-Divergence can even be indefinite -- \citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.
-
-Divergence does not just occur in software engineering; it also appears in end-user workflows.
-Users frequently make copies of structured documents like spreadsheets and then diverge them by altering both data content and schema.
-In the case of spreadsheets schema changes include rearrangements to the structure of rows and columns as well as changes to formulas.
-The problem is that the users want to transfer such data and schema changes between these divergent copies, and they want to pick and choose which of these changes to transfer.
-In practice such transfer is done manually through copy \& paste.
-\citet{Basman19} has documented an ecology of emailed spreadsheets.
-\citet{Burnett14} distilled field observations of these practices in the story of Frieda, which we quote here:
-
-\begin{quotation}
-
-For example, consider ``Frieda'', an office manager in charge of her department's budget tracking.
-(Frieda was a participant in a set of interviews with spreadsheet users that the first author conducted.
-Frieda is not her real name.) Every year, the company she works for produces an updated budget tracking spreadsheet with the newest reporting requirements embedded in its structure and formulas.
-But this spreadsheet is not a perfect fit to the kinds of projects and sub-budgets she manages, so every year Frieda needs to change it.
-She does this by working with four variants of the spreadsheet at once: the one the company sent out last year (we will call it Official-lastYear), the one she derived from that one to fit her department's needs (Dept-lastYear), the one the company sent out this year (Official-thisYear), and the one she is trying to put together for this year (Dept-thisYear).
-
-Using these four variants, Frieda exploratively mixes reverse engineering, reuse, programming, testing, and debugging, mostly by trial-and-error.
-She begins this process by reminding herself of ways she changed last year's by reverse engineering a few of the differences between Official-lastYear and Dept-lastYear.
-She then looks at the same portions of Official-thisYear to see if those same changes can easily be made, given her department's current needs.
-
-She can reuse some of these same changes this year, but copying them into Dept-thisYear is troublesome, with some of the formulas automatically adjusting themselves to refer to Dept-lastYear.
-She patches these up (if she notices them), then tries out some new columns or sections of Dept-thisYear to reflect her new projects.
-She mixes in ``testing'' along the way by entering some of the budget values for this year and eyeballing the values that come out, then debugs if she notices something amiss.
-At some point, she moves on to another set of related columns, repeating the cycle for these.
-Frieda has learned over the years to save some of her spreadsheet variants along the way (using a different filename for each), because she might decide that the way she did some of her changes was a bad idea, and she wants to revert to try a different way she had started before.
-\end{quotation}
-
-\subsection{Example}
-The Divergence Control challenge instantiates the problem of divergent state with the example from the Extract Entity challenge.
-Every month the orders department sends its spreadsheet to the accounting department, which adds the data to a version of the spreadsheet that it has customized for financial tracking.
-But when the orders department migrates its spreadsheet to extract out customers the accounting department does not want to conform.
-They could manually convert incoming data in the new schema into their variant of the old schema.
-But they shouldn't have to.
-It should be possible to run new data through the schema change in reverse, converting it back into the format the accounting department is used to ingesting.
-Note that this is not a matter of synchronizing the divergent spreadsheets -- they maintain differently evolved schema.
-
-\subsection{Challenge}
-What is needed is a user-friendly mechanism for transferring data changes from one schema to another bidirectionally.
-In one direction we want to transfer changes made in the new schema into a variant of the old schema.
-The opposite direction might be needed, for example, if the accounting department makes corrections to customer addresses, which ought to be pushed through into the orders department's new schema.
-
-Bidirectional transfer of changes between divergent copies is similar in some ways to source code version control as in Git~\cite{ProGit}.
-There is \textit{forking} of long-lived divergent copies.
-We want to \textit{diff} these forks to see exactly how they have diverged.
-We want to partially \textit{merge} them by \textit{cherry picking} certain differences.
-Yet there are also many dissimilarities with Git: our data is more richly structured than lines of text; our schema changes are higher-level transformations on these structures than inserting and deleting characters; and we expect that end-users be able to understand it~\cite{gitless}.
-The Divergence Control challenge is in a sense to provide ``version control for schema change'' meeting these criteria, with the key technical challenge being the ability to transfer selected differences bidirectionally through schema changes as independently as possible.
-
-\subsection{Goals}
-This challenge invites a change of perspective in both live programming and local-first software.
-Live programming must move from being a solitary activity to a collaborative workflow, and one where the collaboration is not just on editing source code but on live integrated code and data, as in the Smalltalk/Lisp images of old.\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!} For its part, local-first software must move from automatically converging data replicas to also interactively managing long-term divergence.
-That implies either application state is being directly exposed to the user, or the application code is using an API to present version control affordances.
-We encourage both communities to think outside the box of Git, which has proven to baffle not only end-users but also a substantial fraction of developers.
+~
+\newpage
 
@@
 
-\section{Multiplicity change}
-
-\subsection{Context}
-
-A common kind of schema change is \textit{multiplicity change}: when a field changes from storing a single value to storing multiple values.
-For example, we might start by storing a single address for each contact in an address book, before realizing that we need to store multiple addresses for a single contact; or we might assign each todo in a list to one person before realizing we want the ability to assign a todo to multiple people.
-
-In a relational schema, solving this problem might require extracting a normalized table for the linked entity; some of the challenges of this approach are covered in the Extract Entity challenge.
-In this section we will instead consider a document schema with support for arrays; in this context, we can have a multiplicity change by turning a scalar value into an array value.
-
-Multiplicity change clearly requires changing both the data and the code. It becomes particularly challenging to handle when there is ongoing divergence between multiple versions of the data.
-
-\subsection{Example}
-
-Consider the following schema for a todo list item, in which each item has a single assignee, represented by a user ID:
-
-\begin{lstlisting}[language=ml]
-type Item = { id : id; title : string; assignee : string }
-\end{lstlisting}
-
-Now, we change the schema so that each item has a list of assignees:
-
-\begin{lstlisting}[language=ml]
-type Item = { id : id; title : string; assignees : Array<string> }
-\end{lstlisting}
-
-Our goal is to preserve the ability for actors in the system to read and write to a shared todo list in either the old or new schema -- for example, they should be able to write to either the scalar assignee field or to the list of assignees.
-
-A natural invariant to preserve across these schemas is that the value of the scalar field should equal the first element of the list field (and if the list is empty, the scalar field should be null).
-If we were to write code for a one-time data migration from the scalar to list schema, we could easily satisfy this invariant:
-
-\begin{lstlisting}[language=ml]
-item.assignees = if (item.assignee == null) then [] else [item.assignee]
-\end{lstlisting}
-
-Ongoing edits to the array schema can also be handled in a straightforward way.
-After edits are made, the value of the scalar field should be set to the first element of the new array (or null if the new array is empty.)
-
-However, handling edits to the scalar schema presents more of a challenge.
-Consider the following todo with two assignees, presented in terms of the scalar and array schemas.
-A write is made in the scalar schema to set the new assignee to C.
-
-\begin{lstlisting}[language=ml]
-todoScalar = { id: 1, title: "Foo", assignee: A }
-todoArray = { id: 1, title: "Foo", assignees: [A, B] }
-
-todoScalar.assignee = C
-\end{lstlisting}
-
-What should the array value become after this edit to the scalar schema?
-To satisfy our invariant, we know that C must become the first element of the array, but this leaves open several options with different tradeoffs:
-
-\begin{enumerate}
-  \item \texttt{[C]}: "Only C should be assigned." This option produces an array that corresponds directly to the resulting scalar.
-But it has the downside of deleting data that wasn't even visible in the scalar schema.
-  \item \texttt{[C, B]}: "Replace A with C." If the writer wanted to remove A from the assignment and add C, this option performs that intent.
-However, there is data remaining in the list which was not visible to the scalar writer.
-  \item \texttt{[C, A, B]}: "Add C to the list." Perhaps the scalar writer wanted to add C to the assignment without removing anyone; this option satisfies that intent.
-But it preserves data not visible to the scalar schema.
-\end{enumerate}
-
-Because the intent of the writer to the scalar schema cannot be unambiguously interpreted from the write alone, it is impossible to make a perfect choice among these options.
-
-\subsection{Remarks}
-
-If schemas, code and data can all be updated atomically together, then multiplicity change is relatively straightforward  to handle.
-Existing scalar data can be trivially migrated to a list, and the code using the data will need to change to accommodate the list data.
-
-However, ongoing divergence makes multiplicity change much more difficult, and exposes some general challenges.
-Different schemas may expose partial information, and writers in those schemas have to operate without total knowledge of the information available in other schemas.
-As a result, synchronizing writes across the schemas requires making difficult tradeoffs, such as choosing to either preserve or destroy hidden data not visible to the writer.
-
-Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this problem would give developers or users tools to manage these tradeoffs.
-For example, a system might allow developers to specify the desired behavior for a particular pair of schemas based on the requirements of the domain.
-Or a system might even allow users to manually disambiguate their intent for a given write.
-
 \section{Related Work}
@@
 
-There has been a lot of research on evolving interdependent layers of software, much of it guided by schema and types.
-But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system.
-Nevertheless it is important to understand these batch techniques to envision more interactive ones.
-
 \paragraph{Schema evolution.}
-Schema evolution has long been a major problem for SQL databases because
-the SQL Data Definition Language (DDL) has limited capabilties to redefine existing data.
-It can rename tables and columns.
+Schema evolution has long been a major problem for SQL databases, which provide surprisingly little help.
 SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
@@
 But apart from such special cases, SQL databases have no general purpose support for data migration.
-As a result in practice schema evolution is mostly done by writing custom Data Manipulation Language (DML) code to migrate the data.
+As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data.
 Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.

### 2024-05-28 paper.tex

@@
 
+\begin{figure}[H]
+\begin{tikzcd}[sep = 6em]
+  \textsf{type}_1
+  \arrow[r, "\textsf{constrains}"{above}]
+  \arrow[d, red, "\textsf{change}_1"{name=C, description}]
+  &
+  \textsf{data}_1
+  \arrow[d, "\textsf{migrate}_1"{name=R, description}]
+  \\
+  \textsf{type}_2
+  \arrow[r, "\textsf{constrains}"{above}]
+  \arrow[d, red, "\textsf{change}_2"{name=C2, description}]
+  &
+  \textsf{data}_2
+  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
+  \arrow[d, "\textsf{migrate}_2"{name=R2, description}]
+  \\
+  \textsf{type}_3
+  \arrow[r, "\textsf{constrains}"{above}]
+  &
+  \textsf{data}_3
+  \arrow[Rightarrow, from=C2, to=R2, "\textsf{propagate}"{above}]
+\end{tikzcd}
+\caption{Multiversion type evolution.}
+\label{fig:multiversion}
+\end{figure}
+
 \frameworkbox{
-\begin{wrapfigure}{r}{9em}
-\includegraphics[scale=0.2]{figures/diag-variants.png}
-\end{wrapfigure}
+% wrapfigure doesn't seem to work with tikz
+% \begin{wrapfigure}{r}{9em}
+% \includegraphics[scale=0.2]{figures/diag-variants.png}
+% \end{wrapfigure}
+
 \frameworkboxtitle{Primitive Type Transformations}
-In this example, \emph{type} refers to data tables, but the same scenario applies to
+In this example, \textsf{type} in Figure~\ref{fig:multiversion} refers to data tables, but the same scenario applies to
 objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
@@
 
-Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
-End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
-between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
-the established database products (see Related Work). Acme will need to write code or do a big manual conversion.
-
-TODO: Maybe insert something about methodology in general?
-(ref the table challenges paper? although this paper is no longer so much about challenges)
+\textcolor{red}{Maybe insert something about methodology in general?
+(ref the table challenges paper? although this paper is no longer so much about challenges)}
+
+There has been a lot of research on evolving interdependent layers of software, much of it guided by schema and types.
+But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system.
+Nevertheless it is important to understand these batch techniques to envision more interactive ones.
 
 \paragraph{Schema evolution.}
-Schema evolution has long been a major problem for SQL databases, which provide surprisingly little help.
+Schema evolution has long been a major problem for SQL databases because
+the SQL Data Definition Language (DDL) has limited capabilties to redefine existing data.
+It can rename tables and columns.
 SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.

### 2024-05-29 paper.tex
fix planetscale bib

@@
 code on separate machines, which makes it impossible to update both atomically. To perform
-upgrades, teams must manually perform multi-step zero-downtime deployment processes~\cite{planetscale}
+upgrades, teams must manually perform multi-step zero-downtime deployment processes~\cite{planetscalerails}
 to step the database and the code servers forward gradually such that they are compatible with one

### 2024-05-29 paper.tex
\openframeworkbox \closeframeworkbox to handle math

@@
 \newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}
+\newcommand{\openframeworkbox}{
+\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=gray!15,colframe=gray!15,parbox=false,
+  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
+  outer arc=0pt]\setlength{\parskip}{0.5em}
+}
+\newcommand{\closeframeworkbox}{
+\end{tcolorbox}
+}
 
@@
 
-\begin{figure}[H]
+\openframeworkbox
   \begin{tikzcd}[sep = 6em]
@@
   \end{tikzcd}
-\caption{Multiversion type evolution.}
-\label{fig:multiversion}
-\end{figure}
-
-\frameworkbox{
-% wrapfigure doesn't seem to work with tikz
-% \begin{wrapfigure}{r}{9em}
-% \includegraphics[scale=0.2]{figures/diag-variants.png}
-% \end{wrapfigure}
 
   \frameworkboxtitle{Primitive Type Transformations}
-In this example, \textsf{type} in Figure~\ref{fig:multiversion} refers to data tables, but the same scenario applies to
+In this example, \textsf{type} refers to data tables, but the same scenario applies to
 objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
@@
 programmer, e.g. to a database table schema definition.
-}
+\closeframeworkbox
 

### 2024-05-29 paper.tex
cleanup

@@
 \newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}
-\newcommand{\openframeworkbox}{
-\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=gray!15,colframe=gray!15,parbox=false,
-  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
-  outer arc=0pt]\setlength{\parskip}{0.5em}
-}
-\newcommand{\closeframeworkbox}{
-\end{tcolorbox}
-}
 
@@
 
-\textcolor{red}{[JE. Found this snippet in the related work]
-Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
-End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
-between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
-the established database products (see Related Work). Acme will need to write code or do a big manual conversion.
-}
-
 The \textit{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers,
@@
 
-\openframeworkbox
-  \begin{tikzcd}[sep = 6em]
-    \textsf{type}_1
-    \arrow[r, "\textsf{constrains}"{above}]
-    \arrow[d, red, "\textsf{change}_1"{name=C, description}]
-    &
-    \textsf{data}_1
-    \arrow[d, "\textsf{migrate}_1"{name=R, description}]
-    \\
-    \textsf{type}_2
-    \arrow[r, "\textsf{constrains}"{above}]
-    \arrow[d, red, "\textsf{change}_2"{name=C2, description}]
-    &
-    \textsf{data}_2
-    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
-    \arrow[d, "\textsf{migrate}_2"{name=R2, description}]
-    \\
-    \textsf{type}_3
-    \arrow[r, "\textsf{constrains}"{above}]
-    &
-    \textsf{data}_3
-    \arrow[Rightarrow, from=C2, to=R2, "\textsf{propagate}"{above}]
-  \end{tikzcd}
+\frameworkbox{
+\begin{wrapfigure}{r}{9em}
+\includegraphics[scale=0.2]{figures/diag-variants.png}
+\end{wrapfigure}
 
@@
 programmer, e.g. to a database table schema definition.
-\closeframeworkbox
+}
 
@@
 
-\textcolor{red}{Maybe insert something about methodology in general?
-(ref the table challenges paper? although this paper is no longer so much about challenges)}
-
-There has been a lot of research on evolving interdependent layers of software, much of it guided by schema and types.
+There has been a lot of research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types.
 But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system.

### 2024-05-29 paper.tex
Data wrangling references

@@
 
+\paragraph{Data wrangling.} The evolution of data formats is a concern in Data Science. Data~Diff~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation.
+Wrangler~\cite{kandel11} observes interactive manipulation of sampled data to write data transformation programs.
+\citet{petersohn20} present an algebra of operations on Python dataframes that include shape transformations.
+
 \paragraph{Live programming.}
-Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (mostly solo) programming.
+Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (primarily solo) programming.
 % \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
-% A special case of live programming is \emph{hot reloading} or \emph{hot swapping})~\cite{barenz2020essence, hicks2005dynamic} that installs code changes into a running system preserving state but not attempting to migrate that state through type changes.
+\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
 \citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.

### 2024-05-30 paper.tex
New kind of diagrams, tweaks

--- 2024-05-29 paper.tex
@@
 
-\title{Type Evolution in Programming Systems}
+\title{Type Evolution in Interactive Programming Systems}
 
@@
 \subsection{Types, Data and Code}
-
-As is the case for buildings, making a change at surface layers (skin or stuff) is relatively easy,
+In both buildings and programs, making a change at surface layers (skin or stuff) is relatively easy,
 but a change at a more fundamental layer (structure) has cascading effects on layers that depend on it.
-In the case of programs, a more fundamental layer defines the shape of data and code.
-We refer to this structure as \emph{types}, although we do not use the term in a precise technical
-sense. In our conceptual framework, types may be data types that define the shape of data, they
-may be explicitly declared in code and checked statically, as well as dynamically checked or
-even implicit. The key property of types for us is that they constrain the shape of other layers
+In the case of programs, we refer to the more fundamental layer as \emph{types}, although we do not
+use the term in a precise technical sense. Types is what defines the shape of data and code.
+Types may be database schema, data types that define the shape of data, they may be explicitly
+declared in code and checked statically, as well as dynamically checked or even implicit.
+The key property of types for us is that they constrain the shape of other layers
 of a program, namely data and code. As a result, a change of program types requires a
@@
 
-The situation is illustrated in Figure~\ref{fig:change} using two variations of a diagram that
-we will return to repeatedly in this paper. When a type changes from $\textsf{type}_1$ to $\textsf{type}_2$,
-we need to find and apply a corresponding change to other parts of the program whose shape
-is defined by the changed type. The problem of \emph{type evolution} is thus finding and
-applying changes that will update $\textsf{data}_1$ and $\textsf{code}_1$ into new versions $\textsf{data}_1$ and $\textsf{code}_2$
-that correspond to the new type.
-
-The complexity of the problem varies. The first dimension of challenges that we consider
-is local. The complexity of the challenge depends on how many other parts of the program
-depend on the modified type. If both code and data depend on the type, the complexity of the
-program is greater. If there is no associated code or the existing data can be safely discarded,
-the complexity is lesser.
-
-The second dimension of challenges that we consider is non-local. A program may have multiple
-variants that exist independently and need to be synchronized. The difficulty of the problem
-depends on what kind of synchronization we want to support. If all variants of the program should
-eventually adopt all changes (converge), the complexity is lesser. If we allow user to apply
-only certain changes from other users (divergence), the complexity is greater. A typical case
-may be a combination of the two where types and schema of a program should converge, but data
-may diverge.
-
-\begin{figure}[t]
-\centering
-%\includegraphics[scale=0.175]{figures/diag-data-code.png}
-\begin{tikzcd}[sep = 6em]
-  \textsf{code}_1
-  \arrow[d, "\textsf{refactor}"{name=L, description}]
-  &
-  \textsf{type}_1
-  \arrow[l, "\textsf{constrains}"{above}]
-  \arrow[r, "\textsf{constrains}"{above}]
-  \arrow[d, red, "\textsf{change}"{name=C, description}]
-  &
-  \textsf{data}_1
-  \arrow[d, "\textsf{migrate}"{name=R, description}]
-  \\
-  \textsf{code}_2
-  &
-  \textsf{type}_2
-  \arrow[l, "\textsf{constrains}"{above}]
-  \arrow[r, "\textsf{constrains}"{above}]
-  &
-  \textsf{data}_2
-  \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
-  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
-\end{tikzcd}
-\caption{When a type changes from $\textsf{type}_1$ to $\textsf{type}_2$, a corresponding change needs to take place
-  in data and code constrained by the type. The challenge is finding a suitable
-  corresponding transformation (semi-)automatically.}
-\label{fig:change}
-\end{figure}
+\begin{wrapfigure}{r}{12em}
+\vspace{-1.5em}
+\includegraphics[width=12em]{figures/arr-basic.png}
+\vspace{-1.5em}
+\end{wrapfigure}
+
+The situation is illustrated on the right. We start with code $C_1$ and
+data $D_1$ that have a shape corresponding to type $T_1$. When the type changes from $T_1$ to $T_2$ ($\rightarrow$),
+a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
+of \emph{type evolution} is synthetising ($\Rightarrow$) suitable corresponding transformations
+that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
+to the new type $T_2$.
+%
+The complexity of the problem varies and finding a suitable transformation may require user
+input. If there is code, it typically needs to be transformed, but data can sometimes be
+discarded. We refer to such transformations done by a single user as \emph{local} and draw
+them inside a box.
+
+\begin{wrapfigure}{r}{12em}
+\vspace{-1.5em}
+\includegraphics[width=12em]{figures/arr-forkjoin.png}
+\vspace{-1.5em}
+\end{wrapfigure}
+
+Collaboration adds the \emph{non-local} dimension to the problem. A program may have multiple
+variants that exist independently and changes made to them may need to be synchronized. The
+situation is illustrated on the right. Here, the initial type $T_1$ is transformed by two
+users independently into types $T_2$ and $T_3$. The corresponding data is transformed accordingly,
+producing $D_2$ and $D_3$. The challenge is now merging these two diverging transformation and
+obtaining a type $T_4$ and data $D_4$ that adopt both of the changes.
+
+The difficulty of the problem depends on what kind of synchronization we want to support.
+If all variants of the program should eventually adopt all changes (converge), the complexity
+is lesser. If we allow user to apply only certain changes from other users (divergence), the
+complexity is greater. A common case may be a combination of the two where types and schema
+of a program should converge, but data may diverge.
 
@@
 \noindent
-Some programming systems that use the Elm architecture support hot reloading when the
+Some systems that use the Elm architecture support hot reloading when the
 implementation of the \texttt{render} or \texttt{update} function changes, but they typically
 discard state and restart the application when the \texttt{State} type changes. A more effective
-programming system based on live programming would allow updates to both the code of the two
-functions and the structure of the two types.
+live programming system would allow live updates to the structure of the two types too.
 
@@
 \begin{lstlisting}[language=ml]
-type Item = { id : id; title : string; completed : DateTime option }
+type Item = { id : id; title : string; completed : Maybe DateTime }
 type State = { items : Item list }
 type Event =
-  | .. | SetCompleted of id * DateTime option | ..
+  | .. | SetCompleted of id * Maybe DateTime | ..
 \end{lstlisting}
@@
 to match the new type definition. In this case, this likely cannot be done fully automatically
-and the programmer may need to provide mapping (\texttt{false} to \texttt{None}, \texttt{true}
-to \texttt{Some(DateTime(2024,5,1))}). If the programmer changes \texttt{Item} and \texttt{SetCompleted}
+and the programmer may need to provide mapping (\texttt{false} to \texttt{Nothing}, \texttt{true}
+to \texttt{Just(DateTime(2024,5,1))}). If the programmer changes \texttt{Item} and \texttt{SetCompleted}
 at the same time, the \texttt{update} function likely does not need to change, but the rendering
@@
 \frameworkbox{
-% \begin{wrapfigure}{r}{14em}
-% \vspace{1em}
-% \includegraphics[scale=0.175]{figures/diag-both.png}
-% \end{wrapfigure}
-\frameworkboxtitle{Code-data Co-evolution}
-In this case study, the evolving type in Figure~\ref{fig:change} is the \texttt{State} type. When the type
-evolves, corresponding change needs to be applied to both \emph{code} and \emph{data}.
-The former is the program logic implemented in \texttt{update} and \texttt{render},
-while the latter is the current state (a live value of type \texttt{State}). The scenario
-illustrates that code and data need to co-evolve as illustrated by a revised diagram on the right.
+\begin{wrapfigure}{r}{12em}
+\vspace{0em}
+\includegraphics[width=12em]{figures/arr-steps.png}
+\vspace{0em}
+\end{wrapfigure}
+\frameworkboxtitle{Primitive Type Transformations}
+In this case study, the evolving type is the \texttt{State} type. When the type evolves,
+the program logic (code) and the current live value of the type (data) needs to evolve
+correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
+transformations, such as those in Table~\ref{tbl:elmchanges}, that each has a corresponding
+primitive code and data transformation. The primitive transformations may either be invoked
+through a user interface, or inferred from textual code edits made by the programmer.
 }
@@
 to produce a new state that is as near as possible to the previous state. Migrating
-\emph{code} can be done in various ways (transform code, trigger error condition, etc.),
+\emph{code} can be done in various ways,
 but it is desirable to avoid breaking the \texttt{render} function as this would make the
@@
 \section{Entity Evolution in Data-Centric Systems}
-Many problems studied in the context of database systems equally apply to programming systems
-if we think of them as stateful environments where programs co-exist with their persistent data.
+If we see programming systems as stateful environments where programs co-exist with their persistent
+data, we can gain valuable insights from research on database systems.
 Although schema evolution is well-studied in databses (Section~\ref{sec:related}), past work
@@
 \challenge{Extract Entity}
-Acme needs to evolve the above type and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
+Acme needs to evolve the above schema and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
 one with orders that links to one with customers.  We refer to this as \emph{Extract Entity}
@@
 entities, whether they are uniquely generated primary keys in a database or an abstraction
-of a memory address in a programming language.
+of a memory address in a programming language. As spreadsheets do not support relationships
+between entities, Acme will need to migrate their data to a database and restructure the original
+flat table.
 
@@
 
-More generally, entity evolution can happen through operations such as those listed in
+Entity evolution can happen through operations such as those listed in
 Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
@@
 
-\frameworkbox{
-\begin{wrapfigure}{r}{9em}
-\includegraphics[scale=0.2]{figures/diag-variants.png}
-\end{wrapfigure}
-
-\frameworkboxtitle{Primitive Type Transformations}
-In this example, \textsf{type} refers to data tables, but the same scenario applies to
-objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
-that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
-The evolution can be done through a small set of primitive type transformations (illustrated on the
-right) with corresponding data transformation for each. The system may either expose the
-primitive transformations through a user interface, or infer them from textual edits made by the
-programmer, e.g. to a database table schema definition.
-}
-
 \subsection*{Remarks: Representation and Replication}
-The central problem is to coordinate the type evolution with code edits to keep the system
-executing live and correctly. Solutions may infer the desired type evolution by analysing
+The central problem is to coordinate the schema evolution with code edits to keep the system
+executing live and correctly. Solutions may infer the desired schema evolution by analysing
 source code edits, but would typically offer UI affordances for the commands. They should
@@
 \noindent
-The same code transformation would need to be applied to queries written over other
-data representations such as JSON. The system can also use mechanism such as a database
+In database systems, the system can use a mechanism such as a
 view and execute the original query over the view. When applying code migrations,
@@
 preferably by reducing the need for developer-written code.
+
+\frameworkbox{
+\begin{wrapfigure}{r}{12em}
+\vspace{-0.5em}
+\includegraphics[width=12em]{figures/arr-db.png}
+\vspace{-1em}
+\end{wrapfigure}
+\frameworkboxtitle{Equivalence in Schema Evolution}
+We described our conceptual framework in terms of types, code and data, but this case
+study shows that the same patterns appear elsewhere. In database systems, the schema
+$S_1$ changes into $S_2$ and the data $D_1$ need to transform to corresponding $D_2$.
+Our framing shows that queries $Q_1$ also need to transform, which is not always addressed
+in work on schema evolution. Extract/Absorb Entity operations are also interesting because
+they transform an equivalent pair of schemas and so it should be possible to apply them
+(and the corresponding data and query transformations) automatically.
+}
 
@@
 Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
-Jennings). The first challenge involves merging two independent edits to the data
-(Fig.~\ref{fig:conf-add} and Fig.~\ref{fig:conf-tab}); the second merging structural change
-with code (Fig.~\ref{fig:conf-budget} and Fig.~\ref{fig:conf-tab}). Formulas in the final version
-(Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
+Jennings). The first challenge involves merging a data edit
+(Fig.~\ref{fig:conf-add}) with a type and data edit (Fig.~\ref{fig:conf-tab}); the second involves
+merging a type and data edit (Fig.~\ref{fig:conf-tab}) with edit adding code (Fig.~\ref{fig:conf-budget}).
+Formulas in the final version (Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
 \label{fig:conf}
@@
   tracking which of the organizers should contact the speaker (Figure~\ref{fig:conf-tab}).
-  This changes the document structure (type).
+  This changes the document structure (implicit type).
 \end{enumerate}
@@
 Tomas. The newly added speaker should be reformatted into the new format; for the
-``Organizer'' column, the system may use a suitable default value or ask the user interactively.
+``Organizer'' column of the newly created table, the system may use a suitable default value
+or ask the user interactively.
 
 \frameworkbox{
-\begin{wrapfigure}{r}{8em}
-\includegraphics[scale=0.25]{figures/diag-data-change.png}
+\begin{wrapfigure}{r}{13em}
+\vspace{-0.5em}
+\includegraphics[width=13em]{figures/arr-datafork.png}
+\vspace{-1em}
 \end{wrapfigure}
-\frameworkboxtitle{Independent Data and Code Edits}
+\frameworkboxtitle{Merging Data and Code Edits}
 In this example, the \emph{type} is implicit and represents the structure of the document.
-A change to the \emph{type} done by one organizer needs to be applied to \emph{data} that
-has been modified by the other organizer in the meantime, as illustrated on the right.
-In the next section, we will see similar pattern, but with formulas, or \emph{code}.
-We assume the \emph{decentralized} scenario where \emph{type} evolution can occur in any variant
-and \emph{covergence} model where all variants eventually adopt all changes. As discussed
-later, the system could also operate in the more challenging \emph{divergence} model.}
+It involves two independent edits. In one (top right), the data $D_1$ is
+changed to $D_2'$, but the type $T_1$ remains unchanged. In the other (bottom left), the
+type $T_1$ and data $D_1$ co-evolve into $T_2$ and $D_2$ and the data is further changed
+to $D_3$. The challenge is to merge the two edits and obtain jointly modified data $D_3'$
+of type $T_2$. In the next section, we will see the same pattern, but also involving
+code edits. We assume the \emph{convergence} model in this section, but will discuss
+the more challenging \emph{divergence} model later in the paper.}
 
@@
 code on separate machines, which makes it impossible to update both atomically. To perform
-upgrades, teams must manually perform multi-step zero-downtime deployment processes~\cite{planetscalerails}
+upgrades, teams must manually perform multi-step zero-downtime deployment processes~\cite{planetscale}
 to step the database and the code servers forward gradually such that they are compatible with one
@@
 
-There has been a lot of research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types.
+\textcolor{red}{Maybe insert something about methodology in general?
+(ref the table challenges paper? although this paper is no longer so much about challenges)}
+
+There has been a lot of research on evolving interdependent layers of software, much of it guided by schema and types.
 But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system.

### 2024-05-30 paper.tex
Tweak paragraph size

@@
 If all variants of the program should eventually adopt all changes (converge), the complexity
-is lesser. If we allow user to apply only certain changes from other users (divergence), the
+is lesser. If we allow users to apply only certain changes from other users (divergence), the
 complexity is greater. A common case may be a combination of the two where types and schema
-of a program should converge, but data may diverge.
+of a program converge, but data may diverge.
 

### 2024-05-30 paper.tex
fix diagram reference
remove tikz diagrams

@@
 \usepackage[many]{tcolorbox}
-\usepackage{tikz-cd}
 \usepackage{float}
@@
 \vspace{0.5em}
-\includegraphics[scale=0.25]{figures/diag-divergence.png}
+\includegraphics[scale=0.25]{figures/old/diag-divergence.png}
 \end{wrapfigure}
@@
 Or a system might even allow users to manually disambiguate their intent for a given write.
-
-\newpage
-~
-\newpage
-
-\section{diagrams}
-
-\begin{figure}[H]
-  \centering
-  \begin{tikzcd}[sep = 6em]
-    \textsf{query}
-    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
-    &
-    \textsf{schema}
-    \arrow[l, "\textsf{constrains}"{above}]
-    \arrow[r, "\textsf{constrains}"{above}]
-    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
-    &
-    \textsf{data}
-    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
-    \\
-    \textsf{query}_\alpha
-    &
-    \textsf{schema}_\alpha
-    \arrow[l, "\textsf{constrains}"{above}]
-    \arrow[r, "\textsf{constrains}"{above}]
-    &
-    \textsf{data}_\alpha
-    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
-    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
-  \end{tikzcd}
-  \caption{Extract Entity}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \begin{tikzcd}[column sep = large, row sep = 5em]
-    &
-    \textsf{schema}
-    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
-    \arrow[rr, "\textsf{constrains}"]
-    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
-    &
-    &
-    \textsf{data}
-    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
-    \arrow[Rightarrow, from=L, to=R]
-    \\
-    \textsf{schema}_\alpha
-    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
-    \arrow[rr, crossing over]
-    &
-    &
-    \textsf{data}_\alpha
-    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
-    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
-    &
-    \\
-    &
-    \textsf{schema}_\beta
-    \arrow[ld, "\textsf{change}_\alpha \succ\, \textsf{change}_\beta"{name=L4, near start, description}]
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_\beta
-    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
-    \arrow[Rightarrow, from=L4, to=R4]
-    \\
-    \textsf{schema}_{\alpha\beta}
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_{\alpha\beta}
-    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
-    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
-    &
-  \end{tikzcd}
-  \caption{Mergeing divergent changes with rebase ($\succ$)}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \begin{tikzcd}[column sep = large, row sep = 5em]
-    &
-    \textsf{schema}
-    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
-    \arrow[rr, "\textsf{constrains}"]
-    \arrow[dd, "\textsf{identity}"{near start, description}]
-    &
-    &
-    \textsf{data}
-    \arrow[dd, red, "\textsf{update}_\beta"{near start, description}]
-    \\
-    \textsf{schema}_\alpha
-    \arrow[rr, crossing over]
-    &
-    &
-    \textsf{data}_\alpha
-    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
-    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
-    \arrow[rr, red, crossing over, "\textsf{update}_\gamma"{near start, description}]
-    &
-    &
-    \textsf{data}_{\alpha\gamma}
-    \arrow[dd, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha \succ\, \textsf{update}_\gamma"{near start, description}]
-    \\
-    &
-    \textsf{schema}
-    \arrow[ld, "\textsf{change}_\alpha"{name=L4, near start, description}]
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_\beta
-    \arrow[ld, "\textsf{migrate}_\alpha"{name=R4, near start, description}]
-    \arrow[Rightarrow, from=L4, to=R4]
-    \\
-    \textsf{schema}_\alpha
-    \arrow[rr]
-    &
-    &
-    \textsf{data}_{\alpha\beta}
-    \arrow[from=uu, crossing over, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha"{near start, description}]
-    &
-    &
-    \textsf{data}_{\alpha\beta\gamma}
-  \end{tikzcd}
-  \caption{Conference Organizer merging schema and divergent data updates}
-\end{figure}
-
-\textcolor{red}{[JE] These diagrams are getting hairy. Are they actually helpful?}
-
-\newpage
-
-\newpage
 

### 2024-05-30 paper.tex
tweaks to my stuff

@@
 Acme needs to evolve the above schema and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
-one with orders that links to one with customers.  We refer to this as \emph{Extract Entity}
+one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
 operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
@@
 
-The \textit{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers,
+The \emph{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers,
 and 3) reference these identifiers from the orders. Our example took the simple approach of
@@
 of the customer? To repair that error we need to merge the falsely distinct customers into one
-and fix any references from orders. We call this operation \textit{Merge Entities} -- note that
+and fix any references from orders. We call this operation \emph{Merge Entities} -- note that
 it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.
@@
 Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
-change in requirements that triggered an evolution might change back again. Inverse
-is also required in research on \emph{divergence control}~\cite{Foster2007, herrmann17, chillon22}.
-The inverse of \textit{Extract Entity} is \textit{Absorb Entity}. It raises the question of how
-to handle \textit{one-to-many} relationships. For example should absorbing orders into customers
+change in requirements that triggered an evolution might change back again.
+Invertibility features in other work on schema evolution and \emph{bidirectional transformations}~\cite{Foster2007, herrmann17, chillon22}.
+The inverse of \emph{Extract Entity} is \emph{Absorb Entity}. It raises the question of how
+to handle \emph{one-to-many} relationships. For example should absorbing orders into customers
 do a relational join, returning us back to the starting point, or should it nest orders within
@@
 
-The inverse of \textit{Merge Entities} is \textit{Split Entity}, but that can be awkward.
+The inverse of \emph{Merge Entities} is \emph{Split Entity}, but that can be awkward.
 Imagine that we didn't have addresses for customers to disambiguate them with and only knew
-their non-unique names. Then \textit{Extract Entity} will conflate them and discard the
+their non-unique names. Then \emph{Extract Entity} will conflate them and discard the
 information needed to properly invert the operation. Should this information be retained?
@@
 \noindent
-After applying the \textit{Extract Entity} operation as described earlier, the code should
+After applying the \emph{Extract Entity} operation as described earlier, the code should
 be rewritten into the following query that joins the two linked tables:
@@
 \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}.
-BiDEL appears capable of handling the basic requirements of our \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} challenges, though \textit{Split/Merge Entity} is less clear.
+BiDEL appears capable of handling the basic requirements of our \emph{Extract/Absorb Entity} and \emph{Multiplicity Change} challenges, though \emph{Split/Merge Entity} is less clear.
 It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.
@@
 \citet{scherzinger13} define a set of operators like the relational SMOs discussed above.
-\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} but not \textit{Split/Merge Entity} nor divergence.
+\citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \emph{Multiplicity Change} but not \emph{Split/Merge Entity} nor divergence.
 None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.
@@
 They consider the evolution of OO-style subclass hierarchies which goes beyond most other work.
-Their running example includes both our Extract Entity and Multiplicty Change examples, and would make for a good follow-on problem.
+Their running example includes both our \emph{Extract Entity} and \emph{Multiplicty Change} examples, and would make for a good follow-on problem.
 However they do not address code/query rewriting.

### 2024-05-30 paper.tex
Edits and new diagrams for divergence sections

@@
 a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
-of \emph{type evolution} is synthetising ($\Rightarrow$) suitable corresponding transformations
+of \emph{type evolution} is synthesising ($\Rightarrow$) suitable corresponding transformations
 that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
@@
 \begin{lstlisting}[language=ml]
-type Item = { id : id; title : string; completed : Maybe DateTime }
+type Item = { id : id; title : string; completed : Maybe<DateTime> }
 type State = { items : Item list }
 type Event =
-  | .. | SetCompleted of id * Maybe DateTime | ..
+  | .. | SetCompleted of id * Maybe<DateTime> | ..
 \end{lstlisting}
@@
 \challenge{Extract Entity}
-Acme needs to evolve the above schema and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
+Acme needs to evolve the above schema and data into two tables (Fig.~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
 one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
@@
 
+% ==================================================================================================
+% DIVERGENCE
+% ==================================================================================================
+
 \section{Divergence Control in Spreadsheets and Databases}
@@
 together in an atomic way. In these situations, different versions must coexist and interoperate
-with one another. For example, edits done against data of one type must be applied to
-data of another type, or code written against one type must run against data stored in another.
+with one another. For example, edits done to data of a new type must be applied to
+data of an older type, or code written against older type must run against data of a newer type.
 
@@
 upgrades, teams must manually perform multi-step zero-downtime deployment processes~\cite{planetscalerails}
-to step the database and the code servers forward gradually such that they are compatible with one
+to migrate the database and the code servers forward gradually such that they are compatible with one
 another at any given step. One example would be deploying a change to add a database column
@@
 years, since it is not realistic to force all consumers to upgrade immediately. Divergence can
-even be indefinite -- Cambria \cite{Cambria} addresses collaborative applications where different
+even be indefinite. Cambria \cite{Cambria} addresses collaborative applications where different
 users may want to collaborate on shared data through different tools using different schemas,
@@
 
+\challenge{Divergence Control for Extract Entity}
+
+As a concrete example, we add the problem of divergence control to the
+earlier Extract Entity challenge. Every month the orders department sends its spreadsheet to
+the accounting department, which copies the data to its own version of the spreadsheet.
+This version uses the same structure (type), but includes additional code for financial tracking.
+When the orders department migrates its spreadsheet to a database and extracts out customers,
+the accounting department does not want to conform. They want to continue using their old
+spreadsheet with their additional code.
+
+The accounting department could manually convert incoming data in the new schema into the old schema. But they
+shouldn't have to. It should be possible to run new data through the schema change in reverse,
+converting it back into the format the accounting department is used to ingesting. Note that
+this is not a matter of synchronizing the divergent spreadsheets -- they maintain differently
+evolved schema.
+
 \frameworkbox{
-\begin{wrapfigure}{r}{12em}
-\vspace{0.5em}
-\includegraphics[scale=0.25]{figures/old/diag-divergence.png}
+\begin{wrapfigure}{r}{14em}
+\vspace{-1.5em}
+\includegraphics[width=14em]{figures/arr-backport.png}
+\vspace{-2.5em}
 \end{wrapfigure}
-\frameworkboxtitle{Type Branching with Divergence}
-The divergence scenario adds another dimension to the problem. There are now multiple
-versions of the system with different \emph{types} that may co-exist as multiple branches.
-Edits to \emph{data} are done against a specific type, but they then need to be migrated
-to versions with different types. Consequently, any type evolution needs to be accompanied
-both with forward projection (transforming data to a new version) and backward projection
-(transforming new data to the old version).}
+\frameworkboxtitle{Bidirectional Data Migration}
+The scenario adds the need for transferring data changes from a new type to an older type.
+In the diagram on the right, we only show data and type, ignoring the code added by the
+accounting department. We start with data $D_1$ of type~$T_1$. The accounting deparment
+evolves the type (via Extract Entity), obtaining data $D_2$ of type $T_2$ using an automatically
+generated forward transformation. It further modifies the data, obtiaing $D_3$. When the
+accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to
+$T_1$ using an automatically generated backward transformation. In other words,
+type evolution needs to be accompanied with both forward and backward projection for data
+(and code) migration.}
+
+What is needed is a user-friendly mechanism for transferring data changes from one schema to
+another bidirectionally. In one direction we want to transfer changes made in the new schema
+into a variant of the old schema. The opposite direction might be needed, for example,
+if the accounting department corrects customer addresses, which ought to be pushed
+through into the orders department's new schema.
+
+\subsection*{Remarks: Divergence in End-User Workflows}
 
 Divergence does not just occur in software engineering; it also appears in end-user
-workflows. Users frequently make copies of structured documents like spreadsheets and then
+workflows. Users frequently make copies of documents like spreadsheets and then
 diverge them by altering both data content and schema. In the case of spreadsheets schema
-changes include rearrangements to the structure of rows and columns as well as changes to
-formulas. The problem is that the users want to transfer such data and schema changes between
-these divergent copies, and they want to pick and choose which of these changes to transfer.
+changes include rearrangements to the structure of rows and columns, while code changes affect
+formulas. The users then want to transfer such schema, code and data changes between
+divergent copies, and they want to pick and choose which of the changes to transfer.
 In practice such transfer is done manually through copy \& paste. \citet{Basman19} has
 documented an ecology of emailed spreadsheets. \citet{Burnett14} distilled field observations
-of these practices in the story of Frieda, which we quote here:
+of these practices in the story of Frieda:
 
@@
 \noindent
-\emph{For example, consider ``Frieda'', an office manager in charge of her department's budget
+\emph{[Frieda is] an office manager in charge of her department's budget
 tracking. (\ldots) Every year, the company she works for produces an updated budget tracking
@@
 She can reuse some of these same changes this year, but copying them into Dept-thisYear
-is troublesome, with some of the formulas automatically adjusting themselves to refer to
-Dept-lastYear. (\ldots) Frieda has learned over the years to
-save some of her spreadsheet variants along the way (using a different filename for each),
+is troublesome (\ldots). Frieda has learned over the years to
+save some of her spreadsheet variants along the way (\ldots),
 because she might decide that the way she did some of her changes was a bad idea, and she
@@
 
-\challenge{Divergence Control for Extract Entity}
-
-As a concrete example, we combine the problem of divergence control with the example from the
-earlier Extract Entity challenge. Every month the orders department sends its spreadsheet to
-the accounting department, which adds the data to a version of the spreadsheet that it has
-customized for financial tracking. But when the orders department migrates its spreadsheet to
-extract out customers, the accounting department does not want to conform. They could manually
-convert incoming data in the new schema into their variant of the old schema. But they
-shouldn't have to. It should be possible to run new data through the schema change in reverse,
-converting it back into the format the accounting department is used to ingesting. Note that
-this is not a matter of synchronizing the divergent spreadsheets -- they maintain differently
-evolved schema.
-
-What is needed is a user-friendly mechanism for transferring data changes from one schema (type) to
-another bidirectionally. In one direction we want to transfer changes made in the new schema (type)
-into a variant of the old schema (type). The opposite direction might be needed, for example,
-if the accounting department makes corrections to customer addresses, which ought to be pushed
-through into the orders department's new schema.
+\frameworkbox{
+\begin{wrapfigure}{r}{14.5em}
+\vspace{-0.5em}
+\includegraphics[width=14.5em]{figures/arr-divergence.png}
+\vspace{-1.5em}
+\end{wrapfigure}
+\frameworkboxtitle{Maintaining Divergent Variants}
+The extra complexity in the above scenario arises from the fact that both types and
+code are forked and evolve independently. In one branch, original types $T_1$ evolve
+into $T_2$ and original code $C_1$ into $C_3$. In another branch, original types $T_1$
+evolve into $T'_2$ and original code $C_1$ into $C'_3$. The user then wants to merge
+the independently done changes into a new version of types and code, written as
+$T_2 \diamond T'_2$ and $C_3 \diamond C'_3$. The diverging transformations may make
+conflicting changes that would need to be resolved with assistance from the user.}
 
@@
 %\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!}
-For its part, local-first software must move from automatically converging data replicas
+Local-first software must move from automatically converging data replicas
 to also interactively managing long-term divergence. That implies either application state
@@
 
+% ==================================================================================================
+% MULTIPLICITY
+% ==================================================================================================
+
 \section{Multiplicity Change in Data Formats}
 
-A common kind of schema change is \textit{multiplicity change}: when a field changes from storing
+The challenge of maintaining multiple divergent types in programming systems can
+be well illustrated using \emph{multiplicity change}, which is a common kind of schema change.
+In this case a field changes from storing
 a single value to storing multiple values. For example, we might start by storing a single
 address for each contact in an address book, before realizing that we need to store multiple
-addresses for a single contact; or we might assign each todo in a list to one person before
+addresses for a single contact. Alternatively, we might assign each todo in a list to one person before
 realizing we want the ability to assign a todo to multiple people.
@@
 the linked entity; some of the challenges of this approach are covered in the Extract Entity
-challenge. In this section we will instead consider a document schema with support for arrays;
+challenge. In this section we will instead consider a type structure with support for arrays;
 in this context, we can have a multiplicity change by turning a scalar value into an array value.
-
-Multiplicity change clearly requires changing both the data and the code. It becomes particularly
-challenging to handle when there is ongoing divergence between multiple versions of the data.
-
-\challenge{Multiplicity Change}
-
-Consider the following schema for a todo list item, in which each item has a single assignee, represented by a user ID:
-
-\newpage
+%
+A change of multiplicity in the type requires producing a transformation for both the corresponding
+data and the corresponding code. The challenge becomes particularly difficult to handle when there
+is ongoing divergence between multiple versions of the types and data.
+
+\challenge{Multiplicity Change in Types}
+
+Consider the following schema for a todo list item, in which each item has a single assignee,
+represented by a user ID:
+
 \begin{lstlisting}[language=ml]
@@
 
+\noindent
 Now, we change the schema so that each item has a list of assignees:
@@
 
-Our goal is to preserve the ability for actors in the system to read and write to a shared todo list in either the old or new schema -- for example, they should be able to write to either the scalar assignee field or to the list of assignees.
-
-A natural invariant to preserve across these schemas is that the value of the scalar field should equal the first element of the list field (and if the list is empty, the scalar field should be null).
-If we were to write code for a one-time data migration from the scalar to list schema, we could easily satisfy this invariant:
+\noindent
+Our goal is to preserve the ability for actors in the system to read and write to a shared todo
+list in either the old or the new schema. For example, an actor should be able to write to either
+the scalar assignee field when using the original version or to the list of assignees when using
+the new version.
+
+A natural invariant to preserve across these schemas is that the value of the scalar field should
+equal the first element of the list field (and if the list is empty, the scalar field should be
+\texttt{null}). If we were to write code for a one-time data migration from the scalar to list
+schema, we could easily satisfy this invariant:
 
@@
 
-Ongoing edits to the array schema can also be handled in a straightforward way.
-After edits are made, the value of the scalar field should be set to the first element of the new array (or null if the new array is empty.)
-
-However, handling edits to the scalar schema presents more of a challenge.
-Consider the following todo with two assignees, presented in terms of the scalar and array schemas.
-A write is made in the scalar schema to set the new assignee to C.
+\noindent
+Ongoing edits to the array schema can also be handled in a straightforward way. After edits are
+made, the value of the scalar field should be set to the first element of the new array (or \texttt{null}
+if the new array is empty.)
+
+However, handling edits to the scalar schema presents more of a challenge. Consider the following
+todo with two assignees, presented in terms of the scalar and array schemas. A write is made in
+the scalar schema to set the new assignee to C.
 
@@
 
-What should the array value become after this edit to the scalar schema?
-To satisfy our invariant, we know that C must become the first element of the array, but this leaves open several options with different tradeoffs:
+What should the array value become after this edit to the scalar schema? To satisfy our invariant,
+we know that C must become the first element of the array, but this leaves open several options
+with different tradeoffs:
 
 \begin{enumerate}
-  \item \texttt{[C]}: "Only C should be assigned." This option produces an array that corresponds directly to the resulting scalar.
+  \item \texttt{[C]}: \emph{Only C should be assigned.} This option produces an array that corresponds directly to the resulting scalar.
 But it has the downside of deleting data that wasn't even visible in the scalar schema.
-  \item \texttt{[C, B]}: "Replace A with C." If the writer wanted to remove A from the assignment and add C, this option performs that intent.
+  \item \texttt{[C, B]}: \emph{Replace A with C.} If the writer wanted to remove A from the assignment and add C, this option performs that intent.
 However, there is data remaining in the list which was not visible to the scalar writer.
-  \item \texttt{[C, A, B]}: "Add C to the list." Perhaps the scalar writer wanted to add C to the assignment without removing anyone; this option satisfies that intent.
+  \item \texttt{[C, A, B]}: \emph{Add C to the list.} Perhaps the scalar writer wanted to add C to the assignment without removing anyone; this option satisfies that intent.
 But it preserves data not visible to the scalar schema.
@@
 
-Because the intent of the writer to the scalar schema cannot be unambiguously interpreted from the write alone, it is impossible to make a perfect choice among these options.
+Because the intent of the writer to the scalar schema cannot be unambiguously interpreted from the
+write alone, it is impossible to make a perfect choice among these options. The programming system
+thus needs to consult the user.
+
+\frameworkbox{
+\begin{wrapfigure}{r}{13em}
+\vspace{-1em}
+\includegraphics[width=13em]{figures/arr-backport.png}
+\vspace{-2em}
+\end{wrapfigure}
+\frameworkboxtitle{Bidirectional Data Migration (Revisited)}
+The diagram illustrating the scenario remains the same as in the case of divergence control
+for Extract Entity. The initial data and type $D_1, T_1$ co-evolve into $D_2, T_2$ (adding
+multiplicity) and a new value is then assigned to the array of assignees, resulting in $D_3$.
+The difference from the earlier case is that the backward migration (from $D_3$ to $D'_3$)
+generated for the type evolution (from $T_1$ to $T_2$) cannot be generated fully automatically.
+In case of Extract Entity this was in principle possible.}
 
@@
 
-If schemas, code and data can all be updated atomically together, then multiplicity change is relatively straightforward  to handle.
-Existing scalar data can be trivially migrated to a list, and the code using the data will need to change to accommodate the list data.
-
-However, ongoing divergence makes multiplicity change much more difficult, and exposes some general challenges.
-Different schemas may expose partial information, and writers in those schemas have to operate without total knowledge of the information available in other schemas.
-As a result, synchronizing writes across the schemas requires making difficult tradeoffs, such as choosing to either preserve or destroy hidden data not visible to the writer.
-
-Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this problem would give developers or users tools to manage these tradeoffs.
-For example, a system might allow developers to specify the desired behavior for a particular pair of schemas based on the requirements of the domain.
-Or a system might even allow users to manually disambiguate their intent for a given write.
+If types, code and data can all be updated atomically together, then multiplicity change is
+relatively straightforward  to handle. Existing scalar data can be trivially migrated to a
+list, and the code using the data will need to change to accommodate the list data.
+%
+However, ongoing divergence makes multiplicity change much more difficult, and exposes
+some general challenges. Different types may expose partial information, and writers using
+those types have to operate without total knowledge of the information available in other types.
+As a result, synchronizing writes across the types requires making difficult tradeoffs, such as
+choosing to either preserve or destroy hidden data not visible to the writer.
+
+Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this
+problem would give developers or users tools to manage these tradeoffs. For example, a system
+might allow developers to specify the desired behavior for a particular pair of schemas based
+on the requirements of the domain. Or a system might even allow users to manually
+disambiguate their intent for a given write.
 

### 2024-05-31 paper.tex
leftbar instead of frameworkbox
Move related work to appendix

@@
 \usepackage{wrapfig}
+\usepackage{framed}
 
@@
 
+\leftbar
+\begin{wrapfigure}{r}{12em}
+  \vspace{0em}
+  \includegraphics[width=12em]{figures/arr-steps.png}
+  \vspace{0em}
+\end{wrapfigure}
+\noindent\textbf{Primitive Type Transformations.}\quad
+In this case study, the evolving type is the \texttt{State} type. When the type evolves,
+the program logic (code) and the current live value of the type (data) needs to evolve
+correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
+transformations, such as those in Table~\ref{tbl:elmchanges}, that each has a corresponding
+primitive code and data transformation. The primitive transformations may either be invoked
+through a user interface, or inferred from textual code edits made by the programmer.
+
+\endleftbar
+
+\begin{comment}
 \frameworkbox{
@@
 }
+\end{comment}
 
@@
 data, we can gain valuable insights from research on database systems.
-Although schema evolution is well-studied in databses (Section~\ref{sec:related}), past work
+Although schema evolution is well-studied in databases (Section~\ref{sec:related}), past work
 does not resolve all problems in our second challenge.
@@
 
-\newpage
+%\newpage
 
@@
 
-\section{Related Work}
+\newpage
+\appendix
+
+\section{Literature Review}
 \label{sec:related}
 
-There has been a lot of research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types.
+There has been estensive research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types.
 But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system.
@@
 
-\appendix
+%\appendix
 

### 2024-05-31 paper.tex
typo
Redefine \frameworkbox to use \leftbar

@@
 
+\DeclareRobustCommand{\frameworkbox}[1]{\leftbar#1\endleftbar}
+\newcommand{\frameworkboxtitle}[1]{\noindent{\firamedium #1.}\quad}
+
+\begin{comment}
 \DeclareRobustCommand{\frameworkbox}[2][gray!15]{
@@
 \newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}
+\end{comment}
 
@@
 
-\leftbar
-\begin{wrapfigure}{r}{12em}
-  \vspace{0em}
-  \includegraphics[width=12em]{figures/arr-steps.png}
-  \vspace{0em}
-\end{wrapfigure}
-\noindent\textbf{Primitive Type Transformations.}\quad
-In this case study, the evolving type is the \texttt{State} type. When the type evolves,
-the program logic (code) and the current live value of the type (data) needs to evolve
-correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
-transformations, such as those in Table~\ref{tbl:elmchanges}, that each has a corresponding
-primitive code and data transformation. The primitive transformations may either be invoked
-through a user interface, or inferred from textual code edits made by the programmer.
-
-\endleftbar
-
-\begin{comment}
 \frameworkbox{
@@
 }
-\end{comment}
 
@@
 
-There has been estensive research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types.
+There has been extensive research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types.
 But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system.

### 2024-05-31 paper.tex
WIP tweaking statemachines

@@
 \usepackage{multicol}
+\usepackage{xcolor}
 \usepackage{ccicons}
-\usepackage[many]{tcolorbox}
 \usepackage{float}
@@
 
+% \definecolor{progblue}{RGB}{64,104,224}
+% \renewenvironment{leftbar}[1][\hsize]
+% {%
+%     \def\FrameCommand
+%     {%
+%         {\color{progblue}\vrule width 3pt}%
+%         \hspace{0pt}%must no space.
+%         \fboxsep=\FrameSep\colorbox{white}%
+%     }%
+%     \MakeFramed{\hsize#1\advance\hsize-\width\FrameRestore}%
+% }
+% {\endMakeFramed}
+
 \DeclareRobustCommand{\frameworkbox}[1]{\leftbar#1\endleftbar}
 \newcommand{\frameworkboxtitle}[1]{\noindent{\firamedium #1.}\quad}
-
-\begin{comment}
-\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
-\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
-  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
-  outer arc=0pt]\setlength{\parskip}{0.5em}#2
-\end{tcolorbox}}
-\newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}
-\end{comment}
 
@@
 
-\section{Divergence Control in Spreadsheets and Databases}
+\section{Divergence Control in Spreadsheets}
 Type evolution becomes more challenging when data, code, and types cannot all be updated
@@
 
-%\newpage
-
-\section{Live Modeling Languages Require Run-Time State Migration}
-\label{livemodeling}
-%Related work
-%``Slogan'': editing a program is diverging from run-time schema.
-
-\subsection{Context}
-In stateful live programming, a change in the program can make the run-time state of the execution out of date, hence we want to migrate the run-time state to keep on running.
-
-One can see a program itself as an instance of a static schema (its AST type), that in turn determines (defines/implies/induces) a run-time schema: the run-time structures of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as the structure of the run-time state.
-Whereas the run-time \textit{code} structures do not typically change while using an application, the run-time \textit{state} constantly changes.
-Live programming, however, requires reconciling changes to the program with the run-time structures of both the code and the state.
-This typically means that at a certain point (a quiescent point) during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.
-
-\subsection{Example}
-Consider the example of a simple state machine language with on-entry actions, and typed global variables (loosely inspired by the SML language of \citet{vanRozen19}).
-An example state machine and an excerpt of its abstract syntax schema is shown in Figure~\ref{LST:statemachines}.
-Depicted on the left, an actual statemachine modeling the opening and closing of a door, with one global boolean variable, \lstinline{isClosed}, which is flipped according to transitioning between the \lstinline{closed} and \lstinline{opened} states.
-For the sake of brevity, the abstract syntax schema (similar, to e.g., and Ecore metamodel~\cite{EMF}) shown on the right of the figure omits classes for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).
+% ==================================================================================================
+% MODELING
+% ==================================================================================================
 
 \begin{figure}[t]
-\centering
-\begin{minipage}[t]{0.4\textwidth}
-\begin{lstlisting}[language=java,morekeywords={machine,on,state,var}]
-machine Doors
-
-var isClosed: bool = true
-
-state closed
-  isClosed := true
-  on open => opened
-
-state opened
-  isClosed := false
-  on close => closed
-\end{lstlisting}
-\end{minipage}
-\hspace*{5pt}\vline
-\begin{minipage}[t]{0.4\textwidth}
-\begin{lstlisting}[language=java,morekeywords={on}]
+\begin{adjustwidth}{-1.5em}{-1.5em}
+\begin{minipage}[t]{0.26\textwidth}
+\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
 class Machine
@@
   vars: Decl*
-
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}[t]{0.28\textwidth}
+\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
 class State
@@
   transitions: Trans*
-
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}[t]{0.24\textwidth}
+\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
 class Trans
@@
   target: State
-
+$~$
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}[t]{0.24\textwidth}
+\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
 class Decl
@@
   type: Type
+$~$
 \end{lstlisting}
 \end{minipage}
+\end{adjustwidth}
+\vspace{-1em}
 \caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
-\label{LST:statemachines}
+\label{lst:machinedef}
+\vspace{0.5em}
 \end{figure}
+
+~
+
+\section{Live Modeling Languages}
+\label{sec:livemodeling}
+%Related work
+%``Slogan'': editing a program is diverging from run-time schema.
+
+We considered live programming in the context of the Elm architecture in Section~\ref{sec:elm},
+where the challenge was to migrate run-time state (data) to a new type to keep the program running.
+Work on language modeling offers a new perspective on the problem. In systems such as Rascal
+\cite{klint2011easy}, one can see a program in a domain-specific language not just as \emph{types}
+and \emph{code}, but also as \emph{data}, or an instance of a metamodel (a static type) that
+defines the AST of the domain-specific language. The program in a domain-specific language
+that in turn determines (defines/implies/induces) a run-time schema: the run-time structures
+of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as
+the structure of the run-time state.
+
+In the absence of live programming, the run-time \emph{code} structures do not typically
+change while using an application and the run-time \emph{state} changes constantly.
+
+Live programming, however, requires reconciling changes to the program with the run-time structures
+of both the code and the state. This typically means that at a certain point (a quiescent point)
+during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.
+
+\subsection{Example}
+Consider the example of a simple state machine language with on-entry actions, and typed global
+variables (loosely inspired by the SML language of \citet{vanRozen19}). An example state machine
+and an excerpt of its abstract syntax schema is shown in Figure~\ref{LST:statemachines}. Depicted
+on the left, an actual statemachine modeling the opening and closing of a door, with one global
+boolean variable, \lstinline{isClosed}, which is flipped according to transitioning between the
+\lstinline{closed} and \lstinline{opened} states. For the sake of brevity, the abstract syntax
+schema (similar, to e.g., and Ecore metamodel~\cite{EMF}) shown on the right of the figure
+omits classes for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).
+
+\noindent
+\begin{adjustwidth}{-0.5em}{0em}
+\begin{minipage}[t]{0.32\textwidth}
+\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
+machine Doors
+
+var isClosed: bool = true
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}[t]{0.32\textwidth}
+\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
+state closed
+  isClosed := true
+  on open => opened
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}[t]{0.32\textwidth}
+\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
+state opened
+  isClosed := false
+  on close => closed
+\end{lstlisting}
+\end{minipage}
+\end{adjustwidth}
+
+%\begin{figure}[t]
+%\centering
+%\caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
+%\label{LST:statemachines}
+%\end{figure}
 
@@
 \citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
-\citet{RuntimeConstraint} study the live modeling problem in Section \ref{livemodeling}.
+\citet{RuntimeConstraint} study the live modeling problem in Section \ref{sec:livemodeling}.
 They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution.

### 2024-05-31 paper.tex
Add diagram with very sketchy description

@@
 
+\frameworkbox{
+\begin{wrapfigure}{r}{14.5em}
+\vspace{-0.5em}
+\includegraphics[width=14.5em]{figures/arr-model.png}
+\vspace{-1.5em}
+\end{wrapfigure}
+\frameworkboxtitle{TODO}
+Here, data $D_0$ now represent the state machine definition.
+Live data that we obtain by instantiating the state machine still contain this state machine
+definition, but also the necessary live state defined by the definition.
+
+This is used to synthesize live machine state $+D_0$ conforming to live machine type $+T_0$.
+This then runs producing $+T'_0$ and so on...
+
+If the state machine definition changes from $D_0$ to $D_1$,
+we can synthesize new live machine state $+D_1$ conforming to new live machine type $+T_1$.
+
+The challenge is to merge these changes with the currently running state machine $+T'_0$.
+We have to produce merged live machine type $+T_0 \diamond +T_1$ and also merge the
+live state changes to obtain $+D'_0 \diamond +D_1$ of the new merged live machine type.
+}
+
 \begin{figure}[t]

### 2024-05-31 paper.tex
editing live modeling

@@
 
+% \begin{figure}[t]
+% \begin{adjustwidth}{-1.5em}{-1.5em}
+% \begin{minipage}[t]{0.26\textwidth}
+% \begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
+% class Machine
+%   name: string
+%   states: State*
+%   vars: Decl*
+% \end{lstlisting}
+% \end{minipage}
+% \begin{minipage}[t]{0.28\textwidth}
+% \begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
+% class State
+%   name: string
+%   onEntry: Stmt*
+%   transitions: Trans*
+% \end{lstlisting}
+% \end{minipage}
+% \begin{minipage}[t]{0.24\textwidth}
+% \begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
+% class Trans
+%   event: string
+%   target: State
+% $~$
+% \end{lstlisting}
+% \end{minipage}
+% \begin{minipage}[t]{0.24\textwidth}
+% \begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
+% class Decl
+%   name: string
+%   type: Type
+% $~$
+% \end{lstlisting}
+% \end{minipage}
+% \end{adjustwidth}
+% \vspace{-1em}
+% \caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
+% \label{lst:machinedef}
+% \vspace{0.5em}
+% \end{figure}
+
 \begin{figure}[t]
-\begin{adjustwidth}{-1.5em}{-1.5em}
-\begin{minipage}[t]{0.26\textwidth}
-\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
-class Machine
-  name: string
-  states: State*
-  vars: Decl*
+\centering
+\begin{minipage}{0.3\textwidth}
+\includegraphics[width=\textwidth]{figures/metamodel.pdf}
+\end{minipage}
+\hspace*{2pt}
+\vline
+\begin{minipage}{0.3\textwidth}
+  \includegraphics[width=\textwidth]{figures/doors-ast.pdf}
+  \end{minipage}
+\hspace*{2pt}
+\vline
+\begin{minipage}{0.3\textwidth}
+  \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},basicstyle=\footnotesize,numbers=none]
+machine Doors
+var isClosed: bool = true
+state closed
+  isClosed := true
+  on open => opened
+state opened
+  on close => closed
 \end{lstlisting}
 \end{minipage}
-\begin{minipage}[t]{0.28\textwidth}
-\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
-class State
-  name: string
-  onEntry: Stmt*
-  transitions: Trans*
-\end{lstlisting}
-\end{minipage}
-\begin{minipage}[t]{0.24\textwidth}
-\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
-class Trans
-  event: string
-  target: State
-$~$
-\end{lstlisting}
-\end{minipage}
-\begin{minipage}[t]{0.24\textwidth}
-\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
-class Decl
-  name: string
-  type: Type
-$~$
-\end{lstlisting}
-\end{minipage}
-\end{adjustwidth}
-\vspace{-1em}
-\caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
-\label{lst:machinedef}
-\vspace{0.5em}
+\caption{Definition of the abstract syntax state machine language as a UML metamodel and an example instance, as object diagram and as source code}
+\label{FIG:statemachineUML}
 \end{figure}
@@
 where the challenge was to migrate run-time state (data) to a new type to keep the program running.
-Work on language modeling offers a new perspective on the problem. In systems such as Rascal
-\cite{klint2011easy}, one can see a program in a domain-specific language not just as \emph{types}
-and \emph{code}, but also as \emph{data}, or an instance of a metamodel (a static type) that
-defines the AST of the domain-specific language. The program in a domain-specific language
-that in turn determines (defines/implies/induces) a run-time schema: the run-time structures
-of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as
-the structure of the run-time state.
+Work on language modeling offers a new perspective on the problem.
+In the context of executable domain-specific modeling languages~\cite{}, one can see a program in a domain-specific language as an instance of a metamodel (a static type) that
+defines the AST structure of the domain-specific language. The program in a domain-specific language
+then in turn determines (defines/implies/induces) a run-time schema: the run-time structures
+of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as the structure of the run-time state (the heap, slots for user data, etc.).
+Whereas the example discussed in this section is framed in terms of metamodels and models, this should be seen as an expository device: the principles and challenges apply, e.g., in more grammar-oriented settings.
 
@@
 change while using an application and the run-time \emph{state} changes constantly.
-
-Live programming, however, requires reconciling changes to the program with the run-time structures
+Live programming, however, requires reconciling changes to the program with the induced run-time structures
 of both the code and the state. This typically means that at a certain point (a quiescent point)
@@
 Consider the example of a simple state machine language with on-entry actions, and typed global
-variables (loosely inspired by the SML language of \citet{vanRozen19}). An example state machine
-and an excerpt of its abstract syntax schema is shown in Figure~\ref{LST:statemachines}. Depicted
-on the left, an actual statemachine modeling the opening and closing of a door, with one global
+variables (loosely inspired by the SML language of \citet{vanRozen19}). 
+The abstract syntax (as UML class diagram) is shown on the left of Figure~\ref{FIG:statemachineUML}. 
+The middle of the figure shows an instance: a state machine modeling the opening and closing of a door (in UML object notation). 
+The right shows a possible textual concrete syntax modeling the same state machine. 
+The state machine has one global
 boolean variable, \lstinline{isClosed}, which is flipped according to transitioning between the
 \lstinline{closed} and \lstinline{opened} states. For the sake of brevity, the abstract syntax
-schema (similar, to e.g., and Ecore metamodel~\cite{EMF}) shown on the right of the figure
-omits classes for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).
-
-\noindent
-\begin{adjustwidth}{-0.5em}{0em}
-\begin{minipage}[t]{0.32\textwidth}
-\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
-machine Doors
-
-var isClosed: bool = true
-\end{lstlisting}
-\end{minipage}
-\begin{minipage}[t]{0.32\textwidth}
-\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
-state closed
-  isClosed := true
-  on open => opened
-\end{lstlisting}
-\end{minipage}
-\begin{minipage}[t]{0.32\textwidth}
-\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
-state opened
-  isClosed := false
-  on close => closed
-\end{lstlisting}
-\end{minipage}
-\end{adjustwidth}
+model  shown on the left of the figure
+omits the details for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).
+
+% \noindent
+% \begin{adjustwidth}{-0.5em}{0em}
+% \begin{minipage}[t]{0.32\textwidth}
+% \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
+% machine Doors
+
+% var isClosed: bool = true
+% \end{lstlisting}
+% \end{minipage}
+% \begin{minipage}[t]{0.32\textwidth}
+% \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
+% state closed
+%   isClosed := true
+%   on open => opened
+% \end{lstlisting}
+% \end{minipage}
+% \begin{minipage}[t]{0.32\textwidth}
+% \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
+% state opened
+%   isClosed := false
+%   on close => closed
+% \end{lstlisting}
+% \end{minipage}
+% \end{adjustwidth}
 
@@
 
-In the following we assume that the state machine is executed by an interpreter that process an incoming stream of events, and traversing run-time pointers representing transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program.
-The structure that the interpreter operates on can be described as derived  schema from the abstract syntax scheme shown on the right of Figure~\ref{LST:statemachines}, but specialized for a particular machine, e.g., \lstinline{Doors}.
-This schema will have additional fields and associations modeling the run-time data required for execution.
-For the \lstinline{Doors} statemachine, this means adding the current state to the \lstinline{Machine} type (required for any machine) and the machine specific data fields, namely \lstinline{isClosed}.
+In the following we assume that the state machine is executed by an interpreter that processes incoming events, and traverses run-time pointers representing the transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program.
+The structure that the interpreter operates on can be described as a schema as well, but derived from the abstract syntax scheme shown on the left of Figure~\ref{FIG:statemachineUML}, but decorated with additional properties and associations modeling the run-time state required for execution.
+This is illustrated in Figure~\ref{FIG:doorsRuntime}.
+The left shows an augmented metamodel adding properties \lstinline{vars} (to model variable assigments) and \lstinline{visited}, to the classes \lstinline{Machine} and \lstinline{State}, respectively.
+
+% For the \lstinline{Doors} statemachine, this means adding the current state to the \lstinline{Machine} type (required for any machine) and the machine specific data fields, namely \lstinline{isClosed}.
 The run-time schema for \lstinline{Doors} is then the following\footnote{It is possible to represent the global variables at run-time using an untyped table or hashmap, but for the sake of schema migration complexity, the encoding as fields in classes is more instructive.}:
 
-\begin{lstlisting}[language=java,morekeywords={on},mathescape=true]
-class Machine$^{++}$
-  states: State*
-  current: State
-  isClosed: bool
-\end{lstlisting}
-
-This \textit{extended} class $\text{\lstinline{Machine}}^{++}$ also represents state machines, but this time at run time.
+% \begin{lstlisting}[language=java,morekeywords={on},mathescape=true]
+% class Machine$^{++}$
+%   states: State*
+%   current: State
+%   isClosed: bool
+% \end{lstlisting}
+
+% This \textit{extended} class $\text{\lstinline{Machine}}^{++}$ also represents state machines, but this time at run time.
 
@@
   \centering
-\includegraphics[width=0.5\textwidth]{figures/doorsmachine.pdf}
-\caption{UML object diagram of the run-time structures of the Doors statemachine (excluding code objects)}
+\begin{minipage}{0.5\textwidth}
+\includegraphics[width=\textwidth]{figures/runtime-model.pdf}
+\end{minipage}
+\hspace*{2pt}
+\vline
+\begin{minipage}{0.4\textwidth}  
+\includegraphics[width=\textwidth]{figures/doorsmachine.pdf}
+\end{minipage}
+\caption{Augmented run-time metamodel for state machines and a run-time instance of the Doors state machine during execution.}
 \label{FIG:doorsRuntime}

### 2024-05-31 paper.tex
editing live modeling

@@
 This is illustrated in Figure~\ref{FIG:doorsRuntime}.
-The left shows an augmented metamodel adding properties \lstinline{vars} (to model variable assigments) and \lstinline{visited}, to the classes \lstinline{Machine} and \lstinline{State}, respectively.
+The left shows an augmented metamodel adding properties \lstinline{vars} (to model variable assigments) and \lstinline{visited} (counting number of state visits) to the classes \lstinline{Machine} and \lstinline{State}, respectively.
 
@@
 Live data that we obtain by instantiating the state machine still contain this state machine
-definition, but also the necessary live state defined by the definition.
+definition, but also the necessary run-time state required for execution.
 
@@
 If the state machine definition changes from $D_0$ to $D_1$,
-we can synthesize new live machine state $+D_1$ conforming to new live machine type $+T_1$.
+we can synthesize a new live machine state $+D_1$ conforming to new live machine type $+T_1$.
 
@@
 This object graph is graphically depicted in Figure~\ref{FIG:doorsRuntime} as a UML object diagram, excluding the statement objects representing on-entry actions.
-As the diagram shows, the running machine has a current state (\lstinline{closed}) and the \lstinline{isClosed} field has an actual value (\lstinline{true}).
-It is instructive to note that the diagram encode both static and dynamic aspects of the state machine language.
+As the diagram shows, the running machine has a current state (\lstinline{closed}) and the \lstinline{isClosed} field has an actual value (\lstinline{true}) in the \lstinline{vars} map.
+It is instructive to note that the diagram encodes both static and dynamic aspects of the state machine language.
 Live editing the state machine then requires to \textit{patch}~\cite{SemanticDeltas} this run-time structure without shutting it down, possibly requiring migration of run-time state.
@@
 rename & state & simple\\
-add & variable & migrate class and initialize field \\
+add & variable & add entry to map and initialize \\
 remove & variable & migrate class (NB: assumes variable is unused)\\
 rename & variable & rename in class, preserving value\\
-type change & variable & migrate class and convert or reinitialize value \\
+type change & variable & migrate value in map \\
 add & transition & simple \\
@@
 
-The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{tbl:smchanges}.
+The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references to variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{tbl:smchanges}.
 The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:

### 2024-05-31 paper.tex
more tweaks

@@
 A particular challenge from the end-user perspective is to ``do minimal harm'': it is essential for fluid programming experience that the automatically triggered migrations are in a sense as near as possible to the previous application state, to not surprise or confuse the user/developer.
-One approach  considered in earlier work~\cite{RuntimeConstraint} employs the constraint solver Z3 to find a ``nearest'' run-time instance compatible with a source change.
+% One approach  considered in earlier work~\cite{RuntimeConstraint} employs the constraint solver Z3 to find a ``nearest'' run-time instance compatible with a source change.
 Nevertheless, the patching of the run-time state should be quick enough so as to not disrupt the programmer experience.
@@
 The development of live programming languages, however, is currently still out of reach for all existing language workbenches.
-The aforementioned approach using a constraint solver is an example of such a \textit{language parametric} technique, in that it operates on the (extended) abstract syntax metamodel of a language, and does not assume anything further about the language itself.
+
+The above case study is concerned with changes to the DSL program (e.g., the Doors statemachine).
+However, there is another level of schema change to be considered: what if the metamodel (cf.
+Figure~\ref{FIG:statemachineUML}) itself is edited?
+In this case both the programs (e.g., Doors) possibly need to be migrated (in case of a change that is not backwards compatible), \textit{and} the running instance, including the interpreter, since its structure is, too, governed by the abstract syntax of the language.
+
+%The aforementioned approach using a constraint solver is an example of such a \textit{language parametric} technique, in that it operates on the (extended) abstract syntax metamodel of a language, and does not assume anything further about the language itself.
 Another approach is the \textsc{Cascade} metamodeling formalism which has builtin support for run-time patching~\cite{Cascade}.
@@
 Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program.
-Typically a refactoring should be behavior preserving, but can it also preserve run-time data?
-The minimal example is the consistent variable rename rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.
+Typically, a refactoring should be behavior preserving, but can it also preserve run-time data?
+The minimal example is the consistent variable rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.
 A more complex refactoring is described in Section~\ref{sec:elm}.

### 2024-05-31 paper.tex
tweak

@@
 add & variable & add entry to map and initialize \\
-remove & variable & migrate class (NB: assumes variable is unused)\\
+remove & variable & remove from the variable map (NB: assumes variable is unused)\\
 rename & variable & rename in class, preserving value\\
@@
 In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
-\item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
+% \item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
 \item removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale.

### 2024-05-31 paper.tex
Tweaks in metamodel, conclusions, etc

@@
 
-% \definecolor{progblue}{RGB}{64,104,224}
-% \renewenvironment{leftbar}[1][\hsize]
-% {%
-%     \def\FrameCommand
-%     {%
-%         {\color{progblue}\vrule width 3pt}%
-%         \hspace{0pt}%must no space.
-%         \fboxsep=\FrameSep\colorbox{white}%
-%     }%
-%     \MakeFramed{\hsize#1\advance\hsize-\width\FrameRestore}%
-% }
-% {\endMakeFramed}
-
 \DeclareRobustCommand{\frameworkbox}[1]{\leftbar#1\endleftbar}
@@
 \author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
-\authorinfo{is TBD. Contact him at \email{jonathanmedwards@gmail.com}.}
+\authorinfo{is an independent researcher. Contact him at \email{jonathanmedwards@gmail.com}.}
 \affiliation[a]{Independent, Boston, MA, USA}
@@
 \author[b]{Tomas Petricek}%[0000-0002-7242-2208]
-\authorinfo{is TBD. Contact him at \email{tomas@tomasp.net}.}
+\authorinfo{is an Associate Professor at Charles University. Contact him at \email{tomas@tomasp.net}.}
 \affiliation[b]{Charles University, Prague, Czechia}
@@
 \author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
-\authorinfo{is TBD. Contact him at \email{storm@cwi.nl}.}
+\authorinfo{is a Senior Researcher at Centrum Wiskunde \&\ Informatica (CWI) and a Professor at the University of Groningen. Contact him at \email{storm@cwi.nl}.}
 \affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
@@
 \author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
-\authorinfo{is TBD. Contact him at \email{gklitt@gmail.com}.}
+\authorinfo{is a Senior Researcher at the independent research lab Ink \& Switch. Contact him at \email{gklitt@gmail.com}.}
 \affiliation[e]{Ink \& Switch, Planet Earth}
@@
 Many past improvements in the practice of programming worked by speeding up feedback loops,
-both in collaboration with other programmers (through collaborative development tools
+both in collaboration with other programmers (via collaborative development tools
 \cite{ProGit, goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
 programming environment (through live programming \cite{rein2018exploratory}).
-
+%
 To make programming more effective, future programming systems will need to further tighten
@@
 To conceive of such systems, we need to shift our focus to \emph{programming systems} \cite{techdims}
-where program state is persistent and co-evolves with the behavior and data of the system. This is
-already the case in collaborative applications with state persistent in a database, image-based
+where program state is persistent and co-evolves with the behavior and data of the system. This view
+is already used in collaborative applications with state persistent in a database, image-based
 systems like Smalltalk and live programming environments with hot-reloading where the program can
@@
 future of programming, there are a number of problems that need to be resolved in order for
-such systems to become a reality. A key problem is \emph{type evolution}, i.e., how to
+such systems to become a reality. A key problem is \emph{type evolution}: how to
 gracefully adapt the shape of data used in a program while preserving integrity
@@
 \cite{Cicchetti11,alanen2003} and in more low-level work on dynamic software updating
-\cite{hicks2005dynamic}.
+\cite{hicks2005dynamic}. We draw on extensive literature review, included in Appendix~\ref{sec:related}.
 
 Looking at type evolution from the perspective of programming systems can offer a new perspective that
-unfies the many different aspects of the problem partially explored in other areas. A programming
+unfies the many different aspects of the problem partially explored in other sub-fields. A programming
 system may have permanent data (as in a database) whose structure can change, requiring
@@
 framework for thinking about the relationship between schema, code, and data in a programming
-system. Second, we present five case studies that use the framework to study type evolution in a
+system. Second, we present six case studies that use the framework to study type evolution in a
 highly diverse set of real-world programming scenarios. We hope this will encourage knowledge
@@
 data, we can gain valuable insights from research on database systems.
-Although schema evolution is well-studied in databases (Section~\ref{sec:related}), past work
+Although schema evolution is well-studied in databases (Appendix~\ref{sec:related}), past work
 does not resolve all problems in our second challenge.
@@
 accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to
-$T_1$ using an automatically generated backward transformation. In other words,
+$T_1$ using an automatically generated backward transformation. That is,
 type evolution needs to be accompanied with both forward and backward projection for data
@@
 \begin{figure}[t]
-\centering
-\begin{minipage}{0.3\textwidth}
+\begin{adjustwidth}{-3em}{-3em}
+\begin{minipage}{0.4\textwidth}
 \includegraphics[width=\textwidth]{figures/metamodel.pdf}
@@
 \vline
-\begin{minipage}{0.3\textwidth}
+\begin{minipage}{0.4\textwidth}
   \includegraphics[width=\textwidth]{figures/doors-ast.pdf}
@@
 var isClosed: bool = true
+
 state closed
@@
   on open => opened
+
 state opened
+  isClosed := false
   on close => closed
@@
 \end{minipage}
-\caption{Definition of the abstract syntax state machine language as a UML metamodel and an example instance, as object diagram and as source code}
+\end{adjustwidth}
+\caption{Definition of the abstract syntax state machine language as a UML metamodel (left) and
+an example instance, as object diagram (middle) and as code (right).}
 \label{FIG:statemachineUML}
@@
 Work on language modeling offers a new perspective on the problem.
-In the context of executable domain-specific modeling languages~\cite{}, one can see a program in a domain-specific language as an instance of a metamodel (a static type) that
-defines the AST structure of the domain-specific language. The program in a domain-specific language
-then in turn determines (defines/implies/induces) a run-time schema: the run-time structures
-of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as the structure of the run-time state (the heap, slots for user data, etc.).
-Whereas the example discussed in this section is framed in terms of metamodels and models, this should be seen as an expository device: the principles and challenges apply, e.g., in more grammar-oriented settings.
-
-In the absence of live programming, the run-time \emph{code} structures do not typically
-change while using an application and the run-time \emph{state} changes constantly.
+In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
+program in a domain-specific language as an instance of a metamodel (a static type) that
+defines the AST structure of the domain-specific language.
+
+A program in a domain-specific language in turn determines (defines/implies/induces) a run-time
+schema: the run-time structures of code (e.g., classes, inheritance links, declarations, method
+definitions etc.), as well as the structure of the run-time state (the heap, slots for user data, etc.).
+Whereas the example discussed in this section is framed in terms of metamodels and models, this can
+be seen as an expository device: the principles and challenges apply in more grammar-oriented settings.
+However, the framing also suggests interesting additional challenges.
+
+In the absence of live programming, the run-time \emph{types} and \emph{code} do not change while
+running a program in a given domain-specific language and the run-time \emph{state} changes constantly.
 Live programming, however, requires reconciling changes to the program with the induced run-time structures
 of both the code and the state. This typically means that at a certain point (a quiescent point)
-during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.
-
-\subsection{Example}
-Consider the example of a simple state machine language with on-entry actions, and typed global
-variables (loosely inspired by the SML language of \citet{vanRozen19}). 
-The abstract syntax (as UML class diagram) is shown on the left of Figure~\ref{FIG:statemachineUML}. 
-The middle of the figure shows an instance: a state machine modeling the opening and closing of a door (in UML object notation). 
-The right shows a possible textual concrete syntax modeling the same state machine. 
-The state machine has one global
-boolean variable, \lstinline{isClosed}, which is flipped according to transitioning between the
-\lstinline{closed} and \lstinline{opened} states. For the sake of brevity, the abstract syntax
-model  shown on the left of the figure
-omits the details for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).
+during execution, the code structures need to be updated (hot swapped), and the state needs to be
+\textit{migrated}.
+
+Figure~\ref{FIG:statemachineUML} defines a simple state machine domain-specific language with
+on-entry actions, and typed global variables (loosely inspired by SML \cite{vanRozen19}).
+The figure shows the abstract syntax as UML class diagram (left); an instance modelling the
+opening and closing of a door in UML object notation (middle) and a possible textual concrete
+syntax modeling the same state machine (right).
+%
+The state machine has one global Boolean variable, \lstinline{isClosed}, which is flipped when
+transitioning between the \lstinline{closed} and \lstinline{opened} states. For the sake of brevity,
+the abstract syntax model shown on the left omits the details for the on-entry
+actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).
 
@@
 
-In the following we assume that the state machine is executed by an interpreter that processes incoming events, and traverses run-time pointers representing the transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program.
-The structure that the interpreter operates on can be described as a schema as well, but derived from the abstract syntax scheme shown on the left of Figure~\ref{FIG:statemachineUML}, but decorated with additional properties and associations modeling the run-time state required for execution.
-This is illustrated in Figure~\ref{FIG:doorsRuntime}.
-The left shows an augmented metamodel adding properties \lstinline{vars} (to model variable assigments) and \lstinline{visited} (counting number of state visits) to the classes \lstinline{Machine} and \lstinline{State}, respectively.
+In the following we assume that the state machine is executed by an interpreter that processes
+incoming events, traverses the transitions accordingly, and executes on-entry actions. The notion
+of the current state and the values of the global variables are the collective run-time state of
+the program. The structure that the interpreter operates on can be described by a run-time schema.
+This schema (Figure~\ref{FIG:doorsRuntime}, left) is derived from the abstract syntax
+schema (Figure~\ref{FIG:statemachineUML}, left), but is decorated with additional properties and
+associations modeling the run-time state required for execution. The augmented metamodel includes
+properties \lstinline{vars} (to model variable assigments) and \lstinline{visited} (counting
+number of state visits) to the classes \lstinline{Machine} and \lstinline{State}, respectively.
 
@@
 
-\frameworkbox{
-\begin{wrapfigure}{r}{14.5em}
+\begin{figure}[t]
 \vspace{-0.5em}
-\includegraphics[width=14.5em]{figures/arr-model.png}
-\vspace{-1.5em}
-\end{wrapfigure}
-\frameworkboxtitle{TODO}
-Here, data $D_0$ now represent the state machine definition.
-Live data that we obtain by instantiating the state machine still contain this state machine
-definition, but also the necessary run-time state required for execution.
-
-This is used to synthesize live machine state $+D_0$ conforming to live machine type $+T_0$.
-This then runs producing $+T'_0$ and so on...
-
-If the state machine definition changes from $D_0$ to $D_1$,
-we can synthesize a new live machine state $+D_1$ conforming to new live machine type $+T_1$.
-
-The challenge is to merge these changes with the currently running state machine $+T'_0$.
-We have to produce merged live machine type $+T_0 \diamond +T_1$ and also merge the
-live state changes to obtain $+D'_0 \diamond +D_1$ of the new merged live machine type.
-}
-
-\begin{figure}[t]
 \centering
@@
 \end{minipage}
-\caption{Augmented run-time metamodel for state machines and a run-time instance of the Doors state machine during execution.}
+\caption{Augmented run-time metamodel for state machines (left) and a run-time instance
+of the Doors state machine during execution (right).}
 \label{FIG:doorsRuntime}
+\vspace{-0.5em}
 \end{figure}
 
-An instance of a running state machine consists of an object graph conforming to the run-time schema of state machines, updating the \lstinline{current} state pointer in response to events.
-This object graph is graphically depicted in Figure~\ref{FIG:doorsRuntime} as a UML object diagram, excluding the statement objects representing on-entry actions.
-As the diagram shows, the running machine has a current state (\lstinline{closed}) and the \lstinline{isClosed} field has an actual value (\lstinline{true}) in the \lstinline{vars} map.
-It is instructive to note that the diagram encodes both static and dynamic aspects of the state machine language.
-Live editing the state machine then requires to \textit{patch}~\cite{SemanticDeltas} this run-time structure without shutting it down, possibly requiring migration of run-time state.
-
-\subsection{Challenge}
+An instance of a running state machine consists of an object graph conforming to the run-time
+schema of a state machine, updating the \lstinline{current} state pointer in response to events.
+This object graph is depicted as a UML object diagram in Figure~\ref{FIG:doorsRuntime} (right),
+excluding the statement objects representing on-entry actions. As the diagram shows, the running
+machine has a current state (\lstinline{closed}) and the \lstinline{isClosed} field has a
+value (\lstinline{true}) in the \lstinline{vars} map. It is instructive to note that the
+diagram encodes both static and dynamic aspects of the state machine language.
+
+\challenge{Live Editing Domain-Specific Languages}
+The challenge is editing the state machine (Figure~\ref{FIG:statemachineUML}, right)
+created in the domain-specific language. This requires to \textit{patch}~\cite{SemanticDeltas} the
+run-time structure (Figure~\ref{FIG:doorsRuntime}, right) without discarding and recreating it,
+possibly requiring migration of run-time state.
+
+The space of possible %(well-formed)\footnote{The abstract syntax schema is not expressive enough
+%to define all static invariants of the language, such as: states must by unique by name, references
+%to variables in actions must be declared, etc.})
+changes to a state machine is summarized in Table~\ref{tbl:smchanges}.
+The first column indicates the change category, the second the affected kind of object, and third
+a description of how to deal with the respective change category. In summary:
+\begin{itemize}
+\item All rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
+\item Removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care is required if the subject of removal is the current state.
+In this case, heuristic or user input is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
+% \item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
+\item Removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale.
+Strategies to deal with this situation include dropping the events, or preventing the edit when there are pending events.
+\item Changing the type of a variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion.
+\item Note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
+\end{itemize}
 
 \begin{table}[t]
-  \centering
+\begin{adjustwidth}{-2em}{-2em}
 \begin{tabular}{lll}\toprule
-\textsc{Change} & \textsc{What} & \textsc{How}\\\midrule
-add & state & simple  \\
-remove & state & structurally simple, but heuristic if state is current \\
-rename & state & simple\\
-add & variable & add entry to map and initialize \\
-remove & variable & remove from the variable map (NB: assumes variable is unused)\\
-rename & variable & rename in class, preserving value\\
-type change & variable & migrate value in map \\
-add & transition & simple \\
-remove & transition & structurally simple, but heuristic for pending events\\
-event change & transition & structurally simple, but heuristic for pending events\\
-add/remove & statement & hotswap code at quiescent point of interpreter\\
+{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
+Add/Rename & state & Simple  \\
+Remove & state & Structurally simple, but heuristic if state is current \\
+Add & variable & Add entry to map and initialize \\
+Remove & variable & Remove from the variable map (NB: assumes variable is unused)\\
+Rename & variable & Rename in class, preserving value\\
+Type change & variable & Migrate value in map \\
+Add & transition & Simple \\
+Remove & transition & Structurally simple, but heuristic for pending events\\
+Event change & transition & Structurally simple, but heuristic for pending events\\
+Add/remove & statement & Hotswap code at quiescent point of interpreter\\
 \bottomrule
 \end{tabular}
+\vspace{0.3em}
 \caption{Possible program changes and how to deal with them at run time}
 \label{tbl:smchanges}
+\end{adjustwidth}
+\vspace{-1em}
 \end{table}
 
-The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references to variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{tbl:smchanges}.
-The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:
-\begin{itemize}
-\item all rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
-\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care is required if the subject of removal is the current state.
-In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
-% \item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
-\item removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale.
-Strategies to deal with this situation include: simply dropping the events, or requiring that such edits cannot be patched at run time when there are pending events.
-\item the type change edit of variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion.
-For instance, if the type change is from boolean to integer, then true could become 1, and false could become 0.
-\item note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
-\end{itemize}
-
-\subsection{Goals}
+\frameworkbox{
+\begin{wrapfigure}{r}{14em}
+\vspace{-0.75em}
+\includegraphics[width=14.5em]{figures/arr-model.png}
+\vspace{-1.5em}
+\end{wrapfigure}
+\frameworkboxtitle{Code as Data Migrations}
+We could see this challenge as type, code and data co-evolution as in the case of the Elm
+architecture in Section~\ref{sec:elm}, but a more interesting perspective is to see the
+state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.
+
+When the state machine is instantiated, the metamodel and the definition are used to derive
+the run-time type ${}^+\!T_0$. The run-time data ${}^+\!D_0$ conforming to this type still
+contain the state machine definition, but also the necessary run-time state required for
+its execution, which will produce new versions of the run-time data such as ${}^+\!D_0'$.
+
+If the state machine definition changes from $D_0$ to $D_1$ (top right), we can synthesize
+a new run-time state machine type ${}^+\!T_1$ and new data ${}^+\!D_1$ conforming to it.
+The challenge is to merge the two run-time types instantiated from different definitions
+to produce a type ${}^+\!T_0 \diamond {}+\!T_1$, but also merge the run-time state of the
+running machine with the new instantiated state to obtain ${}^+\!D'_0 \diamond {}^+\!D_1$.
+}
+
+\subsection*{Remarks: Requirements and Extensions}
 
 The goals for this challenge are twofold: from the end-user perspective and from the language engineering perspective.
-A particular challenge from the end-user perspective is to ``do minimal harm'': it is essential for fluid programming experience that the automatically triggered migrations are in a sense as near as possible to the previous application state, to not surprise or confuse the user/developer.
+A challenge from the end-user perspective is again to ``do minimal harm'': it is essential for fluid programming experience that the automatically triggered migrations result in a state as near as possible to the previous application state, to not surprise or confuse the user/developer.
 % One approach  considered in earlier work~\cite{RuntimeConstraint} employs the constraint solver Z3 to find a ``nearest'' run-time instance compatible with a source change.
@@
 From the language engineering perspective the goal is to employ techniques, formalisms, and tools, to make the construction of such languages easier.
-The above example state machine DSL is derived from earlier work~\cite{vanRozen19}, where the authors manually implemented the run-time patch operation and concluded that even for such a simple language (simpler even than the example above) it is a complex and error-pone endeavor.
-Furthermore, the field of software language engineering studies and develops generic and reusable techniques to improve the development of DSLs and programming languages, for instance, in the context of language workbenches~\cite{ERDWEG201524}.
-The development of live programming languages, however, is currently still out of reach for all existing language workbenches.
-
-The above case study is concerned with changes to the DSL program (e.g., the Doors statemachine).
+The above example is derived from earlier work~\cite{vanRozen19}, where the authors manually implemented the run-time patch operation and concluded that even for such a small language (simpler than the example here) it is a complex error-pone endeavor.
+Furthermore, the field of software language engineering studies and develops generic and reusable techniques to improve the development of DSLs and programming languages, e.g., in the context of language workbenches~\cite{ERDWEG201524}.
+The development of live programming languages, however, is currently out of reach for all existing language workbenches.
+
+The above case study is concerned with changes to the DSL program (e.g., the Doors state machine).
 However, there is another level of schema change to be considered: what if the metamodel (cf.
@@
 
-\subsection{Extensions}
-
-While the above example is arguably simple, the problem becomes much more challenging when the programs themselves define data types, classes, records, structures etc.
-Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
+While the above example is arguably simple, the problem becomes more challenging when the programs themselves define data types, classes, records, structures etc.
+Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that the programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
 Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program.
@@
 The minimal example is the consistent variable rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.
-A more complex refactoring is described in Section~\ref{sec:elm}.
+
+\section{Conclusions}
+We believe that effective future programming systems will need to tighten the feedback loop,
+both in making programming more live and in making it more collaborative. One of the key obstacles
+to making progress in this direction is the problem of \emph{type evolution}. When the structure
+of a program evolves, code and data need to co-evolve to match the new structure.
+%
+The problem of type evolution has been explored in a wide range of sub-fields, but rarely in its
+full complexity. Moreover, different sub-fields often use different language, making knowledge
+transfer between them difficult.
+
+In this paper, we developed a conceptual framework
+that made it possible to present six case studies from a wide range of areas in a unified way.
+The areas included live programming, database schema change, model-driven development,
+spreadsheet and document-oriented programming.
+%
+Our work highlights the importance
+of the problem, provides a new uniform way of thinking about it and provides numerous specific
+challenges that can be used to evaluate the capabilities of new programming systems that aim
+to tackle the problem.
+
+We did not present new solutions to the problem, but we hope to inspired the reader to
+do so. Indeed, we look forward to novel solutions to the challenges presented in this paper and
+the resulting more live and collaborative future of programming!
 
@@
 
-\section{\textcolor{red}{[TVDS] Super rough attempt at the paragraph}}
-
-Modern day software development is collaborative. Single user programming does not exist anymore.
-Just as document editing is becoming more collaborative through online systems like pioneered by Google Docs,
-programming would benefit form online collaboration. In a sense this is live programming++: whereas traditional
-live programming systems assume a single developer editing code and see their changes immediately reflected
-in the running system, online collaborative programming is the same but for multiple programmers at the same time.
-The key challenge here, we posit, the *schema change*. We take inspiration from database migration,
-model evolution, and version control, and identify benchmark problems. This is already evident in live programming:
-editing a program, consisting of type definitions defining the structure run-time state and (transient) procedures, functions
-etc. operating on that state, requires run-time state to be migrated, after the programmer edits their program.
-Adding the dimension of multiple programs brings us into the domain local-first, and decentralized version control, where
-copies are created and edited, which eventually have to be reconciled again (e.g., through merging).
-Whereas schema evolution and its associated migration problems has seen a lot of attention in the database world,
-the model-driven world, and the bidirectional transformation (BX) communities, the techniques are still too low-level,
-too manual, and too much focused on a single user perspectives. By identifying the similarities between analogous
-problems in live, multi-user programming systems, we hope to stimulate further research on more comprehensive, automatic,
-high-level, and online solutions. So that the benefits of live programming can be enjoyed in a global and distributed setting
-involving multiple programmers at the same time.
-
-\section{\textcolor{red}{[JE] Type evolution for live programming}}
-
-Many past improvements in the practice of programming worked by speeding up feedback loops.
-Live Programming aspires to the optimum of immediate feedback.
-The problem is that modern software is built out of many layers of different technologies -- thoroughly live programming would propagate changes immediately up and down the entire stack of a running system, from the database to the code, and from the client to the server.
-We call this \textit{full-stack} live programming.
-Another problem is that modern software is built with complex collaborations involving development, testing, deployment, and suppport.
-Such workflows are themselves feedback loops -- thoroughly live programming would propagate changes between participants frictionlessly.
-We call this \textit{end-to-end} live programming.
-
-We propose a research agenda towards this greater vision of live programming.
-The starting point is database \textit{schema evolution}, which propagates database schema changes into data migrations and query rewriting.
-We call for generalized \textit{type evolution} that works immediately across the entire software stack, and automatically across the entire software workflow.
-This paper contributes a set of challenge problems to serve as targets for research on type evolution and as a basis of comparison for the results.
-
-\section{\textcolor{red}{[GL] Rough paragraph}}
-
-Most programming languages analysis treats state as ephemeral: tied to the lifetime of a running program, and reset when the program is edited.
-But when we take a broader view, most \textit{programming systems} (cite) have state that lasts beyond the lifetime of a single execution.
-This includes collaboration applications with state persisted in a database, image-based systems like Smalltalk, and live programming environments where the program can be edited without resetting all state.
-
-Although these programming systems exist in very different contexts, they all encounter similar problems related to the essential difficulty of preserving state across program edits.
-New versions of the program must correctly handle pre-existing data.
-When the data schema or type definitions change, existing data must be evolved to conform to the new schema.
-Changing data representations or types can also require changing corresponding parts of the program.
-These problems have been noticed, named, and studied in these different programming systems: ``schema evolution'' in the databases literature, ``hot reloading'' in live programming, etc.
-
-In this paper, we draw connections between these different contexts, pointing out essential shared structure in the hopes of motivating further research on this important problem and encouraging knowledge sharing across subfields.
-Our contributions are: 1) a shared framework for thinking about the relationship between schema, code, and data in a programming system, 2) Case studies of the framework in a highly diverse set of programming systems, motivated by real-world examples.
-
-\section{\textcolor{red}{[TP] Rough paragraph}}
-
-Programming is increasingly collaborative and interactive. We welcome this trend, because
-greater collaboration and interactivity make programming more effective. Future programming
-environments enable programmers to collaborate at a more fine-grained level, adopting individual
-code changes created by their collaborators as needed. They also let programmers interactively
-edit code, often during program execution.
-
-In order to support this new way of working, programming systems increasingly have to tackle the
-problem of schema evolution. If a programmer imports a change that edits a schema, code and data
-that relies on such schema needs to be adapted to match the new schema. Similarly, if a programmer
-in a live programming environment modifies their type definition, current program state needs to
-be updated to match the new structure of types. This is a well-understood and well-studied problem
-in database community, but it has received little attention in work on programming systems.
-
-The aim of this paper is to provide a conceptual framework for talking about schema evolution in
-collaborative and live programming systems. Taking inspiration from Stewart Brand's analysis of
-pace layers, we propose to think of the problem as the problem of evolution of occuring at three
-layers, the most permanent schema layer, less permanent code layer and least permanent data layer.
-A change at the most permanent layer affects the less permanent layer, which then has to be
-updated to keep correspondence between the layers.
-
-We consider four case studies of such schema evolution in collaborative and live programming
-systems. We present those throught the perspective of our unified conceptual framework. The work
-both sheds a new light at how schema change needs to be handled in programming systems and also
-provides a range of challenges that show interesting problems that future programming systems
-will need to tackle.
-
-\section{Introduction}
-
-\textcolor{red}{[JE] We should adopt the term Schema Evolution, which is favored in the research literature, defined to include schema change, data migration, and query rewriting. }
-
-Schema change won't go away.
-Changing requirements and changing code lead to changing the schema of a database.
-Schema change, also called schema migration, is the problem of migrating existing data from the old schema to the new.
-This often involves custom migration programs or specialized Domain Specific Langiuages.
-The migration must be carefully coordinated with upgrading the application code and associated artifacts that assume the new schema.
-Schema change on database servers is often delegated to Database Administrators and DevOps.
-Live programming~\cite{tanimoto90,Hancock03} and local-first software~\cite{localfirst} both move data away from the server, eliminating those jobs, but not eliminating the need to do schema change, and indeed increasing the need to do it automatically.
-Recently \citet{Cambria} spotlighted these problems and proposed an approach using lenses~\cite{Foster2007}, but otherwise there has been surprisingly little research.
-To promote further progress we offer a set of challenge problems to the live programming and local-first communities, inviting them to propose and compare solutions.
-
-Live programming seeks to erase the boundary between editing and running programs.
-In order to do so program data must be kept around while the program is being edited.
-Classic Lisp and Smalltalk systems integrated code and data into a persistent \textit{image}~\cite{Sandewall78, Goldberg80}.
-An interactive shell or \textit{Read Eval Print Loop (REPL)}~\cite{Deutsch64} is more transient than an image, but still builds up a context of data over long-lived programming sessions the loss of which disrupts the programmer's workflow.
-In all of these environments programs eventually get changed to create and expect data in a form incompatible with extant data.
-This happens whether or not the form of the data is specified explicitly in a type system or database schema, or whether it is left implicit in the behavior of the code.
-Some languages can tolerate a larger set of such changes, most notably Smalltalk which will automatically insert and delete members in existing instances when a class definition is changed~\cite[pp.252-272]{Goldberg80}.\footnote{Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a sophisticated schema change API~\cite{Gemstone}.
-Schema change won't go away.} But there is still a large class of changes that create incompatibilities with existing data, whether it is inside an image, programming environment, REPL, or an external database.
-Some live programming environments generate data with unit tests, but that only shifts the problem of schema change to adapting those tests.
-One way or another, stopping everything to manually deal with schema change contradicts the goals of live programming.
-Schema change won't go away.
-
-Local-first software faces related problems.
-Its goal is to empower users by moving code and data from the cloud to the user's own devices.
-Distributed programming techniques like Convergent Replicated Data Types (CRDTs)~\cite{Shapiro11} are used to coordinate data changes peer-to-peer.
-Unfortunately these techniques so far have not addressed schema change nor code deployment The traditional techniques of schema change used in centrally managed databases are complicated by the distributed and intermittently connected nature of local-first data.
-Schema change won't go away.
-
-In the following sections we present a series of challenge problems dealing with schema change in the context of live programming and local-first software.
-These problems are necessarily expressed using established idioms or conventions, but nevertheless we welcome solutions that translate the spirit of the problem into other contexts.
-
-~
-
-~
-
-\textcolor{red}{TODO: Need to explain that we take the programming system view - program is not something
-built centrally, but something that exists and can be modified}
-
-~
-
-\textcolor{red}{TODO: Maybe the right way to frame this is to say that the challenges are intended
-less for evaluating particular solutions but more for framing and making sense of different
-kinds of schema-change-related problems.  You can read this paper and when you face a schema-change-related
-problem, think about it in terms of our dimensions. Maybe compare it to our challenges.
-We do not necessarilly expect that people will directly solve our challenges though. (Although
-some might and that would be nice..)}
-
-~
-
-\textcolor{red}{TODO: I also guess this is more about just ``change'' than ``schema change''
-which is one particular (most challenging) instance - i.e., change at the most permanent
-pace layer.}
-
 \end{document}

### 2024-05-31 paper.tex
Spellecheck

@@
 % Approach: What was done that unveiled new knowledge?
-In this paper, we develop a unfied conceptual framework for analysing the problem of type evolution.
+In this paper, we develop a unified conceptual framework for analysing the problem of type evolution.
 We use the framework to study type evolution in five diverse case studies.
@@
 Looking at type evolution from the perspective of programming systems can offer a new perspective that
-unfies the many different aspects of the problem partially explored in other sub-fields. A programming
+unifies the many different aspects of the problem partially explored in other sub-fields. A programming
 system may have permanent data (as in a database) whose structure can change, requiring
@@
 Modifying \texttt{State} ranges from relatively simple problems (e.g., adding a new field with a
-default value) to challenging case. In particular, if the programer refactors \texttt{State} to a
+default value) to challenging case. In particular, if the programmer refactors \texttt{State} to a
 semantically equivalent type, the programming system should, in principle, be able to automatically
@@
 one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
-operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
+operation, because it introduces a new \emph{entity}, a referencable holder of mutable attributes.
 The operation is needed when we realize that some attributes of one type of entity should belong to
@@
 change in requirements that triggered an evolution might change back again.
-Invertibility features in other work on schema evolution and \emph{bidirectional transformations}~\cite{Foster2007, herrmann17, chillon22}.
+Invertability features in other work on schema evolution and \emph{bidirectional transformations}~\cite{Foster2007, herrmann17, chillon22}.
 The inverse of \emph{Extract Entity} is \emph{Absorb Entity}. It raises the question of how
@@
 
-The challenge focuses on collaborative conference planning, illustred in Figure~\ref{fig:conf}
+The challenge focuses on collaborative conference planning, illustrated in Figure~\ref{fig:conf}
 and inspired by a typical use of Notion~\cite{notion}. A number of co-organizers are planning
@@
 The ideas in this section also apply to work on computational scientific notebooks
-that combine code and data such as Jupyter \cite{Kluyver2016}. We hope to contribute to ongoinig
+that combine code and data such as Jupyter \cite{Kluyver2016}. We hope to contribute to ongoing
 work on advancing the format as done by Nextjournal \cite{Nextjournal21}, Nota~\cite{Crichton2021}
@@
 apply to live programming. If formulas in the document are evaluated on-the-fly, the system
-needs to identify edits that invalide some of the values (by changing the data that a formula
+needs to identify edits that invalidate some of the values (by changing the data that a formula
 depends on, or by changing the equation). It should then automatically re-evaluate or invalidate
@@
 In the diagram on the right, we only show data and type, ignoring the code added by the
-accounting department. We start with data $D_1$ of type~$T_1$. The accounting deparment
+accounting department. We start with data $D_1$ of type~$T_1$. The accounting department
 evolves the type (via Extract Entity), obtaining data $D_2$ of type $T_2$ using an automatically
-generated forward transformation. It further modifies the data, obtiaing $D_3$. When the
+generated forward transformation. It further modifies the data, obtaining $D_3$. When the
 accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to

### 2024-05-31 paper.tex
Anonymous submission

@@
 
-\author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
-\authorinfo{is an independent researcher. Contact him at \email{jonathanmedwards@gmail.com}.}
-\affiliation[a]{Independent, Boston, MA, USA}
-
-\author[b]{Tomas Petricek}%[0000-0002-7242-2208]
-\authorinfo{is an Associate Professor at Charles University. Contact him at \email{tomas@tomasp.net}.}
-\affiliation[b]{Charles University, Prague, Czechia}
-
-\author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
-\authorinfo{is a Senior Researcher at Centrum Wiskunde \&\ Informatica (CWI) and a Professor at the University of Groningen. Contact him at \email{storm@cwi.nl}.}
-\affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
-\affiliation[d]{University of Groningen, Groningen, Netherlands}
-
-\author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
-\authorinfo{is a Senior Researcher at the independent research lab Ink \& Switch. Contact him at \email{gklitt@gmail.com}.}
-\affiliation[e]{Ink \& Switch, Planet Earth}
-
-\authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}
+\author{Anonymous}
+%\author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
+%\authorinfo{is an independent researcher. Contact him at \email{jonathanmedwards@gmail.com}.}
+%\affiliation[a]{Independent, Boston, MA, USA}
+%
+%\author[b]{Tomas Petricek}%[0000-0002-7242-2208]
+%\authorinfo{is an Associate Professor at Charles University. Contact him at %\email{tomas@tomasp.net}.}
+%\affiliation[b]{Charles University, Prague, Czechia}
+%
+%\author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
+%\authorinfo{is a Senior Researcher at Centrum Wiskunde \&\ Informatica (CWI) and a Professor at the %University of Groningen. Contact him at \email{storm@cwi.nl}.}
+%\affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
+%\affiliation[d]{University of Groningen, Groningen, Netherlands}
+%
+%\author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
+%\authorinfo{is a Senior Researcher at the independent research lab Ink \& Switch. Contact him at %\email{gklitt@gmail.com}.}
+%\affiliation[e]{Ink \& Switch, Planet Earth}
+%
+%\authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}
 

### 2024-06-01 paper.tex
\affiliation is required

--- 2024-05-31 paper.tex
@@
 
-\author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
-\authorinfo{is an independent researcher. Contact him at \email{jonathanmedwards@gmail.com}.}
-\affiliation[a]{Independent, Boston, MA, USA}
-
-\author[b]{Tomas Petricek}%[0000-0002-7242-2208]
-\authorinfo{is an Associate Professor at Charles University. Contact him at \email{tomas@tomasp.net}.}
-\affiliation[b]{Charles University, Prague, Czechia}
-
-\author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
-\authorinfo{is a Senior Researcher at Centrum Wiskunde \&\ Informatica (CWI) and a Professor at the University of Groningen. Contact him at \email{storm@cwi.nl}.}
-\affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
-\affiliation[d]{University of Groningen, Groningen, Netherlands}
-
-\author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
-\authorinfo{is a Senior Researcher at the independent research lab Ink \& Switch. Contact him at \email{gklitt@gmail.com}.}
-\affiliation[e]{Ink \& Switch, Planet Earth}
-
-\authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}
+\author{Anonymous}
+\affiliation{blinded}
+% \author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
+% \authorinfo{is an independent researcher. Contact him at \email{jonathanmedwards@gmail.com}.}
+% \affiliation[a]{Independent, Boston, MA, USA}
+
+% \author[b]{Tomas Petricek}%[0000-0002-7242-2208]
+% \authorinfo{is an Associate Professor at Charles University. Contact him at \email{tomas@tomasp.net}.}
+% \affiliation[b]{Charles University, Prague, Czechia}
+
+% \author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
+% \authorinfo{is a Senior Researcher at Centrum Wiskunde \&\ Informatica (CWI) and a Professor at the University of Groningen. Contact him at \email{storm@cwi.nl}.}
+% \affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
+% \affiliation[d]{University of Groningen, Groningen, Netherlands}
+
+% \author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
+% \authorinfo{is a Senior Researcher at the independent research lab Ink \& Switch. Contact him at \email{gklitt@gmail.com}.}
+% \affiliation[e]{Ink \& Switch, Planet Earth}
+
+% \authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}
 

### 2024-08-22 paper.tex
Rewrote introduction

@@
 
-\title{Type Evolution in Interactive Programming Systems}
+\title{Schema Evolution in Interactive Programming Systems}
 
@@
 
-\keywords{Type Evolution, Schema Evolution, Live Programming, Local-first Programming}
+\keywords{Type Evolution, Schema Evolution, Live Programming, Local-first Programming, Version Control, Hot Reloading}
 
@@
 
+\textcolor{red}{JE: This structure is offputting. We should write a proper abstract. }
+
 % Context: What is the broad context of the work? What is the importance of the general research area?
-Many recent programming improvements have been enabled by shorter feedback loops. To further
-advance the state of programming, we need programming systems that enable rapid collaboration
-and make programming more live across the entire development stack. To do this, programming
-systems need to preserve state during development and interaction with the programmer.
+
+\noindent\textbf{Context:} Many recent improvements to programming have come from shortening feedback loops, for example in Live Programming and Distributed Version Control systems.
+To continue progress in this direction we need to extend Live Programming across the entire development stack, and to extend collaboration beyond compile time.
 
 % Inquiry: What problem or question does the paper address? How has this problem or question been addressed?
-The key obstacle to live collaborative programming systems is the problem of \emph{type evolution}.
-When programmer changes the structure of a type, programming system needs to co-evolve
+\noindent\textbf{Inquiry:} A key obstacle to live and collaborative programming systems is the problem of \emph{schema evolution}.
+When a programmer changes the structure of a datatype, a programming system needs to co-evolve
 existing live data and adapt code that implements the program behaviour to match the new structure.
-This problem has been studied in isolation as schema evolution in databases, hot code reloading
-in live programming and dynamic software updating, but we lack a unified comprehensive perspective.
+This problem has been studied in isolation as schema evolution in databases, but also recurs in hot code reloading in live programming and dynamic software updating.
+We connect these independent efforts in an interdisciplinary inquiry into the problem of schema change through all the layers of the development stack and across all the time-scales of programming.
+%Such schema evolutions must occur immediately during live programming, but also neeed to be conveyed through long-term unplanned collaborations.
 
 % Approach: What was done that unveiled new knowledge?
-In this paper, we develop a unified conceptual framework for analysing the problem of type evolution.
-We use the framework to study type evolution in five diverse case studies.
-%
+\noindent\textbf{Approach:} We distilled a set of case studies from personal experience, the research literature, and common industry practices.
+These cases reveal schema change problems recurring in different contexts in the developer stack and at different time-scales of collaboration, not just in database production upgrades.
+
 % Knowledge: What new facts were uncovered? What new capabilities are enabled by the work?
-By looking at the case studies through a unified perspective, we hope to inspire new thinking
-about the problem of type evolution and also encourage knowledge transfer across sub-fields.
+\noindent\textbf{Knowledge:} The problem of schema change is deeper, more widespread, and more interdisciplinary than commonly appreciated.
+It is a key obstacle to shortening feedback loops in diverse contexts of software development.
 
 % Grounding: What argument, feasibility proof, artifacts, or results and evaluation support this work?
-Our work takes the form of a sequence of challenge problems drawn from real-world programming
-scenario in diverse field such as live programming, database programming and model-driven
+\noindent\textbf{Grounding:} Our work takes the form of a suite of challenge problems drawn from real-world programming
+scenarios in diverse field such as live programming, database programming and model-driven
 development.
 %
+
 % Importance: Why does this work matter?
-Tackling the problem of type evolution in a unified way, enabled by our new conceptual framework,
-will make it possible to design and build a future generation of effective, live and collaborative
-programming systems.
+\noindent\textbf{Importance:} Schema change is a key problem for universal live programming and frictionless collaborative development.
+Our suite of challenge problems serve to ground further research and provide a basis for evaluation and comparison of results.
 
@@
 
-Many past improvements in the practice of programming worked by speeding up feedback loops,
-both in collaboration with other programmers (via collaborative development tools
-\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
-programming environment (through live programming \cite{rein2018exploratory}).
-%
-To make programming more effective, future programming systems will need to further tighten
-the feedback loop. They will need to enable programmers to collaborate at a fine-grained level,
-letting them adopt individual changes created by their collaborators (as in local-first software
-\cite{localfirst}). To provide live feedback, future programming systems will also need to
-transcend treating state as ephemeral and reset when the
-program is edited and rerun.
-
-To conceive of such systems, we need to shift our focus to \emph{programming systems} \cite{techdims}
-where program state is persistent and co-evolves with the behavior and data of the system. This view
-is already used in collaborative applications with state persistent in a database, image-based
-systems like Smalltalk and live programming environments with hot-reloading where the program can
-be edited without resetting all state.
-
-While we are firmly convinced that more collaborative and live programming systems are the
-future of programming, there are a number of problems that need to be resolved in order for
-such systems to become a reality. A key problem is \emph{type evolution}: how to
-gracefully adapt the shape of data used in a program while preserving integrity
-constraints, existing data and the logic of code.
-
-Type evolution is not a new problem. It exists in various forms in a range of existing systems.
-The problem of schema evolution in databases is well-studied \cite{erhard06}.
-It typically considers data integrity, although co-evolution of database schema
-and code has also been studied \cite{Cleve2006,wang19}.
-In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
-although it typically focuses on updating transient state and ignores the problem of more complex
-data migrations. The problem is also studied in the bidirectional transformation (BX)
-community \cite{czarnecki2009bidirectional}, in work on model-driven development
-\cite{Cicchetti11,alanen2003} and in more low-level work on dynamic software updating
-\cite{hicks2005dynamic}. We draw on extensive literature review, included in Appendix~\ref{sec:related}.
-
-Looking at type evolution from the perspective of programming systems can offer a new perspective that
-unifies the many different aspects of the problem partially explored in other sub-fields. A programming
-system may have permanent data (as in a database) whose structure can change, requiring
-corresponding change to data and code. A programming system may have a transient state conforming
-to a type structure that may change during live programming. A programming system may also need
-to allow programmers to independently work on different versions of a program.
-
-In this paper, we draw connections between these different kinds of type evolution, pointing out
-an essential shared structure. First, we present a shared conceptual
-framework for thinking about the relationship between schema, code, and data in a programming
-system. Second, we present six case studies that use the framework to study type evolution in a
-highly diverse set of real-world programming scenarios. We hope this will encourage knowledge
-sharing across different subfields, motivate further research on this important problem and
-offer inspiring challenges that developers of future live and collaborative programming systems
-will attempt to tackle.
+Many improvements to the practice of programming are due to speeding up feedback loops.
+Replacing punched cards with interactive terminals was one of the greatest leaps of productivity in the history of programming.
+Unit tests and interactive type checking provide faster feedback on errors that modern programmers now expect.
+Research on Live Programming~\cite{rein2018exploratory} seeks to provide immediate feedback on runtime behavior while editing.
+Feedback is equally valuable in collaborative programming~\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) to convey changes between coworkers with less friction.
+However there is still much room to improve feedback loops throughout the software development enterprise.
+
+In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development.
+While Schema Change has been extensively studied in the context of databases~\cite{erhard06}, we observe that analogous problems occur in varied and less-studied contexts.
+For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't.
+Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system.
+We discuss further examples in Related Work~\ref{sec:related}.
+In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops.
+To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also types and specifications of all sorts, whether expressed in a formal language or left implicit in code.
+
+Our primary contribution is a suite of challenge problems that exemplify a diverse range of typical schema change scenarios which interrupt feedback loops.
+We believe these problems have not been completely solved in theory or practice (see Related Work~\ref{sec:related}).
+While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches.
+We propose these challenge problems to motivate and facilitate further research.
+We hope they serve as examples to ground such research and as a basis for evaluating results.
+We invite contributions of further challenge problems and look forward to discussing competing solutions.
 

### 2024-08-22 revisions.tex
Rewrote introduction

We thank the anonymous reviewers for their helpful comments. In the following we summarize the changes we have made in response to those comments.

Reviewer A

We rewrote the introduction to remove statements of opinion and to clarify that the goal of our research on schema change is to speed feedback loops in software development. Ultimately that is a value judgement, but we believe that value is important to a substantial community of researchers and practitioners.

Section 2 has been revised to remove the claim of providing a conceptial framework. Instead we define multiple dimensions of schema change that are used throughout the sequel to figuratively constrast different problems.

Reviewer A
-----------------------------------------------------------------
Not a conceptual framework

"our work highlights the importance of.." No, you state that those things are important over and over, but you never make a good argument for it. In particular, you never ever state clearly -why it is important -what is it important for: that is, something is not good or bad in isolation. Something is good or bad instrumentally in order to reach a specific objective. What is the objective that you think is facilitated by having 'better more fluid programming environments'? You leave this implicit across the whole document!!

Section 1: "future programming systems will need to further..." Why? why will they need to do it? again, you do not "need to do" something in isolation. You need to do something in order to obtain or facilitate a goal. What is the goal here? I suspect once you state the goal I would be able to show you that it may not be the case.

So, "future programming systems will need to further..." is incomplete, the goal must be added, but even when the goal is added, this statement must be clearly labelled as an OPINION. I suspect it would be a widely agreed upon opinion, but an opinion; not a fact.

"will also need to transcend.." Again, add goal and flag as OPINION

"While we are firmly convinced that... are the future of programming" So much is implicit here. Do you think this is the future in the sense that you think is good and you think it will happen? Do you think it is the future because you hate it, but you still think it is going to be unavoidable? Do you think it will happen independently of your research work, or that it will happen because of your research work? There are 10 other possible interpretations for this sentence.

Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations. Please, add numbers to those expectations, show examples of corner cases where stuff breaks and is an acceptable evil, and cases were stuff breaks on purpose (in some sense, and intended evil) To make an example: Java is of course not truly "cross-platform": if you wanted to put a formal definition to cross platform, the program should work identically in all environments. This would imply that the program should not be able to discover it's running environment. In Java I can make an 'if' at the beginning of the main and throw error on Windows. This makes Java not truly "cross-platform" in a logical sense, but this is often considered desirable.

In the context of those data transformations, there must be plenty of situations like that, where a change in behavior is not only possible or acceptable, but even desirable. However, we should have some clear formal definition of when/how/why such changes in behavior are introduced.

pg 7: "the key requirement.... do minimal harm" This is an Opinion + Industry convention.

"not commonly supported in data APIs" Define what a data API is and add a citation.

pg 10: "when applying code migrations, the system should minimize... the need for developer -written code" Again, "should" requires an explicit objective we are working toward. Plus, even after you add the objective, this still need to be labelled as an Opinion.

pg 13: Here I'm starting to wonder why you use this model at all. Once we accept that code can be changed and updated, the difference between data and code disappear. We could simply represent all data as hard coded instructions recreating such data. This may produce a much more cohesive mental model for the problem at hand. For example in this new model programs are able to 'restart' and when they restart they will execute this code 'pre-loading' the data in the state it was when the program was terminated.

Once code can be updated, data is just another kind of code. This requires a language where it is easy to declaratively describe any kind of arbitrary reachable object graphs, including circular ones.

Pg 16: "Live programming must move.... affordances" This is again an Opinion, and again 'must' can not exists alone, it needs a GOAL we are trying to obtain. You may think that such goal is really 'obvious', but it is not. What kind of programming you want to facilitate? why you want to facilitate that kind of programming? that kind of programming would in turn facilitate certain human endeavors but maybe not others, and the results may be very hard to predict and quite chaotic.

Pg 18: "no silver bullet to navigate... a good solution... tradeoffs" This assume no perverse incentives. Consider that if an operation is easier such operation will be done more often, and by sloppier and less skilled operators. Assuming the operators will be competent and will care about what they are doing is unrealistic.

Reviewer B
-----------------------------------------------------------------

[Types]

The concept of types as used in the paper was introduced too late (p. 3). And their definition (“we refer to the more fundamental layer as types, although we do not use the term in a precise technical sense”) was surprising, certainly from the point of view of the programming languages community. The “types” referred to in the paper felt closer to schemas / schema definitions (classic data types) than the more mathematical/logical foundations formulated in type systems. Maybe you can try to avoid or at least reduce this potential for confusion by using a different (better?) name. “Schema” might be a good start…

[Problems/Solutions]

I think that the submission being mainly about challenge problems and not about their solution(s) is important to (as mentioned in the abstract) allow the community explore the solution space themselves.

[Collaboration / Local-first]

I liked the discussion of both local/singular and collaborative development and the problems associated with them. And I would have liked to see those two situations discussed separately, because of their technical complexity and also the processes (people) causing them.

The placement of “local-first” felt like it didn’t help me understanding the problem. Maybe you can revisit the intent of those references. My preference would be try without them.

[Structure / Pattern Language]

While re-reading the submission, I noticed quite a similarity to a pattern language (in the sense of Christopher Alexander and/or the software patterns community). All sections covering a challenge problem present the context in which the problems occurs, describe the problem, and provide an example illustrating the problem to better relate to it. Also, they cross-reference other related challenge problems. The only obvious part that is missing (on purpose, see above) is a generative solution.

However, even though those elements are all present to a varying extent, the guiding structure of a pattern and their interconnection through a pattern language might be helpful but is missing. I strongly believe that transforming the material from a simple collection of problems into a pattern-language-like (the “like” because of the missing generative solutions) artifact would be helpful not only for the readers but the community helping provide those generative solutions and extending the (such) system of patterns…

[Literature Review]

To me, this was the weak spot of the submission. While I appreciate the links to related work, I was hoping for a description of (or reference to) the method(s) used to carry out the review. There are certainly several ways to address this. One of them might be to just provide that information. Another one might be to avoid “review” und discuss related work instead…

Reviewer C
-----------------------------------------------------------------

- Live Programming for the Elm Architecture

I agree with the presented arguments but have difficulty understanding the remarks about requirements and implementation. The authors say that the migration needs to be done automatically but do not suggest a solution to the problem. Is it a type-based program synthesis? Generation of lenses? In what scenario does it need to be done automatically? A more straightforward way (but maybe not what the authors were thinking) would be to ask the programmer to write the transformation function. So, what is the goal of the idealised system being thought over?

A reference like https://hazel.org/ should be referenced here.

- Entity Evolution in Data-centric Systems

In this case, the authors miss a classic reference to the problem of updatable views, which deal with the same category of issues when solving a different problem. Migrations here feel a lot like the back translation problem in updatable views.

Also, this is the most common type of evolution problem in the software industry, which is treated by disciplining the software process, versioning, and creating adapters. A contrast with these solutions, pointing again to the ghost system, would be very welcome. This approach is presented in section 6, but it makes sense to refer here as well.

https://www.irisa.fr/lande/lande/icse-proceedings/seams/p86.pdf

- Evolution of Computational Documents

This is a problem that existing approaches solve relatively well, and there is work using, for instance, operation-based CRDTs to solve the problem of collaborative editing and ensure eventual consistency. I think those solutions should be referred to, and their results should be compared to the goals presented in this section.

- Divergence control in Spreadsheets

The problem in this use case is broader than spreadsheets, and the section title may be misleading. Versioning schemes, like semantic versioning or language-based versioning, could solve the problem of divergence control in spreadsheets. This class of solutions could help present this case better.

https://dl.acm.org/doi/10.1145/3479394.3479416
Multiplicity Change in Data Formats (Migrations)

Using lenses to solve this problem would be a very good idea to highlight here, as in some old work of Benjamin Pierce and a system called Unison. This problem was central in transformations and sync operations in XML documents. I have no specific reference to point out, but Unison seems a good place to start. The synthesis of lenses can be done.

- Live Modelling Languages

This case study is, in my opinion, the more solid one in terms of presentation. It almost seems that this is the idealised system that the authors are thinking about. This presentation also gives away the identity of the authors (Programming is not double-blind; I am just referring to this since the authors omitted their names in the paper).

The comment about the presented example being out of reach for all existing language workbenches could be expanded to explain why. Are there any techniques extracted from the remaining case studies that would be useful here?

- Overall

the discussion and conclusions are very brief both at the end and in each case study and do not provide a clear path to solving the challenges presented in the paper. Even a speculative section pointing to combining features from different systems would be welcome to communicate the authors' vision of the future of interactive programming systems.

Meta-review
-----------------------------------------------------------------
Please revise this paper to satisfy the following suggestions:

- clarify if the paper targets a concrete tool under development, present it, and link it to the different aspects
- Make sure that all aspects are related to concrete tools, or explicitly saying that it is not possible or only in research prototypes
- Include an explicitly speculative section when tools do not exist, or the authors want to share some ideas for the future. The authors should take the more detailed changes into account as well.

### 2024-08-22 paper.tex
Related Work section

@@
 Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system.
-We discuss further examples in Related Work~\ref{sec:related}.
+We discuss further examples in \S\ref{sec:related} Related Work.
 In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops.
@@
 Our primary contribution is a suite of challenge problems that exemplify a diverse range of typical schema change scenarios which interrupt feedback loops.
-We believe these problems have not been completely solved in theory or practice (see Related Work~\ref{sec:related}).
+We believe these problems have not been completely solved in theory or practice (see \S\ref{sec:related}  Related Work).
 While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches.
@@
 data, we can gain valuable insights from research on database systems.
-Although schema evolution is well-studied in databases (Appendix~\ref{sec:related}), past work
+Although schema evolution is well-studied in databases (\S\ref{sec:related} Related Work), past work
 does not resolve all problems in our second challenge.
@@
 
-We considered live programming in the context of the Elm architecture in Section~\ref{sec:elm},
+We considered live programming in the context of the Elm architecture in \S\ref{sec:elm},
 where the challenge was to migrate run-time state (data) to a new type to keep the program running.
@@
 We could see this challenge as type, code and data co-evolution as in the case of the Elm
-architecture in Section~\ref{sec:elm}, but a more interesting perspective is to see the
+architecture in \S\ref{sec:elm}, but a more interesting perspective is to see the
 state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.
@@
 
-\section{Conclusions}
-We believe that effective future programming systems will need to tighten the feedback loop,
-both in making programming more live and in making it more collaborative. One of the key obstacles
-to making progress in this direction is the problem of \emph{type evolution}. When the structure
-of a program evolves, code and data need to co-evolve to match the new structure.
-%
-The problem of type evolution has been explored in a wide range of sub-fields, but rarely in its
-full complexity. Moreover, different sub-fields often use different language, making knowledge
-transfer between them difficult.
-
-In this paper, we developed a conceptual framework
-that made it possible to present six case studies from a wide range of areas in a unified way.
-The areas included live programming, database schema change, model-driven development,
-spreadsheet and document-oriented programming.
-%
-Our work highlights the importance
-of the problem, provides a new uniform way of thinking about it and provides numerous specific
-challenges that can be used to evaluate the capabilities of new programming systems that aim
-to tackle the problem.
-
-We did not present new solutions to the problem, but we hope to inspired the reader to
-do so. Indeed, we look forward to novel solutions to the challenges presented in this paper and
-the resulting more live and collaborative future of programming!
-
-\newpage
-\appendix
-
-\section{Literature Review}
+\section{Related Work}
 \label{sec:related}
@@
 \citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
-\citet{RuntimeConstraint} study the live modeling problem in Section \ref{sec:livemodeling}.
+\citet{RuntimeConstraint} study the live modeling problem in \S\ref{sec:livemodeling}.
 They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution.
@@
 
+\section{Conclusions}
+We believe that effective future programming systems will need to tighten the feedback loop,
+both in making programming more live and in making it more collaborative. One of the key obstacles
+to making progress in this direction is the problem of \emph{type evolution}. When the structure
+of a program evolves, code and data need to co-evolve to match the new structure.
+%
+The problem of type evolution has been explored in a wide range of sub-fields, but rarely in its
+full complexity. Moreover, different sub-fields often use different language, making knowledge
+transfer between them difficult.
+
+In this paper, we developed a conceptual framework
+that made it possible to present six case studies from a wide range of areas in a unified way.
+The areas included live programming, database schema change, model-driven development,
+spreadsheet and document-oriented programming.
+%
+Our work highlights the importance
+of the problem, provides a new uniform way of thinking about it and provides numerous specific
+challenges that can be used to evaluate the capabilities of new programming systems that aim
+to tackle the problem.
+
+We did not present new solutions to the problem, but we hope to inspired the reader to
+do so. Indeed, we look forward to novel solutions to the challenges presented in this paper and
+the resulting more live and collaborative future of programming!
+
 %\bibliographystyle{abbrv}

### 2024-08-23 paper.tex
revise section 2: Layers and Dimensions of Schema Evolution

@@
 
-\section{Type Evolution}
+\section{Layers and Dimensions of Schema Evolution}
+
 To paraphrase Stewart Brand \cite{Brand95}, ``almost no programs evolve well. But all programs
@@
 changing constantly.'' Programs evolve both while they are being created and during their operation
-in response to new needs. As with buildings, different aspects of program evolve at different pace
+in response to new needs. As with buildings, different aspects of programs evolve at different paces
 (Figure~\ref{fig:layers}). And ``because of the different rates of change of its components,
@@
 
-\subsection{Types, Data and Code}
+%\subsection{Schema, Data and Code}
+
 In both buildings and programs, making a change at surface layers (skin or stuff) is relatively easy,
 but a change at a more fundamental layer (structure) has cascading effects on layers that depend on it.
-In the case of programs, we refer to the more fundamental layer as \emph{types}, although we do not
-use the term in a precise technical sense. Types is what defines the shape of data and code.
-Types may be database schema, data types that define the shape of data, they may be explicitly
-declared in code and checked statically, as well as dynamically checked or even implicit.
-The key property of types for us is that they constrain the shape of other layers
-of a program, namely data and code. As a result, a change of program types requires a
-corresponding change in program data and code.
+In the case of programs, we refer to the more fundamental layer as \emph{schema}, although we do not
+use the term in a precise technical sense.
+Schema are what define the shape of data and code, either as database schema, data types declared explicitly in code and checked statically, dynamically checked types and specifications, and even expectations implicit in the code.
+The key property of schema for us is that they constrain the shape of other layers
+of a program, namely data and code. As a result, a change of schema requires a
+corresponding change in data and code.
 
@@
 The situation is illustrated on the right. We start with code $C_1$ and
-data $D_1$ that have a shape corresponding to type $T_1$. When the type changes from $T_1$ to $T_2$ ($\rightarrow$),
+data $D_1$ that have a shape corresponding to schema $T_1$. When the schema changes from $T_1$ to $T_2$ ($\rightarrow$),
 a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
-of \emph{type evolution} is synthesising ($\Rightarrow$) suitable corresponding transformations
+of \emph{schema evolution} is synthesising ($\Rightarrow$) suitable corresponding transformations
 that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
-to the new type $T_2$.
+to the new schema $T_2$.
 %
@@
 variants that exist independently and changes made to them may need to be synchronized. The
-situation is illustrated on the right. Here, the initial type $T_1$ is transformed by two
-users independently into types $T_2$ and $T_3$. The corresponding data is transformed accordingly,
+situation is illustrated on the right. Here, the initial schema $T_1$ is transformed by two
+users independently into schema $T_2$ and $T_3$. The corresponding data is transformed accordingly,
 producing $D_2$ and $D_3$. The challenge is now merging these two diverging transformation and
-obtaining a type $T_4$ and data $D_4$ that adopt both of the changes.
+obtaining a schema $T_4$ and data $D_4$ that adopt both of the changes.
 
@@
 
-\subsection{Conceptual Framework}
-
-We now describe a conceptual framework that lets us view a diverse set of real-world
-type evolution problems in a unified way. We distinguish between two dimensions.
-The \emph{local dimension} is concerned with how type evolution affects data and code within
-a single program. The \emph{non-local dimension} is concerned with how type evolution
+%\subsection{Conceptual Framework}
+
+\paragraph{Locality of Change.}
+We can view diverse real-world schema evolution problems from a common perspective by distinguishing between two different dimensions of schema change.
+The \emph{local dimension} is concerned with how schema evolution affects data and code within
+a single program. The \emph{non-local dimension} is concerned with how schema evolution
 propagates across different program variants that may exist independently.
@@
 \paragraph{Program Layers.}
-We consider the more permanent layer of \emph{types} and two less permanent layers of code and
-data that are structured by types. In specific case studies, the layers may map to
-different aspects of programs, some may be not present and some may be implicit.
+We consider the more permanent layer of \emph{schema} and two less permanent layers of code and
+data that are structured by those schema. In specific case studies, the layers may map to
+different aspects of programs, some may be not present, and some may be implicit.
 
 \begin{itemize}
-\item \emph{Types} -- structure program data and determine aspects of code. A change in types
-  requires a corresponding change in data and code. Types may be explicit (database schema, type definition)
-  or implicit (shape of data structures, shape of a document).
+\item \emph{Schema} -- structure program data and determine aspects of code. A change in schema
+  requires a corresponding change in data and code. Schema may be explicit (database schema, type declarations)
+  or implicit (expected shape of data structures or a document).
 \item \emph{Data} -- information stored by the program. Data may be more permanent (data in a
-  database) or less permanent (state in live programming). When types evolve, data needs to be
+  database) or less permanent (state in live programming). When scheam evolve, data needs to be
   updated or (in case of transient state) discarded.
-\item \emph{Code} -- or program logic, but not including source code that defines types.
-  When types evolve, code needs to change to consume/produce data of a correct shape.
+\item \emph{Code} -- or program logic, but not including source code that defines schema.
+  When schema evolve, code needs to change to consume/produce data of the correct shape.
 \end{itemize}
@@
 In programming systems that enable collaboration, multiple variants of a program may exist
-concurrently and their types, data and code may need to be synchronized. Although collaboration
+concurrently and their schema, data and code may need to be synchronized. Although collaboration
 is missing from some of our case studies, it adds an important dimension that is crucial for
-type evolution in future programming systems. We consider two characteristics.
+schema evolution in future programming systems. We consider two characteristics.
 
@@
   particular changes they want to adopt, but keep other aspects of their custom design.
-\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (type
-  evolution) can only originate from a particular source. In the decentralized model,
+\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (schema evolution) can only originate from a particular source. In the decentralized model,
   changes can be done on any of the multiple co-existing variants of a program.
@@
 \paragraph{Challenge Problems.}
-The conceptual framework offers a wide range of configurations for challenge problems, but
-our aim is not to enumerate all options. We use it to recast a number of
+These dimensions, layers, and variants offer a wide range of configurations for challenge problems, but
+our aim is not to enumerate all of them. We use them to recast a number of
 existing challenges in real-world programming systems as instances of the same problem of
-type evolution. The case studies show how well-known solutions or problems in one domain
+schema evolution. The case studies show how well-known solutions or problems in one domain
 map to another domain and they also provide and inspiration and a benchmark for

### 2024-08-23 revisions.tex
Related Work section
revise section 2: Layers and Dimensions of Schema Evolution

@@
+\documentclass{article}
+\usepackage{csquotes}
+\usepackage{xcolor}
+\begin{document}
+
 We thank the anonymous reviewers for their helpful comments. In the following we summarize the changes we have made in response to those comments.
 
-Reviewer A
+\section{Reviewer A}
 
@@
 
-Section 2 has been revised to remove the claim of providing a conceptial framework.
-Instead we define multiple dimensions of schema change that are used throughout the sequel to figuratively constrast different problems.
+Section 2 has been revised to remove the claim of providing a conceptual framework.
+Instead we define a descriptive framework of multiple dimensions and layers of schema change that are used to figuratively contrast the different problems.
+
+\textcolor{red}{I don't know what to do with this:}
+\begin{displayquote}
+  Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations.
+Please, add numbers to those expectations, show examples of corner cases where stuff breaks and is an acceptable evil, and cases were stuff breaks on purpose (in some sense, and intended evil) To make an example: Java is of course not truly "cross-platform": if you wanted to put a formal definition to cross platform, the program should work identically in all environments.
+This would imply that the program should not be able to discover it's running environment.
+In Java I can make an 'if' at the beginning of the main and throw error on Windows.
+This makes Java not truly "cross-platform" in a logical sense, but this is often considered desirable.
+
+  In the context of those data transformations, there must be plenty of situations like that, where a change in behavior is not only possible or acceptable, but even desirable.
+However, we should have some clear formal definition of when/how/why such changes in behavior are introduced.
+\end{displayquote}
+
+\section{Reviewer B}
+
+As suggested we have replaced `type' with `schema' throughout the paper. The diagrams still use the letter `T' but we intend to change those to `S' in the camera-ready verison.
+
+As suggested, the Literature Review was converted into a more appropriately named Related Work section.
+
+\section{Reviewer C}
+
+\end{document}
 
@@
 -----------------------------------------------------------------
-Not a conceptual framework
+I will start from the end.
+In the conclusion you state: 'we developed a conceptual framework....' I disagree.
+You informally spoke about those ideas in interesting and tough provoking ways, but this is very far from developing a conceptual framework, and if I had to interpret your work as a conceptual framework I would find it severely lacking.
+However, I do not need to do so, I think your work can stand on its own as a set of considerations of those problems from a unifying viewpoint.
+This viewpoint is interesting, but is clearly limited and is basically only applicable to layered systems.
+With imperative systems I mean system with a set of predefined datatypes that are in some sense 'superior' to the user defined ones, that are just lists/records/unions of those primitive data types.
+This is not the case for example in lambda calculus, where there are no primitive data types, or in pure OO, where everything truly is an object and you could define string and ints as user defined types using programming patterns not dissimilar to church encoding.
+
+Basically, as soon as 'true abstraction' is into place, most of your considerations stop being applicable.
 

### 2024-08-23 paper.tex
Clarify not a concrete tool

@@
 
-Our primary contribution is a suite of challenge problems that exemplify a diverse range of typical schema change scenarios which interrupt feedback loops.
+Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops.
 We believe these problems have not been completely solved in theory or practice (see \S\ref{sec:related} Related Work).
 While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches.
-We propose these challenge problems to motivate and facilitate further research.
-We hope they serve as examples to ground such research and as a basis for evaluating results.
-We invite contributions of further challenge problems and look forward to discussing competing solutions.
+None of our ongoing efforts yet address the full range of problems.
+Indeed the motivation of this paper is to provide a basis for communicating and collaborating between related efforts, not to mention motivating further research efforts.
+We hope these challange problems serve as examples to ground future research and as a basis for evaluating the results.
+We invite contributions of new challenge problems and look forward to discussing competing solutions.
 

### 2024-08-23 paper.tex
Type->schema

@@
 
-The first challenge problem that we present in terms of our conceptual framework is live programming
+The first challenge problem that we present is live programming
 of user interfaces based on the Elm architecture. In this model, a reactive web application is
@@
 \noindent
-The programming model works as follows:
+The schema involved here are statically checked Elm types. The programming model works as follows:
 
@@
 discard state and restart the application when the \texttt{State} type changes. A more effective
-live programming system would allow live updates to the structure of the two types too.
+live programming system would allow live updates to the structure of the two types.
 
@@
 \end{wrapfigure}
-\frameworkboxtitle{Primitive Type Transformations}
-In this case study, the evolving type is the \texttt{State} type. When the type evolves,
+\frameworkboxtitle{Primitive Schema Transformations}
+In this case study, the evolving schema is the \texttt{State} type. When the type evolves,
 the program logic (code) and the current live value of the type (data) needs to evolve
@@
 data, we can gain valuable insights from research on database systems.
-Although schema evolution is well-studied in databases (\S\ref{sec:related} Related Work), past work
+Although schema evolution is well-studied in databases (\S\ref{sec:related} Related Work), such past work
 does not resolve all problems in our second challenge.
@@
   \end{tabular}
-  \caption{Customers table after type evolution}
+  \caption{Customers table after schema evolution}
   \label{fig:db-customers}
@@
     \end{tabular}
-  \caption{Orders linking to Customers, after type evolution}
+  \caption{Orders linking to Customers, after schema evolution}
  \label{fig:db-orders-link}
@@
 \vspace{0.25em}
-\caption{Types and data before and after performing Extract Entity}
+\caption{Schema and data before and after performing Extract Entity}
 \label{fig:db}
@@
 \frameworkboxtitle{Equivalence in Schema Evolution}
-We described our conceptual framework in terms of types, code and data, but this case
-study shows that the same patterns appear elsewhere. In database systems, the schema
+%We described our conceptual framework in terms of types, code and data, but this case study shows that the same patterns appear elsewhere.
+In database systems, the schema
 $S_1$ changes into $S_2$ and the data $D_1$ need to transform to corresponding $D_2$.
@@
 \section{Evolution of Computational Documents}
-A different kind of type evolution is present in programming systems based on documents.
+A different kind of schema evolution is present in programming systems based on documents.
 The idea, dating back to Boxer \cite{diSessa86}, is to represent programs as documents
@@
 Jennings). The first challenge involves merging a data edit
-(Fig.~\ref{fig:conf-add}) with a type and data edit (Fig.~\ref{fig:conf-tab}); the second involves
-merging a type and data edit (Fig.~\ref{fig:conf-tab}) with edit adding code (Fig.~\ref{fig:conf-budget}).
+(Fig.~\ref{fig:conf-add}) with a schema and data edit (Fig.~\ref{fig:conf-tab}); the second involves
+merging a schema and data edit (Fig.~\ref{fig:conf-tab}) with edit adding code (Fig.~\ref{fig:conf-budget}).
 Formulas in the final version (Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
@@
 \item Jonathan adds an additional speaker to the list and sorts the list of speakers
-  alphabetically by their first name. This changes data, but not the structure (type) of
+  alphabetically by their first name. This changes data, but not the structure (schema) of
   the document (Figure~\ref{fig:conf-add}).
@@
   tracking which of the organizers should contact the speaker (Figure~\ref{fig:conf-tab}).
-  This changes the document structure (implicit type).
+  This changes the document structure (an implicit schema).
 \end{enumerate}
@@
 \frameworkboxtitle{Merging Data and Code Edits}
-In this example, the \emph{type} is implicit and represents the structure of the document.
+In this example, the \emph{schema} is implicit and represents the structure of the document.
 It involves two independent edits. In one (top right), the data $D_1$ is
-changed to $D_2'$, but the type $T_1$ remains unchanged. In the other (bottom left), the
-type $T_1$ and data $D_1$ co-evolve into $T_2$ and $D_2$ and the data is further changed
+changed to $D_2'$, but the schema $T_1$ remains unchanged. In the other (bottom left), the
+schema $T_1$ and data $D_1$ co-evolve into $T_2$ and $D_2$ and the data is further changed
 to $D_3$. The challenge is to merge the two edits and obtain jointly modified data $D_3'$
-of type $T_2$. In the next section, we will see the same pattern, but also involving
+of schema $T_2$. In the next section, we will see the same pattern, but also involving
 code edits. We assume the \emph{convergence} model in this section, but will discuss
@@
 
-The type and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
+The schema and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
 lists edits appearing in our examples (but not a complete list). A system may
@@
 The challenge extends the previous one by adding code. The change to the document structure
-(\emph{type}) thus has to co-evolve with both the \emph{data} and the \emph{code}.
+(\emph{schema}) thus has to co-evolve with both the \emph{data} and the \emph{code}.
 As shown in Figure~\ref{fig:conf-finfin}, the system needs to change the equation
@@
 If the edit is recorded as a list of high-level edits shown in Figure~\ref{tbl:docchanges}
-(wrapping element, changing their types, etc.), then each edit needs to perform corresponding
+(wrapping element, changing their schema, etc.), then each edit needs to perform corresponding
 edit to selectors in formulas in the document.
@@
 Split & Selected element(s) & Split matching elements using a rule (regular expression)\\
-Add & New item to a list & Add data without affecting document structure (type)\\
-Reorder & List items & Reorder data without affecting document structure (type)\\
+Add & New item to a list & Add data without affecting document structure (schema)\\
+Reorder & List items & Reorder data without affecting document structure (schema)\\
 \bottomrule
@@
 \vspace{0.3em}
-\caption{Changes to type and data and how the programming system should respond}
+\caption{Changes to schema and data and how the programming system should respond}
 \label{tbl:docchanges}
@@
 \emph{permanent data} (information in the document) and transient \emph{data} (computed by code).
-Whereas permanent data should be synchronized and co-evolve with the document structure (type),
+Whereas permanent data should be synchronized and co-evolve with the document structure (schema),
 transient data can be discarded.
@@
 \section{Divergence Control in Spreadsheets}
-Type evolution becomes more challenging when data, code, and types cannot all be updated
+Schema evolution becomes more challenging when data, code, and Schema cannot all be updated
 together in an atomic way. In these situations, different versions must coexist and interoperate
-with one another. For example, edits done to data of a new type must be applied to
-data of an older type, or code written against older type must run against data of a newer type.
+with one another. For example, edits done to data of a new schema must be applied to
+data of an older schema, or code written against older schema must run against data of a newer schema.
 
@@
 the accounting department, which copies the data to its own version of the spreadsheet.
-This version uses the same structure (type), but includes additional code for financial tracking.
+This version uses the same structure (schema), but includes additional code for financial tracking.
 When the orders department migrates its spreadsheet to a database and extracts out customers,
@@
 \frameworkboxtitle{Bidirectional Data Migration}
-The scenario adds the need for transferring data changes from a new type to an older type.
-In the diagram on the right, we only show data and type, ignoring the code added by the
-accounting department. We start with data $D_1$ of type~$T_1$. The accounting department
-evolves the type (via Extract Entity), obtaining data $D_2$ of type $T_2$ using an automatically
+The scenario adds the need for transferring data changes from a new schema to an older schema.
+In the diagram on the right, we only show data and schema, ignoring the code added by the
+accounting department. We start with data $D_1$ of schema~$T_1$. The accounting department
+evolves the schema (via Extract Entity), obtaining data $D_2$ of schema $T_2$ using an automatically
 generated forward transformation. It further modifies the data, obtaining $D_3$. When the
-accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to
+accounting department receives data $D_3$, it needs to migrate it back from schema $T_2$ to
 $T_1$ using an automatically generated backward transformation. That is,
-type evolution needs to be accompanied with both forward and backward projection for data
+schema evolution needs to be accompanied with both forward and backward projection for data
 (and code) migration.}
@@
 \frameworkboxtitle{Maintaining Divergent Variants}
-The extra complexity in the above scenario arises from the fact that both types and
-code are forked and evolve independently. In one branch, original types $T_1$ evolve
-into $T_2$ and original code $C_1$ into $C_3$. In another branch, original types $T_1$
+The extra complexity in the above scenario arises from the fact that both schema and
+code are forked and evolve independently. In one branch, original schema $T_1$ evolve
+into $T_2$ and original code $C_1$ into $C_3$. In another branch, original schema $T_1$
 evolve into $T'_2$ and original code $C_1$ into $C'_3$. The user then wants to merge
-the independently done changes into a new version of types and code, written as
+the independently done changes into a new version of schema and code, written as
 $T_2 \diamond T'_2$ and $C_3 \diamond C'_3$. The diverging transformations may make
@@
 
-The challenge of maintaining multiple divergent types in programming systems can
+The challenge of maintaining multiple divergent schema in programming systems can
 be well illustrated using \emph{multiplicity change}, which is a common kind of schema change.
@@
 the linked entity; some of the challenges of this approach are covered in the Extract Entity
-challenge. In this section we will instead consider a type structure with support for arrays;
+challenge. In this section we will instead consider a schema structure with support for arrays;
 in this context, we can have a multiplicity change by turning a scalar value into an array value.
 %
-A change of multiplicity in the type requires producing a transformation for both the corresponding
+A change of multiplicity in the schema requires producing a transformation for both the corresponding
 data and the corresponding code. The challenge becomes particularly difficult to handle when there
-is ongoing divergence between multiple versions of the types and data.
-
-\challenge{Multiplicity Change in Types}
+is ongoing divergence between multiple versions of the schema and data.
+
+\challenge{Multiplicity Change in Schema}
 
@@
 The diagram illustrating the scenario remains the same as in the case of divergence control
-for Extract Entity. The initial data and type $D_1, T_1$ co-evolve into $D_2, T_2$ (adding
+for Extract Entity. The initial data and schema $D_1, T_1$ co-evolve into $D_2, T_2$ (adding
 multiplicity) and a new value is then assigned to the array of assignees, resulting in $D_3$.
 The difference from the earlier case is that the backward migration (from $D_3$ to $D'_3$)
-generated for the type evolution (from $T_1$ to $T_2$) cannot be generated fully automatically.
+generated for the schema evolution (from $T_1$ to $T_2$) cannot be generated fully automatically.
 In case of Extract Entity this was in principle possible.}
@@
 
-If types, code and data can all be updated atomically together, then multiplicity change is
+If schema, code and data can all be updated atomically together, then multiplicity change is
 relatively straightforward  to handle. Existing scalar data can be trivially migrated to a
@@
 However, ongoing divergence makes multiplicity change much more difficult, and exposes
-some general challenges. Different types may expose partial information, and writers using
-those types have to operate without total knowledge of the information available in other types.
-As a result, synchronizing writes across the types requires making difficult tradeoffs, such as
+some general challenges. Different schema may expose partial information, and writers using
+those schema have to operate without total knowledge of the information available in other schema.
+As a result, synchronizing writes across the schema requires making difficult tradeoffs, such as
 choosing to either preserve or destroy hidden data not visible to the writer.
@@
 We considered live programming in the context of the Elm architecture in \S\ref{sec:elm},
-where the challenge was to migrate run-time state (data) to a new type to keep the program running.
+where the challenge was to migrate run-time state (data) to a new schema to keep the program running.
 Work on language modeling offers a new perspective on the problem.
 In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
-program in a domain-specific language as an instance of a metamodel (a static type) that
+program in a domain-specific language as an instance of a metamodel (a static type AKA schema) that
 defines the AST structure of the domain-specific language.
@@
 
-In the absence of live programming, the run-time \emph{types} and \emph{code} do not change while
+In the absence of live programming, the run-time \emph{schema} and \emph{code} do not change while
 running a program in a given domain-specific language and the run-time \emph{state} changes constantly.
@@
 \frameworkboxtitle{Code as Data Migrations}
-We could see this challenge as type, code and data co-evolution as in the case of the Elm
+We could see this challenge as schema, code and data co-evolution as in the case of the Elm
 architecture in \S\ref{sec:elm}, but a more interesting perspective is to see the
@@
 There has been extensive research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types.
-But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system.
+But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action and feedback in an interactive programming system.
 Nevertheless it is important to understand these batch techniques to envision more interactive ones.
@@
 both in making programming more live and in making it more collaborative. One of the key obstacles
-to making progress in this direction is the problem of \emph{type evolution}. When the structure
+to making progress in this direction is the problem of \emph{schema evolution}. When the structure
 of a program evolves, code and data need to co-evolve to match the new structure.
 %
-The problem of type evolution has been explored in a wide range of sub-fields, but rarely in its
+The problem of schema evolution has been explored in a wide range of sub-fields, but rarely in its
 full complexity. Moreover, different sub-fields often use different language, making knowledge

### 2024-08-24 paper.tex
Reviewer A nitpicks

@@
 
-Many improvements to the practice of programming are due to speeding up feedback loops.
+Many improvements to the practice of programming have come from speeding up feedback loops.
 Replacing punched cards with interactive terminals was one of the greatest leaps of productivity in the history of programming.
@@
 Feedback is equally valuable in collaborative programming~\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) to convey changes between coworkers with less friction.
-However there is still much room to improve feedback loops throughout the software development enterprise.
+However there is still much room to improve feedback loops throughout software development.
 
 In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development.
-While Schema Change has been extensively studied in the context of databases~\cite{erhard06}, we observe that analogous problems occur in varied and less-studied contexts.
+Schema Change has been extensively studied in the context of databases where it necesitates manual and error-prone data migrations.~\cite{erhard06} We observe analogous problems occuring in varied and less-studied contexts.
 For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't.
@@
 In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops.
-To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also types and specifications of all sorts, whether expressed in a formal language or left implicit in code.
+To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.
 
@@
 None of our ongoing efforts yet address the full range of problems.
-Indeed the motivation of this paper is to provide a basis for communicating and collaborating between related efforts, not to mention motivating further research efforts.
-We hope these challange problems serve as examples to ground future research and as a basis for evaluating the results.
+Indeed the motivation of this paper is to provide a basis for communicating and collaborating between such related efforts, as well as to facilitate further research efforts.
+We hope these challange problems serve as examples to ground future research and as a basis for evaluating its results.
 We invite contributions of new challenge problems and look forward to discussing competing solutions.
@@
 \paragraph{Challenge Problems.}
-These dimensions, layers, and variants offer a wide range of configurations for challenge problems, but
-our aim is not to enumerate all of them. We use them to recast a number of
+These dimensions, layers, and variants generate a wide range of configurations for challenge problems, but
+our aim is not to cover all of them. Instead we use them to recast a number of
 existing challenges in real-world programming systems as instances of the same problem of
@@
 \subsection*{Remarks: Requirements and Implementation}
-The key requirement from the live programming perspective is to ``do minimal harm''.
+The key requirement from the live programming perspective is to ``do minimal harm'' to the experience of liveness.
 The migration of \emph{data} needs to be done automatically and the system should strive
@@
 and fix any references from orders. We call this operation \emph{Merge Entities} -- note that
-it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.
+it is not a schema change but a data change, yet nevertheless not commonly supported as an operaton in database APIs or DMLs.
 
@@
 view and execute the original query over the view. When applying code migrations,
-the system should minimize the need for users to manaually intervene. It should also
+the system should minimize the need for users to manually intervene, which interrupts feedback from the migration. It should also
 minimize the possibilities that developer-written code is subtly incorrect in edge cases,
-preferably by reducing the need for developer-written code.
+preferably by reducing the need for developer-written code in the first place.
 
@@
 
-This challenge invites a change of perspective in both live programming and local-first software.
-Live programming must move from being a solitary activity to a collaborative workflow, and one
-where the collaboration is not just on editing source code but on live integrated code and data,
+This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to docuuments, replacing the hellscape of manual copy \& paste.
+But meeting this challenge may require a basic change of perspective in both live programming and local-first software.
+Live programming must move from being a solitary activity to a collaborative workflow, and further where the collaboration is not just on editing source code but on live integrated code and data,
 as in the Smalltalk/Lisp images of old.
@@
 to also interactively managing long-term divergence. That implies either application state
-is being directly exposed to the user, or the application code is using an API to present
-version control affordances. We encourage both communities to think outside the box of Git,
+is being directly exposed to the user, or the application itself offers
+version control affordances to the user. We encourage both communities to think outside the box of Git,
 which has proven to baffle not only end-users but also a substantial fraction of developers.
+The Divergence challenge may be as much a technical challenge as a conceptual challenge, asking us to reconsider the entrenched boundaries between technologies in the stack.
 
@@
 disambiguate their intent for a given write.
+Whatever the solution, it must be easy to learn and simple to use otherwise it won't be used in practice.
 

### 2024-08-24 revisions.tex
Clarify not a concrete tool
Reviewer A nitpicks

@@
 
+\section{Summary review}
+
+We have clarified that this paper is not about a tool under development.
+Some of the problems are extracted from our different research projects but none of these projects span the rull range of problems.
+
+\textrcolor{red}{explicitly speculative section?}
+
 \section{Reviewer A}
 
-We rewrote the introduction to remove statements of opinion and to clarify that the goal of our research on schema change is to speed feedback loops in software development.
+We rewrote the introduction to clarify that the goal of our research on schema change is to speed feedback loops in software development.
 Ultimately that is a value judgement, but we believe that value is important to a substantial community of researchers and practitioners.
+We have qualified statements of opinion throughout the paper.
 
@@
 
+We clarified that the goal of Divergence in End-User Workflows is to create a new kind of feedback channel for end users rather than speeding up existing channels.
+
+In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of oferring solutions simple enough to be actually used.
+
 \section{Reviewer B}
@@
 \end{document}
-
-Reviewer A
------------------------------------------------------------------
-I will start from the end.
-In the conclusion you state: 'we developed a conceptual framework....' I disagree.
-You informally spoke about those ideas in interesting and tough provoking ways, but this is very far from developing a conceptual framework, and if I had to interpret your work as a conceptual framework I would find it severely lacking.
-However, I do not need to do so, I think your work can stand on its own as a set of considerations of those problems from a unifying viewpoint.
-This viewpoint is interesting, but is clearly limited and is basically only applicable to layered systems.
-With imperative systems I mean system with a set of predefined datatypes that are in some sense 'superior' to the user defined ones, that are just lists/records/unions of those primitive data types.
-This is not the case for example in lambda calculus, where there are no primitive data types, or in pure OO, where everything truly is an object and you could define string and ints as user defined types using programming patterns not dissimilar to church encoding.
-
-Basically, as soon as 'true abstraction' is into place, most of your considerations stop being applicable.
-
-"our work highlights the importance of.." No, you state that those things are important over and over, but you never make a good argument for it.
-In particular, you never ever state clearly -why it is important -what is it important for: that is, something is not good or bad in isolation.
-Something is good or bad instrumentally in order to reach a specific objective.
-What is the objective that you think is facilitated by having 'better more fluid programming environments'?
-You leave this implicit across the whole document!!
-
-Section 1: "future programming systems will need to further..." Why?
-why will they need to do it?
-again, you do not "need to do" something in isolation.
-You need to do something in order to obtain or facilitate a goal.
-What is the goal here?
-I suspect once you state the goal I would be able to show you that it may not be the case.
-
-So, "future programming systems will need to further..." is incomplete, the goal must be added, but even when the goal is added, this statement must be clearly labelled as an OPINION.
-I suspect it would be a widely agreed upon opinion, but an opinion; not a fact.
-
-"will also need to transcend.." Again, add goal and flag as OPINION
-
-"While we are firmly convinced that...
-are the future of programming" So much is implicit here.
-Do you think this is the future in the sense that you think is good and you think it will happen?
-Do you think it is the future because you hate it, but you still think it is going to be unavoidable?
-Do you think it will happen independently of your research work, or that it will happen because of your research work?
-There are 10 other possible interpretations for this sentence.
-
-Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations.
-Please, add numbers to those expectations, show examples of corner cases where stuff breaks and is an acceptable evil, and cases were stuff breaks on purpose (in some sense, and intended evil) To make an example: Java is of course not truly "cross-platform": if you wanted to put a formal definition to cross platform, the program should work identically in all environments.
-This would imply that the program should not be able to discover it's running environment.
-In Java I can make an 'if' at the beginning of the main and throw error on Windows.
-This makes Java not truly "cross-platform" in a logical sense, but this is often considered desirable.
-
-In the context of those data transformations, there must be plenty of situations like that, where a change in behavior is not only possible or acceptable, but even desirable.
-However, we should have some clear formal definition of when/how/why such changes in behavior are introduced.
-
-pg 7: "the key requirement.... do minimal harm" This is an Opinion + Industry convention.
-
-"not commonly supported in data APIs" Define what a data API is and add a citation.
-
-pg 10: "when applying code migrations, the system should minimize...
-the need for developer -written code" Again, "should" requires an explicit objective we are working toward.
-Plus, even after you add the objective, this still need to be labelled as an Opinion.
-
-pg 13: Here I'm starting to wonder why you use this model at all.
-Once we accept that code can be changed and updated, the difference between data and code disappear.
-We could simply represent all data as hard coded instructions recreating such data.
-This may produce a much more cohesive mental model for the problem at hand.
-For example in this new model programs are able to 'restart' and when they restart they will execute this code 'pre-loading' the data in the state it was when the program was terminated.
-
-Once code can be updated, data is just another kind of code. This requires a language where it is easy to declaratively describe any kind of arbitrary reachable object graphs, including circular ones.
-
-Pg 16: "Live programming must move....
-affordances" This is again an Opinion, and again 'must' can not exists alone, it needs a GOAL we are trying to obtain.
-You may think that such goal is really 'obvious', but it is not.
-What kind of programming you want to facilitate?
-why you want to facilitate that kind of programming?
-that kind of programming would in turn facilitate certain human endeavors but maybe not others, and the results may be very hard to predict and quite chaotic.
-
-Pg 18: "no silver bullet to navigate...
-a good solution...
-tradeoffs" This assume no perverse incentives.
-Consider that if an operation is easier such operation will be done more often, and by sloppier and less skilled operators.
-Assuming the operators will be competent and will care about what they are doing is unrealistic.
 

### 2024-08-24 paper.tex
Reviewer B nitpicks

@@
 
-\keywords{Type Evolution, Schema Evolution, Live Programming, Local-first Programming, Version Control, Hot Reloading}
+\keywords{Type Evolution, Schema Evolution, Live Programming, Local-first Software, Version Control, Hot Reloading, Image-based Programming}
 
@@
 Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops.
-We believe these problems have not been completely solved in theory or practice (see \S\ref{sec:related} Related Work).
+We believe these problems have not been completely solved in theory or practice, which we discuss further in \S\ref{sec:related} Related Work.
 While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches.
@@
 These dimensions, layers, and variants generate a wide range of configurations for challenge problems, but
-our aim is not to cover all of them. Instead we use them to recast a number of
+our aim is not to cover all of them. Instead we use this perspective to recast a number of
 existing challenges in real-world programming systems as instances of the same problem of

### 2024-08-24 revisions.tex
Reviewer B nitpicks

@@
 
-We have clarified that this paper is not about a tool under development.
+In the introduction we clarify that this paper is not about a tool under development.
 Some of the problems are extracted from our different research projects but none of these projects span the rull range of problems.
 
-\textrcolor{red}{explicitly speculative section?}
+We have clarified in the introduction that we avoid speculating about possible solutions to the challenge problems so as to not bias them towards our preferred approaches (and to avoid the issues where we disagree).
+A primary goal of this paper is to establish a neutral common ground where researchers can meet to discuss their alternative approaches.
+\textcolor{red}{Or do we want to add a speculative section?}
 
@@
 
-As suggested we have replaced `type' with `schema' throughout the paper. The diagrams still use the letter `T' but we intend to change those to `S' in the camera-ready verison.
+As suggested we have replaced `type' with `schema' throughout the paper. Some diagrams still use the letter `T' for type but we intend to change those to `S' in the camera-ready version.
 
-As suggested, the Literature Review was converted into a more appropriately named Related Work section.
-
-\section{Reviewer C}
-
-\end{document}
-
-Reviewer B
------------------------------------------------------------------
-
-[Types]
-
-The concept of types as used in the paper was introduced too late (p.
-3).
-And their definition (“we refer to the more fundamental layer as types, although we do not use the term in a precise technical sense”) was surprising, certainly from the point of view of the programming languages community.
-The “types” referred to in the paper felt closer to schemas / schema definitions (classic data types) than the more mathematical/logical foundations formulated in type systems.
-Maybe you can try to avoid or at least reduce this potential for confusion by using a different (better?) name.
-“Schema” might be a good start…
-
-[Problems/Solutions]
-
-I think that the submission being mainly about challenge problems and not about their solution(s) is important to (as mentioned in the abstract) allow the community explore the solution space themselves.
-
-[Collaboration / Local-first]
-
+\textcolor{red}{I don't know what to do with this:}
+\begin{displayquote}
 I liked the discussion of both local/singular and collaborative development and the problems associated with them.
@@
 The placement of “local-first” felt like it didn’t help me understanding the problem. Maybe you can revisit the intent of those references. My preference would be try without them.
+\end{displayquote}
 
-[Structure / Pattern Language]
-
+\textcolor{red}{I don't know what to do with this:}
+\begin{displayquote}
   While re-reading the submission, I noticed quite a similarity to a pattern language (in the sense of Christopher Alexander and/or the software patterns community).
@@
 I strongly believe that transforming the material from a simple collection of problems into a pattern-language-like (the “like” because of the missing generative solutions) artifact would be helpful not only for the readers but the community helping provide those generative solutions and extending the (such) system of patterns…
+\end{displayquote}
 
-[Literature Review]
+As suggested, the Literature Review was converted into a more appropriately named Related Work section.
 
-To me, this was the weak spot of the submission.
-While I appreciate the links to related work, I was hoping for a description of (or reference to) the method(s) used to carry out the review.
-There are certainly several ways to address this.
-One of them might be to just provide that information.
-Another one might be to avoid “review” und discuss related work instead…
+\section{Reviewer C}
+
+\end{document}
 
@@
 Even a speculative section pointing to combining features from different systems would be welcome to communicate the authors' vision of the future of interactive programming systems.
-
-Meta-review
------------------------------------------------------------------
-Please revise this paper to satisfy the following suggestions:
-
-- clarify if the paper targets a concrete tool under development, present it, and link it to the different aspects
-- Make sure that all aspects are related to concrete tools, or explicitly saying that it is not possible or only in research prototypes
-- Include an explicitly speculative section when tools do not exist, or the authors want to share some ideas for the future. The authors should take the more detailed changes into account as well.

### 2024-08-24 paper.tex
Reviewer C nitpicks

@@
 We connect these independent efforts in an interdisciplinary inquiry into the problem of schema change through all the layers of the development stack and across all the time-scales of programming.
-%Such schema evolutions must occur immediately during live programming, but also neeed to be conveyed through long-term unplanned collaborations.
+%Such schema evolutions must occur immediately during live programming, but also need to be conveyed through long-term unplanned collaborations.
 
@@
 the Extract Entity challenge discussed below. Finally, a programming system tackling this
-challenge may rely on structure editing so that the system has access to a high-level logical
+challenge may rely on structure editing~\cite{Teitelbaum81} so that the system has access to a high-level logical
 description of the edits performed by the user.
@@
 the document structure and data) and the order in which the edits are applied should also not
-matter. This may not always be possible and the system may ask the user for resolution or,
+matter.
+One approach is to use CRDTs~\cite{Shapiro11} which guarantee order-independent convergence over a restricted set of edit operations.
+This may not always be possible and the system may ask the user for resolution or,
 possibly, use a default behavior specified by the user.
@@
 
-\section{Divergence Control in Spreadsheets}
-Schema evolution becomes more challenging when data, code, and Schema cannot all be updated
+\section{Divergence Control}
+
+Schema evolution becomes more challenging when data, code, and schema cannot all be updated
 together in an atomic way. In these situations, different versions must coexist and interoperate
@@
 Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code.
-\citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
+Schema evolution is related to the classical view-update problem~\cite{Bancilhon81} which was approached from a language perspective by lenses~\cite{Foster2007}.
+Generalizing from lenses \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
 SMOs can also rewrite queries to accomodate schema changes.
@@
 \paragraph{Live programming.}
-Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (primarily solo) programming.
+Live programming~\cite{tanimoto90, rein2018exploratory, HazelnutLive19} seeks to speed up feedback in (primarily solo) programming.
 % \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}

### 2024-08-24 revisions.tex
Reviewer C nitpicks

@@
 In the introduction we clarify that this paper is not about a tool under development.
-Some of the problems are extracted from our different research projects but none of these projects span the rull range of problems.
+Some of the problems are extracted from our different research projects but none of these projects span the full range of problems.
 
@@
 
-In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of oferring solutions simple enough to be actually used.
+In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of offering solutions simple enough to be actually used.
 
@@
 
+in \S\emph{Remarks: Requirements and Implementation} we clarify that the goal is to adapt to schema changes automatically to avoid interrupting live programming.
+
+We added to Related Work the history of how the view-update problem led to lenses which led to DELs.
+
+We cited Hazel in Related Work.
+
+We now mention the possibility of using CRDTs for evolving computational documents.
+
+Divergence Control for Spreadsheets has been corrected to just Divergence Control.
+
+As mentioned above, we have added an explanation in the introduction of why we maintain neutrality by avoiding speculation on our preferred solutions.
+
 \end{document}
-
-Reviewer C
------------------------------------------------------------------
-
-- Live Programming for the Elm Architecture
-
-I agree with the presented arguments but have difficulty understanding the remarks about requirements and implementation.
-The authors say that the migration needs to be done automatically but do not suggest a solution to the problem.
-Is it a type-based program synthesis?
-Generation of lenses?
-In what scenario does it need to be done automatically?
-A more straightforward way (but maybe not what the authors were thinking) would be to ask the programmer to write the transformation function.
-So, what is the goal of the idealised system being thought over?
-
-A reference like https://hazel.org/ should be referenced here.
-
-- Entity Evolution in Data-centric Systems
-
-In this case, the authors miss a classic reference to the problem of updatable views, which deal with the same category of issues when solving a different problem.
-Migrations here feel a lot like the back translation problem in updatable views.
-
-Also, this is the most common type of evolution problem in the software industry, which is treated by disciplining the software process, versioning, and creating adapters.
-A contrast with these solutions, pointing again to the ghost system, would be very welcome.
-This approach is presented in section 6, but it makes sense to refer here as well.
-
-https://www.irisa.fr/lande/lande/icse-proceedings/seams/p86.pdf
-
-- Evolution of Computational Documents
-
-This is a problem that existing approaches solve relatively well, and there is work using, for instance, operation-based CRDTs to solve the problem of collaborative editing and ensure eventual consistency.
-I think those solutions should be referred to, and their results should be compared to the goals presented in this section.
-
-- Divergence control in Spreadsheets
-
-The problem in this use case is broader than spreadsheets, and the section title may be misleading.
-Versioning schemes, like semantic versioning or language-based versioning, could solve the problem of divergence control in spreadsheets.
-This class of solutions could help present this case better.
-
-https://dl.acm.org/doi/10.1145/3479394.3479416
-Multiplicity Change in Data Formats (Migrations)
-
-Using lenses to solve this problem would be a very good idea to highlight here, as in some old work of Benjamin Pierce and a system called Unison.
-This problem was central in transformations and sync operations in XML documents.
-I have no specific reference to point out, but Unison seems a good place to start.
-The synthesis of lenses can be done.
-
-- Live Modelling Languages
-
-This case study is, in my opinion, the more solid one in terms of presentation.
-It almost seems that this is the idealised system that the authors are thinking about.
-This presentation also gives away the identity of the authors (Programming is not double-blind; I am just referring to this since the authors omitted their names in the paper).
-
-The comment about the presented example being out of reach for all existing language workbenches could be expanded to explain why.
-Are there any techniques extracted from the remaining case studies that would be useful here?
-
-- Overall
-
-the discussion and conclusions are very brief both at the end and in each case study and do not provide a clear path to solving the challenges presented in the paper.
-Even a speculative section pointing to combining features from different systems would be welcome to communicate the authors' vision of the future of interactive programming systems.

### 2024-08-24 paper.tex
typos

@@
 In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development.
-Schema Change has been extensively studied in the context of databases where it necesitates manual and error-prone data migrations.~\cite{erhard06} We observe analogous problems occuring in varied and less-studied contexts.
+Schema Change has been extensively studied in the context of databases where it necessitates manual and error-prone data migrations.~\cite{erhard06} We observe analogous problems occuring in varied and less-studied contexts.
 For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't.
@@
 Indeed the motivation of this paper is to provide a basis for communicating and collaborating between such related efforts, as well as to facilitate further research efforts.
-We hope these challange problems serve as examples to ground future research and as a basis for evaluating its results.
+We hope these challenge problems serve as examples to ground future research and as a basis for evaluating its results.
 We invite contributions of new challenge problems and look forward to discussing competing solutions.
@@
 use a structure editor that provides high-level commands for triggering those edits.
-Recognising user intent from plain text editing of source code may make the challenge harder.
+Recognizing user intent from plain text editing of source code may make the challenge harder.
 
@@
 
-This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to docuuments, replacing the hellscape of manual copy \& paste.
+This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to documents, replacing the hellscape of manual copy \& paste.
 But meeting this challenge may require a basic change of perspective in both live programming and local-first software.
@@
 associations modeling the run-time state required for execution. The augmented metamodel includes
-properties \lstinline{vars} (to model variable assigments) and \lstinline{visited} (counting
+properties \lstinline{vars} (to model variable assignments) and \lstinline{visited} (counting
 number of state visits) to the classes \lstinline{Machine} and \lstinline{State}, respectively.
@@
 Schema evolution has long been a major problem for SQL databases because
-the SQL Data Definition Language (DDL) has limited capabilties to redefine existing data.
+the SQL Data Definition Language (DDL) has limited capabilities to redefine existing data.
 It can rename tables and columns.
@@
 Generalizing from lenses \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''.
-SMOs can also rewrite queries to accomodate schema changes.
+SMOs can also rewrite queries to accommodate schema changes.
 \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}.
@@
 Model Driven Engineering must also handle evolution.
-Unlike the textual artifacts involved in other domains, models typically assign unique indentifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
+Unlike the textual artifacts involved in other domains, models typically assign unique identifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
 Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}.
@@
 They consider the evolution of OO-style subclass hierarchies which goes beyond most other work.
-Their running example includes both our \emph{Extract Entity} and \emph{Multiplicty Change} examples, and would make for a good follow-on problem.
+Their running example includes both our \emph{Extract Entity} and \emph{Multiplicity Change} examples, and would make for a good follow-on problem.
 However they do not address code/query rewriting.

### 2024-08-24 revisions.tex
typos

@@
 
-We thank the anonymous reviewers for their helpful comments. In the following we summarize the changes we have made in response to those comments.
+We thank the anonymous reviewers for their helpful comments. In the following we summarize the changes we have made in response.
 
@@
 
-In the introduction we clarify that this paper is not about a tool under development.
+\begin{itemize}
+\item In the introduction we clarify that this paper is not about a tool under development.
 Some of the problems are extracted from our different research projects but none of these projects span the full range of problems.
 
-We have clarified in the introduction that we avoid speculating about possible solutions to the challenge problems so as to not bias them towards our preferred approaches (and to avoid the issues where we disagree).
+\item We have clarified in the introduction that we avoid speculating about possible solutions to the challenge problems so as to not bias them towards our preferred approaches (and to avoid the issues where we disagree).
 A primary goal of this paper is to establish a neutral common ground where researchers can meet to discuss their alternative approaches.
 \textcolor{red}{Or do we want to add a speculative section?}
+\end{itemize}
 
@@
 
-We rewrote the introduction to clarify that the goal of our research on schema change is to speed feedback loops in software development.
+\begin{itemize}
+  \item We rewrote the introduction to clarify that the goal of our research on schema change is to speed feedback loops in software development.
 Ultimately that is a value judgement, but we believe that value is important to a substantial community of researchers and practitioners.
@@
 
-Section 2 has been revised to remove the claim of providing a conceptual framework.
+  \item Section 2 has been revised to remove the claim of providing a conceptual framework.
 Instead we define a descriptive framework of multiple dimensions and layers of schema change that are used to figuratively contrast the different problems.
 
-\textcolor{red}{I don't know what to do with this:}
+  \item \textcolor{red}{I don't know what to do with this:}
 \begin{displayquote}
@@
 
-We clarified that the goal of Divergence in End-User Workflows is to create a new kind of feedback channel for end users rather than speeding up existing channels.
+\item We clarified that the goal of Divergence in End-User Workflows is to create a new kind of feedback channel for end users rather than speeding up existing channels.
 
-In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of offering solutions simple enough to be actually used.
+\item In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of offering solutions simple enough to be actually used.
+\end{itemize}
 
 \section{Reviewer B}
+\begin{itemize}
 
-As suggested we have replaced `type' with `schema' throughout the paper. Some diagrams still use the letter `T' for type but we intend to change those to `S' in the camera-ready version.
+  \item As suggested we have replaced `type' with `schema' throughout the paper. Some diagrams still use the letter `T' for type but we intend to change those to `S' in the camera-ready version.
 
-\textcolor{red}{I don't know what to do with this:}
+  \item \textcolor{red}{I don't know what to do with this:}
 \begin{displayquote}
@@
 
-\textcolor{red}{I don't know what to do with this:}
+\item \textcolor{red}{I don't know what to do with this:}
 \begin{displayquote}
@@
 
-As suggested, the Literature Review was converted into a more appropriately named Related Work section.
+\item As suggested, the Literature Review was converted into a more appropriately named Related Work section.
+\end{itemize}
 
 \section{Reviewer C}
+\begin{itemize}
+  \item in \S\emph{Remarks: Requirements and Implementation} we clarify that the goal is to adapt to schema changes automatically to avoid interrupting live programming.
 
-in \S\emph{Remarks: Requirements and Implementation} we clarify that the goal is to adapt to schema changes automatically to avoid interrupting live programming.
+  \item Added to Related Work the history of how the view-update problem led to lenses which led to DELs.
 
-We added to Related Work the history of how the view-update problem led to lenses which led to DELs.
+  \item Cited Hazel in Related Work.
 
-We cited Hazel in Related Work.
+  \item Mention the possibility of using CRDTs for evolving computational documents.
 
-We now mention the possibility of using CRDTs for evolving computational documents.
+  \item Divergence Control for Spreadsheets has been corrected to just Divergence Control.
 
-Divergence Control for Spreadsheets has been corrected to just Divergence Control.
-
-As mentioned above, we have added an explanation in the introduction of why we maintain neutrality by avoiding speculation on our preferred solutions.
-
+  \item As mentioned above, we have added an explanation in the introduction of why we maintain neutrality by avoiding speculation on our preferred solutions.
+\end{itemize}
 \end{document}

### 2024-08-25 paper.tex
DIvergence creates new feedback channels

@@
 
-This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to documents, replacing the hellscape of manual copy \& paste.
-But meeting this challenge may require a basic change of perspective in both live programming and local-first software.
+This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to documents, replacing manual copy \& paste.
+But meeting this challenge may require a  change of perspective in both live programming and local-first software.
 Live programming must move from being a solitary activity to a collaborative workflow, and further where the collaboration is not just on editing source code but on live integrated code and data,

### 2024-08-26 paper.tex
new abstract

@@
 
-\textcolor{red}{JE: This structure is offputting. We should write a proper abstract. }
-
+Many improvements to programming have come from shortening feedback loops, for example with Integrated Development Environments, Unit Testing, Live Programming, and Distributed Version Control.
+\emph{Schema evolution} is a roadblock to feedback in many situations that deserves greater attention.
+When requirements on the shape of data change then existing data must be migrated into the new shape, and existing code must be modified to suit.
+In many cases these adaptations must be performed manually, or with ad hoc code.
+Manual schema evolution not only delays feedback but since it occurs outside the purvey of version control tools it also interrupts collaboration.
+
+Schema evolution has long been studied in databases.
+We observe that the problem also occurs in non-database contexts that have been less studied.
+We present a suite of challenge problems exemplifying this range of contexts, including traditional database programming as well as live front-end programming, model-driven development, and collaboration in computational documents.
+We systematize these various contexts by defining a set of layers and dimensions of schema evolution.
+
+We offer these challenge problems to ground future research on the general problem of schema evolution in interactive programming systems and to serve as a basis for evaluating the results of that research.
+We hope that better support for schema evolution will make programming more live and collaboration more fluid.
+
+\begin{comment}
+Old abstract.
 % Context: What is the broad context of the work? What is the importance of the general research area?
-
-\noindent\textbf{Context:} Many recent improvements to programming have come from shortening feedback loops, for example in Live Programming and Distributed Version Control systems.
-To continue progress in this direction we need to extend Live Programming across the entire development stack, and to extend collaboration beyond compile time.
-
+Many recent programming improvements have been enabled by shorter feedback loops.
+To further advance the state of programming, we need to enable rapid collaboration and make programming more live across the entire development stack.
+To do this, programming systems need to preserve state during development and interaction with the programmer.
 % Inquiry: What problem or question does the paper address? How has this problem or question been addressed?
-\noindent\textbf{Inquiry:} A key obstacle to live and collaborative programming systems is the problem of \emph{schema evolution}.
-When a programmer changes the structure of a datatype, a programming system needs to co-evolve
-existing live data and adapt code that implements the program behaviour to match the new structure.
-This problem has been studied in isolation as schema evolution in databases, but also recurs in hot code reloading in live programming and dynamic software updating.
-We connect these independent efforts in an interdisciplinary inquiry into the problem of schema change through all the layers of the development stack and across all the time-scales of programming.
-%Such schema evolutions must occur immediately during live programming, but also need to be conveyed through long-term unplanned collaborations.
-
+The key obstacle to live collaborative programming systems is the problem of \emph{type evolution}.
+When programmer changes the structure of a type, programming system needs to co-evolve.
+This problem has been studied in isolation as schema evolution in databases, hot code reloading in live programming and dynamic software updating, but we lack a unified comprehensive perspective.
 % Approach: What was done that unveiled new knowledge?
-\noindent\textbf{Approach:} We distilled a set of case studies from personal experience, the research literature, and common industry practices.
-These cases reveal schema change problems recurring in different contexts in the developer stack and at different time-scales of collaboration, not just in database production upgrades.
-
+In this paper, we develop a unified conceptual framework for analysing the problem of type evolution. We use the framework to study type evolution in five diverse case studies.
 % Knowledge: What new facts were uncovered? What new capabilities are enabled by the work?
-\noindent\textbf{Knowledge:} The problem of schema change is deeper, more widespread, and more interdisciplinary than commonly appreciated.
-It is a key obstacle to shortening feedback loops in diverse contexts of software development.
-
+By looking at the case studies through a unified perspective, we hope to inspire new thinking about the problem of type evolution and also encourage knowledge transfer across sub-fields.
 % Grounding: What argument, feasibility proof, artifacts, or results and evaluation support this work?
-\noindent\textbf{Grounding:} Our work takes the form of a suite of challenge problems drawn from real-world programming
-scenarios in diverse field such as live programming, database programming and model-driven
-development.
-%
-
+Our work takes the form of a sequence of challenge problems drawn from real-world programming scenario in diverse field such as live programming, database programming and model-driven development.
 % Importance: Why does this work matter?
-\noindent\textbf{Importance:} Schema change is a key problem for universal live programming and frictionless collaborative development.
-Our suite of challenge problems serve to ground further research and provide a basis for evaluation and comparison of results.
+Tackling the problem of type evolution in a unified way, enabled by our new conceptual framework, will make it possible to design and build a future generation of effective, live and collaborative programming systems.
+\end{comment}
 

### 2024-08-28 paper.tex
tweak abstract
clarify that manual evolution isn't a requirement in principle, it's just the status quo

@@
 Many improvements to programming have come from shortening feedback loops, for example with Integrated Development Environments, Unit Testing, Live Programming, and Distributed Version Control.
-\emph{Schema evolution} is a roadblock to feedback in many situations that deserves greater attention.
+A barrier to feedback that deserves greater attention is \emph{Schema Evolution}.
 When requirements on the shape of data change then existing data must be migrated into the new shape, and existing code must be modified to suit.
-In many cases these adaptations must be performed manually, or with ad hoc code.
+Currently these adaptations are often performed manually, or with ad hoc code.
 Manual schema evolution not only delays feedback but since it occurs outside the purvey of version control tools it also interrupts collaboration.

### 2024-08-28 paper.tex
simplify wording
move citation before period

@@
 
-In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development.
-Schema Change has been extensively studied in the context of databases where it necessitates manual and error-prone data migrations.~\cite{erhard06} We observe analogous problems occuring in varied and less-studied contexts.
+In this paper we identify a common problem, \emph{Schema Change}, that interferes with feedback loops in diverse contexts of software development.
+Schema Change has been extensively studied in the context of databases where it necessitates manual and error-prone data migrations ~\cite{erhard06}.
+We observe analogous problems occuring in varied and less-studied contexts.
 For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't.

### 2024-08-28 paper.tex
fix typos

@@
 \begin{itemize}
-\item \emph{Schema} -- structure program data and determine aspects of code. A change in schema
+\item \emph{Schema} -- structures program data and determines aspects of code. A change in schema
   requires a corresponding change in data and code. Schema may be explicit (database schema, type declarations)
@@
 \item \emph{Data} -- information stored by the program. Data may be more permanent (data in a
-  database) or less permanent (state in live programming). When scheam evolve, data needs to be
+  database) or less permanent (state in live programming). When schemas evolve, data needs to be
   updated or (in case of transient state) discarded.
 \item \emph{Code} -- or program logic, but not including source code that defines schema.
-  When schema evolve, code needs to change to consume/produce data of the correct shape.
+  When schemas evolve, code needs to change to consume/produce data of the correct shape.
 \end{itemize}

### 2024-08-28 paper.tex
hedge claim about being inspiring

@@
 schema evolution. The case studies show how well-known solutions or problems in one domain
-map to another domain and they also provide and inspiration and a benchmark for
+map to another domain, and we hope they provide an inspiration and a benchmark for
 developers of future programming systems.

### 2024-08-28 paper.tex
clean up first mention of local-first

@@
 
-The Extract Entity challenge in the non-local case (for local-first software) is similar to the
-local live programming challenge except that the context is an application that offers similar
-functionality with peer-to-peer collaboration. In this context the schema change does not need
+We can also consider the Extract Entity challenge in a local-first ~\cite{localfirst} case, where the context is an application that offers peer-to-peer collaboration.
+In this context the schema change does not need
 to be performed interactively -- a developer can take some time to build, test, and package a

### 2024-08-28 paper.tex
fix pluralization typo

@@
 
-The challenge of maintaining multiple divergent schema in programming systems can
+The challenge of maintaining multiple divergent schemas in programming systems can
 be well illustrated using \emph{multiplicity change}, which is a common kind of schema change.

### 2024-08-29 revisions.tex
DIvergence creates new feedback channels
new abstract
more tweaks

@@
 
-We thank the anonymous reviewers for their helpful comments. In the following we summarize the changes we have made in response.
+We thank the anonymous reviewers for their helpful comments. In the following we summarize the revisions we have made in response.
 
-\section{Summary review}
+\paragraph{Summary review}{}
 
 \begin{itemize}
-\item In the introduction we clarify that this paper is not about a tool under development.
+\item We rewrote the introduction section and clarified that this paper is not about a tool under development.
 Some of the problems are extracted from our different research projects but none of these projects span the full range of problems.
 
-\item We have clarified in the introduction that we avoid speculating about possible solutions to the challenge problems so as to not bias them towards our preferred approaches (and to avoid the issues where we disagree).
+\item The new introduction explains that we avoid speculating about possible solutions to the challenge problems so as to not bias them towards our preferred approaches (and to avoid the issues where we disagree).
 A primary goal of this paper is to establish a neutral common ground where researchers can meet to discuss their alternative approaches.
-\textcolor{red}{Or do we want to add a speculative section?}
+
 \end{itemize}
 
-\section{Reviewer A}
+\paragraph{Reviewer A}{}
 
@@
 Ultimately that is a value judgement, but we believe that value is important to a substantial community of researchers and practitioners.
-We have qualified statements of opinion throughout the paper.
+We have qualified statements of opinion throughout the paper and tied everything back to the goal of improving feedback loops.
+Thanks for helping us to get clearer on our goals.
 
@@
 
-  \item \textcolor{red}{I don't know what to do with this:}
-\begin{displayquote}
-  Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations.
+% I don't know what to do with this:
+% \begin{displayquote}
+%   Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations.
 Please, add numbers to those expectations, show examples of corner cases where stuff breaks and is an acceptable evil, and cases were stuff breaks on purpose (in some sense, and intended evil) To make an example: Java is of course not truly "cross-platform": if you wanted to put a formal definition to cross platform, the program should work identically in all environments.
@@
 
-  In the context of those data transformations, there must be plenty of situations like that, where a change in behavior is not only possible or acceptable, but even desirable.
+%   In the context of those data transformations, there must be plenty of situations like that, where a change in behavior is not only possible or acceptable, but even desirable.
 However, we should have some clear formal definition of when/how/why such changes in behavior are introduced.
-\end{displayquote}
+% \end{displayquote}
 
@@
 
-\section{Reviewer B}
+\paragraph{Reviewer B}{}
 \begin{itemize}
 
-  \item As suggested we have replaced `type' with `schema' throughout the paper. Some diagrams still use the letter `T' for type but we intend to change those to `S' in the camera-ready version.
+  \item As suggested we have replaced `type' with `schema' throughout the paper, defined to include types and specifications of all sorts.
+The Elm discussion still refers to types as they are the kind of schema involved there.
+Some diagrams still use the letter `T' for type but we intend to change those to `S' in the camera-ready version.
 
-  \item \textcolor{red}{I don't know what to do with this:}
-\begin{displayquote}
-I liked the discussion of both local/singular and collaborative development and the problems associated with them.
+% I don't know what to do with this:
+% \begin{displayquote}
+% I liked the discussion of both local/singular and collaborative development and the problems associated with them.
 And I would have liked to see those two situations discussed separately, because of their technical complexity and also the processes (people) causing them.
 
-The placement of “local-first” felt like it didn’t help me understanding the problem. Maybe you can revisit the intent of those references. My preference would be try without them.
-\end{displayquote}
+% The placement of “local-first” felt like it didn’t help me understanding the problem. Maybe you can revisit the intent of those references. My preference would be try without them.
+% \end{displayquote}
 
-\item \textcolor{red}{I don't know what to do with this:}
-\begin{displayquote}
-  While re-reading the submission, I noticed quite a similarity to a pattern language (in the sense of Christopher Alexander and/or the software patterns community).
+% I don't know what to do with this:
+% \begin{displayquote}
+%   While re-reading the submission, I noticed quite a similarity to a pattern language (in the sense of Christopher Alexander and/or the software patterns community).
 All sections covering a challenge problem present the context in which the problems occurs, describe the problem, and provide an example illustrating the problem to better relate to it.
@@
 
-  However, even though those elements are all present to a varying extent, the guiding structure of a pattern and their interconnection through a pattern language might be helpful but is missing.
+%   However, even though those elements are all present to a varying extent, the guiding structure of a pattern and their interconnection through a pattern language might be helpful but is missing.
 I strongly believe that transforming the material from a simple collection of problems into a pattern-language-like (the “like” because of the missing generative solutions) artifact would be helpful not only for the readers but the community helping provide those generative solutions and extending the (such) system of patterns…
-\end{displayquote}
+% \end{displayquote}
 
-\item As suggested, the Literature Review was converted into a more appropriately named Related Work section.
+\item As suggested, the Literature Review Appendix was converted into a more appropriately named Related Work section.
+
 \end{itemize}
 
-\section{Reviewer C}
+\paragraph{Reviewer C}{}
 \begin{itemize}
@@
 
-  \item Divergence Control for Spreadsheets has been corrected to just Divergence Control.
+  \item Divergence Control for Spreadsheets has been corrected to just Divergence Control. The goal is clarified as seeking to create new feedback channels for end users to communicate changes.
 

### 2024-10-09 paper.tex
Update figures, final version layout tweaks

@@
-\documentclass[english,submission]{programming}
+\documentclass[english,crc]{programming}
 
@@
 
-\begin{document}
-
 % define natbib citet
@@
 
+\begin{document}
+
 \title{Schema Evolution in Interactive Programming Systems}
 
-\author{Anonymous}
-\affiliation{blinded}
-% \author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
-% \authorinfo{is an independent researcher. Contact him at \email{jonathanmedwards@gmail.com}.}
-% \affiliation[a]{Independent, Boston, MA, USA}
-
-% \author[b]{Tomas Petricek}%[0000-0002-7242-2208]
-% \authorinfo{is an Associate Professor at Charles University. Contact him at \email{tomas@tomasp.net}.}
-% \affiliation[b]{Charles University, Prague, Czechia}
-
-% \author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
-% \authorinfo{is a Senior Researcher at Centrum Wiskunde \&\ Informatica (CWI) and a Professor at the University of Groningen. Contact him at \email{storm@cwi.nl}.}
-% \affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
-% \affiliation[d]{University of Groningen, Groningen, Netherlands}
-
-% \author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
-% \authorinfo{is a Senior Researcher at the independent research lab Ink \& Switch. Contact him at \email{gklitt@gmail.com}.}
-% \affiliation[e]{Ink \& Switch, Planet Earth}
-
-% \authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}
+\author[a]{Jonathan Edwards}[0000-0003-1958-7967]
+\authorinfo{is an independent researcher working on drastically simplifying programming. He is known
+for his Subtext series of programming language experiments and his blog at \href{https://alarmingdevelopment.org}{\texttt{alarmingdevelopment.org}}.
+He has been a researcher at MIT CSAIL and CDG/HARC. He tweets \href{https://x.com/jonathoda}{\texttt{@jonathoda}} and can be reached at
+\email{jonathanmedwards@gmail.com}.}
+\affiliation[a]{Independent, Boston, MA, USA}
+
+\author[b]{Tomas Petricek}[0000-0002-7242-2208]
+\authorinfo{is an assistant professor at Charles University. He
+is interested in finding easier and more accessible ways of thinking
+about programming. To do so, he combines technical work on
+programming systems and tools with research into history and
+philosophy of science. His work can be found at \href{https://tomasp.net}{\texttt{tomasp.net}} and
+he can be reached at \email{tomas@tomasp.net}.}
+\affiliation[b]{Charles University, Prague, Czechia}
+
+\author[c,d]{Tijs van der Storm}[0000-0001-8853-7934]
+\authorinfo{is a senior researcher in the Software Analysis and Transformation (SWAT) group at Centrum Wiskunde \&\ Informatica (CWI) in Amsterdam,
+and full professor in Software Engineering at the University of Groningen in Groningen. His research focuses on improving programmer
+experience through new and better software languages and developing the tools and techniques to engineer them in a modular and interactive fashion. Contact him at \email{storm@cwi.nl}.}
+\affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
+\affiliation[d]{University of Groningen, Groningen, Netherlands}
+
+\author[e]{Geoffrey Litt}[0000-0003-0858-5165]
+\authorinfo{is a senior researcher at the independent research lab Ink \& Switch.
+Previously, he completed a PhD at MIT CSAIL in the Software Design Group advised by Daniel Jackson.
+Before that, he spent five years as an early engineer and designer building the edtech startup Panorama Education (YC S13).
+Contact him at \email{gklitt@gmail.com}.}
+\affiliation[e]{Ink \& Switch, Planet Earth}
+
+\authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}
 
@@
 
-% TODO: Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification System
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007.10011006.10011008</concept_id>
+       <concept_desc>Software and its engineering~General programming languages</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011024.10011028</concept_id>
+       <concept_desc>Software and its engineering~Data types and structures</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011073</concept_id>
+       <concept_desc>Software and its engineering~Software maintenance tools</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10002951.10002952.10003219</concept_id>
+       <concept_desc>Information systems~Information integration</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[300]{Software and its engineering~General programming languages}
+\ccsdesc[500]{Software and its engineering~Data types and structures}
+\ccsdesc[300]{Software and its engineering~Software maintenance tools}
+\ccsdesc[300]{Information systems~Information integration}
 
@@
 When requirements on the shape of data change then existing data must be migrated into the new shape, and existing code must be modified to suit.
-Currently these adaptations are often performed manually, or with ad hoc code.
-Manual schema evolution not only delays feedback but since it occurs outside the purvey of version control tools it also interrupts collaboration.
+Currently these adaptations are often performed manually, or with ad hoc scripts.
+Manual schema evolution not only delays feedback but since it occurs outside the purview of version control tools it also interrupts collaboration.
 
@@
 Research on Live Programming~\cite{rein2018exploratory} seeks to provide immediate feedback on runtime behavior while editing.
-Feedback is equally valuable in collaborative programming~\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) to convey changes between coworkers with less friction.
+Feedback is equally valuable in collaborative programming~\cite{ProGit, goldman2011real,kurniawan2015coder,replit} to convey changes between coworkers with less friction.
 However there is still much room to improve feedback loops throughout software development.
@@
 Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system.
-We discuss further examples in \S\ref{sec:related} Related Work.
-In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops.
-To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.
+We discuss further examples in Section~\ref{sec:related}.
+In general, when the shape of data becomes a dependency, changes to that data's shape can require manual intervention which interrupts feedback loops.
+To coordinate research on these related problems we shift the focus to \emph{Interactive Programming Systems}~\cite{techdims} and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.
 
 Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops.
-We believe these problems have not been completely solved in theory or practice, which we discuss further in \S\ref{sec:related} Related Work.
-While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches.
+We believe these problems have not been completely solved in theory or practice.
+While some of them come from our individual research efforts we have attempted to make neutral problem statements independent of our preferred approaches.
 None of our ongoing efforts yet address the full range of problems.
-Indeed the motivation of this paper is to provide a basis for communicating and collaborating between such related efforts, as well as to facilitate further research efforts.
-We hope these challenge problems serve as examples to ground future research and as a basis for evaluating its results.
+Indeed, the aim of this paper is to provide a basis for collaboration between such related efforts, facilitate further research, as well as to serve as a basis for evaluating research results.
 We invite contributions of new challenge problems and look forward to discussing competing solutions.
@@
 \begin{wrapfigure}{r}{12em}
-\vspace{-1.5em}
+\vspace{-1.6em}
 \includegraphics[width=12em]{figures/arr-basic.png}
-\vspace{-1.5em}
+\vspace{-1.6em}
 \end{wrapfigure}
@@
 The situation is illustrated on the right. We start with code $C_1$ and
-data $D_1$ that have a shape corresponding to schema $T_1$. When the schema changes from $T_1$ to $T_2$ ($\rightarrow$),
+data $D_1$ that have a shape corresponding to schema $S_1$. When the schema changes from $S_1$ to $S_2$ ($\rightarrow$),
 a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
@@
 that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
-to the new schema $T_2$.
+to the new schema $S_2$.
 %
@@
 \includegraphics[width=12em]{figures/arr-forkjoin.png}
-\vspace{-1.5em}
+\vspace{-2em}
 \end{wrapfigure}
@@
 variants that exist independently and changes made to them may need to be synchronized. The
-situation is illustrated on the right. Here, the initial schema $T_1$ is transformed by two
-users independently into schema $T_2$ and $T_3$. The corresponding data is transformed accordingly,
+situation is illustrated on the right. Here, the initial schema $S_1$ is transformed by two
+users independently into schema $S_2$ and $S_3$. The corresponding data is transformed accordingly,
 producing $D_2$ and $D_3$. The challenge is now merging these two diverging transformation and
-obtaining a schema $T_4$ and data $D_4$ that adopt both of the changes.
+obtaining a schema $S_4$ and data $D_4$ that adopt both of the changes.
 
@@
 
+\vspace{-0.4em}
 \paragraph{Locality of Change.}
@@
 
+\vspace{-0.4em}
 \paragraph{Program Layers.}
@@
 \item \emph{Code} -- or program logic, but not including source code that defines schema.
-  When schemas evolve, code needs to change to consume/produce data of the correct shape.
+  When schemas evolve, code needs to change to work with data of the correct shape.
 \end{itemize}
 
+\vspace{-0.4em}
 \paragraph{Program Variants.}
@@
 
+\vspace{-0.4em}
 \paragraph{Challenge Problems.}
-These dimensions, layers, and variants generate a wide range of configurations for challenge problems, but
-our aim is not to cover all of them. Instead we use this perspective to recast a number of
+These dimensions, layers, and variants generate a wide range of problem configuration.
+Our aim is not to cover all of them. Instead we use this perspective to recast a number of
 existing challenges in real-world programming systems as instances of the same problem of
 schema evolution. The case studies show how well-known solutions or problems in one domain
-map to another domain, and we hope they provide an inspiration and a benchmark for
-developers of future programming systems.
+map to another domain. We hope they provide an inspiration and a benchmark for
+developers of future systems.
 
@@
 live programming system would allow live updates to the structure of the two types.
+
+\challenge{Live State Type Evolution}
+Assume that we have a running todo list with the above state and events. To test the application,
+the programmer has already created a number of items and so there is a value of the
+\texttt{State} type that represents the current state of the application such as:
+
+\begin{lstlisting}[language=ml,morekeywords={true,false}]
+{ items : [
+   { id = 1; title = "Check Twitter"; completed = true }
+   { id = 2; title = "Write the paper"; completed = false } ] }
+\end{lstlisting}
 
@@
 
-\challenge{Live State Type Evolution}
-Assume that we have a running todo list with the above state and events. To test the application,
-the programmer has already created a number of items and so there is a value of the
-\texttt{State} type that represents the current state of the application such as:
-
-\begin{lstlisting}[language=ml,morekeywords={true,false}]
-{ items : [
-   { id = 1; title = "Check Twitter"; completed = true }
-   { id = 2; title = "Write the paper"; completed = false } ] }
-\end{lstlisting}
-
-\noindent
 There is a number of edits to the types that the programmer may now want to do without
@@
 type State = { items : Item list }
-type Event =
-  | .. | SetCompleted of id * Maybe<DateTime> | ..
+type Event = .. | SetCompleted of id * Maybe<DateTime> | ..
 \end{lstlisting}
@@
 \frameworkboxtitle{Primitive Schema Transformations}
-In this case study, the evolving schema is the \texttt{State} type. When the type evolves,
-the program logic (code) and the current live value of the type (data) needs to evolve
+Here, the evolving schema is the \texttt{State} type. When the type evolves,
+the program logic (code) and the current live value (data) needs to evolve
 correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
@@
 primitive code and data transformation. The primitive transformations may either be invoked
-through a user interface, or inferred from textual code edits made by the programmer.
+through a UI, or inferred from textual code edits made by the programmer.
 }
@@
 data, we can gain valuable insights from research on database systems.
-Although schema evolution is well-studied in databases (\S\ref{sec:related} Related Work), such past work
+Although schema evolution is well-studied in databases (Section~\ref{sec:related}), such past work
 does not resolve all problems in our second challenge.
@@
 it is not a schema change but a data change, yet nevertheless not commonly supported as an operaton in database APIs or DMLs.
-
-Entity evolution can happen through operations such as those listed in
-Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
-change in requirements that triggered an evolution might change back again.
-Invertability features in other work on schema evolution and \emph{bidirectional transformations}~\cite{Foster2007, herrmann17, chillon22}.
-The inverse of \emph{Extract Entity} is \emph{Absorb Entity}. It raises the question of how
-to handle \emph{one-to-many} relationships. For example should absorbing orders into customers
-do a relational join, returning us back to the starting point, or should it nest orders within
-customers, as a NoSQL database might do?
-
-The inverse of \emph{Merge Entities} is \emph{Split Entity}, but that can be awkward.
-Imagine that we didn't have addresses for customers to disambiguate them with and only knew
-their non-unique names. Then \emph{Extract Entity} will conflate them and discard the
-information needed to properly invert the operation. Should this information be retained?
-One could object that we shouldn't have taken orders for people we can't uniquely identify
-yet. And even if that was necessary it means the customers really aren't distinct entities
-and shouldn't have been extracted. These are arguments that we shouldn't need \textit{Split
-Entity}, contradicting the principle that evolution operations should be invertible.
-It is interesting to observe that evolution in nature can also fail to invert, leaving
-behind vestigial features. We consider this to be an open issue.
 
@@
 
+Entity evolution can happen through operations such as those listed in
+Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
+change in requirements that triggered an evolution might change back again.
+Invertability features in other work on schema evolution and \emph{bidirectional transformations}~\cite{Foster2007, herrmann17, chillon22}.
+The inverse of \emph{Extract Entity} is \emph{Absorb Entity}. It raises the question of how
+to handle \emph{one-to-many} relationships. For example should absorbing orders into customers
+do a relational join, returning us back to the starting point, or should it nest orders within
+customers, as a NoSQL database might do?
+
+The inverse of \emph{Merge Entities} is \emph{Split Entity}, but that can be awkward.
+Imagine that we didn't have addresses for customers to disambiguate them with and only knew
+their non-unique names. Then \emph{Extract Entity} will conflate them and discard the
+information needed to properly invert the operation. Should this information be retained?
+One could object that we shouldn't have taken orders for people we can't uniquely identify
+yet. And even if that was necessary it means the customers really aren't distinct entities
+and shouldn't have been extracted. These are arguments that we shouldn't need \textit{Split
+Entity}, contradicting the principle that evolution operations should be invertible.
+It is interesting to observe that evolution in nature can also fail to invert, leaving
+behind vestigial features. We consider this to be an open issue.
+
 \subsection*{Remarks: Representation and Replication}
@@
 \begin{wrapfigure}{r}{12em}
-\vspace{-0.5em}
+\vspace{-2em}
 \includegraphics[width=12em]{figures/arr-db.png}
-\vspace{-1em}
+\vspace{-1.5em}
 \end{wrapfigure}
@@
 Our framing shows that queries $Q_1$ also need to transform, which is not always addressed
-in work on schema evolution. Extract/Absorb Entity operations are also interesting because
+in work on schema evolution. Extract/Absorb Entity are also interesting because
 they transform an equivalent pair of schemas and so it should be possible to apply them
@@
 %
-The ideas in this section also apply to work on computational scientific notebooks
-that combine code and data such as Jupyter \cite{Kluyver2016}. We hope to contribute to ongoing
-work on advancing the format as done by Nextjournal \cite{Nextjournal21}, Nota~\cite{Crichton2021}
-and Living Papers~\cite{Heer2023}.
+The ideas in this section also apply to computational scientific notebooks
+that combine code and data \cite{Kluyver2016}. We hope to contribute to ongoing
+work advancing the format \cite{Nextjournal21,Crichton2021,Heer2023}.
 
@@
 
+\subsection*{Remarks: Requirements and Representation}
+It should be possible to automatically merge any two sequences of edits (changing both
+the document structure and data) and the order in which the edits are applied should also not
+matter.
+One approach is to use CRDTs~\cite{Shapiro11} which guarantee order-independent convergence over a restricted set of edit operations.
+This may not always be possible and the system may ask the user for resolution or,
+possibly, use a default behavior specified by the user.
+
+The schema and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
+lists edits appearing in our examples (but not a complete list). A system may
+use a structure editor that provides high-level commands for triggering those edits.
+Recognizing user intent from plain text editing of source code may make the challenge harder.
+
 \frameworkbox{
@@
 It involves two independent edits. In one (top right), the data $D_1$ is
-changed to $D_2'$, but the schema $T_1$ remains unchanged. In the other (bottom left), the
-schema $T_1$ and data $D_1$ co-evolve into $T_2$ and $D_2$ and the data is further changed
+changed to $D_2'$, but the schema $S_1$ remains unchanged. In the other (bottom left), the
+schema $S_1$ and data $D_1$ co-evolve into $S_2$ and $D_2$ and the data is further changed
 to $D_3$. The challenge is to merge the two edits and obtain jointly modified data $D_3'$
-of schema $T_2$. In the next section, we will see the same pattern, but also involving
+of schema $S_2$. In the next section, we will see the same pattern, but also involving
 code edits. We assume the \emph{convergence} model in this section, but will discuss
 the more challenging \emph{divergence} model later in the paper.}
-
-\subsection*{Remarks: Requirements and Representation}
-It should be possible to automatically merge any two sequences of edits (changing both
-the document structure and data) and the order in which the edits are applied should also not
-matter.
-One approach is to use CRDTs~\cite{Shapiro11} which guarantee order-independent convergence over a restricted set of edit operations.
-This may not always be possible and the system may ask the user for resolution or,
-possibly, use a default behavior specified by the user.
-
-The schema and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
-lists edits appearing in our examples (but not a complete list). A system may
-use a structure editor that provides high-level commands for triggering those edits.
-Recognizing user intent from plain text editing of source code may make the challenge harder.
 
@@
 In the diagram on the right, we only show data and schema, ignoring the code added by the
-accounting department. We start with data $D_1$ of schema~$T_1$. The accounting department
+accounting department. We start with data $D_1$ of schema~$S_1$. The accounting department
 evolves the schema (via Extract Entity), obtaining data $D_2$ of schema $T_2$ using an automatically
 generated forward transformation. It further modifies the data, obtaining $D_3$. When the
-accounting department receives data $D_3$, it needs to migrate it back from schema $T_2$ to
-$T_1$ using an automatically generated backward transformation. That is,
+accounting department receives data $D_3$, it needs to migrate it back from schema $S_2$ to
+$S_1$ using an automatically generated backward transformation. That is,
 schema evolution needs to be accompanied with both forward and backward projection for data
@@
 The extra complexity in the above scenario arises from the fact that both schema and
-code are forked and evolve independently. In one branch, original schema $T_1$ evolve
-into $T_2$ and original code $C_1$ into $C_3$. In another branch, original schema $T_1$
-evolve into $T'_2$ and original code $C_1$ into $C'_3$. The user then wants to merge
+code are forked and evolve independently. In one branch, original schema $S_1$ evolve
+into $S_2$ and original code $C_1$ into $C_3$. In another branch, original schema $S_1$
+evolve into $S'_2$ and original code $C_1$ into $C'_3$. The user then wants to merge
 the independently done changes into a new version of schema and code, written as
-$T_2 \diamond T'_2$ and $C_3 \diamond C'_3$. The diverging transformations may make
+$S_2 \diamond S'_2$ and $C_3 \diamond C'_3$. The diverging transformations may make
 conflicting changes that would need to be resolved with assistance from the user.}
@@
 Because the intent of the writer to the scalar schema cannot be unambiguously interpreted from the
-write alone, it is impossible to make a perfect choice among these options. The programming system
-thus needs to consult the user.
+write alone, it is impossible to make a perfect choice among these options.
+A programming system could consult the user to disambiguate their intent, as well as choose (and explain) a default behavior.
 
@@
 The diagram illustrating the scenario remains the same as in the case of divergence control
-for Extract Entity. The initial data and schema $D_1, T_1$ co-evolve into $D_2, T_2$ (adding
+for Extract Entity. The initial data and schema $D_1, S_1$ co-evolve into $D_2, S_2$ (adding
 multiplicity) and a new value is then assigned to the array of assignees, resulting in $D_3$.
 The difference from the earlier case is that the backward migration (from $D_3$ to $D'_3$)
-generated for the schema evolution (from $T_1$ to $T_2$) cannot be generated fully automatically.
+generated for the schema evolution (from $S_1$ to $S_2$) cannot be generated fully automatically.
 In case of Extract Entity this was in principle possible.}
@@
 
-Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this
-problem would give developers or users tools to manage these tradeoffs. For example, a system
+Although there is likely no silver bullet to navigate these tradeoffs, a good solution
+would give developers or users tools to manage these tradeoffs. For example, a system
 might allow developers to specify the desired behavior for a particular pair of schemas based
@@
 
+\section{Live Modeling Languages}
+\label{sec:livemodeling}
+%Related work
+%``Slogan'': editing a program is diverging from run-time schema.
+
+We considered live programming in the context of Elm in Section~\ref{sec:elm},
+where the challenge was to migrate run-time state (data) to a new schema to keep the program running.
+Work on language modeling offers a new perspective on the problem.
+In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
+program in a domain-specific language as an instance of a metamodel (a static type, or, a schema) that
+defines the AST structure of the domain-specific language.
+
+A program in a domain-specific language in turn determines (defines/implies/induces) a run-time
+schema: the run-time structures of code (e.g., classes, inheritance links, declarations, method
+definitions etc.), as well as the structure of the run-time state (the heap, slots for user data, etc.).
+Whereas the example discussed in this section is framed in terms of metamodels and models, this can
+be seen as an expository device: the principles and challenges apply in more grammar-oriented settings.
+However, the framing also suggests interesting additional challenges.
+
+In the absence of live programming, the run-time \emph{schema} and \emph{code} do not change while
+running a program in a given domain-specific language and the run-time \emph{state} changes constantly.
+Live programming, however, requires reconciling changes to the program with the induced run-time structures
+of both the code and the state. This typically means that at a certain point (a quiescent point)
+during execution, the code structures need to be updated (hot swapped), and the state needs to be
+\textit{migrated}.
+
 \begin{figure}[t]
@@
 \end{figure}
-
-~
-
-\section{Live Modeling Languages}
-\label{sec:livemodeling}
-%Related work
-%``Slogan'': editing a program is diverging from run-time schema.
-
-We considered live programming in the context of the Elm architecture in \S\ref{sec:elm},
-where the challenge was to migrate run-time state (data) to a new schema to keep the program running.
-Work on language modeling offers a new perspective on the problem.
-In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
-program in a domain-specific language as an instance of a metamodel (a static type AKA schema) that
-defines the AST structure of the domain-specific language.
-
-A program in a domain-specific language in turn determines (defines/implies/induces) a run-time
-schema: the run-time structures of code (e.g., classes, inheritance links, declarations, method
-definitions etc.), as well as the structure of the run-time state (the heap, slots for user data, etc.).
-Whereas the example discussed in this section is framed in terms of metamodels and models, this can
-be seen as an expository device: the principles and challenges apply in more grammar-oriented settings.
-However, the framing also suggests interesting additional challenges.
-
-In the absence of live programming, the run-time \emph{schema} and \emph{code} do not change while
-running a program in a given domain-specific language and the run-time \emph{state} changes constantly.
-Live programming, however, requires reconciling changes to the program with the induced run-time structures
-of both the code and the state. This typically means that at a certain point (a quiescent point)
-during execution, the code structures need to be updated (hot swapped), and the state needs to be
-\textit{migrated}.
 
@@
 
+\begin{table}[b]
+\begin{tabular}{lll}\toprule
+{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
+Add/Rename & state & Simple  \\
+Remove & state & Structurally simple, but heuristic if state is current \\
+Add & variable & Add entry to map and initialize \\
+Remove & variable & Remove from the variable map (assuming it is unused)\\
+Rename & variable & Rename in class, preserving value\\
+Type change & variable & Migrate value in map \\
+Add & transition & Simple \\
+Remove & transition & Structurally simple, but heuristic for pending events\\
+Event change & transition & Structurally simple, but heuristic for pending events\\
+Add/remove & statement & Hotswap code at quiescent point of interpreter\\
+\bottomrule
+\end{tabular}
+\vspace{0.3em}
+\caption{Possible program changes and how to deal with them at run time}
+\label{tbl:smchanges}
+\vspace{-1em}
+\end{table}
+
 \challenge{Live Editing Domain-Specific Languages}
@@
 
-\begin{table}[t]
-\begin{adjustwidth}{-2em}{-2em}
-\begin{tabular}{lll}\toprule
-{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
-Add/Rename & state & Simple  \\
-Remove & state & Structurally simple, but heuristic if state is current \\
-Add & variable & Add entry to map and initialize \\
-Remove & variable & Remove from the variable map (NB: assumes variable is unused)\\
-Rename & variable & Rename in class, preserving value\\
-Type change & variable & Migrate value in map \\
-Add & transition & Simple \\
-Remove & transition & Structurally simple, but heuristic for pending events\\
-Event change & transition & Structurally simple, but heuristic for pending events\\
-Add/remove & statement & Hotswap code at quiescent point of interpreter\\
-\bottomrule
-\end{tabular}
-\vspace{0.3em}
-\caption{Possible program changes and how to deal with them at run time}
-\label{tbl:smchanges}
-\end{adjustwidth}
-\vspace{-1em}
-\end{table}
-
 \frameworkbox{
@@
 We could see this challenge as schema, code and data co-evolution as in the case of the Elm
-architecture in \S\ref{sec:elm}, but a more interesting perspective is to see the
-state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.
+architecture in Section~\ref{sec:elm}, but a more interesting perspective is to see the
+state machine definition as data $D_0$ conforming to a type $S$ defined by the metamodel.
 
 When the state machine is instantiated, the metamodel and the definition are used to derive
-the run-time type ${}^+\!T_0$. The run-time data ${}^+\!D_0$ conforming to this type still
+the run-time type ${}^+\!S_0$. The run-time data ${}^+\!D_0$ conforming to this type still
 contain the state machine definition, but also the necessary run-time state required for
@@
 If the state machine definition changes from $D_0$ to $D_1$ (top right), we can synthesize
-a new run-time state machine type ${}^+\!T_1$ and new data ${}^+\!D_1$ conforming to it.
+a new run-time state machine type ${}^+\!S_1$ and new data ${}^+\!D_1$ conforming to it.
 The challenge is to merge the two run-time types instantiated from different definitions
-to produce a type ${}^+\!T_0 \diamond {}+\!T_1$, but also merge the run-time state of the
+to produce a type ${}^+\!S_0 \diamond {}+\!S_1$, but also merge the run-time state of the
 running machine with the new instantiated state to obtain ${}^+\!D'_0 \diamond {}^+\!D_1$.
@@
 \citet{Cambria} applies lenses~\cite{Foster2007} to this problem, specifically discussing the \emph{Multiplicity Change} problem.
-
+%
 Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}.
@@
 \citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
-\citet{RuntimeConstraint} study the live modeling problem in \S\ref{sec:livemodeling}.
+\citet{RuntimeConstraint} study the live modeling problem in Section~\ref{sec:livemodeling}.
 They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution.
@@
 
+\acks
+We are thank reviewers and attendees of the LIVE 2023 workshop who provided invaluable
+feedback on earlier version of the paper. We are also grateful to the anonymous reviewers
+of the journal version of the paper whose feedback greatly improved the quality of the final
+version. The second author has been supported by Charles University grant PRIMUS/24/SCI/021.
+
 %\bibliographystyle{abbrv}

### 2024-10-25 paper.tex
More final version
Make conference publishing happier

@@
 \usepackage{xcolor}
-\usepackage{ccicons}
+%\usepackage{ccicons}
 \usepackage{float}
@@
   aboveskip=1em,
-  framesep=0.5em,
+  %framesep=0.5em,
   basicstyle=\normalsize\ttfamily,
@@
   numbersep=1em,
-  xleftmargin=1.5em,
+  %xleftmargin=1.5em,
   escapechar=`,}
@@
 \DeclareRobustCommand{\frameworkbox}[1]{\leftbar#1\endleftbar}
-\newcommand{\frameworkboxtitle}[1]{\noindent{\firamedium #1.}\quad}
+\newcommand{\frameworkboxtitle}[1]{\noindent{\firamedium #1}\quad}
 
@@
   area={Programming Systems},
+}
+
+\paperdetails{
+  submitted=2024-05-31,
+  published=2024-10-15,
+  year=2025,
+  volume=9,
+  issue=1,
+  articlenumber=2,
 }
@@
 Contact him at \email{gklitt@gmail.com}.}
-\affiliation[e]{Ink \& Switch, Planet Earth}
+\affiliation[e]{Ink \& Switch, USA}
 
@@
 \vspace{-0.4em}
-\paragraph{Locality of Change.}
+\paragraph{Locality of Change}
 We can view diverse real-world schema evolution problems from a common perspective by distinguishing between two different dimensions of schema change.
@@
 \vspace{-0.4em}
-\paragraph{Program Layers.}
+\paragraph{Program Layers}
 We consider the more permanent layer of \emph{schema} and two less permanent layers of code and
@@
 \vspace{-0.4em}
-\paragraph{Program Variants.}
+\paragraph{Program Variants}
 In programming systems that enable collaboration, multiple variants of a program may exist
@@
 \vspace{-0.4em}
-\paragraph{Challenge Problems.}
+\paragraph{Challenge Problems}
 These dimensions, layers, and variants generate a wide range of problem configuration.
@@
 \begin{table}[t]
+\caption{Changes to types and how a programming system should handle them}
+\label{tbl:elmchanges}
 \begin{tabular}{lll}\toprule
@@
 \end{tabular}
-\vspace{0.3em}
-\caption{Changes to types and how a programming system should handle them}
-\label{tbl:elmchanges}
-\vspace{-1em}
+\vspace{-0.5em}
 \end{table}
@@
 \begin{table}[t]
+\caption{Changes to schema and how a programming system should handle them}
+\label{tbl:dbchanges}
 \begin{tabular}{lll}\toprule
@@
 \end{tabular}
-\vspace{0.3em}
-\caption{Changes to types and how a programming system should handle them}
-\label{tbl:dbchanges}
-\vspace{-1em}
+\vspace{-0.5em}
 \end{table}
@@
 \begin{table}[t]
+\caption{Changes to schema and data and how the programming system should respond}
+\label{tbl:docchanges}
 \begin{adjustwidth}{-2em}{-2em}
@@
 \end{tabular}
-\vspace{0.3em}
-\caption{Changes to schema and data and how the programming system should respond}
-\label{tbl:docchanges}
 \end{adjustwidth}
-\vspace{-1em}
+\vspace{-0.5em}
 \end{table}
@@
 \begin{table}[b]
+\caption{Possible program changes and how to deal with them at run time}
+\label{tbl:smchanges}
 \begin{tabular}{lll}\toprule
@@
 \end{tabular}
-\vspace{0.3em}
-\caption{Possible program changes and how to deal with them at run time}
-\label{tbl:smchanges}
-\vspace{-1em}
+\vspace{-0.5em}
 \end{table}
@@
 
-\paragraph{Schema evolution.}
+\paragraph{Schema Evolution}
 Schema evolution has long been a major problem for SQL databases because
@@
 
-\paragraph{Type-driven transformations.}
+\paragraph{Type-Driven Transformations}
 Programming language theory undergirds research on type-driven transformations.
@@
 
-\paragraph{Model Driven Engineering.}
+\paragraph{Model Driven Engineering}
 Model Driven Engineering must also handle evolution.
@@
 
-\paragraph{Migration by example.}
+\paragraph{Migration by Example}
 Programming by example has been used to infer schema migration programs from examples specified by the developer~\cite{wang20, Alexe11}.
@@
 
-\paragraph{Data wrangling.} The evolution of data formats is a concern in Data Science. Data~Diff~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation.
+\paragraph{Data Wrangling} The evolution of data formats is a concern in Data Science. Data~Diff~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation.
 Wrangler~\cite{kandel11} observes interactive manipulation of sampled data to write data transformation programs.
@@
 
-\paragraph{Live programming.}
+\paragraph{Live Programming}
 Live programming~\cite{tanimoto90, rein2018exploratory, HazelnutLive19} seeks to speed up feedback in (primarily solo) programming.

### 2024-11-06 sources/paper.tex
Cleanup, final tweaks, sources

--- 2024-10-25 paper.tex
@@
 
-\authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}
-
 \keywords{Type Evolution, Schema Evolution, Live Programming, Local-first Software, Version Control, Hot Reloading, Image-based Programming}
@@
 \challenge{Extract Entity}
-Acme needs to evolve the above schema and data into two tables (Fig.~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
+Acme needs to evolve the above schema and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
 one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
@@
 The operation is needed when we realize that some attributes of one type of entity should belong to
-a distinct type of entity that will be associated with the first one.
+a distinct type of entity that will be associated with the~first~one.
 
@@
 Jennings). The first challenge involves merging a data edit
-(Fig.~\ref{fig:conf-add}) with a schema and data edit (Fig.~\ref{fig:conf-tab}); the second involves
-merging a schema and data edit (Fig.~\ref{fig:conf-tab}) with edit adding code (Fig.~\ref{fig:conf-budget}).
-Formulas in the final version (Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
+(Figure~\ref{fig:conf-add}) with a schema and data edit (Figure~\ref{fig:conf-tab}); the second involves
+merging a schema and data edit (Figure~\ref{fig:conf-tab}) with edit adding code (Figure~\ref{fig:conf-budget}).
+Formulas in the final version (Figure~\ref{fig:conf-finfin}) are updated to match the new structure.}
 \label{fig:conf}
@@
 \begin{figure}[t]
-\begin{adjustwidth}{-3em}{-3em}
+\begin{adjustwidth}{-2.5em}{-2.5em}
 \begin{minipage}{0.4\textwidth}
@@
 \end{minipage}
-\hspace*{2pt}
 \vline
@@
   \end{minipage}
-\hspace*{2pt}
 \vline
 \begin{minipage}{0.3\textwidth}
-  \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},basicstyle=\footnotesize,numbers=none]
+  \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},basicstyle=\footnotesize,numbers=none,xleftmargin=1.5em]
 machine Doors
@@
 
-\paragraph{Data Wrangling} The evolution of data formats is a concern in Data Science. Data~Diff~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation.
+\paragraph{Data Wrangling} The evolution of data formats is a concern in Data Science.
+The Data~Diff tool~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation.
+AI~assistants~\cite{AIassistants} provide a mechanism for interactively correcting
+transformations inferred by Data~Diff and similar automatic tools.
 Wrangler~\cite{kandel11} observes interactive manipulation of sampled data to write data transformation programs.
